### 2021-03-26 README.Rmd
Initial

---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# dialpastrepository

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/isakro/dialpastrepository/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). *`r Title`*. Name of journal/book
> <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). *`r Title`*. Name of journal/book, Accessed
> `r format(Sys.Date(), "%d %b %Y")`. Online at
> <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). *Compendium of R code and
> data for `r Title`*. Accessed `r format(Sys.Date(), "%d %b %Y")`.
> Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

-   [:file_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
-   [:file_folder: data](/analysis/data): Data used in the analysis.
-   [:file_folder: figures](/analysis/figures): Plots and other
    illustrations
-   [:file_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents prepared
    and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical
programming language R. To work with the compendium, you will need
installed on your computer the [R
software](https://cloud.r-project.org/) itself and optionally [RStudio
Desktop](https://rstudio.com/products/rstudio/download/).

The simplest way to explore the text, code and data is to click on
[binder](https://mybinder.org/v2/gh/isakro/dialpastrepository/master?urlpath=rstudio)
to open an instance of RStudio in your browser, which will have the
compendium files ready to work with. Binder uses rocker-project.org
Docker images to ensure a consistent and reproducible computational
environment. These Docker images can also be used locally.

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:

-   open the `.Rproj` file in RStudio
-   run `devtools::install()` to ensure you have the packages this
    analysis depends on (also listed in the [DESCRIPTION](/DESCRIPTION)
    file).
-   finally, open `analysis/paper/paper.Rmd` and knit to produce the
    `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")`
    in the R console

### Licenses

**Text and figures :**
[CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/)
attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please
see our [contributor guidelines](CONTRIBUTING.md). Please note that this
project is released with a [Contributor Code of Conduct](CONDUCT.md). By
participating in this project you agree to abide by its terms.

### 2021-03-26 analysis/paper/paper.Rmd
Initial

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2021-03-26 analysis/paper/paper.Rmd
Friday afternoon session. Attempts at CA and PCA analysis

@@
 ---
-title: "Title Goes Here"
+title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
 author:
-  - Myfirstname Mylastname:
-      email: fl@oneplace.edu
-      institute: [UofO]
+  - Isak Roalkvam:
+      email: isak.roalkvam@iakh.uio.no
+      institute: [UiO]
       correspondence: true
-  - Yourfirstname Yourlastname:
-      email: fl@another.edu
-      institute: [UofA]
-      correspondence: false
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - UiO: University of Oslo
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-abstract: |
-  Text of abstract
-keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
 ---
@@
 <!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-<!-- With the following code you can access and display values from the yml header above. -->
-
-Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
 
@@
 
-<!-- The actual document text starts here: -->
-
-# Introduction
-
-Here is a citation [@Marwick2017]
-
-# Background
-
-# Methods
-
-# Results
+```{r libraries, eval = TRUE}
+library(tidyverse)
+library(FactoMineR)
+library(factoextra)
+library(patchwork)
+```
 
@@
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
+# Plots
+
+```{r get-data, eval = TRUE}
+sites <- read.csv(here::here('analysis/data/raw_data/Roalkvam_DialPast_March2021_data.csv'))
 ```
 
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
+```{r wrangling}
+lithics <- sites %>%
+  # Combine flint projectiles into a single variable
+  mutate(
+    projectile = microlith + tanged_point + transverse_point + 
+    single_edged_point + hognipen_point + projectile_undefined,
+    small_tool = scraper + knife + burin + drill + blade_retouch + 
+      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
+    macro_tool = axe + chisel, 
+    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
+    small_tool_other = knife_other + scraper_other + drill_other + 
+      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
+      flake_retouch_other,
+    .keep = "unused") %>%
+    # Find mean age for each site (to be used in the plot below)
+  rowwise() %>% 
+  mutate(
+    mean_date = mean(c(reported_start_bp, reported_end_bp)),
+    .keep = "unused") %>% 
+  ungroup() %>% 
+  # Make site names the rownames
+  column_to_rownames(var = "site") %>% 
+  # Deselect a few of the columns not be used
+  select(-one_of(c("n", "excavation_project", "excavated_m3")))
 ```
 
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
+```{r cor, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 7, fig.width = 10}
+# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
+# so is excluded here
+calithics <- CA(select(lithics, -mean_date), graph = FALSE)
 
-```{r demo-inline-code}
-x <- round(pi, 2)
+# Find date to hold the central tendency date
+centdate <- mean(lithics$mean_date)
+
+# Call to plot
+fviz_ca_row(calithics, col.row = lithics$mean_date, title = "",
+                     repel = TRUE) +
+  scale_color_gradient2(name = "Years cal. BP", low="blue", mid="yellow", 
+                        high="red", midpoint = centdate)
+
+fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "")
+
 ```
 
-Here is an example of inline code `r x` in the middle of a sentence.
+Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data. \@ref(fig:cor)
 
-# Discussion
+```{r pcadata}
+# Tried to do this to avoid having to type the variable names into both mutate()
+# and select(). The general syntax select(-(vector_of_variable_names)) worked,
+# but not: mutate(new_varaible = artefact_count / sum(vector_of_variable_names))
+primary_lithics <- c("core", "core_fragment", "blade", "microblade", "chip",
+                     "fragment", "flake", "flake_other", "fragment_other", 
+                     "chip_other", "blade_other",  "microblade_other", 
+                     "core_other", "core_fragment_other")
 
-# Conclusion
+secondary_lithics <- c("axe_debitage", "axe", "chisel", "microlith", 
+"tanged_point", "transverse_point", "single_edged_point", "hognipen_point",
+"projectile_undefined", "microburin", "scraper", "knife", "burin", "drill", 
+"blade_retouch", "microblade_retouch", "chip_retouch", "fragment_retouch",
+"flake_retouch", "axe_other", "axe_debitage_other", "chisel_other",
+"club_hatchet_other", "grinding_slab_stone", "projectile_point_other",
+"knife_other", "scraper_other", "drill_other", "microblade_retouch_other", 
+"blade_retouch_other", "fragment_retouch_other", "flake_retouch_other")
 
-# Acknowledgements
+lithics2 <- sites %>% 
+  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
+  rowwise() %>% 
+  # Create a new column with mutate() called "artefact_count" which is to hold 
+  # the sum() of the artefact columns for each row. 
+  mutate(
+    artefact_count = sum(c_across(core:flake_retouch_other))) %>% 
+  mutate(
+    mean_date = mean(c(reported_start_bp, reported_end_bp)),
+    .keep = "unused") %>% 
+  ungroup() %>% 
+  mutate(chip_freq = artefact_count / sum(chip, chip_other),
+         core_freq = artefact_count / sum(core, core_other),
+         primary_freq = artefact_count / sum(core, core_fragment, blade, 
+                     microblade, chip, fragment, flake, flake_other, 
+                     fragment_other, chip_other, blade_other,  microblade_other, 
+                     core_other, core_fragment_other),
+         secondary_freq = artefact_count / sum(axe_debitage, axe, chisel, 
+                microlith, tanged_point, transverse_point, single_edged_point,
+                hognipen_point, projectile_undefined, microburin, scraper, 
+                knife, burin, drill, blade_retouch, microblade_retouch, 
+                chip_retouch, fragment_retouch, flake_retouch, axe_other,
+                axe_debitage_other, chisel_other, club_hatchet_other, 
+                grinding_slab_stone, projectile_point_other, knife_other,
+                scraper_other, drill_other, microblade_retouch_other, 
+                blade_retouch_other, fragment_retouch_other,
+                flake_retouch_other),
+         lithic_dens = artefact_count / excavated_m3) %>%
+  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
+  column_to_rownames("site")
+```
+
+```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
+pcalithics <- PCA(select(lithics2, -mean_date), graph = FALSE)
+fviz_pca(pcalithics, col.ind = lithics2$mean_date, col.var = "black", 
+         repel = TRUE) +  scale_color_gradient2(name = "Years cal. BP", 
+         low="blue", mid="yellow", high="red", midpoint = centdate)
+
+```
+
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using the continuous measures as defined by @Bicho2020
 

### 2021-03-26 analysis/paper/paper.tex
Friday afternoon session. Attempts at CA and PCA analysis

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway},
  pdfauthor={Isak Roalkvam1,},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway}
\author{Isak Roalkvam\textsuperscript{1,*}}
\date{26 March, 2021}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\textsuperscript{1} University of Oslo

\textsuperscript{*} Correspondence: \href{mailto:isak.roalkvam@iakh.uio.no}{Isak Roalkvam \textless{}\href{mailto:isak.roalkvam@iakh.uio.no}{\nolinkurl{isak.roalkvam@iakh.uio.no}}\textgreater{}}

\hypertarget{plots}{%
\section{Plots}\label{plots}}

\includegraphics{../figures/cor-1.pdf} \includegraphics{../figures/cor-2.pdf}

Figure \ref{fig:cor} displays a correspondence analysis using the lithic count data. \ref{fig:cor}

\begin{figure}
\centering
\includegraphics{../figures/pca-analysis-1.pdf}
\caption{\label{fig:pca-analysis}PCA.}
\end{figure}

Figure \ref{fig:pca-analysis} displays a principle components analysis using the continuous measures as defined by Bicho and Cascalheira (2020)

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Bicho2020}{}%
Bicho, N., Cascalheira, J., 2020. Use of lithic assemblages for the definition of short-term occupations in hunter-gatherer prehistory, in: Cascalheira, J., Picin, A. (Eds.),. Springer, Cham, p. 1938. \url{https://doi.org/10.1007/978-3-030-27403-0₂}

\end{CSLReferences}

\newpage

\hypertarget{colophon}{%
\subsubsection{Colophon}\label{colophon}}

This report was generated on 2021-03-26 16:50:59 using the following computational environment and dependencies:

\begin{verbatim}
#> - Session info ---------------------------------------------------------------
#>  setting  value                       
#>  version  R version 4.0.4 (2021-02-15)
#>  os       Linux Mint 19.3             
#>  system   x86_64, linux-gnu           
#>  ui       X11                         
#>  language en_US                       
#>  collate  en_US.UTF-8                 
#>  ctype    en_US.UTF-8                 
#>  tz       Europe/Oslo                 
#>  date     2021-03-26                  
#> 
#> - Packages -------------------------------------------------------------------
#>  package       * version  date       lib source        
#>  abind           1.4-5    2016-07-21 [1] CRAN (R 4.0.3)
#>  assertthat      0.2.1    2019-03-21 [1] CRAN (R 4.0.3)
#>  backports       1.2.0    2020-11-02 [1] CRAN (R 4.0.3)
#>  bookdown        0.21     2020-10-13 [1] CRAN (R 4.0.3)
#>  broom           0.7.3    2020-12-16 [1] CRAN (R 4.0.3)
#>  callr           3.5.1    2020-10-13 [1] CRAN (R 4.0.3)
#>  car             3.0-10   2020-09-29 [1] CRAN (R 4.0.3)
#>  carData         3.0-4    2020-05-22 [1] CRAN (R 4.0.3)
#>  cellranger      1.1.0    2016-07-27 [1] CRAN (R 4.0.3)
#>  cli             2.2.0    2020-11-20 [1] CRAN (R 4.0.3)
#>  cluster         2.1.1    2021-02-14 [4] CRAN (R 4.0.3)
#>  colorspace      1.4-1    2019-03-18 [1] CRAN (R 4.0.3)
#>  crayon          1.3.4    2017-09-16 [1] CRAN (R 4.0.3)
#>  curl            4.3      2019-12-02 [1] CRAN (R 4.0.3)
#>  data.table      1.13.4   2020-12-08 [1] CRAN (R 4.0.3)
#>  DBI             1.1.0    2019-12-15 [1] CRAN (R 4.0.3)
#>  dbplyr          2.0.0    2020-11-03 [1] CRAN (R 4.0.3)
#>  desc            1.2.0    2018-05-01 [1] CRAN (R 4.0.3)
#>  devtools        2.3.2    2020-09-18 [1] CRAN (R 4.0.3)
#>  digest          0.6.27   2020-10-24 [1] CRAN (R 4.0.3)
#>  dplyr         * 1.0.2    2020-08-18 [1] CRAN (R 4.0.3)
#>  DT              0.16     2020-10-13 [1] CRAN (R 4.0.3)
#>  ellipsis        0.3.1    2020-05-15 [1] CRAN (R 4.0.3)
#>  evaluate        0.14     2019-05-28 [1] CRAN (R 4.0.3)
#>  factoextra    * 1.0.7    2020-04-01 [1] CRAN (R 4.0.3)
#>  FactoMineR    * 2.4      2020-12-11 [1] CRAN (R 4.0.3)
#>  fansi           0.4.1    2020-01-08 [1] CRAN (R 4.0.3)
#>  farver          2.0.3    2020-01-16 [1] CRAN (R 4.0.3)
#>  flashClust      1.01-2   2012-08-21 [1] CRAN (R 4.0.3)
#>  forcats       * 0.5.0    2020-03-01 [1] CRAN (R 4.0.3)
#>  foreign         0.8-81   2020-12-22 [4] CRAN (R 4.0.3)
#>  fs              1.5.0    2020-07-31 [1] CRAN (R 4.0.3)
#>  generics        0.1.0    2020-10-31 [1] CRAN (R 4.0.3)
#>  ggplot2       * 3.3.2    2020-06-19 [1] CRAN (R 4.0.3)
#>  ggpubr          0.4.0    2020-06-27 [1] CRAN (R 4.0.3)
#>  ggrepel         0.9.1    2021-01-15 [1] CRAN (R 4.0.3)
#>  ggsignif        0.6.0    2019-08-08 [1] CRAN (R 4.0.3)
#>  glue            1.4.2    2020-08-27 [1] CRAN (R 4.0.3)
#>  gtable          0.3.0    2019-03-25 [1] CRAN (R 4.0.3)
#>  haven           2.3.1    2020-06-01 [1] CRAN (R 4.0.3)
#>  here            1.0.0    2020-11-15 [1] CRAN (R 4.0.3)
#>  highr           0.8      2019-03-20 [1] CRAN (R 4.0.3)
#>  hms             0.5.3    2020-01-08 [1] CRAN (R 4.0.3)
#>  htmltools       0.5.0    2020-06-16 [1] CRAN (R 4.0.3)
#>  htmlwidgets     1.5.2    2020-10-03 [1] CRAN (R 4.0.3)
#>  httr            1.4.2    2020-07-20 [1] CRAN (R 4.0.3)
#>  jsonlite        1.7.1    2020-09-07 [1] CRAN (R 4.0.3)
#>  knitr           1.30     2020-09-22 [1] CRAN (R 4.0.3)
#>  labeling        0.4.2    2020-10-20 [1] CRAN (R 4.0.3)
#>  lattice         0.20-41  2020-04-02 [1] CRAN (R 4.0.3)
#>  leaps           3.1      2020-01-16 [1] CRAN (R 4.0.3)
#>  lifecycle       0.2.0    2020-03-06 [1] CRAN (R 4.0.3)
#>  lubridate       1.7.9.2  2020-11-13 [1] CRAN (R 4.0.3)
#>  magrittr        2.0.1    2020-11-17 [1] CRAN (R 4.0.3)
#>  MASS            7.3-53.1 2021-02-12 [4] CRAN (R 4.0.3)
#>  memoise         1.1.0    2017-04-21 [1] CRAN (R 4.0.3)
#>  modelr          0.1.8    2020-05-19 [1] CRAN (R 4.0.3)
#>  munsell         0.5.0    2018-06-12 [1] CRAN (R 4.0.3)
#>  openxlsx        4.2.3    2020-10-27 [1] CRAN (R 4.0.3)
#>  patchwork     * 1.1.0    2020-11-09 [1] CRAN (R 4.0.3)
#>  pillar          1.4.7    2020-11-20 [1] CRAN (R 4.0.3)
#>  pkgbuild        1.1.0    2020-07-13 [1] CRAN (R 4.0.3)
#>  pkgconfig       2.0.3    2019-09-22 [1] CRAN (R 4.0.3)
#>  pkgload         1.1.0    2020-05-29 [1] CRAN (R 4.0.3)
#>  prettyunits     1.1.1    2020-01-24 [1] CRAN (R 4.0.3)
#>  processx        3.4.4    2020-09-03 [1] CRAN (R 4.0.3)
#>  ps              1.4.0    2020-10-07 [1] CRAN (R 4.0.3)
#>  purrr         * 0.3.4    2020-04-17 [1] CRAN (R 4.0.3)
#>  R6              2.5.0    2020-10-28 [1] CRAN (R 4.0.3)
#>  Rcpp            1.0.5    2020-07-06 [1] CRAN (R 4.0.3)
#>  readr         * 1.4.0    2020-10-05 [1] CRAN (R 4.0.3)
#>  readxl          1.3.1    2019-03-13 [1] CRAN (R 4.0.3)
#>  remotes         2.2.0    2020-07-21 [1] CRAN (R 4.0.3)
#>  reprex          0.3.0    2019-05-16 [1] CRAN (R 4.0.3)
#>  rio             0.5.26   2021-03-01 [1] CRAN (R 4.0.4)
#>  rlang           0.4.10   2020-12-30 [1] CRAN (R 4.0.4)
#>  rmarkdown       2.5      2020-10-21 [1] CRAN (R 4.0.3)
#>  rprojroot       2.0.2    2020-11-15 [1] CRAN (R 4.0.3)
#>  rstatix         0.6.0    2020-06-18 [1] CRAN (R 4.0.3)
#>  rstudioapi      0.13     2020-11-12 [1] CRAN (R 4.0.3)
#>  rvest           0.3.6    2020-07-25 [1] CRAN (R 4.0.3)
#>  scales          1.1.1    2020-05-11 [1] CRAN (R 4.0.3)
#>  scatterplot3d   0.3-41   2018-03-14 [1] CRAN (R 4.0.3)
#>  sessioninfo     1.1.1    2018-11-05 [1] CRAN (R 4.0.3)
#>  stringi         1.5.3    2020-09-09 [1] CRAN (R 4.0.3)
#>  stringr       * 1.4.0    2019-02-10 [1] CRAN (R 4.0.3)
#>  testthat        3.0.0    2020-10-31 [1] CRAN (R 4.0.3)
#>  tibble        * 3.0.4    2020-10-12 [1] CRAN (R 4.0.3)
#>  tidyr         * 1.1.2    2020-08-27 [1] CRAN (R 4.0.3)
#>  tidyselect      1.1.0    2020-05-11 [1] CRAN (R 4.0.3)
#>  tidyverse     * 1.3.0    2019-11-21 [1] CRAN (R 4.0.3)
#>  usethis         2.0.1    2021-02-10 [1] CRAN (R 4.0.4)
#>  vctrs           0.3.5    2020-11-17 [1] CRAN (R 4.0.3)
#>  withr           2.3.0    2020-09-22 [1] CRAN (R 4.0.3)
#>  xfun            0.19     2020-10-30 [1] CRAN (R 4.0.3)
#>  xml2            1.3.2    2020-04-23 [1] CRAN (R 4.0.3)
#>  yaml            2.2.1    2020-02-01 [1] CRAN (R 4.0.3)
#>  zip             2.1.1    2020-08-27 [1] CRAN (R 4.0.3)
#> 
#> [1] /home/isak/R/x86_64-pc-linux-gnu-library/4.0
#> [2] /usr/local/lib/R/site-library
#> [3] /usr/lib/R/site-library
#> [4] /usr/lib/R/library
\end{verbatim}

The current Git commit details are:

\begin{verbatim}
#> Local:    master /home/isak/phd/dialpast_r/dialpastrepository
#> Remote:   master @ origin (https://github.com/isakro/dialpastrepository.git)
#> Head:     [75c1135] 2021-03-26: Initial
\end{verbatim}

\end{document}

### 2021-03-26 analysis/paper/paper.Rmd
Slight changes to the data prep for PCA

@@
 library(factoextra)
-library(patchwork)
 ```
@@
 
-Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data. \@ref(fig:cor)
+Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data.
 
@@
   ungroup() %>% 
-  mutate(chip_freq = artefact_count / sum(chip, chip_other),
-         core_freq = artefact_count / sum(core, core_other),
-         primary_freq = artefact_count / sum(core, core_fragment, blade, 
+  # The reason for not using .keep below is for the option to retain artefact_count
+  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
+         core_freq = sum(core, core_other) / artefact_count,
+         primary_freq = sum(core, core_fragment, blade, 
                      microblade, chip, fragment, flake, flake_other, 
                      fragment_other, chip_other, blade_other,  microblade_other, 
-                     core_other, core_fragment_other),
-         secondary_freq = artefact_count / sum(axe_debitage, axe, chisel, 
+                     core_other, core_fragment_other) / artefact_count,
+         secondary_freq =  sum(axe_debitage, axe, chisel, 
                 microlith, tanged_point, transverse_point, single_edged_point,
@@
                 blade_retouch_other, fragment_retouch_other,
-                flake_retouch_other),
+                flake_retouch_other) / artefact_count,
          lithic_dens = artefact_count / excavated_m3) %>%
@@
 ```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
-pcalithics <- PCA(select(lithics2, -mean_date), graph = FALSE)
+pcalithics <- PCA(select(lithics2, -mean_date), scale.unit = TRUE, graph = FALSE)
 fviz_pca(pcalithics, col.ind = lithics2$mean_date, col.var = "black", 
@@
 
-Figure \@ref(fig:pca-analysis) displays a principle components analysis using the continuous measures as defined by @Bicho2020
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using continuous measures as defined by @Bicho2020
 

### 2021-03-26 analysis/paper/paper.Rmd
Removed sites with artefact count < 100 from the analysis

--- 2021-03-26 analysis/paper/paper.Rmd
@@
 
-<!-- The actual document text starts here: -->
-
-# Introduction
-
-Here is a citation [@Marwick2017]
-
-# Background
-
-# Methods
-
-# Results
+```{r libraries, eval = TRUE}
+library(tidyverse)
+library(FactoMineR)
+library(factoextra)
+```
 
@@
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
+# Plots
+
+```{r get-data, eval = TRUE}
+sites <- read.csv(here::here('analysis/data/raw_data/Roalkvam_DialPast_March2021_data.csv'))
+
 ```
 
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
+```{r wrangling}
+sites <- sites %>% 
+      # Find mean age for each site (to be used in the plot below)
+  rowwise() %>% 
+  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp)),
+         .keep = "unused") %>% 
+  mutate(
+    artefact_count = sum(c_across(core:flake_retouch_other))) %>% 
+  filter(artefact_count > 100)
+
+lithics <- sites %>%
+  # Combine flint projectiles into a single variable
+  mutate(
+    projectile = microlith + tanged_point + transverse_point + 
+    single_edged_point + hognipen_point + projectile_undefined,
+    small_tool = scraper + knife + burin + drill + blade_retouch + 
+      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
+    macro_tool = axe + chisel, 
+    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
+    small_tool_other = knife_other + scraper_other + drill_other + 
+      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
+      flake_retouch_other,
+    .keep = "unused") %>%
+  # Make site names the rownames
+  column_to_rownames(var = "site") %>% 
+  # Deselect a few of the columns not be used
+  select(-one_of(c("n", "excavation_project", "excavated_m3")))
 ```
 
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
+```{r cor, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 5, fig.width = 10}
+# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
+# so is excluded here
+calithics <- CA(select(lithics, -mean_date), graph = FALSE)
 
-```{r demo-inline-code}
-x <- round(pi, 2)
+# Find date to hold the central tendency date
+centdate <- mean(lithics$mean_date) / 1000
+
+# Call to plot
+fviz_ca_row(calithics, col.row = lithics$mean_date / 1000, title = "",
+                     repel = TRUE) +
+  scale_color_gradient2(name = "Years cal. BP", low="blue", mid="yellow", 
+                        high="red", midpoint = centdate) +
+  theme(legend.position = "bottom")
+
+fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "")
+
 ```
 
-Here is an example of inline code `r x` in the middle of a sentence.
+Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data.
 
-# Discussion
+```{r pcadata}
+# Tried to do this to avoid having to type the variable names into both mutate()
+# and select(). The general syntax select(-(vector_of_variable_names)) worked,
+# but not: mutate(new_varaible = sum(vector_of_variable_names) / artefact_count)
+primary_lithics <- c("core", "core_fragment", "blade", "microblade", "chip",
+                     "fragment", "flake", "flake_other", "fragment_other", 
+                     "chip_other", "blade_other",  "microblade_other", 
+                     "core_other", "core_fragment_other")
 
-# Conclusion
+secondary_lithics <- c("axe_debitage", "axe", "chisel", "microlith", 
+"tanged_point", "transverse_point", "single_edged_point", "hognipen_point",
+"projectile_undefined", "microburin", "scraper", "knife", "burin", "drill", 
+"blade_retouch", "microblade_retouch", "chip_retouch", "fragment_retouch",
+"flake_retouch", "axe_other", "axe_debitage_other", "chisel_other",
+"club_hatchet_other", "grinding_slab_stone", "projectile_point_other",
+"knife_other", "scraper_other", "drill_other", "microblade_retouch_other", 
+"blade_retouch_other", "fragment_retouch_other", "flake_retouch_other")
 
-# Acknowledgements
+lithics2 <- sites %>% 
+  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
+  # The reason for not using .keep below is for the option to retain artefact_count
+  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
+         core_freq = sum(core, core_other) / artefact_count,
+         primary_freq = sum(core, core_fragment, blade, 
+                     microblade, chip, fragment, flake, flake_other, 
+                     fragment_other, chip_other, blade_other,  microblade_other, 
+                     core_other, core_fragment_other) / artefact_count,
+         secondary_freq =  sum(axe_debitage, axe, chisel, 
+                microlith, tanged_point, transverse_point, single_edged_point,
+                hognipen_point, projectile_undefined, microburin, scraper, 
+                knife, burin, drill, blade_retouch, microblade_retouch, 
+                chip_retouch, fragment_retouch, flake_retouch, axe_other,
+                axe_debitage_other, chisel_other, club_hatchet_other, 
+                grinding_slab_stone, projectile_point_other, knife_other,
+                scraper_other, drill_other, microblade_retouch_other, 
+                blade_retouch_other, fragment_retouch_other,
+                flake_retouch_other) / artefact_count,
+         lithic_dens = artefact_count / excavated_m3) %>%
+  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
+  column_to_rownames("site")
+```
+
+```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
+pcalithics <- PCA(select(lithics2, -mean_date), scale.unit = TRUE, graph = FALSE)
+fviz_pca(pcalithics, col.ind = lithics2$mean_date / 1000, col.var = "black", 
+         repel = TRUE) +  scale_color_gradient2(name = "Years cal. BP", 
+         low="blue", mid="yellow", high="red", midpoint = centdate) +
+  theme(legend.position = "bottom")
+
+```
+
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using continuous measures as defined by @Bicho2020
 

### 2021-03-27 analysis/paper/paper.Rmd
Adding some initial interpretations of the CA and PCA results.

@@
 
-```{r cor, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 5, fig.width = 10}
+```{r cor, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 8, fig.width = 10}
 # Perform correspondence analysis on the data. 'mean_age' is meant for colouring
 # so is excluded here
-calithics <- CA(select(lithics, -mean_date), graph = FALSE)
+calithics <- CA(select(lithics, -mean_date, -artefact_count), graph = FALSE)
+
+# Retrieve eigenvalues
+eigval <- get_eigenvalue(calithics)
+explained_cumul_dim2 <- eigval[2,3]
 
@@
                      repel = TRUE) +
-  scale_color_gradient2(name = "Years cal. BP", low="blue", mid="yellow", 
+  scale_color_gradient2(name = "ka BP", low="blue", mid="yellow", 
                         high="red", midpoint = centdate) +
@@
 
-Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data.
+Figure \@ref(fig:cor) displays a CA using the lithic count data.
+While there is a considerable degree of mixing, and no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension appears associated with the patterning in the data.
+This is indicated by the general transition across the colour scale in the row plot, combined with the fact that the two first dimensions of the CA-analysis account for as much as `r round(explained_cumul_dim2, 2)` % of the inertia in the data.
+The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites being pulled towards the bottom, along the second dimension.
+The sites from the later parts of the Mesolithic are mainly drawn out along the first dimension of the plot.
+
+The earliest sites are characterised by microburins, projectiles, macro tools and associated debitage in flint.
+It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, while complete cores appear to be a trait more defining of assemblages with later dates.
+The non-flint material on the earliest sites also appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades appear to these sites.
+
+The first dimension of the plot is mainly defined by macro tools and associated debitage in non-flint materials that stand in contrast to more flint dominated assemblages and non-flint projectiles.
+While the outer end of the first dimensions is dominated by later Mesolithic sites such as Stokke/Polland 3, Vallermyrene 4 and Gunnarsrød 10, which are all associated with axe production in non-flint materials, the later Mesolithic sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
+In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
+The use of metarhyolite is typically seen as a feature of the Middle Mesolithic, but is evidently not as prominent a part of the assemblages on the sites treated here, with the exception of for Nedre Hobekk 2.
+T
 
@@
 fviz_pca(pcalithics, col.ind = lithics2$mean_date / 1000, col.var = "black", 
-         repel = TRUE) +  scale_color_gradient2(name = "Years cal. BP", 
+         repel = TRUE) +  scale_color_gradient2(name = "ka BP", 
          low="blue", mid="yellow", high="red", midpoint = centdate) +
@@
 
-Figure \@ref(fig:pca-analysis) displays a principle components analysis using continuous measures as defined by @Bicho2020
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020.
+While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the fequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here.
+If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into Mesolithic Norway, indicating that earlier sites are associated with higher degree of mobility that sites from later phases.
 

### 2021-03-27 analysis/paper/paper.Rmd
Removed site labels from CA and PCA plots. Some additional plots and tables exploring the components of the PCA

@@
 output: 
-    bookdown::word_document2:
+    bookdown::html_document2:
       fig_caption: yes
@@
   comment = "#>",
+  fig.width = 9,
   fig.path = "../figures/",
@@
 library(tidyverse)
+library(patchwork)
+library(gt)
 library(FactoMineR)
@@
 
-```{r cor, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 8, fig.width = 10}
+```{r cor1, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 8, fig.width = 10}
 # Perform correspondence analysis on the data. 'mean_age' is meant for colouring
@@
 # Find date to hold the central tendency date
-centdate <- mean(lithics$mean_date) / 1000
+centdate <- (sum(range(lithics$mean_date)) / 2) / 1000
 
-# Call to plot
-fviz_ca_row(calithics, col.row = lithics$mean_date / 1000, title = "",
-                     repel = TRUE) +
-  scale_color_gradient2(name = "ka BP", low="blue", mid="yellow", 
+# Plot with site names tags (probably most immediately relevant for an audience 
+# familiar with the sites)
+fviz_ca_row(calithics, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "ka BP", low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
   theme(legend.position = "bottom") 
+```
 
-fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "")
+```{r cor, fig.cap = "Correspondence analysis using artefact count data."}
+# Plot without site names tags
+fviz_ca_row(calithics, geom = "point", col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "ka BP", low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  theme(legend.position = "bottom") 
+#guide(fill = guide_legend(reverse = TRUE)) # Reverses the legend but makes it
+                                            # discrete
+
+# Plot variables
+fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "B")
 
@@
 Figure \@ref(fig:cor) displays a CA using the lithic count data.
-While there is a considerable degree of mixing, and no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension appears associated with the patterning in the data.
+While no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension is associated with the patterning in the data.
 This is indicated by the general transition across the colour scale in the row plot, combined with the fact that the two first dimensions of the CA-analysis account for as much as `r round(explained_cumul_dim2, 2)` % of the inertia in the data.
-The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites being pulled towards the bottom, along the second dimension.
-The sites from the later parts of the Mesolithic are mainly drawn out along the first dimension of the plot.
+The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
+The sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension.
 
-The earliest sites are characterised by microburins, projectiles, macro tools and associated debitage in flint.
+The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage.
 It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, while complete cores appear to be a trait more defining of assemblages with later dates.
-The non-flint material on the earliest sites also appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades appear to these sites.
+The non-flint material on the earliest sites also appear to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
 
-The first dimension of the plot is mainly defined by macro tools and associated debitage in non-flint materials that stand in contrast to more flint dominated assemblages and non-flint projectiles.
+The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
 While the outer end of the first dimensions is dominated by later Mesolithic sites such as Stokke/Polland 3, Vallermyrene 4 and Gunnarsrød 10, which are all associated with axe production in non-flint materials, the later Mesolithic sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
 In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
-The use of metarhyolite is typically seen as a feature of the Middle Mesolithic, but is evidently not as prominent a part of the assemblages on the sites treated here, with the exception of for Nedre Hobekk 2.
+The use of metarhyolite is typically seen as a feature of the Middle Mesolithic, but is evidently not as prominent a part of the assemblages on the other sites that are contemporaneous with Nedre Hobekk 2.
 T
@@
 pcalithics <- PCA(select(lithics2, -mean_date), scale.unit = TRUE, graph = FALSE)
-fviz_pca(pcalithics, col.ind = lithics2$mean_date / 1000, col.var = "black", 
-         repel = TRUE) +  scale_color_gradient2(name = "ka BP", 
-         low="blue", mid="yellow", high="red", midpoint = centdate) +
+fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithics2$mean_date / 1000, 
+          repel = TRUE, col.var = "black", pointshape = 21,  pointsize = 2.9,
+         title = "") + 
+  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
   theme(legend.position = "bottom")
@@
 Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020.
-While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the fequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here.
+While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the frequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here.
 If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into Mesolithic Norway, indicating that earlier sites are associated with higher degree of mobility that sites from later phases.
+
+```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
+# Scree plot indicates an 'elbow' at the second dimension. 
+# The two first dimension account for 68 % and the first three
+# for 87 % (see table below)
+# fviz_screeplot(pcalithics, addlabels=TRUE)
+
+# Create a table showing the eigenvalues of the components
+round(as.tibble(get_eig(pcalithics)), 2) %>% 
+  rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
+         "Cumulative variance (%)" = cumulative.variance.percent) %>% 
+  rownames_to_column(var = "Component") %>% 
+gt(rowname_col = "row", auto_align = FALSE)
+```
+
+```{r var_contrib, fig.cap = "Contribution of variables to components."}
+# The plots below investigates the contribution of the variables to the
+# components. The code has in large part been taken from 
+# Bicho & Cascalheira (2020).
+var <- get_pca_var(pcalithics)
+dim1 <- fviz_contrib(pcalithics, choice = "var", axes = 1,
+                     color = "black", fill = "gold2", 
+                     title = "Contribution of variables to Component 1")
+dim2 <- fviz_contrib(pcalithics, choice = "var", axes = 2, 
+                     color = "black", fill = "gold2", 
+                     title = "Contribution of variables to Component 2")
+dim3 <- fviz_contrib(pcalithics, choice = "var", axes = 3,
+                     color = "black", fill = "gold2", 
+                     title = "Contribution of variables to Component 3")
+dim4 <- fviz_contrib(pcalithics, choice = "var", axes = 4,
+                     color = "black", fill = "gold2", 
+                     title = "Contribution of variables to Component 4")
+
+# Patchwork for arranging plots
+(dim1 | dim2) /
+(dim3 | dim4)
+```
+
+```{r var_cor, fig.cap = "Plots showing the correlation between the variables"}
+plt1 <- fviz_pca_var(pcalithics,
+                  col.var = "black",
+                  title = "Correlation plot: Component 1/Component 2",
+                  repel = TRUE
+                  )
+plt2 <- fviz_pca_var(pcalithics,
+                  col.var = "black",
+                  axes = 3:4,
+                  title = "Correlation plot: Component 3/Component 4",
+                  repel = TRUE
+                  )
+
+plt1 + plt2
+```
 

### 2021-03-28 analysis/paper/paper.Rmd
Adjustments to plots (still need to reverse the legend without making it display discrete values)

@@
 
-```{r cor, fig.cap = "Correspondence analysis using artefact count data."}
+```{r cor, fig.cap = "Correspondence analysis using artefact count data. A) Object map, B) Variable map.", fig.height = 13}
 # Plot without site names tags
-fviz_ca_row(calithics, geom = "point", col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+plt1 <- fviz_ca_row(calithics, geom = "point", col.row = "black", fill = lithics$mean_date / 1000, title = "A",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 # Plot variables
-fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "B")
+plt2 <- fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "B")
+
+plt1 /
+plt2
 
@@
 While no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot, combined with the fact that the two first dimensions of the CA-analysis account for as much as `r round(explained_cumul_dim2, 2)` % of the inertia in the data.
+This is indicated by the general transition across the colour scale in the row plot, combined with the fact that the two first dimensions of the CA-analysis account for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
@@
 While the outer end of the first dimensions is dominated by later Mesolithic sites such as Stokke/Polland 3, Vallermyrene 4 and Gunnarsrød 10, which are all associated with axe production in non-flint materials, the later Mesolithic sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
+Although the sample size is quite strained and the discussion of finer chronological points might as a consequence not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
+This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research.
+
 In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
-The use of metarhyolite is typically seen as a feature of the Middle Mesolithic, but is evidently not as prominent a part of the assemblages on the other sites that are contemporaneous with Nedre Hobekk 2.
+The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2 treated here.
 T
@@
 While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the frequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here.
-If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into Mesolithic Norway, indicating that earlier sites are associated with higher degree of mobility that sites from later phases.
+If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility that sites from later phases.
 

### 2021-03-30 analysis/paper/paper.Rmd
Adding some text and references from the analysis proposal to the workshop to test the Zotero functionality in Rmd

@@
 
-# Plots
+# Introduction
+
+Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories -- for example as related to the production of blades [@damlien2016].
+Some studies have attempted more comprehensive evaluations of a number of lithic inventories to get at potential site types [@viken2018; @breivik2020], but these have been conducted without support from quantitative frameworks that can help structure such analyses.
+In sum, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods.
+My analysis would aim to employ multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites.
+This would be done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
+However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g.
+@dibble2017], ranging from the impact of natural formation processes to various and intermixed behavioural aspects such as purpose, duration, frequency, and group sizes at visits to the sites.
+Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
+Finally, analytical and methodological dimensions relating to survey, excavation and classification practices will also influence how the assemblages are defined.
+Consequently, the proposed analysis would be conducted from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
+In this light, the following sections will present the data to be analysed, a few potential sources of bias that might affect the comparisons to be conducted, while also highlighting a few ways in which idiosyncrasies in the chosen material might alleviate some of these issues.
 

### 2021-03-31 analysis/paper/paper.Rmd
Some more writing. NTS: Terming the density/retouch relationship WABI as Bicho and Cascalheira do might not be entirely in line with Clark & Barton?

--- 2021-03-28 analysis/paper/paper.Rmd
@@
 
-# Plots
+# Introduction
+
+Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories -- for example as related to the production of blades or axes [@damlien2016].
+Some studies have attempted more comprehensive evaluations of a number of lithic inventories to get at potential site types [@viken2018; @breivik2020], but these have been conducted without support from quantitative frameworks that can help structure such analyses, and have treated site type as a categorical variable with three categories.
+In sum therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods.
+The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites.
+This is done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
+However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g.
+@dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency, and group sizes at visits to the sites.
+Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
+Finally, analytical and methodological dimensions relating to survey, excavation and classification practices will also influence how the assemblages are defined.
+Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
+In this light, the following sections will present the data to be analysed, a few potential sources of bias that might affect the comparisons to be conducted, while also highlighting a few ways in which idiosyncrasies in the chosen material might alleviate some of these issues.
+
+## Material
 
@@
 The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage.
-It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, while complete cores appear to be a trait more defining of assemblages with later dates.
-The non-flint material on the earliest sites also appear to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, while the cores themselves appear to be a trait more defining of assemblages with later dates.
+The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
 
@@
 While the outer end of the first dimensions is dominated by later Mesolithic sites such as Stokke/Polland 3, Vallermyrene 4 and Gunnarsrød 10, which are all associated with axe production in non-flint materials, the later Mesolithic sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
-Although the sample size is quite strained and the discussion of finer chronological points might as a consequence not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
+Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
 This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research.
@@
 In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
-The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2 treated here.
+The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2.
 T
@@
 
+```{r wabi, fig.cap = "Whole-assemblage behavioural index", fig.height = 5, fig.width = 10}
+
+ggplot(lithics2, aes(log10(lithic_dens), log10(secondary_freq), 
+                     fill = mean_date / 1000)) + 
+  geom_point(shape = 21, size = 3) +
+  geom_smooth(method = "lm", col = "black", se = FALSE) +
+  labs(x = "Lithic volumetric density (log10)", 
+       y = "Relative frequency of secondarily worked lithics (log10)") +
+  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  theme_classic()
+
+wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
+```
+
+Figure \@ref(fig:wabi) displays the WABI as found for the site data.
+This indicates a negative correlation between the two variables (pearsons r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+
 ```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
@@
 If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility that sites from later phases.
+@Bicho2020 also found that WABI, as introduced by, gave a reasonable estimation of mobility in their data.
+The same is evident here where fig indicates a
 
@@
 
+The consistent nature of the WABI between this study (r = -0.50), @clark2017 (r = -0.50), @Bicho2020 (r = -0.48) is striking, and should speak to the cross-cultural and cross-temporal applicability of the WABI measure for lithic assemblages stemming from hunter-gatherer(-fisher) societies, and the potential it holds for comparative studies.
+The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.
+However, while this can perhaps support the of the .
+The question then becomes what is the significance of the negatively correlated variables of relative frequency of cores and chips?
+Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these.
+As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, another possibility is that these are indicative of site that are temporally ubiquitous.
+In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020.
+This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to elements, and failed to identify any diachronic variation in this pattern.
+As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.
+
+This is arguably more established than the variables included bysimilarities between the variable plots of @Bicho2020 and the variables indicated in is striking.
+This pertains to the negative correlation between core and chip frequency, and the negative correlation between primary and secondarily worked lithics.
+This last
+
 <!-- The following line inserts a page break  -->

### 2021-04-03 analysis/paper/paper.Rmd
Writing some more. Not limiting y-axis in WABI plot

@@
 Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories -- for example as related to the production of blades or axes [@damlien2016].
-Some studies have attempted more comprehensive evaluations of a number of lithic inventories to get at potential site types [@viken2018; @breivik2020], but these have been conducted without support from quantitative frameworks that can help structure such analyses, and have treated site type as a categorical variable with three categories.
+Some studies have attempted more comprehensive evaluations of a number of lithic inventories to get at potential site types and associated mobility patterns [@viken2018; @breivik2020], but these have been conducted without support from quantitative frameworks that can help structure such analyses.
 In sum therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods.
@@
 The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
-While the outer end of the first dimensions is dominated by later Mesolithic sites such as Stokke/Polland 3, Vallermyrene 4 and Gunnarsrød 10, which are all associated with axe production in non-flint materials, the later Mesolithic sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
+While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
@@
 The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2.
-T
+
+# Hunter-gatherer mobility and whole assemblage behavioral indicators
+
+The results of the correspondence analysis does appear to align reasonably well with previous research that is based on more informal analyses of, indicating that the employed artefact categories are behavourally meaningful.
+Aligning the signficance of this chronological patterning to specific behavioural dimensions, is, however, another task entirely, and a task that is severely hampered by fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, and consequently how the different debitage and tool categories should be treated @dibble2017.
+For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tools could have had various different purposes and had a multitude of shapes throughout their use-life.
+Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis and an overly naive trust in the fact that these units reflect
 
@@
 
-ggplot(lithics2, aes(log10(lithic_dens), log10(secondary_freq), 
+ggplot(lithics2, aes(log10(lithic_dens), log10(secondary_freq * 100), 
                      fill = mean_date / 1000)) + 
@@
 This indicates a negative correlation between the two variables (pearsons r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.
 
@@
 
-The consistent nature of the WABI between this study (r = -0.50), @clark2017 (r = -0.50), @Bicho2020 (r = -0.48) is striking, and should speak to the cross-cultural and cross-temporal applicability of the WABI measure for lithic assemblages stemming from hunter-gatherer(-fisher) societies, and the potential it holds for comparative studies.
+The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking, speaks to the cross-cultural and cross-temporal applicability of the measure, the potential it holds for comparative studies, and is to be expected given the .
 The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.

### 2021-04-04 analysis/paper/paper.Rmd
Minor edits

@@
 While no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot, combined with the fact that the two first dimensions of the CA-analysis account for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA-analysis account for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
@@
 
-The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage.
+The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage.
 It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, while the cores themselves appear to be a trait more defining of assemblages with later dates.
@@
 
-The results of the correspondence analysis does appear to align reasonably well with previous research that is based on more informal analyses of, indicating that the employed artefact categories are behavourally meaningful.
-Aligning the signficance of this chronological patterning to specific behavioural dimensions, is, however, another task entirely, and a task that is severely hampered by fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, and consequently how the different debitage and tool categories should be treated @dibble2017.
-For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tools could have had various different purposes and had a multitude of shapes throughout their use-life.
-Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis and an overly naive trust in the fact that these units reflect
+The results of the correspondence analysis does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturng some culturally meaningful patterning.
+Aligning the signficance of these chronological patterns with any specific behavioural dimensions, for example as related to various forms of mobiliy patterns or modes of resource exploitation  is, however, another task entirely.
+This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, and consequently how the different debitage and tool categories should be treated [@dibble2017].
+For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
+While use-wear analysis can help identify what an artefact was used fo towards the end of its use-file, these are typically only conducted on artefacts already grouped into a single category based on on .
+Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis, and an overly naive trust in the fact that these units reflect any behaviourally meaningful categories.
+This way of viewing lithic assembalges is a fairly established tradition outside Scandinavian Mesolithic research, meaning a range of empirical  measures associated with various mobility patterns are available.
+Of these the most well-estbalished is the.
+base don the assumption that, this measure and has been shown to display a meaningful measure across a wide range of temporal, geographical and cultural contexts.
 

### 2021-04-09 analysis/paper/paper.Rmd
More writing

@@
 library(factoextra)
+library(sf)
+```
+
+```{r get-data, eval = TRUE}
+assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
+
 ```
@@
 
-Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories -- for example as related to the production of blades or axes [@damlien2016].
-Some studies have attempted more comprehensive evaluations of a number of lithic inventories to get at potential site types and associated mobility patterns [@viken2018; @breivik2020], but these have been conducted without support from quantitative frameworks that can help structure such analyses.
-In sum therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods.
+Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016].
+Some studies have attempted more comprehensive evaluations of entire lithic assemblages to get at potential site types and associated mobility patterns [@viken2018a; @breivik2020], but these have predominantly been conducted without support from quantitative frameworks that can help structure such analyses.
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods.
 The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites.
@@
 Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
-Finally, analytical and methodological dimensions relating to survey, excavation and classification practices will also influence how the assemblages are defined.
+Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
 Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
-In this light, the following sections will present the data to be analysed, a few potential sources of bias that might affect the comparisons to be conducted, while also highlighting a few ways in which idiosyncrasies in the chosen material might alleviate some of these issues.
 
-## Material
+## Archaeological context and material
 
-```{r get-data, eval = TRUE}
-sites <- read.csv(here::here('analysis/data/raw_data/Roalkvam_DialPast_March2021_data.csv'))
+A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas.
+Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
+In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
+@sørensen1979; @romundset2018].
+The fairly rapid shoreline displacement means that the location of the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf.
+@perreault2019, 47].
+The shore-bound settlement combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of shoreline displacement allows for a relatively good chronological control of these accumulation events.
+In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
+However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
+Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.
 
-```
+The `r nrow(assemblages)` sites chosen for analysis here have a relatively limited geographical distribution.
+The sites were excavated as part of four larger projects all that took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
+The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
+This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the projects.
+Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified.
+As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g.
+@clark2006; @dibble2017].
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014].
+Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences.
+While certainly difficult to fully account for, I do believe that the relative contemporaneity of the projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis aimed at revealing behavioural rather than analytical patterning.
+
+## Methods
+
+Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without having to impose overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
+Structure that can be revealed from considering all of the assemblages in aggregate should be of value to any subsequent in-depth studies of individual sites, and as a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages.
+The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
+As such, the main aim of this analysis is to compare the results with findings reported in the literature, and have the generation of new hypotheses as a possible outcome.
+To this end, the analysis follows two analytical avenues.
+
+The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified in the original excavation reports.
+The purpose of this exercise is to evaluate the degree to which the assemblages follow a chronological structure as has been suggested by previous literature based on more informally driven methods.
+While some artefact categories will be combined in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
+However, given the vast inferential leap that this assumption involves, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with this limitation in mind.
+These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at more specific behavioural dimensions relating to mobility patterns among hunter-gatherers.
+Through a series of studies, show that the relationship between lithic volumetric density and relative frequency of retouched artefacts provides consistent and comparable results across a wide range of chronological and cultural context.
+These measures have, however, to my knowledge not seen application in Scandinavian archaeology as of yet.
+Following the continuous operationalisations of these measures, principal components analysis (PCA) is more suitable than CA for this part of the analysis [e.g.
+@baxter1994, 100; @Bicho2020].
+
+# Theory/calculation
+
+# Results
 
 ```{r wrangling}
-sites <- sites %>% 
+sites <- assemblages %>% 
       # Find mean age for each site (to be used in the plot below)
@@
 
-Figure \@ref(fig:cor) displays a CA using the lithic count data.
+Figure \@ref(fig:cor) displays the CA using the lithic count data.
 While no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension is associated with the patterning in the data.
@@
 
+In sum, the results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time.
+However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
+This is evident in the present data as well, highlighted by the influence of the later Mesolithic axe production sites, which it would be reasonable to assume corresponds to a more pervasive cultural change than a purely typo-chronological development of artefact shape.
+Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
+This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the significance of the employed units of analysis unclear [@dibble2017].
+
 In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
@@
 
-# Hunter-gatherer mobility and whole assemblage behavioral indicators
+#
 
-The results of the correspondence analysis does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturng some culturally meaningful patterning.
-Aligning the signficance of these chronological patterns with any specific behavioural dimensions, for example as related to various forms of mobiliy patterns or modes of resource exploitation  is, however, another task entirely.
-This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, and consequently how the different debitage and tool categories should be treated [@dibble2017].
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
-While use-wear analysis can help identify what an artefact was used fo towards the end of its use-file, these are typically only conducted on artefacts already grouped into a single category based on on .
+While use-wear analysis can help identify what an artefact was used for towards the end of its use-file, these are typically only conducted on artefacts already grouped into a single category based on on .
+
 Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis, and an overly naive trust in the fact that these units reflect any behaviourally meaningful categories.
-This way of viewing lithic assembalges is a fairly established tradition outside Scandinavian Mesolithic research, meaning a range of empirical  measures associated with various mobility patterns are available.
+This realisation has led to a wide range of outside Scandinavian Mesolithic research, meaning a range of empirical measures associated with various mobility patterns are available.
+
+Within this framework, retouch is viewed as a means of extending the use-life of lithics by repeated rejuvination and modification of edges.
+At logistical basecamps the extended time spent in a single location would imply a better control of raw material availability, and allow for longer trips to retrieve these.
+Higher availability of lithic raw materials should in turn reduce the necessity for extensive retouch, meaning the material expectation for logistical basecamps would be relatively high density of lithics and a low relative frequency of retouched artefacts.
+Higher degree of mobility, on the other hand, would mean a higher dependency on artefacts and material groups could bring with them, and dimensions such as weight, reliabilty, repairability and the degree to which artefacts could be manipulated to fulfill a wide range of tasks would likely have been of concern.
+From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
+
 Of these the most well-estbalished is the.
@@
 
-lithics2 <- sites %>% 
+lithics2 <- assemblages %>% 
   select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
@@
 
-The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking, speaks to the cross-cultural and cross-temporal applicability of the measure, the potential it holds for comparative studies, and is to be expected given the .
+# Discussion
+
+The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking.
+While this result is perhaps not surprising, given the previous indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
 The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.
@@
 Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these.
-As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, another possibility is that these are indicative of site that are temporally ubiquitous.
+As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, another possibility is that these are indicative of site types that are temporally ubiquitous.
 In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020.
-This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to elements, and failed to identify any diachronic variation in this pattern.
+This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to the elements, and failed to identify any diachronic variation in this pattern.
 As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.

### 2021-04-10 analysis/paper/paper.Rmd
Some more writing and introducing spatial data for map overview.

@@
 library(sf)
+library(rnaturalearth)
 ```
@@
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
-
+site_pts <- st_read('analysis/data/raw_data/sites.gpkg')
+countries <- ne_download(scale = 110, type = 'land', category = 'physical',
+                         returnclass = "sf")
 ```
@@
 
-Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016].
+Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
+@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016].
 Some studies have attempted more comprehensive evaluations of entire lithic assemblages to get at potential site types and associated mobility patterns [@viken2018a; @breivik2020], but these have predominantly been conducted without support from quantitative frameworks that can help structure such analyses.
@@
 
+Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
+Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to any subsequent in-depth studies of individual sites.
+The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
+As such, the main aim of this analysis is to compare the results with findings reported in previous literature, and have the generation of new hypotheses as a possible outcome.
+To this end, the analysis follows two analytical avenues.
+
 ## Archaeological context and material
 
+The `r nrow(assemblages)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway.
+The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
+The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
+This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations.
+Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the Museum of Cultural History.
+As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [@clark2006; @dibble2017].
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014].
+Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
+While certainly difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.
+
 A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas.
-Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
+Furthermore, coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
@@
 @perreault2019, 47].
-The shore-bound settlement combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of shoreline displacement allows for a relatively good chronological control of these accumulation events.
-In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
+The shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
+In other parts of the world a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
 However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
+In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, which is reduced to around 0.5 cm/year in the Late Mesolithic [@sørensen2014].
 Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.
 
-The `r nrow(assemblages)` sites chosen for analysis here have a relatively limited geographical distribution.
-The sites were excavated as part of four larger projects all that took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
-The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
-This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the projects.
-Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified.
-As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g.
-@clark2006; @dibble2017].
-In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014].
-Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences.
-While certainly difficult to fully account for, I do believe that the relative contemporaneity of the projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis aimed at revealing behavioural rather than analytical patterning.
+In a series of case studies from Northern Norway, @bølviken1982 were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles [@baxter1994].
+In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project.
+As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at
+
+There have been a few studies that have conducted comprehensive analyses of artefact inventories and site features to get at variation in Mesolithic site types [@viken2018; @breivik2020].
+However, the narratively driven analyses of these data makes the weighting of different variables unclear and subjectively defined, and several of the studies cited here ultimately draw their conclusions based on an arguably over-constraining trichotomy of site types.
+As Binford originally clarified, which is a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.
 
@@
 
-Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without having to impose overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
-Structure that can be revealed from considering all of the assemblages in aggregate should be of value to any subsequent in-depth studies of individual sites, and as a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages.
-The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
-As such, the main aim of this analysis is to compare the results with findings reported in the literature, and have the generation of new hypotheses as a possible outcome.
-To this end, the analysis follows two analytical avenues.
+The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+The purpose of this exercise is to evaluate the degree to which the assemblages align with patterns that have been suggested by earlier application of more informally driven methods.
+The expectation is that chronological trends will be the main driver of While some artefact categories will be combined in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
 
-The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified in the original excavation reports.
-The purpose of this exercise is to evaluate the degree to which the assemblages follow a chronological structure as has been suggested by previous literature based on more informally driven methods.
-While some artefact categories will be combined in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
 However, given the vast inferential leap that this assumption involves, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with this limitation in mind.
@@
 
-lithics2 <- assemblages %>% 
+lithics2 <- sites %>% 
   select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
@@
 While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the frequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here.
-If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility that sites from later phases.
+If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
 @Bicho2020 also found that WABI, as introduced by, gave a reasonable estimation of mobility in their data.
@@
 The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking.
-While this result is perhaps not surprising, given the previous indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
+While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for larger scale comparative studies.
 The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.
-However, while this can perhaps support the of the .
-The question then becomes what is the significance of the negatively correlated variables of relative frequency of cores and chips?
+However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips?
 Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these.
-As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, another possibility is that these are indicative of site types that are temporally ubiquitous.
+As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.
 In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020.

### 2021-04-11 analysis/paper/paper.Rmd
Changes to wabi scatter plots and some more writing

@@
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
-site_pts <- st_read('analysis/data/raw_data/sites.gpkg')
+site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
 countries <- ne_download(scale = 110, type = 'land', category = 'physical',
@@
 The `r nrow(assemblages)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway.
-The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
+The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b] .
 The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
@@
 A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas.
-Furthermore, coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
+Furthermore, coastal Mesolithic sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
 @sørensen1979; @romundset2018].
-The fairly rapid shoreline displacement means that the location of the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf.
+The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf.
 @perreault2019, 47].
@@
 However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
-In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, which is reduced to around 0.5 cm/year in the Late Mesolithic [@sørensen2014].
+In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014].
 Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.
@@
 In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project.
-As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at
+As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at identifying potential variation in the
 
@@
 However, the narratively driven analyses of these data makes the weighting of different variables unclear and subjectively defined, and several of the studies cited here ultimately draw their conclusions based on an arguably over-constraining trichotomy of site types.
-As Binford originally clarified, which is a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.
+As Binford originally clarified, a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.
 
@@
 The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-The purpose of this exercise is to evaluate the degree to which the assemblages align with patterns that have been suggested by earlier application of more informally driven methods.
-The expectation is that chronological trends will be the main driver of While some artefact categories will be combined in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
+The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier application of more informally driven methods.
+The expectation is that chronological trends will be the main driver of this part of the analysis.
+While some artefact categories from the original reports are collapsed in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
 
@@
 Through a series of studies, show that the relationship between lithic volumetric density and relative frequency of retouched artefacts provides consistent and comparable results across a wide range of chronological and cultural context.
-These measures have, however, to my knowledge not seen application in Scandinavian archaeology as of yet.
 Following the continuous operationalisations of these measures, principal components analysis (PCA) is more suitable than CA for this part of the analysis [e.g.
@@
 # Theory/calculation
+
+Higher frequency of secondarily worked lithics
 
@@
   mutate(
-    artefact_count = sum(c_across(core:flake_retouch_other))) %>% 
+    artefact_count = sum(c_across(core:flake_retouch_other)),
+    flint_count = sum(c_across(core:flake_retouch))) %>% 
   filter(artefact_count > 100)
@@
 
-```{r cor1, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 8, fig.width = 10}
+```{r co_data}
 # Perform correspondence analysis on the data. 'mean_age' is meant for colouring
 # so is excluded here
-calithics <- CA(select(lithics, -mean_date, -artefact_count), graph = FALSE)
+calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count), graph = FALSE)
 
@@
 centdate <- (sum(range(lithics$mean_date)) / 2) / 1000
+```
 
+```{r cor1, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 8, fig.width = 10, eval = FALSE}
 # Plot with site names tags (probably most immediately relevant for an audience 
@@
 While no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA-analysis account for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
@@
 Within this framework, retouch is viewed as a means of extending the use-life of lithics by repeated rejuvination and modification of edges.
-At logistical basecamps the extended time spent in a single location would imply a better control of raw material availability, and allow for longer trips to retrieve these.
+At logistical basecamps, the extended time spent in a single location would imply a better control of raw material availability, and allow for longer trips to retrieve these.
 Higher availability of lithic raw materials should in turn reduce the necessity for extensive retouch, meaning the material expectation for logistical basecamps would be relatively high density of lithics and a low relative frequency of retouched artefacts.
-Higher degree of mobility, on the other hand, would mean a higher dependency on artefacts and material groups could bring with them, and dimensions such as weight, reliabilty, repairability and the degree to which artefacts could be manipulated to fulfill a wide range of tasks would likely have been of concern.
+Higher degree of mobility, on the other hand, would mean a higher dependency on the artefacts and material people could bring with them, and dimensions such as weight, reliabilty, repairability and the degree to which artefacts could be manipulated to fulfill a wide range of tasks would likely have been of concern.
 From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
@@
 # but not: mutate(new_varaible = sum(vector_of_variable_names) / artefact_count)
-primary_lithics <- c("core", "core_fragment", "blade", "microblade", "chip",
-                     "fragment", "flake", "flake_other", "fragment_other", 
+primary_flint <- c("core", "core_fragment", "blade", "microblade", "chip",
+                     "fragment", "flake")
+
+primary_lithics <- c(primary_flint, "flake_other", "fragment_other", 
                      "chip_other", "blade_other",  "microblade_other", 
@@
 
-secondary_lithics <- c("axe_debitage", "axe", "chisel", "microlith", 
-"tanged_point", "transverse_point", "single_edged_point", "hognipen_point",
-"projectile_undefined", "microburin", "scraper", "knife", "burin", "drill", 
-"blade_retouch", "microblade_retouch", "chip_retouch", "fragment_retouch",
-"flake_retouch", "axe_other", "axe_debitage_other", "chisel_other",
-"club_hatchet_other", "grinding_slab_stone", "projectile_point_other",
-"knife_other", "scraper_other", "drill_other", "microblade_retouch_other", 
-"blade_retouch_other", "fragment_retouch_other", "flake_retouch_other")
+secondary_flint <- c("axe_debitage", "axe", "chisel", "microlith", 
+                    "tanged_point", "transverse_point", "single_edged_point",
+                    "hognipen_point", "projectile_undefined", "microburin",
+                    "scraper", "knife", "burin", "drill", "blade_retouch",
+                    "microblade_retouch", "chip_retouch", "fragment_retouch",
+                    "flake_retouch")
+
+secondary_lithics <- c(secondary_flint, "axe_other", "axe_debitage_other",
+                       "chisel_other", "club_hatchet_other", 
+                       "grinding_slab_stone", "projectile_point_other", 
+                       "knife_other", "scraper_other", "drill_other", 
+                       "microblade_retouch_other", "blade_retouch_other",
+                       "fragment_retouch_other", "flake_retouch_other")
 
@@
          lithic_dens = artefact_count / excavated_m3) %>%
-  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
+  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
+         -flint_count) %>% 
   column_to_rownames("site")
+
+flint2 <- sites %>% 
+  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
+  select(!ends_with("other")) %>% 
+  # The reason for not using .keep below is for the option to retain artefact_count
+  mutate(chip_freq =  sum(chip) / artefact_count,
+         core_freq = sum(core) / artefact_count,
+         primary_freq = sum(core, core_fragment, blade, 
+                     microblade, chip, fragment, flake) / artefact_count,
+         secondary_freq =  sum(axe_debitage, axe, chisel, 
+                microlith, tanged_point, transverse_point, single_edged_point,
+                hognipen_point, projectile_undefined, microburin, scraper, 
+                knife, burin, drill, blade_retouch, microblade_retouch, 
+                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
+         flint_dens = flint_count / excavated_m3) %>%
+  select(-primary_flint, -secondary_flint, -excavated_m3, -artefact_count,
+         -flint_count) %>% 
+  column_to_rownames("site")
+
+wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
+wabicorflint <- cor(log10(flint2$flint_dens), log10(flint2$secondary_freq))
 ```
 
+Figure \@ref(fig:wabi) displays the WABI as found for the site data.
+This indicates a negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.
+Variable raw-material availability can also impact these measures as [@manninen2014].
+Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies.
+However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically.
+Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly chronologically, the general pattern is the same.
+
 ```{r wabi, fig.cap = "Whole-assemblage behavioural index", fig.height = 5, fig.width = 10}
-
-ggplot(lithics2, aes(log10(lithic_dens), log10(secondary_freq * 100), 
+w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
                      fill = mean_date / 1000)) + 
@@
   geom_smooth(method = "lm", col = "black", se = FALSE) +
-  labs(x = "Lithic volumetric density (log10)", 
+  labs(x = "Volumetric density of lithics (log10)", 
        y = "Relative frequency of secondarily worked lithics (log10)") +
@@
                         high="red", midpoint = centdate) +
+  scale_y_log10() +
+  scale_x_log10() +
   theme_classic()
 
-wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
+w2 <- ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
+                     fill = mean_date / 1000)) + 
+  geom_point(shape = 21, size = 3) +
+  geom_smooth(method = "lm", col = "black", se = FALSE) +
+  labs(x = "Volumetric density of flint (log10)", 
+       y = "Relative frequency of secondarily worked flint (log10)") +
+  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  scale_y_log10() +
+  scale_x_log10() +
+  theme_classic()
+
+w1 | w2
 ```
-
-Figure \@ref(fig:wabi) displays the WABI as found for the site data.
-This indicates a negative correlation between the two variables (pearsons r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
-Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.
 

### 2021-04-12 analysis/paper/paper.Rmd
Assembling map overview

@@
 
-```{r get-data, eval = TRUE}
+```{r get-data, eval = TRUE, message = FALSE, warning = FALSE}
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
 site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
-countries <- ne_download(scale = 110, type = 'land', category = 'physical',
+countries <- ne_download(scale = 10, type = 'countries', category = 'cultural',
                          returnclass = "sf")
@@
 Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
-@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016].
-Some studies have attempted more comprehensive evaluations of entire lithic assemblages to get at potential site types and associated mobility patterns [@viken2018a; @breivik2020], but these have predominantly been conducted without support from quantitative frameworks that can help structure such analyses.
-In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods.
+@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016; @solheim2020].
+Some studies have attempted more comprehensive evaluations of entire lithic assemblages, mainly to get at potential site types and associated mobility patterns among Early Mesolithic sites [@breivik2016; @viken2018a; @breivik2020].
+Furthermore, only a single study has employed multivariate statistics in the treatment of the assemblages quantitative frameworks that can help structure such analyses [cf.
+@naeroy2018, 905].
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to qualitatively and subjectively driven methods.
 The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites.
@@
 ## Archaeological context and material
+```{r wrangling}
+sites <- assemblages %>% 
+      # Find mean age for each site (to be used in the plot below)
+  rowwise() %>% 
+  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp))) %>% 
+  mutate(
+    artefact_count = sum(c_across(core:flake_retouch_other)),
+    flint_count = sum(c_across(core:flake_retouch))) %>% 
+  filter(artefact_count > 100)
+```
 
-The `r nrow(assemblages)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway.
-The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b] .
+The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway.
+The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
+The sites included here are all Mesolithic sites with artefact counts above 100 excavated in conjuction with the projects.
 The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
@@
 While certainly difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.
+
+```{r overview}
+# Create a bounding box around site points to limit the plot
+sitbbox <- st_bbox(site_pts)
+sitbbox[1:2] <- boundingbox[1:2] - 500000
+sitbbox[3:4] <- boundingbox[3:4] + 500000
+boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
+
+bound_reproj <- st_transform(boundingpoly, st_crs(countries))
+mapcountries <- countries %>% 
+  filter(st_intersects(., bound_reproj, sparse = FALSE))
+count_reproj <- st_transform(mapcountries, st_crs(site_pts))
+
+overview <-
+  ggplot() +
+  geom_sf(data = count_reproj, fill = "grey", colour = NA) +
+  geom_sf(data = st_as_sf(st_as_sfc(st_bbox(site_pts))),
+          fill = NA, colour = "black", size = 1) +
+  #geom_sf(data = site_pts, size = 2, shape=21, fill = "red", 
+  #        colour = "black") +
+  #ggsn::scalebar(data = site_pts, dist = 150, dist_unit = "m", transform = FALSE,
+  #               st.size = 2.75, height = 0.015, border.size = 0.1) +
+coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
+         expand = FALSE) +
+ theme_bw() + theme(axis.title=element_blank(),
+                     axis.text.y = element_blank(),
+                     axis.text.x = element_blank(),
+                     rect = element_rect(),
+                     axis.ticks = element_blank(),
+                     panel.grid.major = element_blank())
+
+sitbbox <- st_bbox(site_pts)
+sitbbox[1:2] <- boundingbox[1:2] - 5000
+sitbbox[3:4] <- boundingbox[3:4] + 5000
+boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
+
+bound_reproj <- st_transform(boundingpoly, st_crs(countries))
+mapcountries <- countries %>% 
+  filter(st_intersects(., bound_reproj, sparse = FALSE))
+count_reproj <- st_transform(mapcountries, st_crs(site_pts))
+
+anc <- as.numeric(c(sitbbox$ymin, sitbbox$xmax))
+spat <-
+  ggplot() +
+  geom_sf(data = count_reproj, fill = "grey", colour = NA) +
+  geom_sf(data = site_pts, size = 2, shape=21, fill = "red", 
+          colour = "black") +
+  ggsn::scalebar(data = site_pts, dist = 15, dist_unit = "km", 
+                 transform = FALSE, st.size = 4, height = 0.1, 
+                 border.size = 0.05, anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
+coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
+         expand = FALSE) +
+ theme_bw() + theme(axis.title=element_blank(),
+                     axis.text.y = element_blank(),
+                     axis.text.x = element_blank(),
+                     rect = element_rect(),
+                     axis.ticks = element_blank(),
+                     panel.grid.major = element_blank())
+
+yrs <- seq(11500, 5000, -500)
+temp <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
+  geom_segment(aes(x = reported_start_bp, xend = reported_end_bp,
+                   yend = reorder(site, mean_date),
+                   colour = excavation_project), size = 1.7) +
+  xlab("Cal. BP") +
+  scale_x_reverse() +
+  scale_color_manual(values = c("springgreen4", "royalblue2", "gold1", "firebrick3")) +
+  theme_classic() +
+  theme(axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank(),
+        legend.position = "none")
+```
 
@@
 
-```{r wrangling}
-sites <- assemblages %>% 
-      # Find mean age for each site (to be used in the plot below)
-  rowwise() %>% 
-  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp)),
-         .keep = "unused") %>% 
-  mutate(
-    artefact_count = sum(c_across(core:flake_retouch_other)),
-    flint_count = sum(c_across(core:flake_retouch))) %>% 
-  filter(artefact_count > 100)
-
+```{r lithics}
 lithics <- sites %>%

### 2021-04-13 analysis/paper/paper.Rmd
Creating map overview

@@
 Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
-@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016; @solheim2020].
-Some studies have attempted more comprehensive evaluations of entire lithic assemblages, mainly to get at potential site types and associated mobility patterns among Early Mesolithic sites [@breivik2016; @viken2018a; @breivik2020].
-Furthermore, only a single study has employed multivariate statistics in the treatment of the assemblages quantitative frameworks that can help structure such analyses [cf.
+@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016; @solheim2020], or a handful of sites.
+Some studies have attempted more comprehensive evaluations of entire lithic assemblages associated with larger groups of sites, mainly to get at potential site types and associated mobility patterns among Early Mesolithic sites [@breivik2016; @viken2018a; @breivik2020].
+However, only a single study has employed a multivariate quantitative framework in the treatment of the assemblages [cf.
 @naeroy2018, 905].
@@
 
-Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
-Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to any subsequent in-depth studies of individual sites.
+Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
+Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
 The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
@@
 sites <- assemblages %>% 
-      # Find mean age for each site (to be used in the plot below)
+      # Find mean age for each site (to be used in the plots below)
   rowwise() %>% 
@@
   mutate(
+    # Find total artefact count and flint count for each site
     artefact_count = sum(c_across(core:flake_retouch_other)),
     flint_count = sum(c_across(core:flake_retouch))) %>% 
+  # Exclude sites with less than 100 artefacts
   filter(artefact_count > 100)
@@
 The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
-The sites included here are all Mesolithic sites with artefact counts above 100 excavated in conjuction with the projects.
+The sites included here consists of all Mesolithic sites excavated in conjunction with the projects with assemblages holding more than 100 artefacts.
 The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
@@
 Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
-While certainly difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.
+While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.
 
 ```{r overview}
-# Create a bounding box around site points to limit the plot
+# Create a bounding box around site points to limit the overview map
 sitbbox <- st_bbox(site_pts)
-sitbbox[1:2] <- boundingbox[1:2] - 500000
-sitbbox[3:4] <- boundingbox[3:4] + 500000
+sitbbox[1:2] <- sitbbox[1:2] - 500000
+sitbbox[3:4] <- sitbbox[3:4] + 500000
 boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
 
+# Reproject the bounding box to match world map, and crop the world map
+# with the bounding box polygon
 bound_reproj <- st_transform(boundingpoly, st_crs(countries))
@@
 
+# Create bounding box around the sites, both for extent indicator on the
+# overview map and to create basemap for the site overview.
+sitbbox2 <- st_bbox(site_pts)
+sitbbox2[1:2] <- sitbbox2[1:2] - 15000
+sitbbox2[3:4] <- sitbbox2[3:4] + 15000
+boundingpoly2 <- st_as_sf(st_as_sfc(sitbbox2))
+
+# same as above, using the smaller bounding box
+bound_reproj2 <- st_transform(boundingpoly2, st_crs(countries))
+mapcountries2 <- countries %>% 
+  filter(st_intersects(., bound_reproj2, sparse = FALSE))
+count_reproj2 <- st_transform(mapcountries2, st_crs(site_pts))
+
+# Assemble overview map
 overview <-
@@
   geom_sf(data = count_reproj, fill = "grey", colour = NA) +
-  geom_sf(data = st_as_sf(st_as_sfc(st_bbox(site_pts))),
+  geom_sf(data = boundingpoly2,
           fill = NA, colour = "black", size = 1) +
-  #geom_sf(data = site_pts, size = 2, shape=21, fill = "red", 
-  #        colour = "black") +
-  #ggsn::scalebar(data = site_pts, dist = 150, dist_unit = "m", transform = FALSE,
-  #               st.size = 2.75, height = 0.015, border.size = 0.1) +
   coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
@@
 
-sitbbox <- st_bbox(site_pts)
-sitbbox[1:2] <- boundingbox[1:2] - 5000
-sitbbox[3:4] <- boundingbox[3:4] + 5000
-boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
+# Combine spatial point data with assemblage data (holds excavation project
+# for colouring points in the maps)
+sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id"))
 
-bound_reproj <- st_transform(boundingpoly, st_crs(countries))
-mapcountries <- countries %>% 
-  filter(st_intersects(., bound_reproj, sparse = FALSE))
-count_reproj <- st_transform(mapcountries, st_crs(site_pts))
-
-anc <- as.numeric(c(sitbbox$ymin, sitbbox$xmax))
-spat <-
+# Assemble site map. anc defined first is used for placing the scalebar,
+# which required some tweaking to get right.
+anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
+sitemap <-
   ggplot() +
-  geom_sf(data = count_reproj, fill = "grey", colour = NA) +
-  geom_sf(data = site_pts, size = 2, shape=21, fill = "red", 
+  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
+  geom_sf(data = sitejoin, aes(fill = excavation_project),
+          size = 2.25, shape=21, 
           colour = "black") +
-  ggsn::scalebar(data = site_pts, dist = 15, dist_unit = "km", 
-                 transform = FALSE, st.size = 4, height = 0.1, 
-                 border.size = 0.05, anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
-coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
+  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
+                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
+                    values = c("springgreen4", "royalblue2",
+                                "gold1", "firebrick3")) +
+  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
+                 transform = FALSE, st.size = 4, height = 0.05, 
+                 border.size = 0.05, st.dist = 0.05,
+                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
+  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
+           ylim = c(sitbbox2[2], sitbbox2[4]),
          expand = FALSE) +
+  labs(fill = "Excavation project:") +
  theme_bw() + theme(axis.title=element_blank(),
@@
                      axis.ticks = element_blank(),
-                     panel.grid.major = element_blank())
+                     panel.grid.major = element_blank(),
+                    legend.position = "bottom")
 
@@
   scale_x_reverse() +
-  scale_color_manual(values = c("springgreen4", "royalblue2", "gold1", "firebrick3")) +
+  scale_color_manual(values = c("springgreen4", "royalblue2",
+                                "gold1", "firebrick3")) +
   theme_classic() +
@@
 A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas.
-Furthermore, coastal Mesolithic sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
+Furthermore, these coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
@@
 @perreault2019, 47].
-The shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
-In other parts of the world a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
+Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
+In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
 However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
@@
 Higher availability of lithic raw materials should in turn reduce the necessity for extensive retouch, meaning the material expectation for logistical basecamps would be relatively high density of lithics and a low relative frequency of retouched artefacts.
-Higher degree of mobility, on the other hand, would mean a higher dependency on the artefacts and material people could bring with them, and dimensions such as weight, reliabilty, repairability and the degree to which artefacts could be manipulated to fulfill a wide range of tasks would likely have been of concern.
+Higher degree of mobility, on the other hand, would mean a higher dependency on the artefacts and the raw-material people could bring with them, and dimensions such as weight, reliabilty, repairability and the degree to which artefacts could be manipulated to fulfill a wide range of tasks would likely have been of concern.
 From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
+
+It has, however, been suggested that access to raw-materials will be a more important decider for assemblage composition than mobility patterns [@andrefsky1994], and that variable workability of these materials will impact factors such as the relative frequency of tools to debitage in the assemblages [@manninen2014].
+Raw-material accessibility in south-eastern Norway is arguably understudied,.
+In addition to attempts at correcting for this in the analysis conducted here (see below), the limited geographical distribution of the sites should alleviate the issue somewhat, as the overall availability of raw materials could be expected to be more comparable through time than if a wider region was under consideration.
 

### 2021-04-14 analysis/paper/paper.Rmd
Maps done, some additional writing

@@
 
-```{r get-data, eval = TRUE, message = FALSE, warning = FALSE}
+```{r get-data, eval = TRUE, results = 'hide'}
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
@@
 
-The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway.
+The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref@(fig:dist)A].
 The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
@@
 As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [@clark2006; @dibble2017].
-In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014].
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref@(fig:dist)B].
 Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
@@
 
-```{r overview}
+```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
+The date ranges follow those given in the original excavation reports", fig.height = 10, fig.width = 12}
 # Create a bounding box around site points to limit the overview map
 sitbbox <- st_bbox(site_pts)
-sitbbox[1:2] <- sitbbox[1:2] - 500000
-sitbbox[3:4] <- sitbbox[3:4] + 500000
+sitbbox[1:2] <- sitbbox[1:2] - 600000
+sitbbox[3:4] <- sitbbox[3:4] + 600000
 boundingpoly <- st_as_sf(st_as_sfc(sitbbox))
@@
          expand = FALSE) +
+  labs(title = "A") +
   theme_bw() + theme(axis.title=element_blank(),
@@
           colour = "black") +
+  labs(fill = "Excavation project") +
   scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
@@
                  transform = FALSE, st.size = 4, height = 0.05, 
-                 border.size = 0.05, st.dist = 0.05,
+                 border.size = 0.05, st.dist = 0.07,
                  anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
@@
          expand = FALSE) +
-  labs(fill = "Excavation project:") +
  theme_bw() + theme(axis.title=element_blank(),
@@
                      panel.grid.major = element_blank(),
-                    legend.position = "bottom")
+                     legend.position = "none") +
+  guides(fill = guide_legend(ncol = 2,
+                             title.position = "top", title.hjust = 0.5))
 
 yrs <- seq(11500, 5000, -500)
-temp <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
+temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
   geom_segment(aes(x = reported_start_bp, xend = reported_end_bp,
@@
   scale_x_reverse() +
-  scale_color_manual(values = c("springgreen4", "royalblue2",
+  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
+                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
+                     values = c("springgreen4", "royalblue2",
                                 "gold1", "firebrick3")) +
+  labs(title = "B", colour = "Excavation project") +
   theme_classic() +
@@
         axis.ticks.y=element_blank(),
-        legend.position = "none")
+        legend.position = "bottom") +
+  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5))
+
+(overview + sitemap) /
+(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))
+
 ```
@@
 @perreault2019, 47].
-Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
-In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
+Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
+In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
 However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
@@
 Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.
+
+The data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of 48 variables representing different debitage and tool types.
+While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions.
+Some categories in the original reports have therefore been combined.
+This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm.
+In the reports, this was only separated from (macro-)blades (width $\leq$ 12 mm) and micro-blades (width $\leq$ 8 mm) for some of the sites.
+This was combined with the blade category here.
+Furthermore, the artefact data is here divided into flint and non-flint materials.
+Flint does not occur indigenously in Norway, and is only available as nodules that have been transported and deposited by retreating and drifting ice [e.g.
+@berg-hansen1999].
+This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46].
+Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material accessibility and technological choices [@eigeland2015, 45-53].
+Furthermore, reducing the number of dimensions by lumping all non-flint materials together also has some issues.
+A general sentiment is that an increase in the use of non-flint materials signifies an increased adaption to local environments, and possibly a reduction in mobility.
+However, the various non-flint raw materials can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+Given this differentiated use, the retained debitage and tool categories are expected to still reflect the differentiated use associated with different raw-materials.
+An important benefit of combining all non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised [cf.
+@frivoll2017].
 
@@
 The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier application of more informally driven methods.
-The expectation is that chronological trends will be the main driver of this part of the analysis.
-While some artefact categories from the original reports are collapsed in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
+The expectation was that chronological trends would be the main driver of this part of the analysis - this appears to be largely verfied.
+While some artefact categories from the original reports are collapsed in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this largely involved an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
 
@@
 # so is excluded here
-calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count), graph = FALSE)
+calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count,
+                       -reported_start_bp, -reported_end_bp, -askeladden_id),
+                graph = FALSE)
 
@@
 The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage.
-It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, while the cores themselves appear to be a trait more defining of assemblages with later dates.
+It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, as opposed to the cores themselves.
 The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
@@
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
-This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research.
+This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [@eigeland2015, 370].
 
@@
 
-It has, however, been suggested that access to raw-materials will be a more important decider for assemblage composition than mobility patterns [@andrefsky1994], and that variable workability of these materials will impact factors such as the relative frequency of tools to debitage in the assemblages [@manninen2014].
-Raw-material accessibility in south-eastern Norway is arguably understudied,.
-In addition to attempts at correcting for this in the analysis conducted here (see below), the limited geographical distribution of the sites should alleviate the issue somewhat, as the overall availability of raw materials could be expected to be more comparable through time than if a wider region was under consideration.
+It has, however, been suggested that access to raw-materials can be a more important decider for assemblage composition than mobility patterns , and that variable workability of these materials will impact factors such as the relative frequency of tools to debitage in the assemblages [e.g.
+@andrefsky1994; @manninen2014].
+In addition to attempts at accounting for this analytically (see below), the limited geographical distribution of the sites should alleviate the issue somewhat, as the overall availability of raw materials could be expected to be more comparable through time than if a wider region was under consideration.
 
@@
 lithics2 <- sites %>% 
-  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
-  # The reason for not using .keep below is for the option to retain artefact_count
+  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
+         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
+  # The reason for not using .keep below is for the option to 
+  #retain artefact_count
   mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
@@
 
+# Repeat but for flint only
 flint2 <- sites %>% 
-  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
+  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
+         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
   select(!ends_with("other")) %>% 
-  # The reason for not using .keep below is for the option to retain artefact_count
   mutate(chip_freq =  sum(chip) / artefact_count,
@@
 
-```{r wabi, fig.cap = "Whole-assemblage behavioural index", fig.height = 5, fig.width = 10}
+```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint.
+The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
 w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
@@
   geom_smooth(method = "lm", col = "black", se = FALSE) +
-  labs(x = "Volumetric density of lithics (log10)", 
+  labs(title = "A",
+       x = "Volumetric density of lithics (log10)", 
        y = "Relative frequency of secondarily worked lithics (log10)") +
@@
   geom_smooth(method = "lm", col = "black", se = FALSE) +
-  labs(x = "Volumetric density of flint (log10)", 
+  labs(title = "B", 
+       x = "Volumetric density of flint (log10)", 
        y = "Relative frequency of secondarily worked flint (log10)") +
@@
 Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020.
-While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the frequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here.
-If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
-@Bicho2020 also found that WABI, as introduced by, gave a reasonable estimation of mobility in their data.
-The same is evident here where fig indicates a
+While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data.
+hese last dimensions are capturing the most variation in the dataset presented here.
+Thus, if frequency of secondarily worked artefacts is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
 

### 2021-04-16 analysis/paper/paper.Rmd
Writing, expedient vs. curated

@@
 ## Archaeological context and material
+
 ```{r wrangling}
@@
 The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
-The sites included here consists of all Mesolithic sites excavated in conjunction with the projects with assemblages holding more than 100 artefacts.
+The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects with assemblages holding more than 100 artefacts.
 The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
@@
 
-The data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of 48 variables representing different debitage and tool types.
+The data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
 While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions.
-Some categories in the original reports have therefore been combined.
+Some categories in the original reports have therefore been combined in the dataset.
 This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm.
-In the reports, this was only separated from (macro-)blades (width $\leq$ 12 mm) and micro-blades (width $\leq$ 8 mm) for some of the sites.
+In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
 This was combined with the blade category here.
-Furthermore, the artefact data is here divided into flint and non-flint materials.
+Furthermore, the artefact data has been divided into flint and non-flint materials.
 Flint does not occur indigenously in Norway, and is only available as nodules that have been transported and deposited by retreating and drifting ice [e.g.
@@
 Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material accessibility and technological choices [@eigeland2015, 45-53].
-Furthermore, reducing the number of dimensions by lumping all non-flint materials together also has some issues.
-A general sentiment is that an increase in the use of non-flint materials signifies an increased adaption to local environments, and possibly a reduction in mobility.
-However, the various non-flint raw materials can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+Furthermore, a general sentiment is that an increase in the use of non-flint materials signifies an increased adaption to local environments, and possibly a reduction in mobility.
+However, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
 Given this differentiated use, the retained debitage and tool categories are expected to still reflect the differentiated use associated with different raw-materials.
@@
 The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier application of more informally driven methods.
-The expectation was that chronological trends would be the main driver of this part of the analysis - this appears to be largely verfied.
-While some artefact categories from the original reports are collapsed in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this largely involved an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
+The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by studies employing more informally driven methods.
+This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
+However, the approach taken is partially informed by research founded on the so-called Frison Effect.
+These scholars have argued that most morphological variation in retouched lithics is not the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part related to variable degrees of modification through use and rejuvination.
+Consequently, some artefact categories have here been collapased into a single category for the CA.
+This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch.
+That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as analytic units potentially entirely misleading.
+These have all been combined into the single category "small flint tools" here.
+(A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
+While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis.
+An underlying assumption is therefore that the retained categories represent artefacts categories that have fulfilled different purposes or are related to different technological processes.
+While ultimately intuitive in nature, it does seem reasonable to assume that large non-flint stone tools (axes, adzes, chisels, clubs and hatches) have fulfilled different purposes than for example the previously mentioned small flint tools.
 
-However, given the vast inferential leap that this assumption involves, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with this limitation in mind.
-These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at more specific behavioural dimensions relating to mobility patterns among hunter-gatherers.
-Through a series of studies, show that the relationship between lithic volumetric density and relative frequency of retouched artefacts provides consistent and comparable results across a wide range of chronological and cultural context.
-Following the continuous operationalisations of these measures, principal components analysis (PCA) is more suitable than CA for this part of the analysis [e.g.
-@baxter1994, 100; @Bicho2020].
+However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for [@dibble2017], does have major implications that the above-outlined analysis does not take into account.
+For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
+While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g.
+@solheim2018].
+Thus, while these analyses can potentially get at in-group variation among artefacts that we assume can be treated in aggregate, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
+As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind.
+The logic behind these measures are founded in an understanding of lithic technology as being organised along a continuum ranging between curated and expedient [@binford1979].
+An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of replacement.
+Curated technological organisation has been defined as the manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@bamforth1986].
+The empirical correspondent to a curated technological organisation is consequently high degrees of retouch, as this is commonly seen as a means of extending the use-life of lithics by repeated rejuvination and modification of edges.
+The continuum between curated and expedient technology has in turn been related to settlement and mobility strategies, as well as raw-material quality and availability [@smith2015].
+Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
+From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
+More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability, and allow for the accumulation of material resources.
+Higher raw-material availability should in turn reduce the necessity for extensive retouch, meaning the material expectation for a lower degree of mobility is relatively high density of lithics and a low relative frequency of retouched artefacts.
+Additional empirical expectation for
+
+have shown through a series of studies that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent and comparable results across a wide range of chronological and cultural context, and is the main measure employed here.
+Some additional variables suggested by @Bicho2020 to be related to mobility patterns are are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
+@baxter1994, 100].
+
+A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here.
+This pertains to number of site features such as fireplaces, cooking pits and dwelling structures.
+The omission of feature count from the analysis follows from the fact that taphonomic loss is likely to lead to a chronological bias in the preservation and associated diffuclty of discovering of features.
+Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
+This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied.
+This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jorgensen2017; @darmark2018e].
+
+Within this framework, amount of retouch has been seen as a empirical indicator of degree of curation.
+While various methods for measuring the curation associated with individual tools have been forwarded, the approach taken here focuses on the assemblage considered as a whole.
+
+Here retouch i At logistical basecamps, the extended time spent in a single location would imply a better control of raw material availability, and allow for longer trips to retrieve these.
+Higher availability of lithic raw materials should in turn reduce the necessity for extensive retouch, meaning the material expectation for logistical basecamps would be relatively high density of lithics and a low relative frequency of retouched artefacts.
+From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
 
@@
 Figure \@ref(fig:cor) displays the CA using the lithic count data.
-While no clear-cut clusters can be readily delineated, the general impression from the plots is still that a chronological dimension is associated with the patterning in the data.
+While no clear-cut clusters can be readily delineated, the general impression from the plots is that a chronological dimension is associated with the patterning in the data.
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
-The sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension.
+Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
 
@@
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
-This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [@eigeland2015, 370].
+This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
+@eigeland2015, 370].
 
@@
 
-For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
-While use-wear analysis can help identify what an artefact was used for towards the end of its use-file, these are typically only conducted on artefacts already grouped into a single category based on on .
-
 Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis, and an overly naive trust in the fact that these units reflect any behaviourally meaningful categories.
 This realisation has led to a wide range of outside Scandinavian Mesolithic research, meaning a range of empirical measures associated with various mobility patterns are available.
-
-Within this framework, retouch is viewed as a means of extending the use-life of lithics by repeated rejuvination and modification of edges.
-At logistical basecamps, the extended time spent in a single location would imply a better control of raw material availability, and allow for longer trips to retrieve these.
-Higher availability of lithic raw materials should in turn reduce the necessity for extensive retouch, meaning the material expectation for logistical basecamps would be relatively high density of lithics and a low relative frequency of retouched artefacts.
-Higher degree of mobility, on the other hand, would mean a higher dependency on the artefacts and the raw-material people could bring with them, and dimensions such as weight, reliabilty, repairability and the degree to which artefacts could be manipulated to fulfill a wide range of tasks would likely have been of concern.
-From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
 
@@
 
+These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at behavioural dimensions relating to mobility patterns among hunter-gatherers.
+
 <!-- The following line inserts a page break  -->

### 2021-04-17 analysis/paper/paper.Rmd
Set up Zotero web with RStudio

@@
 The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by studies employing more informally driven methods.
+The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods.
 This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
-However, the approach taken is partially informed by research founded on the so-called Frison Effect.
-These scholars have argued that most morphological variation in retouched lithics is not the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part related to variable degrees of modification through use and rejuvination.
-Consequently, some artefact categories have here been collapased into a single category for the CA.
+However, the approach taken is partially informed by the so-called Frison Effect, which highlights the fact that the shape of lithics studied by archaeologists when they where discarded and could have had long and complex use-lifes in which they took on a multitude of different shapes.
+Starting with , several scholars have argued that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
+@dibble1995] .
+Consequently, some artefact categories have here been collapased for the CA.
 This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch.
 That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as analytic units potentially entirely misleading.
-These have all been combined into the single category "small flint tools" here.
+These have all been combined into the single category "small flint tools".
 (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
@@
 An underlying assumption is therefore that the retained categories represent artefacts categories that have fulfilled different purposes or are related to different technological processes.
-While ultimately intuitive in nature, it does seem reasonable to assume that large non-flint stone tools (axes, adzes, chisels, clubs and hatches) have fulfilled different purposes than for example the previously mentioned small flint tools.
+While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.
 
-However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for [@dibble2017], does have major implications that the above-outlined analysis does not take into account.
+However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not manage to take into account [@dibble2017].
+This renders it difficult to align any of the results with specific behavioural dimensions.
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
@@
 @solheim2018].
-Thus, while these analyses can potentially get at in-group variation among artefacts that we assume can be treated in aggregate, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
+Thus, while these analyses can potentially get at in-group variation pertaining to the end state of artefacts that we assume can be treated in aggregate, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
 As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind.
 The logic behind these measures are founded in an understanding of lithic technology as being organised along a continuum ranging between curated and expedient [@binford1979].
-An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of replacement.
-Curated technological organisation has been defined as the manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@bamforth1986].
+An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources for modification and rejuvination, resulting in high rates of replacement.
+Curated technological organisation, on the other hand, has been defined as the manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979, @bamforth1986].
 The empirical correspondent to a curated technological organisation is consequently high degrees of retouch, as this is commonly seen as a means of extending the use-life of lithics by repeated rejuvination and modification of edges.
-The continuum between curated and expedient technology has in turn been related to settlement and mobility strategies, as well as raw-material quality and availability [@smith2015].
+
+The continuum between curated and expedient technology has in turn been related to settlement and mobility strategies, as well as raw-material quality and availability [@andrefsky1994; @smith2015].
 Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
@@
 This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied.
-This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jorgensen2017; @darmark2018e].
+This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].
 

### 2021-04-19 analysis/paper/paper.Rmd
Writing on curation vs. expediency

@@
 
-Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
-@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016; @solheim2020], or a handful of sites.
-Some studies have attempted more comprehensive evaluations of entire lithic assemblages associated with larger groups of sites, mainly to get at potential site types and associated mobility patterns among Early Mesolithic sites [@breivik2016; @viken2018a; @breivik2020].
-However, only a single study has employed a multivariate quantitative framework in the treatment of the assemblages [cf.
-@naeroy2018, 905].
-In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to qualitatively and subjectively driven methods.
-The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites.
+The analyses conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
 This is done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
@@
 The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
-As such, the main aim of this analysis is to compare the results with findings reported in previous literature, and have the generation of new hypotheses as a possible outcome.
+As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway that have been based on more informally driven methods, and have the generation of new hypotheses as a possible outcome.
 To this end, the analysis follows two analytical avenues.
@@
 
+Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
+@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016; @solheim2020], or in-depth analyses of handful of sites.
+Some studies have attempted more comprehensive evaluations of entire lithic assemblages associated with larger groups of sites, mainly to get at potential site types and associated mobility patterns among Early Mesolithic sites [@breivik2016; @viken2018a; @breivik2020].
+However, only a single study has employed a multivariate quantitative framework in the treatment of the assemblages [cf.
+@naeroy2018, 905].
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to qualitatively and subjectively driven methods.
+
 There have been a few studies that have conducted comprehensive analyses of artefact inventories and site features to get at variation in Mesolithic site types [@viken2018; @breivik2020].
@@
 This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
-However, the approach taken is partially informed by the so-called Frison Effect, which highlights the fact that the shape of lithics studied by archaeologists when they where discarded and could have had long and complex use-lifes in which they took on a multitude of different shapes.
-Starting with , several scholars have argued that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
-@dibble1995] .
-Consequently, some artefact categories have here been collapased for the CA.
+However, the approach taken is partially informed by the so-called Frison Effect, which highlights the fact that the shape of lithics studied by archaeologists is that which they had when they where discarded - the artefacts can have had long and complex use-lives in which they took on a multitude of different shapes.
+Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
+@barton1991; @dibble1995] .
+Consequently, several artefact categories have here been collapased for the CA.
 This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch.
-That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as analytic units potentially entirely misleading.
+That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose, is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially entirely misleading.
 These have all been combined into the single category "small flint tools".
@@
 While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis.
-An underlying assumption is therefore that the retained categories represent artefacts categories that have fulfilled different purposes or are related to different technological processes.
+An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
 While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.
 
-However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not manage to take into account [@dibble2017].
-This renders it difficult to align any of the results with specific behavioural dimensions.
+However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take into account, and renders it difficult to align any identified patterns with specific behavioural dimensions [@dibble2017].
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
@@
 @solheim2018].
-Thus, while these analyses can potentially get at in-group variation pertaining to the end state of artefacts that we assume can be treated in aggregate, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
+Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
 As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind.
-The logic behind these measures are founded in an understanding of lithic technology as being organised along a continuum ranging between curated and expedient [@binford1979].
-An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources for modification and rejuvination, resulting in high rates of replacement.
-Curated technological organisation, on the other hand, has been defined as the manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979, @bamforth1986].
-The empirical correspondent to a curated technological organisation is consequently high degrees of retouch, as this is commonly seen as a means of extending the use-life of lithics by repeated rejuvination and modification of edges.
+The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979].
+An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of replacement.
+Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].
+Although precisely how degrees of retouch is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, as this is commonly seen as a means of extending the use-life or realising the potential utility of a tool by repeated rejuvination and modification of edges [e.g.
+@bamforth1986; @dibble1995; @shott2005].
 
-The continuum between curated and expedient technology has in turn been related to settlement and mobility strategies, as well as raw-material quality and availability [@andrefsky1994; @smith2015].
+The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies [e.g.
+@parry1987; @clark2017], as well as raw-material quality and availability [e.g.
+@andrefsky1994; @smith2015].
 Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
 From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
-More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability, and allow for the accumulation of material resources.
-Higher raw-material availability should in turn reduce the necessity for extensive retouch, meaning the material expectation for a lower degree of mobility is relatively high density of lithics and a low relative frequency of retouched artefacts.
-Additional empirical expectation for
+More time spent in a single location, on the other hand, is in this framework assumed to lead to better control of raw-material availability, and to allow for the accumulation of material resources, which should in turn lead to an expedient technological organisation and reduce the necessity for extensive retouch.
+The material expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts.
+Additional empirical expectations following from this is a higher number of cores and unretouched flakes and blades [@clark2017].
 
-have shown through a series of studies that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent and comparable results across a wide range of chronological and cultural context, and is the main measure employed here.
-Some additional variables suggested by @Bicho2020 to be related to mobility patterns are are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
+While the concept of curation and its empirical and behavioural implications have been widely discussed [e.g.
+@bamforth1986, @shott1996] , following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested.
+Through a series of studies, have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent and comparable results across a wide range of chronological and cultural context which are taken to reflect .
+This is the main measure employed here.
+Some additional variables suggested by @Bicho2020 to be related to mobility patterns are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
 @baxter1994, 100].
@@
 A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here.
-This pertains to number of site features such as fireplaces, cooking pits and dwelling structures.
-The omission of feature count from the analysis follows from the fact that taphonomic loss is likely to lead to a chronological bias in the preservation and associated diffuclty of discovering of features.
+This pertains to number of site features such as fireplaces, cooking pits and dwelling structures which has been omitted as taphonomic loss is likely to lead to a chronological bias in their preservation.
 Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
@@
 
-The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as macro tools and associated debitage.
-It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, as opposed to the cores themselves.
+The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
+It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves.
 The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
@@
 Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
-This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the significance of the employed units of analysis unclear [@dibble2017].
+This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).
 

### 2021-04-20 analysis/paper/paper.Rmd
Implications of orthogonal relationship between WABI and core/chip rel freq

@@
 ```{r get-data, eval = TRUE, results = 'hide'}
+# gis <- read.csv(here::here('../../site_data.csv'))
+# gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
+
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
@@
 This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
-However, the approach taken is partially informed by the so-called Frison Effect, which highlights the fact that the shape of lithics studied by archaeologists is that which they had when they where discarded - the artefacts can have had long and complex use-lives in which they took on a multitude of different shapes.
+However, the approach taken is partially informed by the so-called Frison Effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes, and that the shapes studied by archaeologists are the shape they took upon discard.
 Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
-@barton1991; @dibble1995] .
+@barton1991; @dibble1995].
 Consequently, several artefact categories have here been collapased for the CA.
@@
 
-Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020.
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for settlement strategies as operationalised by @clark2017 and @Bicho2020.
 While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data.
@@
 
+The results of the CA align more or less with results of previous research that has considered the chronological .
+This would indicate that in general, meaningful patterning is associated with the employed artefact groups.
+
 The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking.
@@
 However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips?
-Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these.
+Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pairs is not correlated with these.
 As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.
-In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020.
-This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to the elements, and failed to identify any diachronic variation in this pattern.
-As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.
+In attempt to determine if this might be the case, it was decided to introduce site locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020.
+Of the variables considered, this study found that the most consistently important environmental covariates for the location of sites in the region was related to the exposure of the sites, and failed to identify any diachronic variation in this pattern.
+As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similarly temporary , one could envisage that variablity in site functionality might be reflected in degree of exposure.
+However, introducing these variables does not clarify the picture, as both variables show negligible degrees of correlation with any of the variables pertaining to the assemblage data.
 
@@
 These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at behavioural dimensions relating to mobility patterns among hunter-gatherers.
+
+# Conclusion
+
+The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway.
+This would indicate that in general, meaningful chronological patterning is associated with the employed artefact groups.
+Precisely what behavioural implication the development in the use of these types have, however, is unclear
+
+The results of considering variables that are have been associated with mobility patterns gave rise to two competing hypotheses.
+These pertain to whether the degree of curation associated with the assemblages as captured by the changing ratio of primary to secondarily worked lithics over time is the result of changing raw material availability or mobility patterns.
+The orthogonal relationship between core to chip frequency would in this framework be expected to be capturing the opposite relationship.
+If degree of curation mainly reflects changes in raw-material quality and availability, this would indicate that mobility patterns are comparable throughout the Mesolithic as the core to chip frequency is not time-dependent.
+While this would be in line with findings from a recent evaluation of the locational patterns associated with the sites [@roalkvam2020], these findings are the results of an exploratory analysis, and the results presented here are therefore perhaps best seen as potentially laying the framework for later devising a targeted strategic test of these competing hypotheses.
 

### 2021-04-22 analysis/paper/paper.Rmd
Minor adjustments to text

@@
 
-Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
+However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
 Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
@@
 To this end, the analysis follows two analytical avenues.
+The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports.
+The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.
 
@@
     artefact_count = sum(c_across(core:flake_retouch_other)),
-    flint_count = sum(c_across(core:flake_retouch))) %>% 
+    flint_count = sum(c_across(core:flake_retouch)),
+    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
   # Exclude sites with less than 100 artefacts
@@
 
-The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref@(fig:dist)A].
+The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref\@(fig:dist)A].
 The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
-The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects with assemblages holding more than 100 artefacts.
-The institution responsible for all of these excavations was the Museum of Cultural History in Oslo.
+The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts.
+The institution responsible for these excavations was the Museum of Cultural History in Oslo.
 This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations.
-Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the Museum of Cultural History.
-As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [@clark2006; @dibble2017].
-In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref@(fig:dist)B].
+Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum.
+As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g.
+@clark2006; @dibble2017].
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref\@(fig:dist)B].
 Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
@@
 In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014].
-Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.
+Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.
 
@@
 This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46].
-Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material accessibility and technological choices [@eigeland2015, 45-53].
-Furthermore, a general sentiment is that an increase in the use of non-flint materials signifies an increased adaption to local environments, and possibly a reduction in mobility.
-However, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material quality and accessibility as well as technological choices [@eigeland2015, 45-53].
+Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of adzes and other macro tools.
 Given this differentiated use, the retained debitage and tool categories are expected to still reflect the differentiated use associated with different raw-materials.
@@
 @frivoll2017].
+
+The date ranges provided for the sites are taken from the original excavation reports, and are the result of a consideration of typological indicators, dating with reference to shoreline displacement, and radiocarbon dates where these are available.
+Employing these date-ranges was deemed sufficient, as the main purpose is to get at general tendencies in the material, and not fine-grained chronological developments.
 
@@
 Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
-@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016; @solheim2020], or in-depth analyses of handful of sites.
-Some studies have attempted more comprehensive evaluations of entire lithic assemblages associated with larger groups of sites, mainly to get at potential site types and associated mobility patterns among Early Mesolithic sites [@breivik2016; @viken2018a; @breivik2020].
-However, only a single study has employed a multivariate quantitative framework in the treatment of the assemblages [cf.
-@naeroy2018, 905].
-In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to qualitatively and subjectively driven methods.
+@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g.
+@damlien2016; @solheim2020].
+Some studies have involved in-depth analyses of handful of sites, either in terms of intrasite spatial organisation [@naeroy2018], or in terms of technological organisation.
+There are also some comprehensive evaluations of entire lithic assemblages associated with a larger group of sites to get at potential site types and associated mobility patterns [@breivik2016; @viken2018c; @breivik2020].
+However, only a single study concerned with eight Middle-Mesolithic assemblages has employed a multivariate quantitative framework to help structure such a treatment [@naeroy2018, 905].
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively driven methods.
 
@@
 
-The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
 The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods.
-This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
-However, the approach taken is partially informed by the so-called Frison Effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes, and that the shapes studied by archaeologists are the shape they took upon discard.
+This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
+However, the approach taken is also partially informed by the so-called Frison effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
 Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
@@
 Consequently, several artefact categories have here been collapased for the CA.
-This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch.
-That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose, is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially entirely misleading.
+This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
+That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially misleading.
 These have all been combined into the single category "small flint tools".
 (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
-While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis.
+While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
 An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
@@
 
-However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take into account, and renders it difficult to align any identified patterns with specific behavioural dimensions [@dibble2017].
+However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take properly into account, and renders it difficult to align any identified patterns with specific behavioural dimensions [@dibble2017].
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
@@
 Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].
-Although precisely how degrees of retouch is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, as this is commonly seen as a means of extending the use-life or realising the potential utility of a tool by repeated rejuvination and modification of edges [e.g.
+Although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, this is commonly seen as a means of extending the use-life, or realising the potential utility of a tool by repeated rejuvination and modification of edges [e.g.
 @bamforth1986; @dibble1995; @shott2005].
+The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies, as well as raw-material quality and availability [e.g.
+@parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].
 
-The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies [e.g.
-@parry1987; @clark2017], as well as raw-material quality and availability [e.g.
-@andrefsky1994; @smith2015].
+While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g.
+@bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested.
+Through a series of studies @barton1999 and collegues [e.g.
+@barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context.
+This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014].
 Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
-From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
-More time spent in a single location, on the other hand, is in this framework assumed to lead to better control of raw-material availability, and to allow for the accumulation of material resources, which should in turn lead to an expedient technological organisation and reduce the necessity for extensive retouch.
-The material expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts.
-Additional empirical expectations following from this is a higher number of cores and unretouched flakes and blades [@clark2017].
+From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
+More time spent in a single location, on the other hand, is in this framework assumed to lead to better control of raw-material availability and to allow for its accumulation.
+This should in turn lead to an expedient technological organisation with reduced necessity for conservation of lithics and extensive use of retouch.
+The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts.
+From this is a higher number of cores and unretouched flakes and blades [@clark2017].
 
-While the concept of curation and its empirical and behavioural implications have been widely discussed [e.g.
-@bamforth1986, @shott1996] , following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested.
-Through a series of studies, have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent and comparable results across a wide range of chronological and cultural context which are taken to reflect .
 This is the main measure employed here.
@@
 
-In sum, the results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time.
-However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
-This is evident in the present data as well, highlighted by the influence of the later Mesolithic axe production sites, which it would be reasonable to assume corresponds to a more pervasive cultural change than a purely typo-chronological development of artefact shape.
-Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
-This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).
-
 In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
@@
 # but not: mutate(new_varaible = sum(vector_of_variable_names) / artefact_count)
-primary_flint <- c("core", "core_fragment", "blade", "microblade", "chip",
-                     "fragment", "flake")
+blanks_flint <- c("blade", "microblade", "fragment", "flake")
 
-primary_lithics <- c(primary_flint, "flake_other", "fragment_other", 
-                     "chip_other", "blade_other",  "microblade_other", 
-                     "core_other", "core_fragment_other")
+blanks_lithics <- c(blanks_flint, "flake_other", "fragment_other", 
+                     "blade_other",  "microblade_other")
 
@@
          core_freq = sum(core, core_other) / artefact_count,
-         primary_freq = sum(core, core_fragment, blade, 
-                     microblade, chip, fragment, flake, flake_other, 
-                     fragment_other, chip_other, blade_other,  microblade_other, 
-                     core_other, core_fragment_other) / artefact_count,
+         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
+                     fragment_other, blade_other,  microblade_other) / artefact_count,
          secondary_freq =  sum(axe_debitage, axe, chisel, 
@@
                 flake_retouch_other) / artefact_count,
+         nonflint_freq = nonflint_count / artefact_count,
          lithic_dens = artefact_count / excavated_m3) %>%
-  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
-         -flint_count) %>% 
+  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
+         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
+         -chip_other, -core_other, -core_fragment_other) %>% 
   column_to_rownames("site")
@@
          core_freq = sum(core) / artefact_count,
-         primary_freq = sum(core, core_fragment, blade, 
-                     microblade, chip, fragment, flake) / artefact_count,
+         blanks_freq = sum( blade, microblade, fragment, flake) / artefact_count,
          secondary_freq =  sum(axe_debitage, axe, chisel, 
@@
          flint_dens = flint_count / excavated_m3) %>%
-  select(-primary_flint, -secondary_flint, -excavated_m3, -artefact_count,
-         -flint_count) %>% 
+  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
+         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
+         -core_fragment, -chip) %>% 
   column_to_rownames("site")
@@
 
-Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for settlement strategies as operationalised by @clark2017 and @Bicho2020.
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020.
 While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data.
@@
 
-The results of the CA align more or less with results of previous research that has considered the chronological .
-This would indicate that in general, meaningful patterning is associated with the employed artefact groups.
+The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time.
+However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
+This is evident in the present data as well, highlighted by the influence of the later Mesolithic axe production sites, which it would be reasonable to assume corresponds to a more pervasive cultural change than a purely typo-chronological development of artefact morphology.
+Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
+This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).
 
 The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking.
-While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for larger scale comparative studies.
+While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
 The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.
 However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips?
-Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pairs is not correlated with these.
+Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pair is not correlated with these.
 As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.
@@
 The orthogonal relationship between core to chip frequency would in this framework be expected to be capturing the opposite relationship.
-If degree of curation mainly reflects changes in raw-material quality and availability, this would indicate that mobility patterns are comparable throughout the Mesolithic as the core to chip frequency is not time-dependent.
-While this would be in line with findings from a recent evaluation of the locational patterns associated with the sites [@roalkvam2020], these findings are the results of an exploratory analysis, and the results presented here are therefore perhaps best seen as potentially laying the framework for later devising a targeted strategic test of these competing hypotheses.
+If degree of curation mainly reflects changes in raw-material quality and availability, this would then indicate that mobility patterns are comparable throughout the Mesolithic as the core to chip frequency is not time-dependent.
+And vice verca --- if the changing mobility patterns is the main driver of the transition from curated to expedient technological organisation, as operationalised here, While this would be in line with findings from a recent evaluation of the locational patterns associated with the sites [@roalkvam2020], there is no readily available empirical evidence, no smoking gun, se findings are the results of an exploratory analysis, and are therefore perhaps best seen as potentially laying the framework for later targeted strategic test of these competing hypotheses.
 

### 2021-04-25 analysis/paper/paper.Rmd
Exploring the data some more - setting up for correlation plots.

@@
 library(rnaturalearth)
+```
+```{r get-functions, include = FALSE}
+#source("plot_functions.R", local = knitr::knit_global())
 ```
@@
 
+lithics3 <- sites %>% 
+  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
+         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
+  # The reason for not using .keep below is for the option to 
+  #retain artefact_count
+  mutate(chip_freq =  sum(chip, chip_other),
+         core_freq = sum(core, core_other),
+         blanks = sum(blade, microblade, fragment, flake, flake_other, 
+                     fragment_other, blade_other,  microblade_other, -core, -core_fragment, -chip, 
+         -chip_other, -core_other, -core_fragment_other),
+         secondary =  sum(axe_debitage, axe, chisel, 
+                microlith, tanged_point, transverse_point, single_edged_point,
+                hognipen_point, projectile_undefined, microburin, scraper, 
+                knife, burin, drill, blade_retouch, microblade_retouch, 
+                chip_retouch, fragment_retouch, flake_retouch, axe_other,
+                axe_debitage_other, chisel_other, club_hatchet_other, 
+                grinding_slab_stone, projectile_point_other, knife_other,
+                scraper_other, drill_other, microblade_retouch_other, 
+                blade_retouch_other, fragment_retouch_other,
+                flake_retouch_other),
+         lithic_dens = artefact_count / excavated_m3) %>%
+  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
+         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
+         -chip_other, -core_other, -core_fragment_other) %>% 
+  column_to_rownames("site")
+
 # Repeat but for flint only
@@
          core_freq = sum(core) / artefact_count,
-         blanks_freq = sum( blade, microblade, fragment, flake) / artefact_count,
+         blanks_freq = sum(blade, microblade, fragment, flake) / artefact_count,
          secondary_freq =  sum(axe_debitage, axe, chisel, 
@@
 
-wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
+wabicor <- cor.test(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
 wabicorflint <- cor(log10(flint2$flint_dens), log10(flint2$secondary_freq))
@@
 ```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
-pcalithics <- PCA(select(lithics2, -mean_date), scale.unit = TRUE, graph = FALSE)
+pcalithics <- PCA(select(lithics2, -mean_date, -nonflint_freq), scale.unit = TRUE, graph = FALSE)
 fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithics2$mean_date / 1000, 

### 2021-04-29 analysis/paper/paper.Rmd
Correlation plot and transformation of variables for PCA

@@
 library(tidyverse)
+library(GGally)
 library(patchwork)
@@
 ```
+
 ```{r get-functions, include = FALSE}
-#source("plot_functions.R", local = knitr::knit_global())
+source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
 ```
@@
 site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
-countries <- ne_download(scale = 10, type = 'countries', category = 'cultural',
-                         returnclass = "sf")
+
+# ne_download() is currently failing
+#countries <- ne_download(scale = 10, type = 'countries', category = 'physical',
+#                         returnclass = "sf")
+countries <- st_read(here::here('analysis/data/raw_data/naturalearth_countries.gpkg'))
 ```
@@
 The analyses conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
-This is done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
+This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
 However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g.
-@dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency, and group sizes at visits to the sites.
-Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
+@dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites.
+The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials and seasonality.
 Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
@@
 
-However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
+Even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
 Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
 The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
-As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway that have been based on more informally driven methods, and have the generation of new hypotheses as a possible outcome.
+As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway, that are typically based on more informally driven methods, and have the generation of new hypotheses as a possible outcome.
 To this end, the analysis follows two analytical avenues.
@@
 A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas.
-Furthermore, these coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
+Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
@@
 In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
-This was combined with the blade category here.
-Furthermore, the artefact data has been divided into flint and non-flint materials.
-Flint does not occur indigenously in Norway, and is only available as nodules that have been transported and deposited by retreating and drifting ice [e.g.
+Narrow-blades were combined with the blade category here.
+Furthermore, the artefact data have here been divided into flint and non-flint materials.
+Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g.
 @berg-hansen1999].
 This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46].
-Consequently, although flint is treated as a unified category here, there is reason to believe that this variability has had consequences for raw-material quality and accessibility as well as technological choices [@eigeland2015, 45-53].
-Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of adzes and other macro tools.
-Given this differentiated use, the retained debitage and tool categories are expected to still reflect the differentiated use associated with different raw-materials.
-An important benefit of combining all non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised [cf.
+An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites.
+However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here [@eigeland2015, 45-53].
+Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories.
+An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf.
 @frivoll2017].
+While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [@eigeland2015, 370].
 
@@
 @damlien2016; @solheim2020].
-Some studies have involved in-depth analyses of handful of sites, either in terms of intrasite spatial organisation [@naeroy2018], or in terms of technological organisation.
+Some studies have involved in-depth analyses of handful of sites, either in terms of intrasite spatial organisation [@nærøy2018], or in terms of technological organisation.
 There are also some comprehensive evaluations of entire lithic assemblages associated with a larger group of sites to get at potential site types and associated mobility patterns [@breivik2016; @viken2018c; @breivik2020].
-However, only a single study concerned with eight Middle-Mesolithic assemblages has employed a multivariate quantitative framework to help structure such a treatment [@naeroy2018, 905].
+However, the subjeticely driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of site types.
+Each site could have had complex life histories and been associated with a wide range of tasks that in sum make up the assemblages.
+Finally, only a single study concerned with eight Middle-Mesolithic assemblages that are also treated here has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013].
 In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively driven methods.
@@
 There have been a few studies that have conducted comprehensive analyses of artefact inventories and site features to get at variation in Mesolithic site types [@viken2018; @breivik2020].
-However, the narratively driven analyses of these data makes the weighting of different variables unclear and subjectively defined, and several of the studies cited here ultimately draw their conclusions based on an arguably over-constraining trichotomy of site types.
 As Binford originally clarified, a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.
@@
 This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
-That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially misleading.
+That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading.
 These have all been combined into the single category "small flint tools".
@@
 Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
-As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind.
+As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind [see @clark2017, and below].
 The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979].
-An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of replacement.
+An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement.
 Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].
-Although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, this is commonly seen as a means of extending the use-life, or realising the potential utility of a tool by repeated rejuvination and modification of edges [e.g.
+Although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool or extending its use-life by the repeated rejuvination and modification of edges [e.g.
 @bamforth1986; @dibble1995; @shott2005].
@@
 From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
-More time spent in a single location, on the other hand, is in this framework assumed to lead to better control of raw-material availability and to allow for its accumulation.
-This should in turn lead to an expedient technological organisation with reduced necessity for conservation of lithics and extensive use of retouch.
+More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation.
+This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
 The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts.
-From this is a higher number of cores and unretouched flakes and blades [@clark2017].
+A further empirical expectation of an expendient tehcnological organisation is a higher number of cores and unretouched flakes and blades [@clark2017].
 
@@
 A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here.
-This pertains to number of site features such as fireplaces, cooking pits and dwelling structures which has been omitted as taphonomic loss is likely to lead to a chronological bias in their preservation.
+This pertains to number of site features such as fireplaces, cooking pits and dwelling structures.
+Feature count has been omitted as taphonomic loss is likely to lead to a chronological bias in their preservation.
 Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
@@
 calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count, 
-                       -reported_start_bp, -reported_end_bp, -askeladden_id),
+                       -nonflint_count, -reported_start_bp, -reported_end_bp,
+                       -askeladden_id),
                 graph = FALSE)
@@
 
-wabicor <- cor.test(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
+wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
 wabicorflint <- cor(log10(flint2$flint_dens), log10(flint2$secondary_freq))
@@
 
-```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
-pcalithics <- PCA(select(lithics2, -mean_date, -nonflint_freq), scale.unit = TRUE, graph = FALSE)
-fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithics2$mean_date / 1000, 
+```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
+# Transform the skewed variables - sqrt() is used where there are zero values,
+# and seems to provide an adequate result.
+lithicstr <- lithics2 %>%
+  # Transform skewed variables and reorder
+  mutate("Mean date cal. BP*-1" = mean_date * -1,
+          "log10(lithic_dens)" = log10(lithic_dens),
+         "log10(secondary_freq)" = log10(secondary_freq),
+         "core_freq^(1/3)" = core_freq^(1/3),
+         "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
+  relocate("Mean date cal. BP*-1",
+    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
+    "chip_freq", "core_freq^(1/3)" 
+  )
+
+corr_plot(lithicstr)
+```
+
+```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 7, fig.width = 10}
+# Rename the transformed variables again to avoid clutter in the plot. 
+# Specify in figure caption that the transformed variables are used 
+lithicspca <- lithicstr %>% 
+  mutate(mean_date =  `Mean date cal. BP*-1` * -1, .keep = "unused") %>% 
+  rename(
+   lithic_dens = "log10(lithic_dens)",
+   secondary_freq = "log10(secondary_freq)",
+   core_freq = "core_freq^(1/3)",
+   nonflint_freq = "nonflint_freq^(1/3)"
+  )
+
+pcalithics <- PCA(select(lithicspca, -mean_date),
+                  scale.unit = TRUE, graph = FALSE)
+fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithicspca$mean_date / 1000, 
           repel = TRUE, col.var = "black", pointshape = 21,  pointsize = 2.9,
@@
 hese last dimensions are capturing the most variation in the dataset presented here.
-Thus, if frequency of secondarily worked artefacts is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+Thus, if frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+Furthermore, the fact that these
 
@@
 
-```{r var_contrib, fig.cap = "Contribution of variables to components."}
+```{r var_contrib, fig.cap = "Contribution of variables to components.", fig.height = 3}
 # The plots below investigates the contribution of the variables to the
@@
 # Patchwork for arranging plots
-(dim1 | dim2) /
-(dim3 | dim4)
+(dim1 | dim2) 
+#(dim3 | dim4)
 ```
 
-```{r var_cor, fig.cap = "Plots showing the correlation between the variables"}
+```{r var_cor, fig.cap = "Plots showing the correlation between the variables", eval = FALSE}
 plt1 <- fviz_pca_var(pcalithics,

### 2021-04-29 analysis/script/plot_functions.R

@@ -1,0 +1,59 @@
+##### Functions used for correlation plot #####
+
+# Histogram function to show univariate distribution
+hist_func <- function(data, mapping, ...) {
+  ggplot(data = data, mapping = mapping) +
+    geom_histogram() +
+    theme(panel.background = element_rect(fill = "white", colour = "black"))
+}
+
+# Function to retrieve correlation
+# and color according to pearson's r
+cor_func <- function(data, mapping, ...){
+
+  # Retrieve data
+  x <- eval_data_col(data, mapping$x)
+  y <- eval_data_col(data, mapping$y)
+
+  # calculate correlation
+  corr <- cor(x, y, method = 'pearson', use = 'complete.obs')
+  pearsr <-  paste('r =' , as.character(round(corr, 2)))
+
+  colFn <- colorRampPalette(c("coral1", "white", "palegreen3"),
+                            interpolate ='spline')
+  fill <- colFn(100)[findInterval(corr, seq(-1, 1, length=100))]
+
+  ggally_text(
+    label = pearsr,
+    mapping = aes(),
+    xP = 0.5, yP = 0.5,
+    color = 'black',
+    ...) +
+    theme(panel.background = element_rect(fill = fill, colour = "black"))
+}
+
+# Function to plot bivariate distributions with fitted OLS regression
+smooth_func <- function(data, mapping, ...){
+
+  x <- eval_data_col(data, mapping$x)
+  y <- eval_data_col(data, mapping$y)
+
+  ggplot(data = data, aes (x = x, y = y))+
+    geom_point(shape = 16, colour = 'black') +
+    geom_smooth(method = 'lm', se = FALSE, colour = 'red') +
+    theme(panel.background = element_blank(),
+          panel.border = element_rect(fill = NA, colour = 'black'))
+}
+
+# Function to combine the above functions into a correlation plot
+corr_plot <- function(data, title = NULL, ...){
+
+  ggpairs(data, columns = 1: ncol(data), title = title, switch = 'y',
+          upper = list(continuous = wrap(cor_func)),
+          lower = list(continuous = smooth_func),
+          diag = list(continuous = hist_func),
+          axisLabels = "none") +
+    theme(strip.background = element_rect(fill = "white", colour = "black"),
+          panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank())
+}

### 2021-04-29 analysis/paper/paper.Rmd
Minor text edits

@@
 ```{r get-data, eval = TRUE, results = 'hide'}
-# gis <- read.csv(here::here('../../site_data.csv'))
-# gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
-
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
@@
 countries <- st_read(here::here('analysis/data/raw_data/naturalearth_countries.gpkg'))
+
+gis <- read.csv(here::here('../../site_data.csv'))
+gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
+gisjoin <- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
 ```
@@
 @damlien2016; @solheim2020].
-Some studies have involved in-depth analyses of handful of sites, either in terms of intrasite spatial organisation [@nærøy2018], or in terms of technological organisation.
-There are also some comprehensive evaluations of entire lithic assemblages associated with a larger group of sites to get at potential site types and associated mobility patterns [@breivik2016; @viken2018c; @breivik2020].
-However, the subjeticely driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of site types.
+Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;].
+There are also some more comprehensive evaluations of a larger number of entire lithic assemblages to get at potential site types and associated mobility patterns [@breivik2016; @viken2018c; @breivik2020].
+However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of site types.
 Each site could have had complex life histories and been associated with a wide range of tasks that in sum make up the assemblages.
-Finally, only a single study concerned with eight Middle-Mesolithic assemblages that are also treated here has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013].
-In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively driven methods.
+Finally, only a single study concerned with eight Middle-Mesolithic assemblages --- assemblages that are also treated here --- has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013].
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods.
 
@@
 
-lithics3 <- sites %>% 
+lithics3 <- gisjoin %>% 
   select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
@@
   #retain artefact_count
-  mutate(chip_freq =  sum(chip, chip_other),
-         core_freq = sum(core, core_other),
-         blanks = sum(blade, microblade, fragment, flake, flake_other, 
-                     fragment_other, blade_other,  microblade_other, -core, -core_fragment, -chip, 
-         -chip_other, -core_other, -core_fragment_other),
-         secondary =  sum(axe_debitage, axe, chisel, 
+  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
+         core_freq = sum(core, core_other) / artefact_count,
+         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
+                     fragment_other, blade_other,  microblade_other) / artefact_count,
+         secondary_freq =  sum(axe_debitage, axe, chisel, 
                 microlith, tanged_point, transverse_point, single_edged_point,
@@
                 blade_retouch_other, fragment_retouch_other,
-                flake_retouch_other),
-         lithic_dens = artefact_count / excavated_m3) %>%
+                flake_retouch_other) / artefact_count,
+         nonflint_freq = nonflint_count / artefact_count,
+         lithic_dens = artefact_count / excavated_m3,
+         viewsize_5k = log10(viewsize_5k),
+         viewsize_10k = log10(viewsize_10k),
+         fetch = log10(avg_fetch)) %>%
   select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
          -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
-         -chip_other, -core_other, -core_fragment_other) %>% 
+         -chip_other, -core_other, -core_fragment_other, -cat, -municipality, -quality,
+         -elev_minimum, -site_size, -cent_elev, -phase, -slope_average, -aspect_average,
+         -infiltration_mode,-emergence_50, -emergence_500, -emergence_1k, -location,
+         -island_size) %>% 
   column_to_rownames("site")
@@
 
-The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking.
-While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
+these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
 The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.
@@
 The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway.
-This would indicate that in general, meaningful chronological patterning is associated with the employed artefact groups.
-Precisely what behavioural implication the development in the use of these types have, however, is unclear
+This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
+These tendencies are already well-established when it comes to formal tool types, but the have been given less focus in terms of entire assemblages these notions have previously Precisely what behavioural implication the development in the use of these types have, however, is unclear
 

### 2021-05-03 analysis/paper/paper.Rmd
Introducing oxcal and more proper treatment of dates

@@
 library(rnaturalearth)
+library(oxcAAR)
+quickSetupOxcal()
 ```
@@
 source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
+
+#setOxcalExecutablePath("/tmp/RtmpFVOcBJ")
 ```
@@
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
+radiocarbon <- read.csv(here::here('analysis/data/raw_data/radiocarbon_data.csv'))
 site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
@@
 
-gis <- read.csv(here::here('../../site_data.csv'))
-gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
-gisjoin <- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
+# gis <- read.csv(here::here('../../site_data.csv'))
+# gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
+# gisjoin <- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
 ```
-
-<!-- Here's some example analysis code: -->
 
@@
 sites <- assemblages %>% 
-      # Find mean age for each site (to be used in the plots below)
+      # Find mean reported date for each site (to be used in the plots below)
   rowwise() %>% 
@@
   filter(artefact_count > 100)
+
+h3 <- radiocarbon %>% 
+  filter(site == "Hovland 3")
+
+h3dat <- R_Date(h3$lab_code, h3$c14_bp, h3$error)
+h3_sum <- oxcal_Sum(h3dat)
+h3_run <- executeOxcalScript(h3_sum)
+h3_text <- readOxcalOutput(h3_run)
+h3_result <- parseFullOxcalOutput(h3_text)
+
+oxcal_code <- paste(' Plot()
+   {
+    Sequence()
+    {
+     Boundary("Start");
+     Phase()
+     {',
+      h3dat,
+     '};
+     Boundary("End");
+    };
+  };')
+my_result_file <- executeOxcalScript(oxcal_code)
+my_result_text <- readOxcalOutput(my_result_file)
+my_result_data <- parseOxcalOutput(my_result_text)
+
+my_oxcal_code <- ' Plot()
+ {
+  Sequence("Sequence1")
+  {
+   Boundary("Beginn");
+   Phase("Phase1")
+   {
+    R_Date("Lab-1",5000,25);
+    R_Date("Lab-2",4900,37);
+   };
+   Boundary("Between");
+   Phase("Phase2")
+   {
+    R_Date("Lab-3",4800,43);
+   };
+   Boundary("End");
+  };
+ };'
+
 ```
@@
 
-A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas.
+A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@glørstad2010].
 Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
@@
 The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods.
+The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods when the entirety of the assemblages are taken into account.
 This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
-However, the approach taken is also partially informed by the so-called Frison effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
-Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
+However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
+Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
 @barton1991; @dibble1995].
@@
 @bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested.
-Through a series of studies @barton1999 and collegues [e.g.
-@barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context.
+Through a series of studies Barton and collegues [e.g.
+-@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context.
 This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014].
@@
   # Deselect a few of the columns not be used
-  select(-one_of(c("n", "excavation_project", "excavated_m3")))
+  select(-n, -excavation_project, -excavated_m3, -dating_method)
 ```
@@
 The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+These findings are in large part in line with previous research [e.g.
+@bang-andersen2003], although possibly highlighting the importance of the non-flint materials to a larger extent than what has traditionally been the case.
 
@@
 This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
-@eigeland2015, 370].
+@eigeland2015, 370, @glørstad2011].
 
@@
          -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
-         -chip_other, -core_other, -core_fragment_other) %>% 
-  column_to_rownames("site")
-
-lithics3 <- gisjoin %>% 
-  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
-         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
-  # The reason for not using .keep below is for the option to 
-  #retain artefact_count
-  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
-         core_freq = sum(core, core_other) / artefact_count,
-         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
-                     fragment_other, blade_other,  microblade_other) / artefact_count,
-         secondary_freq =  sum(axe_debitage, axe, chisel, 
-                microlith, tanged_point, transverse_point, single_edged_point,
-                hognipen_point, projectile_undefined, microburin, scraper, 
-                knife, burin, drill, blade_retouch, microblade_retouch, 
-                chip_retouch, fragment_retouch, flake_retouch, axe_other,
-                axe_debitage_other, chisel_other, club_hatchet_other, 
-                grinding_slab_stone, projectile_point_other, knife_other,
-                scraper_other, drill_other, microblade_retouch_other, 
-                blade_retouch_other, fragment_retouch_other,
-                flake_retouch_other) / artefact_count,
-         nonflint_freq = nonflint_count / artefact_count,
-         lithic_dens = artefact_count / excavated_m3,
-         viewsize_5k = log10(viewsize_5k),
-         viewsize_10k = log10(viewsize_10k),
-         fetch = log10(avg_fetch)) %>%
-  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
-         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
-         -chip_other, -core_other, -core_fragment_other, -cat, -municipality, -quality,
-         -elev_minimum, -site_size, -cent_elev, -phase, -slope_average, -aspect_average,
-         -infiltration_mode,-emergence_50, -emergence_500, -emergence_1k, -location,
-         -island_size) %>% 
+         -chip_other, -core_other, -core_fragment_other, -dating_method) %>% 
   column_to_rownames("site")
@@
          -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
-         -core_fragment, -chip) %>% 
+         -core_fragment, -chip, dating_method) %>% 
   column_to_rownames("site")
@@
 wabicorflint <- cor(log10(flint2$flint_dens), log10(flint2$secondary_freq))
-```
-
-Figure \@ref(fig:wabi) displays the WABI as found for the site data.
-This indicates a negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
-Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.
-Variable raw-material availability can also impact these measures as [@manninen2014].
-Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies.
-However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically.
-Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly chronologically, the general pattern is the same.
-
-```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint.
-The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
-w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
-                     fill = mean_date / 1000)) + 
-  geom_point(shape = 21, size = 3) +
-  geom_smooth(method = "lm", col = "black", se = FALSE) +
-  labs(title = "A",
-       x = "Volumetric density of lithics (log10)", 
-       y = "Relative frequency of secondarily worked lithics (log10)") +
-  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
-                        high="red", midpoint = centdate) +
-  scale_y_log10() +
-  scale_x_log10() +
-  theme_classic()
-
-w2 <- ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
-                     fill = mean_date / 1000)) + 
-  geom_point(shape = 21, size = 3) +
-  geom_smooth(method = "lm", col = "black", se = FALSE) +
-  labs(title = "B", 
-       x = "Volumetric density of flint (log10)", 
-       y = "Relative frequency of secondarily worked flint (log10)") +
-  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
-                        high="red", midpoint = centdate) +
-  scale_y_log10() +
-  scale_x_log10() +
-  theme_classic()
-
-w1 | w2
 ```
@@
 
+To visualise this more clearly, a curation index based on the two variables was devised by perfoming a min-max normalisation on the two variables, scaling them to take on values between 0 and 1.
+The values for artefact density was then made negative to reflect its relationship with degree of curation.
+The mean was then found for each site on these two normalised values.
+Plotting the results indicates a higher degree of curation among the earliest sites, which then falls and settles without any further major fluctuations.
+
 Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020.
@@
 ```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
-# Scree plot indicates an 'elbow' at the second dimension. 
-# The two first dimension account for 68 % and the first three
-# for 87 % (see table below)
-# fviz_screeplot(pcalithics, addlabels=TRUE)
-
 # Create a table showing the eigenvalues of the components
@@
 
+Figure \@ref(fig:wabi) displays the WABI as found for the site data.
+This indicates a negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.
+Variable raw-material availability can also impact these measures as [@manninen2014].
+Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies.
+However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically.
+Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly chronologically, the general pattern is the same.
+
+```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint.
+The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
+w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
+                     fill = mean_date / 1000)) + 
+  geom_point(shape = 21, size = 3) +
+  geom_smooth(method = "lm", col = "black", se = FALSE) +
+  labs(title = "A",
+       x = "Volumetric density of lithics (log10)", 
+       y = "Relative frequency of secondarily worked lithics (log10)") +
+  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  scale_y_log10() +
+  scale_x_log10() +
+  theme_classic()
+
+w2 <- ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
+                     fill = mean_date / 1000)) + 
+  geom_point(shape = 21, size = 3) +
+  geom_smooth(method = "lm", col = "black", se = FALSE) +
+  labs(title = "B", 
+       x = "Volumetric density of flint (log10)", 
+       y = "Relative frequency of secondarily worked flint (log10)") +
+  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  scale_y_log10() +
+  scale_x_log10() +
+  theme_classic()
+
+w1 | w2
+```
+
+```{r index}
+# Create curation index
+minmax <- lithics2 %>% 
+  mutate(dens =  ((lithic_dens - min(lithic_dens)) / (max(lithic_dens) - min(lithic_dens))),
+         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))
+
+indexdata <- minmax %>% 
+  rowwise() %>% 
+  mutate(index = mean(c(dens * -1, secondary)))
+
+w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
+  geom_point() +
+  labs(title = "C",
+       x = "Mean date cal. BP",
+       y = "Curation index") +
+  scale_x_reverse() + 
+  geom_smooth(method = "loess", col = "red") +
+  theme_classic() 
+w3
+```
 # Discussion
@@
 However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
-This is evident in the present data as well, highlighted by the influence of the later Mesolithic axe production sites, which it would be reasonable to assume corresponds to a more pervasive cultural change than a purely typo-chronological development of artefact morphology.
+The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites.
 Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.

### 2021-05-04 analysis/paper/paper.Rmd
Working on implementing oxcaar for calibration and modelling of radiocarbon dates

@@
 h3 <- radiocarbon %>% 
-  filter(site == "Hovland 3")
+  filter(site == "Vallermyrene 1")
 
 h3dat <- R_Date(h3$lab_code, h3$c14_bp, h3$error)
-h3_sum <- oxcal_Sum(h3dat)
-h3_run <- executeOxcalScript(h3_sum)
-h3_text <- readOxcalOutput(h3_run)
-h3_result <- parseFullOxcalOutput(h3_text)
 
-oxcal_code <- paste(' Plot()
+oxcal_code <- paste0('Plot()
    {
@@
      Phase()
-     {',
+     {
+    ',
       h3dat,
-     '};
+     '
+     Sum("Vallermyrene 1");
+     };
      Boundary("End");
@@
 my_result_text <- readOxcalOutput(my_result_file)
-my_result_data <- parseOxcalOutput(my_result_text)
+my_result_data <- parseOxcalOutput(my_result_text, only.R_Date = FALSE)
+plot(my_result_data$` Vallermyrene 1 `$posterior_probabilities$dates, my_result_data$` Vallermyrene 1 `$posterior_probabilities$probabilities, type = "l")
 
-my_oxcal_code <- ' Plot()
- {
-  Sequence("Sequence1")
-  {
-   Boundary("Beginn");
-   Phase("Phase1")
-   {
-    R_Date("Lab-1",5000,25);
-    R_Date("Lab-2",4900,37);
-   };
-   Boundary("Between");
-   Phase("Phase2")
-   {
-    R_Date("Lab-3",4800,43);
-   };
-   Boundary("End");
-  };
- };'
+pst <- get_posterior_probabilities(my_result_data)
+# h3_sum <- oxcal_Sum(h3dat)
+# h3_run <- executeOxcalScript(h3_sum)
+# h3_text <- readOxcalOutput(h3_run)
+# h3_result <- parseFullOxcalOutput(h3_text)
 

### 2021-05-05 analysis/paper/paper.Rmd
Set up the function for modelling radiocarbon dates and summing the posterior distributions. Struggling with ordering the dates properly for the plot

@@
 The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
-As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway, that are typically based on more informally driven methods, and have the generation of new hypotheses as a possible outcome.
+As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome.
 To this end, the analysis follows two analytical avenues.
@@
   rowwise() %>% 
-  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp))) %>% 
+  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
   mutate(
@@
   filter(artefact_count > 100)
+```
 
-h3 <- radiocarbon %>% 
-  filter(site == "Vallermyrene 1")
+```{r c14-modelling}
+# define function to model radiocarbon dates. Only assumption is that 
+# dates from a single site belong to a single group
+model_rcarb <- function(data){
 
-h3dat <- R_Date(h3$lab_code, h3$c14_bp, h3$error)
+  # Format individual dates for R_Date in Oxcal
+  dates <- R_Date(data[["lab_code"]], data[["c14_bp"]], data[["error"]])
 
-oxcal_code <- paste0('Plot()
+  # Manually compile rest of OxCal code
+  oxcal_code <- paste0('
+  Plot()
    {
@@
     ',
-      h3dat,
+     dates,
      '
-     Sum("Vallermyrene 1");
+     Sum("Sum");
      };
@@
   };')
+
+  # Run OxCal code - wrapped this in some functions to  not print to 
+  # rmarkdown
+  invisible(capture.output(oxcal_exec <- executeOxcalScript(oxcal_code)))
+  oxcal_read <- readOxcalOutput(oxcal_exec)
+  # Note that parseOxcalOutput() defaults to only returning individual dates
+  oxcal_data <- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)
+
+  # Assemble data frame to be returned
+  posterior <- data.frame(
+    probabilities = oxcal_data$Sum$posterior_probabilities$probabilities,
+    date_grid = oxcal_data$Sum$posterior_probabilities$dates,
+    site = data[["site"]][1],
+    excavation_project = data[["excavation_project"]][1])
+
+  # Return data
+  return(posterior)
+}
+
+# Apply modelling function to the radiocarbon dates, grouped by site name
+posteriors <- radiocarbon %>% 
+  group_by(site) %>%
+  group_map(~ model_rcarb(.), .keep = TRUE)
+
+# The data is returned as a list of lists that needs to be unpacked
+rcarb_data <- do.call(rbind, posteriors) 
+  group_by(site) %>%
+  mutate(meandate = mean(date_grid * -1)) %>% 
+  ungroup() %>% 
+  arrange(meandate)
+
+h3 <- radiocarbon %>% 
+  filter(site == "Vallermyrene 1" | site == "Stokke/Polland 5")
+
+v1 <- model_rcarb(h3)
+
+h3dat <- R_Date(h3$lab_code, h3$c14_bp, h3$error)
+
 my_result_file <- executeOxcalScript(oxcal_code)
@@
 my_result_data <- parseOxcalOutput(my_result_text, only.R_Date = FALSE)
-plot(my_result_data$` Vallermyrene 1 `$posterior_probabilities$dates, my_result_data$` Vallermyrene 1 `$posterior_probabilities$probabilities, type = "l")
+plot(my_result_data$`Vallermyrene 1`$posterior_probabilities$dates, my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities, type = "l")
+
+posterior <- data.frame(probabilities = my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities,
+                        date_grid = my_result_data$`Vallermyrene 1`$posterior_probabilities$dates,
+                        site = "Vallermyrene 1",
+                        excavation_project = "Vestfoldbaneprosjektet")
+
+ggplot() +
+  geom_line(aes(x = yrs, y = my_result_data$` Vallermyrene 1 `$posterior_probabilities$probabilities))
+
+yrs <- seq(11500, 5000, -500)
+plt1 <- ggplot(sites, aes(x = yrs)) +
+  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
+                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
+                   colour = excavation_project), size = 1.7) +
+  xlab("Cal. BC") +
+  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
+                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
+                     values = c("springgreen4", "royalblue2",
+                                "gold1", "firebrick3")) +
+  labs(title = "B", colour = "Excavation project") +
+  theme_classic() +
+  theme(axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank(),
+        legend.position = "bottom") +
+  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
+    scale_x_reverse()
+
+ggplot(sites, aes(x = yrs)) +
+      ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
+                                                      y = fct_reorder(site, date_grid * -1, .fun = mean),
+                                      height = probabilities * 100,
+                                      fill = excavation_project)) +
+    geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
+                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
+                   colour = excavation_project), size = 1.7) +
+  xlab("Cal. BC") +
+  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
+                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
+                     values = c("springgreen4", "royalblue2",
+                                "gold1", "firebrick3")) +
+  labs(title = "B", colour = "Excavation project") +
+  theme_classic() +
+  theme(axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank(),
+        legend.position = "bottom") +
+  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
+    scale_x_reverse()
+
+    ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
+                                      y = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
+                                      height = probabilities * 100,
+                                      group = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
+                                      fill = excavation_project)) +
+  scale_fill_manual(labels = c("E18 Rugtvedt-Dørdal", 
+                                "Vestfoldbaneprosjektet", "E18 Bommestad-Sky"),
+                     values = c("springgreen4",
+                                "gold1", "firebrick3"),
+                    guide = FALSE) +
 
@@
 temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
-  geom_segment(aes(x = reported_start_bp, xend = reported_end_bp,
+  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                    yend = reorder(site, mean_date),
                    colour = excavation_project), size = 1.7) +
-  xlab("Cal. BP") +
+  xlab("Cal. BC") +
   scale_x_reverse() +
@@
 calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count, 
-                       -nonflint_count, -reported_start_bp, -reported_end_bp,
+                       -nonflint_count, -reported_start_bc, -reported_end_bc,
                        -askeladden_id),
@@
   select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
-         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
+         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
   # The reason for not using .keep below is for the option to 
@@
   select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
-         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
+         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
   select(!ends_with("other")) %>% 
@@
   # Transform skewed variables and reorder
-  mutate("Mean date cal. BP*-1" = mean_date * -1,
+  mutate("Mean date cal. BC*-1" = mean_date * -1,
           "log10(lithic_dens)" = log10(lithic_dens),
@@
          "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
-  relocate("Mean date cal. BP*-1",
+  relocate("Mean date cal. BC*-1",
     "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
@@
 lithicspca <- lithicstr %>% 
-  mutate(mean_date =  `Mean date cal. BP*-1` * -1, .keep = "unused") %>% 
+  mutate(mean_date =  `Mean date cal. BC*-1` * -1, .keep = "unused") %>% 
   rename(
@@
   labs(title = "C",
-       x = "Mean date cal. BP",
+       x = "Mean date cal. BC",
        y = "Curation index") +

### 2021-05-07 analysis/paper/paper.Rmd
Started assembling probabilistic sampling of dates for curation index plot

@@
 library(GGally)
+library(ggridges)
 library(patchwork)
@@
 
-The analyses conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
+This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
 This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
@@
 @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites.
-The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials and seasonality.
+The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
 Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
@@
 
-Even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
+However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
 Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
@@
 
-```{r c14-modelling}
+```{r c14-modelling, cache = TRUE}
 # define function to model radiocarbon dates. Only assumption is that 
@@
   oxcal_read <- readOxcalOutput(oxcal_exec)
-  # Note that parseOxcalOutput() defaults to only returning individual dates
+  # Note that parseOxcalOutput() defaults to only returning individual dates,
+  # specified with only.R_Date
   oxcal_data <- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)
@@
 rcarb_data <- do.call(rbind, posteriors)
-  group_by(site) %>%
-  mutate(meandate = mean(date_grid * -1)) %>% 
-  ungroup() %>% 
-  arrange(meandate)
-
-h3 <- radiocarbon %>% 
-  filter(site == "Vallermyrene 1" | site == "Stokke/Polland 5")
-
-v1 <- model_rcarb(h3)
-
-h3dat <- R_Date(h3$lab_code, h3$c14_bp, h3$error)
-
-my_result_file <- executeOxcalScript(oxcal_code)
-my_result_text <- readOxcalOutput(my_result_file)
-my_result_data <- parseOxcalOutput(my_result_text, only.R_Date = FALSE)
-plot(my_result_data$`Vallermyrene 1`$posterior_probabilities$dates, my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities, type = "l")
-
-posterior <- data.frame(probabilities = my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities,
-                        date_grid = my_result_data$`Vallermyrene 1`$posterior_probabilities$dates,
-                        site = "Vallermyrene 1",
-                        excavation_project = "Vestfoldbaneprosjektet")
-
-ggplot() +
-  geom_line(aes(x = yrs, y = my_result_data$` Vallermyrene 1 `$posterior_probabilities$probabilities))
-
-yrs <- seq(11500, 5000, -500)
-plt1 <- ggplot(sites, aes(x = yrs)) +
-  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
-                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
-                   colour = excavation_project), size = 1.7) +
-  xlab("Cal. BC") +
-  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
-                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
-                     values = c("springgreen4", "royalblue2",
-                                "gold1", "firebrick3")) +
-  labs(title = "B", colour = "Excavation project") +
-  theme_classic() +
-  theme(axis.title.y=element_blank(),
-        axis.text.y=element_blank(),
-        axis.ticks.y=element_blank(),
-        legend.position = "bottom") +
-  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
-    scale_x_reverse()
-
-ggplot(sites, aes(x = yrs)) +
-      ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
-                                                      y = fct_reorder(site, date_grid * -1, .fun = mean),
-                                      height = probabilities * 100,
-                                      fill = excavation_project)) +
-    geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
-                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
-                   colour = excavation_project), size = 1.7) +
-  xlab("Cal. BC") +
-  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
-                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
-                     values = c("springgreen4", "royalblue2",
-                                "gold1", "firebrick3")) +
-  labs(title = "B", colour = "Excavation project") +
-  theme_classic() +
-  theme(axis.title.y=element_blank(),
-        axis.text.y=element_blank(),
-        axis.ticks.y=element_blank(),
-        legend.position = "bottom") +
-  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
-    scale_x_reverse()
-
-    ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
-                                      y = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
-                                      height = probabilities * 100,
-                                      group = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
-                                      fill = excavation_project)) +
-  scale_fill_manual(labels = c("E18 Rugtvedt-Dørdal", 
-                                "Vestfoldbaneprosjektet", "E18 Bommestad-Sky"),
-                     values = c("springgreen4",
-                                "gold1", "firebrick3"),
-                    guide = FALSE) +
-
-pst <- get_posterior_probabilities(my_result_data)
-# h3_sum <- oxcal_Sum(h3dat)
-# h3_run <- executeOxcalScript(h3_sum)
-# h3_text <- readOxcalOutput(h3_run)
-# h3_result <- parseFullOxcalOutput(h3_text)
 
@@
 ```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
-The date ranges follow those given in the original excavation reports", fig.height = 10, fig.width = 12}
+Radiocarbon probability density functions are given as the posterior sum.
+Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
+These follow the original reports.", fig.height = 11, fig.width = 12}
 # Create a bounding box around site points to limit the overview map
@@
 
-yrs <- seq(11500, 5000, -500)
-temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
+# Combine radiocarbon data and reported (effectively shoreline/typology) dates,
+# removing the reported date if there are relevant 14C-dates.
+datedata <- full_join(sites, rcarb_data, by = "site") %>% 
+  mutate(reported_start_bc = ifelse(dating_method != "c14", reported_start_bc, NA),
+         reported_end_bc = ifelse(dating_method != "c14", reported_end_bc, NA))
+
+project_colors <- c("bommestad_sky" = "springgreen4",
+                  "brunlanes" = "royalblue2", 
+                  "rugtvedt_dordal" = "gold1",
+                  "vestfoldbaneprosjektet" = "firebrick3")
+
+# Assemble plot of date ranges
+temp_dist <- ggplot(datedata, aes(x = mean_date, y = reorder(site, mean_date))) +
   geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
-                   yend = reorder(site, mean_date),
-                   colour = excavation_project), size = 1.7) +
+                   yend = site, colour = excavation_project.x,
+                   fill = excavation_project.x), size = 1.8) +
+  geom_ridgeline(aes(x = date_grid * -1, height = probabilities * 150,
+                     fill = excavation_project.y, 
+                     colour = excavation_project.x)) +
+  scale_x_reverse() +
+  scale_fill_manual(values = project_colors,
+                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
+                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
+  scale_colour_manual(values = project_colors) +
+  labs(title = "B", fill = "Excavation project") +
   xlab("Cal. BC") +
-  scale_x_reverse() +
-  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
-                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
-                     values = c("springgreen4", "royalblue2",
-                                "gold1", "firebrick3")) +
-  labs(title = "B", colour = "Excavation project") +
   theme_classic() +
@@
         legend.position = "bottom") +
-  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5))
+  guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
+         colour = FALSE)
 
@@
 
-The date ranges provided for the sites are taken from the original excavation reports, and are the result of a consideration of typological indicators, dating with reference to shoreline displacement, and radiocarbon dates where these are available.
-Employing these date-ranges was deemed sufficient, as the main purpose is to get at general tendencies in the material, and not fine-grained chronological developments.
-
-In a series of case studies from Northern Norway, @bølviken1982 were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles [@baxter1994].
-In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project.
-As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at identifying potential variation in the
+Where available radiocarbon dates associated with the sites have been subjected to bayesian modelling employing Oxcal through the oxcAAR package in R.
+The only constraint imposed in the model was that the dates from each site are assumed to belong to a group ---
+
+title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
+
+author:
+
+-   Isak Roalkvam:
+
+email: isak.roalkvam\@iakh.uio.no
+
+institute: [UiO]
+
+correspondence: true
+
+institute:
+
+-   UiO: University of Oslo
+
+date: "\`r format(Sys.time(), '%d %B, %Y')\`"
+
+output:
+
+bookdown::html_document2:
+
+fig_caption: yes
+
+reference_docx: "../templates/template.docx" \# Insert path for the DOCX file
+
+pandoc_args:
+
+-   --lua-filter=../templates/scholarly-metadata.lua
+
+-   --lua-filter=../templates/author-info-blocks.lua
+
+-   --lua-filter=../templates/pagebreak.lua
+
+bibliography: references.bib
+
+csl: "../templates/journal-of-archaeological-science.csl" \# Insert path for the bib-style
+
+------------------------------------------------------------------------
+
+<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
+
+<!-- The following code chunk defines some general settings how code chunks should behave. -->
+
+\`\`\`{r setup, echo = FALSE}
+
+knitr::opts_chunk\$set(
+
+collapse = TRUE,
+
+warning = FALSE,
+
+message = FALSE,
+
+echo = FALSE,
+
+comment = "\#\>",
+
+fig.width = 9,
+
+fig.path = "../figures/",
+
+dpi = 300
+
+)
+
+\`\`\`
+
+\`\`\`{r libraries, eval = TRUE}
+
+library(tidyverse)
+
+library(GGally)
+
+library(ggridges)
+
+library(patchwork)
+
+library(gt)
+
+library(FactoMineR)
+
+library(factoextra)
+
+library(sf)
+
+library(rnaturalearth)
+
+library(oxcAAR)
+
+quickSetupOxcal()
+
+\`\`\`
+
+\`\`\`{r get-functions, include = FALSE}
+
+source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
+
+\#setOxcalExecutablePath("/tmp/RtmpFVOcBJ")
+
+\`\`\`
+
+\`\`\`{r get-data, eval = TRUE, results = 'hide'}
+
+assemblages \<- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
+
+radiocarbon \<- read.csv(here::here('analysis/data/raw_data/radiocarbon_data.csv'))
+
+site_pts \<- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
+
+\# ne_download() is currently failing
+
+\#countries \<- ne_download(scale = 10, type = 'countries', category = 'physical',
+
+\# returnclass = "sf")
+
+countries \<- st_read(here::here('analysis/data/raw_data/naturalearth_countries.gpkg'))
+
+\# gis \<- read.csv(here::here('../../site_data.csv'))
+
+\# gis\$askeladden_id \<- as.integer(str_replace_all(gis\$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
+
+\# gisjoin \<- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
+
+\`\`\`
+
+\# Introduction
+
+The analyses conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
+This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
+However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g.
+\@dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites.
+The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials and seasonality.
+Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
+Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
+
+Even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [\@bevan2015].
+Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
+The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
+As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome.
+To this end, the analysis follows two analytical avenues.
+The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports.
+The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.
+
+\#\# Archaeological context and material
+
+\`\`\`{r wrangling}
+
+sites \<- assemblages %\>%
+
+\# Find mean reported date for each site (to be used in the plots below)
+
+rowwise() %\>%
+
+mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %\>%
+
+mutate(
+
+\# Find total artefact count and flint count for each site
+
+artefact_count = sum(c_across(core:flake_retouch_other)),
+
+flint_count = sum(c_across(core:flake_retouch)),
+
+nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %\>%
+
+\# Exclude sites with less than 100 artefacts
+
+filter(artefact_count \> 100)
+
+\`\`\`
+
+\`\`\`{r c14-modelling, cache = TRUE}
+
+\# define function to model radiocarbon dates. Only assumption is that
+
+\# dates from a single site belong to a single group
+
+model_rcarb \<- function(data){
+
+\# Format individual dates for R_Date in Oxcal
+
+dates \<- R_Date(data[["lab_code"]], data[["c14_bp"]], data[["error"]])
+
+\# Manually compile rest of OxCal code
+
+oxcal_code \<- paste0('
+
+Plot()
+
+{
+
+Sequence()
+
+{
+
+Boundary("Start");
+
+Phase()
+
+{
+
+',
+
+dates,
+
+'
+
+Sum("Sum");
+
+};
+
+Boundary("End");
+
+};
+
+};')
+
+\# Run OxCal code - wrapped this in some functions to not print to
+
+\# rmarkdown
+
+invisible(capture.output(oxcal_exec \<- executeOxcalScript(oxcal_code)))
+
+oxcal_read \<- readOxcalOutput(oxcal_exec)
+
+\# Note that parseOxcalOutput() defaults to only returning individual dates,
+
+\# specified with only.R_Date
+
+oxcal_data \<- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)
+
+\# Assemble data frame to be returned
+
+posterior \<- data.frame(
+
+probabilities = oxcal_data\$Sum\$posterior_probabilities\$probabilities,
+
+date_grid = oxcal_data\$Sum\$posterior_probabilities\$dates,
+
+site = data[["site"]][1],
+
+excavation_project = data[["excavation_project"]][1])
+
+\# Return data
+
+return(posterior)
+
+}
+
+\# Apply modelling function to the radiocarbon dates, grouped by site name
+
+posteriors \<- radiocarbon %\>%
+
+group_by(site) %\>%
+
+group_map(\~ model_rcarb(.), .keep = TRUE)
+
+\# The data is returned as a list of lists that needs to be unpacked
+
+rcarb_data \<- do.call(rbind, posteriors)
+
+\`\`\`
+
+The \`r nrow(sites)\` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \\ref\\\@(fig:dist)A].
+The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [\@solheim2013b; \@jaksland2014a; \@melvold2014b; \@reitan2014l; \@solheim2017b].
+The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts.
+The institution responsible for these excavations was the Museum of Cultural History in Oslo.
+This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations.
+Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum.
+As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g.
+\@clark2006; \@dibble2017].
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [\@solheim2013b; \@jaksland2014, see also \\ref\\\@(fig:dist)B].
+Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
+While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.
+
+\`\`\`{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
+Radiocarbon probability density functions are given as the posterior sum.
+Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
+These follow the original reports.", fig.height = 12, fig.width = 12}
+
+\# Create a bounding box around site points to limit the overview map
+
+sitbbox \<- st_bbox(site_pts)
+
+sitbbox[1:2] \<- sitbbox[1:2] - 600000
+
+sitbbox[3:4] \<- sitbbox[3:4] + 600000
+
+boundingpoly \<- st_as_sf(st_as_sfc(sitbbox))
+
+\# Reproject the bounding box to match world map, and crop the world map
+
+\# with the bounding box polygon
+
+bound_reproj \<- st_transform(boundingpoly, st_crs(countries))
+
+mapcountries \<- countries %\>%
+
+filter(st_intersects(., bound_reproj, sparse = FALSE))
+
+count_reproj \<- st_transform(mapcountries, st_crs(site_pts))
+
+\# Create bounding box around the sites, both for extent indicator on the
+
+\# overview map and to create basemap for the site overview.
+
+sitbbox2 \<- st_bbox(site_pts)
+
+sitbbox2[1:2] \<- sitbbox2[1:2] - 15000
+
+sitbbox2[3:4] \<- sitbbox2[3:4] + 15000
+
+boundingpoly2 \<- st_as_sf(st_as_sfc(sitbbox2))
+
+\# same as above, using the smaller bounding box
+
+bound_reproj2 \<- st_transform(boundingpoly2, st_crs(countries))
+
+mapcountries2 \<- countries %\>%
+
+filter(st_intersects(., bound_reproj2, sparse = FALSE))
+
+count_reproj2 \<- st_transform(mapcountries2, st_crs(site_pts))
+
+\# Assemble overview map
+
+overview \<-
+
+ggplot() +
+
+geom_sf(data = count_reproj, fill = "grey", colour = NA) +
+
+geom_sf(data = boundingpoly2,
+
+fill = NA, colour = "black", size = 1) +
+
+coord_sf(xlim = c(sitbbox[1], sitbbox[3]), ylim = c(sitbbox[2], sitbbox[4]),
+
+expand = FALSE) +
+
+labs(title = "A") +
+
+theme_bw() + theme(axis.title=element_blank(),
+
+axis.text.y = element_blank(),
+
+axis.text.x = element_blank(),
+
+rect = element_rect(),
+
+axis.ticks = element_blank(),
+
+panel.grid.major = element_blank())
+
+\# Combine spatial point data with assemblage data (holds excavation project
+
+\# for colouring points in the maps)
+
+sitejoin \<- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id"))
+
+\# Assemble site map. anc defined first is used for placing the scalebar,
+
+\# which required some tweaking to get right.
+
+anc \<- as.numeric(c(sitbbox2\$ymin, sitbbox2\$xmax))
+
+sitemap \<-
+
+ggplot() +
+
+geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
+
+geom_sf(data = sitejoin, aes(fill = excavation_project),
+
+size = 2.25, shape=21,
+
+colour = "black") +
+
+labs(fill = "Excavation project") +
+
+scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
+
+"E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
+
+values = c("springgreen4", "royalblue2",
+
+"gold1", "firebrick3")) +
+
+ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km",
+
+transform = FALSE, st.size = 4, height = 0.05,
+
+border.size = 0.05, st.dist = 0.07,
+
+anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
+
+coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]),
+
+ylim = c(sitbbox2[2], sitbbox2[4]),
+
+expand = FALSE) +
+
+theme_bw() + theme(axis.title=element_blank(),
+
+axis.text.y = element_blank(),
+
+axis.text.x = element_blank(),
+
+rect = element_rect(),
+
+axis.ticks = element_blank(),
+
+panel.grid.major = element_blank(),
+
+legend.position = "none") +
+
+guides(fill = guide_legend(ncol = 2,
+
+title.position = "top", title.hjust = 0.5))
+
+\# Combine radiocarbon data and reported (effectively shoreline/typology) dates,
+
+\# removing the reported date if there are relevant 14C-dates.
+
+datedata \<- full_join(sites, rcarb_data, by = "site") %\>%
+
+mutate(reported_start_bc = ifelse(dating_method != "c14", reported_start_bc, NA),
+
+reported_end_bc = ifelse(dating_method != "c14", reported_end_bc, NA))
+
+project_colors \<- c("bommestad_sky" = "springgreen4",
+
+"brunlanes" = "royalblue2",
+
+"rugtvedt_dordal" = "gold1",
+
+"vestfoldbaneprosjektet" = "firebrick3")
+
+\# Assemble plot of date ranges
+
+temp_dist \<- ggplot(datedata, aes(x = mean_date, y = reorder(site, mean_date))) +
+
+geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
+
+yend = site, colour = excavation_project.x,
+
+fill = excavation_project.x), size = 1.8) +
+
+geom_ridgeline(aes(x = date_grid \* -1, height = probabilities \* 150,
+
+fill = excavation_project.y,
+
+colour = excavation_project.x)) +
+
+scale_x\_reverse() +
+
+scale_fill_manual(values = project_colors,
+
+labels = c("E18 Bommestad-Sky","E18 Brunlanes",
+
+"E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
+
+scale_colour_manual(values = project_colors) +
+
+labs(title = "B", fill = "Excavation project") +
+
+xlab("Cal. BC") +
+
+theme_classic() +
+
+theme(axis.title.y=element_blank(),
+
+axis.text.y=element_blank(),
+
+axis.ticks.y=element_blank(),
+
+legend.position = "bottom") +
+
+guides(fill = guide_legend(title.position = "top", title.hjust = 0.5),
+
+colour = FALSE)
+
+(overview + sitemap) /
+
+(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))
+
+\`\`\`
+
+A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [\@glørstad2010].
+Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [\@breivik2018; \@solheim2020].
+In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
+\@sørensen1979; \@romundset2018].
+The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf.
+\@perreault2019, 47].
+Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of \@bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
+In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [\@bailey2007].
+However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
+In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [\@sørensen2014].
+Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.
+
+Where available, radiocarbon dates associated with the sites have been subjected to bayesian modelling employing OxCal through the oxcAAR package in R.
+The only constraint imposed in the models was that the dates from each site are assumed to belong to the same group employing the Phase function from OxCal.
+The posterior distributions were then summed for each site.
+Radiocarbon dates considered irrelevant or erroneous in the original reports were disregarded.
+Where relevant radiocarbon dates are not available, the date ranges provided for the sites are taken from the original excavation reports.
+These are the result of a consideration of typological indicators as well as shoreline displacement and local topography.
+
+The assemblage data analysed here is based on the classification of the \`r nrow(sites)\` site assemblages done for the original excavation reports, and consists of \`r ncol(select(sites, core:flake_retouch_other))\` variables representing different debitage and tool types.
+While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions.
+Some categories in the original reports have therefore been combined in the dataset.
+This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm.
+In the reports, this was only separated from (macro-)blades (width \$\\ge\$ 12 mm) and micro-blades (width \$\\le\$ 8 mm) for some of the sites.
+Narrow-blades were combined with the blade category here.
+Furthermore, the artefact data have here been divided into flint and non-flint materials.
+Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g.
+\@berg-hansen1999].
+This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [\@eigeland2015, 46].
+An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites.
+However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here [\@eigeland2015, 45-53].
+Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories.
+An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf.
+\@frivoll2017].
+While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [\@eigeland2015, 370].
 
@@
 In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods.
+
+In a series of case studies from Northern Norway, @bølviken1982 were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles [@baxter1994].
+In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project.
+As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at identifying potential variation in the
 
@@
 The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts.
-A further empirical expectation of an expendient tehcnological organisation is a higher number of cores and unretouched flakes and blades [@clark2017].
+A further empirical expectation of an expendient tehcnological organisation is as a consequence also a higher number of cores and unretouched flakes and blades [@clark2017].
 
@@
   select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
-         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
+          -askeladden_id) %>% 
   # The reason for not using .keep below is for the option to 
@@
          -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
-         -chip_other, -core_other, -core_fragment_other, -dating_method) %>% 
+         -chip_other, -core_other, -core_fragment_other) %>% 
   column_to_rownames("site")
@@
 
-corr_plot(lithicstr)
+corr_plot(select(lithicstr, -reported_start_bc, -reported_end_bc, -dating_method))
 ```
@@
 
-pcalithics <- PCA(select(lithicspca, -mean_date),
+pcalithics <- PCA(select(lithicspca, -reported_start_bc, -dating_method,
+                         -reported_end_bc, -mean_date),
                   scale.unit = TRUE, graph = FALSE)
@@
 minmax <- lithics2 %>% 
+   rownames_to_column(var = "site") %>% 
   mutate(dens =  ((lithic_dens - min(lithic_dens)) / (max(lithic_dens) - min(lithic_dens))),
@@
 ```
+
+```{r aoristic}
+aordata <- full_join(indexdata, rcarb_data, by = "site")
+
+# Defined function to draw sample from either 14C probability distribution,
+# or uniform typological/displacement-inferred date range.
+sample_date <- function(site){
+
+  # Extract all rows pertaining to site of interest
+  # site <- as.data.frame(filter(data, site == sitename))
+
+  # If the dating method is 14C, draw a sample using the associate posterior
+  # probability distribution
+  if(site["dating_method"][1,] == "c14"){
+    samplingframe <- list(x = site[["date_grid"]], y = site[["probabilities"]])
+    sdate <- with(samplingframe, sample(x, size = 1, prob = y))
+  # If not draw a random year from the reported date range (uniform probability)
+  } else {
+    sdate <- sample(as.integer(site["reported_start_bc"]*-1):as.integer(site["reported_end_bc"]*-1), 1)
+  }
+  # Return sampled date
+  return(sdate)
+}
+
+# Create loop to draw samples
+nsamp = 10
+
+for (i in 1:nsamp){
+  tst <- aordata %>% 
+    group_by(site) %>% 
+    group_map(~ sample_date(.), .keep = TRUE) %>% 
+    do.call(rbind, .)
+}
+
+```
+
 # Discussion

### 2021-05-10 analysis/paper/paper.Rmd
Switching to word document for now

--- 2021-05-05 analysis/paper/paper.Rmd
@@
 output: 
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
@@
 ```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
-The date ranges follow those given in the original excavation reports", fig.height = 10, fig.width = 12}
+Radiocarbon probability density functions are given as the posterior sum.
+Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
+These follow the original reports.", fig.height = 11, fig.fullwidth = TRUE}
 # Create a bounding box around site points to limit the overview map
@@
 
-yrs <- seq(11500, 5000, -500)
-temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
+# Combine radiocarbon data and reported (effectively shoreline/typology) dates,
+# removing the reported date if there are relevant 14C-dates.
+datedata <- full_join(sites, rcarb_data, by = "site") %>% 
+  mutate(reported_start_bc = ifelse(dating_method != "c14", reported_start_bc, NA),
+         reported_end_bc = ifelse(dating_method != "c14", reported_end_bc, NA))
+
+project_colors <- c("bommestad_sky" = "springgreen4",
+                  "brunlanes" = "royalblue2", 
+                  "rugtvedt_dordal" = "gold1",
+                  "vestfoldbaneprosjektet" = "firebrick3")
+
+# Assemble plot of date ranges
+temp_dist <- ggplot(datedata, aes(x = mean_date, y = reorder(site, mean_date))) +
   geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
-                   yend = reorder(site, mean_date),
-                   colour = excavation_project), size = 1.7) +
-  xlab("Cal. BC") +
+                   yend = site, colour = excavation_project.x,
+                   fill = excavation_project.x), size = 1.8) +
+  geom_ridgeline(aes(x = date_grid * -1, height = probabilities * 150,
+                     fill = excavation_project.y, 
+                     colour = excavation_project.x)) +
   scale_x_reverse() +
-  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
-                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
-                     values = c("springgreen4", "royalblue2",
-                                "gold1", "firebrick3")) +
-  labs(title = "B", colour = "Excavation project") +
+  scale_fill_manual(values = project_colors,
+                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
+                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
+  scale_colour_manual(values = project_colors) +
+  labs(title = "B", fill = "Excavation project") +
+  xlab("Cal. BCE") +
   theme_classic() +
@@
 (temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))
+
+#plot.margin = unit(c(0,135,0,135)
+# top right, borrom , l
 
@@
 
-The date ranges provided for the sites are taken from the original excavation reports, and are the result of a consideration of typological indicators, dating with reference to shoreline displacement, and radiocarbon dates where these are available.
-Employing these date-ranges was deemed sufficient, as the main purpose is to get at general tendencies in the material, and not fine-grained chronological developments.
-
-In a series of case studies from Northern Norway, @bølviken1982 were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles [@baxter1994].
-In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project.
-As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at identifying potential variation in the
-
 Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
@@
 
-While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g.
-@bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested.
+While the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g.
+@bamforth1986; @shott1996; @surovell2009,9--13], following not least from the ambiguous definition first put forward by @binford1979.
+Some concrete operationalisation of the terms have, however, been suggested.
 Through a series of studies Barton and collegues [e.g.
--@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context.
-This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014].
+-@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context, and is measure taken to reflect degree of curation associated with the assemblages.
+which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014].
 Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
@@
   # Transform skewed variables and reorder
-  mutate("Mean date cal. BC*-1" = mean_date * -1,
+  mutate("Mean date cal. BCE*-1" = mean_date * -1,
           "log10(lithic_dens)" = log10(lithic_dens),
@@
          "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
-  relocate("Mean date cal. BC*-1",
+  relocate("Mean date cal. BCE*-1",
     "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
@@
 lithicspca <- lithicstr %>% 
-  mutate(mean_date =  `Mean date cal. BC*-1` * -1, .keep = "unused") %>% 
+  mutate(mean_date =  `Mean date cal. BCE*-1` * -1, .keep = "unused") %>% 
   rename(
@@
 
-To visualise this more clearly, a curation index based on the two variables was devised by perfoming a min-max normalisation on the two variables, scaling them to take on values between 0 and 1.
-The values for artefact density was then made negative to reflect its relationship with degree of curation.
-The mean was then found for each site on these two normalised values.
-Plotting the results indicates a higher degree of curation among the earliest sites, which then falls and settles without any further major fluctuations.
-
 Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020.
@@
 # Create a table showing the eigenvalues of the components
-round(as.tibble(get_eig(pcalithics)), 2) %>% 
+round(as_tibble(get_eig(pcalithics)), 2) %>% 
   rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
@@
 
+To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
+The values for artefact density was then made negative to reflect its relationship with degree of curation.
+The mean was then found for each site on these two normalised values.
+To account for the temporal uncertainty associated with the sites, a simulation based approach was also adopted [e.g.
+@crema2012; @baxter2016; @orton2017].
+A LOESS curve was fit to the data for each simulation, where the age of each site was drawn from the date ranges associated with the sites.
+For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while dates for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date ranges.
+This simulation was repeated `r nsim` times, the results of which is visualed in .
+Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE before remaining relatively stable until the end of the Mesolithic.
+It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site types and associated mobility patterns.
+
 ```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint.
@@
 
+# Curation index plot simply using mean site age (not used)
 w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
@@
   labs(title = "C",
-       x = "Mean date cal. BC",
+       x = "Mean date cal. BCE",
        y = "Curation index") +
@@
   theme_classic() 
-w3
+#w3
 ```
+
+```{r simage, cache = TRUE}
+agedata <- full_join(indexdata, rcarb_data, by = "site")
+
+# Define function to draw sample from either 14C probability distribution,
+# or uniform typological/displacement-inferred date range.
+sample_date <- function(site){
+
+  # If the dating method is 14C, draw a sample using the associated posterior
+  # pdf
+  if(site["dating_method"][1,] == "c14"){
+    samplingframe <- list(x = site[["date_grid"]], y = site[["probabilities"]])
+    sdate <- with(samplingframe, sample(x, size = 1, prob = y))
+  # If not, draw a random year from the reported date range (uniform probability)
+  } else {
+    sdate <- sample(as.integer(site["reported_start_bc"]*-1):as.integer(site["reported_end_bc"]*-1), 1)
+  }
+  # Return sampled date with site name as rowname. 
+  sdate <- data.frame(sdate)
+  row.names(sdate) <- site[["site"]][1]
+  return(sdate)
+}
+
+# Specify number of simulations
+nsim <- 1000
+
+# Create data frame to hold simulation results.
+results <- data.frame(matrix(NA, ncol = nsim, nrow = nrow(sites))) 
+# Insert the curation index value for each site
+results[,"index"] <- agedata %>% 
+  group_by(site) %>% 
+  slice(1) %>% 
+  ungroup() %>% 
+  select(index) %>% 
+  c()
+
+# Simulation loop
+for (i in 1:nsim){
+  agesamples <- agedata %>% 
+    group_by(site) %>% 
+    group_map(~ sample_date(.), .keep = TRUE) %>% 
+    do.call(rbind, .)
+
+  results[,i] <- agesamples
+  row.names(results) <- row.names(agesamples)
+}
+
+# Apply pivot_longer to make the results amenable to plotting
+resultslong <- results %>% 
+  rownames_to_column(var = "site") %>% 
+  pivot_longer(
+    cols = starts_with("X"),
+    names_to = "sim", 
+    values_to = "age"
+  )
+
+# Call to plot, grouping loess fitting by simulation runs
+ggplot(resultslong, aes(x = age, y = index)) + 
+  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black", alpha = 0.05, se = TRUE) +
+  geom_point(size = 0.5, alpha = 0.005) +
+  xlab("Cal. BCE") +
+  ylab("Curation index") +
+  theme_classic()
+
+```
+
 # Discussion
@@
 The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites.
-Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
+Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
 This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).

### 2021-05-10 analysis/paper/paper.tex
Switching to word document for now

@@
 \setcounter{secnumdepth}{5}
+\usepackage{amsmath}
+\usepackage{booktabs}
+\usepackage{caption}
+\usepackage{longtable}
 \ifluatex
@@
 \author{Isak Roalkvam\textsuperscript{1,*}}
-\date{26 March, 2021}
+\date{09 May, 2021}
 
@@
 
-\hypertarget{plots}{%
-\section{Plots}\label{plots}}
+\begin{verbatim}
+#> NULL
+\end{verbatim}
 
-\includegraphics{../figures/cor-1.pdf} \includegraphics{../figures/cor-2.pdf}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-Figure \ref{fig:cor} displays a correspondence analysis using the lithic count data. \ref{fig:cor}
+This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
+This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
+However, the composition of the assemblages can be expected to be determined by a multitude of factors (e.g.
+Dibble et al., 2017), ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites.
+The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
+Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
+Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
+
+However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities (Bevan, 2015).
+Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
+The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
+As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome.
+To this end, the analysis follows two analytical avenues.
+The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports.
+The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.
+
+\hypertarget{archaeological-context-and-material}{%
+\subsection{Archaeological context and material}\label{archaeological-context-and-material}}
+
+The 54 coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref@(fig:dist)A{]}.
+The sites were excavated as part of four larger excavation projects that all took place within the last 15 years (\textbf{solheim2013b?}; \textbf{jaksland2014a?}; \textbf{melvold2014b?}; \textbf{reitan2014l?}; \textbf{solheim2017b?}).
+The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts.
+The institution responsible for these excavations was the Museum of Cultural History in Oslo.
+This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations.
+Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum.
+As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation (e.g.
+Clark and Riel-Salvatore, 2006; Dibble et al., 2017).
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates (Jaksland, 2014, see also \ref@(fig:dist)B; \textbf{solheim2013b?}).
+Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
+While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.
 
@@
 \centering
-\includegraphics{../figures/pca-analysis-1.pdf}
-\caption{\label{fig:pca-analysis}PCA.}
+\includegraphics{../figures/dist-1.pdf}
+\caption{\label{fig:dist}A) Spatial and B) temporal distribution of the sites chosen for analysis.
+Radiocarbon probability density functions are given as the posterior sum.
+Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
+These follow the original reports.}
 \end{figure}
 
-Figure \ref{fig:pca-analysis} displays a principle components analysis using the continuous measures as defined by Bicho and Cascalheira (2020)
+A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas (Glørstad, 2010).
+Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use (Breivik et al., 2018; Solheim et al., 2020).
+In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound (Romundset et al., 2018; e.g.
+Sørensen, 1979).
+The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time (cf.
+Perreault, 2019, p.
+47).
+Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of Bailey (2007), while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
+In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost (Bailey, 2007).
+However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
+In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic (Sørensen et al., 2014).
+Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.
+
+The data analysed here is based on the classification of the 54 site assemblages done for the original excavation reports, and consists of 48 variables representing different debitage and tool types.
+While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions.
+Some categories in the original reports have therefore been combined in the dataset.
+This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm.
+In the reports, this was only separated from (macro-)blades (width \(\ge\) 12 mm) and micro-blades (width \(\le\) 8 mm) for some of the sites.
+Narrow-blades were combined with the blade category here.
+Furthermore, the artefact data have here been divided into flint and non-flint materials.
+Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice (e.g.
+Berg-Hansen, 1999).
+This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents (Eigeland, 2015, p.
+46).
+An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites.
+However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here (Eigeland, 2015, pp.
+45--53).
+Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories.
+An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports (cf.
+Frivoll, 2017).
+While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway (Eigeland, 2015, p.
+370).
+
+Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types (e.g.
+Helskog et al., 1976; Reitan, 2016), involved refit-studies of the entirety of individual assemblages (Skar and Coulson, 1986), or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes (e.g.
+Damlien, 2016; Solheim et al., 2020).
+Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation (Nærøy, 2018), and technological organisation {[}Eigeland (2015); Mansrud and Eymundsson (2016);{]}.
+There are also some more comprehensive evaluations of a larger number of entire lithic assemblages to get at potential site types and associated mobility patterns (Breivik, 2020; \textbf{breivik2016?}; \textbf{viken2018c?}).
+However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of site types.
+Each site could have had complex life histories and been associated with a wide range of tasks that in sum make up the assemblages.
+Finally, only a single study concerned with eight Middle-Mesolithic assemblages --- assemblages that are also treated here --- has employed a multivariate quantitative framework to help structure such a treatment (\textbf{solheim2013?}).
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods.
+
+In a series of case studies from Northern Norway, Bølviken et al.
+(1982) were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles (Baxter, 1994).
+In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to (\textbf{solheim2013a?}) use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project.
+As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at identifying potential variation in the
+
+There have been a few studies that have conducted comprehensive analyses of artefact inventories and site features to get at variation in Mesolithic site types (Breivik, 2020; Viken, 2018).
+As Binford originally clarified, a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.
+
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
+
+The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods when the entirety of the assemblages are taken into account.
+This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
+However, the approach taken is also partially informed by the so-called Frison effect (Jelinek, 1976), which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
+Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination (e.g.
+Barton, 1991; Dibble, 1995).
+Consequently, several artefact categories have here been collapased for the CA.
+This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
+That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading.
+These have all been combined into the single category ``small flint tools.'' (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
+While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
+An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
+While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.
+
+However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take properly into account, and renders it difficult to align any identified patterns with specific behavioural dimensions (Dibble et al., 2017).
+For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
+While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape (e.g.
+Solheim et al., 2018).
+Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place (Dibble et al., 2017).
+As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind (see Clark and Barton, 2017, and below).
+The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient (Binford, 1979).
+An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement.
+Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations (Binford, 1979).
+Although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool or extending its use-life by the repeated rejuvination and modification of edges (e.g.
+Bamforth, 1986; Dibble, 1995; Shott and Sillitoe, 2005).
+The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies, as well as raw-material quality and availability (Andrefsky, 1994; Clark and Barton, 2017; Kuhn, 1992; e.g.
+Parry and Kelly, 1987; Smith, 2015).
+
+While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed Shott (1996), following not least from the ambiguous definition first put forward by Binford (1979), some concrete operationalisation of the terms have been suggested.
+Through a series of studies Barton and collegues (e.g.
+1999; Barton and Riel-Salvatore, 2014; Clark and Barton, 2017) have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context.
+This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites (Barton and Riel-Salvatore, 2014).
+Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
+From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics (Clark and Barton, 2017).
+More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation.
+This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
+The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts.
+A further empirical expectation of an expendient tehcnological organisation is as a consequence also a higher number of cores and unretouched flakes and blades (Clark and Barton, 2017).
+
+This is the main measure employed here.
+Some additional variables suggested by Bicho and Cascalheira (2020) to be related to mobility patterns are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures (cf.
+Baxter, 1994, p.
+100).
+
+A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here.
+This pertains to number of site features such as fireplaces, cooking pits and dwelling structures.
+Feature count has been omitted as taphonomic loss is likely to lead to a chronological bias in their preservation.
+Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
+This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied.
+This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here (Jørgensen, 2017; \textbf{darmark2018e?}).
+
+Within this framework, amount of retouch has been seen as a empirical indicator of degree of curation.
+While various methods for measuring the curation associated with individual tools have been forwarded, the approach taken here focuses on the assemblage considered as a whole.
+
+Here retouch i At logistical basecamps, the extended time spent in a single location would imply a better control of raw material availability, and allow for longer trips to retrieve these.
+Higher availability of lithic raw materials should in turn reduce the necessity for extensive retouch, meaning the material expectation for logistical basecamps would be relatively high density of lithics and a low relative frequency of retouched artefacts.
+From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics (Clark and Barton, 2017).
+
+\hypertarget{theorycalculation}{%
+\section{Theory/calculation}\label{theorycalculation}}
+
+Higher frequency of secondarily worked lithics
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/cor-1.pdf}
+\caption{\label{fig:cor}Correspondence analysis using artefact count data. A) Object map, B) Variable map.}
+\end{figure}
+
+Figure \ref{fig:cor} displays the CA using the lithic count data.
+While no clear-cut clusters can be readily delineated, the general impression from the plots is that a chronological dimension is associated with the patterning in the data.
+This is indicated by the general transition across the colour scale in the row plot (Figure \ref{fig:cor}A), combined with the fact that the two first dimensions of the CA accounts for as much as 80.53 \% of the inertia or variance in the data.
+The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
+Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
+
+The column plot (Figure \ref{fig:cor}B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
+It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves.
+The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+These findings are in large part in line with previous research (e.g.
+Bang-Andersen, 2003), although possibly highlighting the importance of the non-flint materials to a larger extent than what has traditionally been the case.
+
+The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
+While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
+Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
+This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research Glørstad (2011).
+
+In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
+The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2.
+
+\hypertarget{section}{%
+\section{}\label{section}}
+
+Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis, and an overly naive trust in the fact that these units reflect any behaviourally meaningful categories.
+This realisation has led to a wide range of outside Scandinavian Mesolithic research, meaning a range of empirical measures associated with various mobility patterns are available.
+
+It has, however, been suggested that access to raw-materials can be a more important decider for assemblage composition than mobility patterns , and that variable workability of these materials will impact factors such as the relative frequency of tools to debitage in the assemblages (e.g.
+Andrefsky, 1994; Manninen and Knutsson, 2014).
+In addition to attempts at accounting for this analytically (see below), the limited geographical distribution of the sites should alleviate the issue somewhat, as the overall availability of raw materials could be expected to be more comparable through time than if a wider region was under consideration.
+
+Of these the most well-estbalished is the.
+base don the assumption that, this measure and has been shown to display a meaningful measure across a wide range of temporal, geographical and cultural contexts.
+
+\begin{figure}
+\centering
+\includegraphics{../figures/cor-plot-1.pdf}
+\caption{\label{fig:cor-plot}Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these}
+\end{figure}
+
+\includegraphics{../figures/pca-analysis-1.pdf}
+Figure \ref{fig:pca-analysis} displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by Clark and Barton (2017) and Bicho and Cascalheira (2020).
+While the investigation performed by Bicho and Cascalheira (2020) indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data.
+hese last dimensions are capturing the most variation in the dataset presented here.
+Thus, if frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+Furthermore, the fact that these
+
+\captionsetup[table]{labelformat=empty,skip=1pt}
+\begin{longtable}{cccc}
+\toprule
+Component & Eigenvalue & Variance (\%) & Cumulative variance (\%) \\ 
+\midrule
+1 & 2.29 & 38.21 & 38.21 \\ 
+2 & 1.79 & 29.90 & 68.11 \\ 
+3 & 0.95 & 15.85 & 83.96 \\ 
+4 & 0.62 & 10.32 & 94.28 \\ 
+5 & 0.32 & 5.39 & 99.66 \\ 
+6 & 0.02 & 0.34 & 100.00 \\ 
+\bottomrule
+\end{longtable}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/var_contrib-1.pdf}
+\caption{(\#fig:var\_contrib)Contribution of variables to components.}
+\end{figure}
+
+Figure \ref{fig:wabi} displays the WABI as found for the site data.
+This indicates a negative correlation between the two variables (r = -0.5) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+Following Clark and Barton (2017), this would be in line with a general transition from a expedient to curated lithic technology.
+Variable raw-material availability can also impact these measures as (Manninen and Knutsson, 2014).
+Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies.
+However, as is evident in when considering both plots in \ref{fig:wabi}, the pattern is evident for both the lithics considered in aggregate and for the flint data specifically.
+Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly chronologically, the general pattern is the same.
+
+To visualise this temporal trend more clearly, a curation index based on the two variables was devised by performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
+The values for artefact density was then made negative to reflect its relationship with degree of curation.
+The mean was then found for each site on these two normalised values.
+To account for the temporal uncertainty associated with the sites, a simulation based approach was also adopted.
+A LOESS curve was fit to the data for each simulation, where the age of each site was drawn from the date ranges associated with the sites.
+For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date ranges.
+This simulation was repeated 5000 times, the results of which is visualed in .
+Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, which then has a marked drop around 8000 BCE before remaining relatively stable until the end of the Mesolithic.
+It is also interseting that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site types and associated mobility patterns.
+Plotting the results indicates a higher degree of curation among the earliest sites, which then falls and settles without any further major fluctuations.
+
+\begin{figure}
+\centering
+\includegraphics{../figures/wabi-1.pdf}
+\caption{\label{fig:wabi}Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m\textsuperscript{3}) for A) All lithics, B) Flint.
+The logarithm is taken to base 10 on all axes.}
+\end{figure}
+
+\includegraphics{../figures/simage-1.pdf}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time.
+However, as several authors have pointed out in relation to the application of CA for seriation purposes, `time is not the only dimension.' The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites.
+Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
+This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).
+
+these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
+The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.
+However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips?
+Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pair is not correlated with these.
+As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.
+In attempt to determine if this might be the case, it was decided to introduce site locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here Roalkvam (2020).
+Of the variables considered, this study found that the most consistently important environmental covariates for the location of sites in the region was related to the exposure of the sites, and failed to identify any diachronic variation in this pattern.
+As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similarly temporary , one could envisage that variablity in site functionality might be reflected in degree of exposure.
+However, introducing these variables does not clarify the picture, as both variables show negligible degrees of correlation with any of the variables pertaining to the assemblage data.
+
+This is arguably more established than the variables included bysimilarities between the variable plots of Bicho and Cascalheira (2020) and the variables indicated in is striking.
+This pertains to the negative correlation between core and chip frequency, and the negative correlation between primary and secondarily worked lithics.
+This last
+
+These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at behavioural dimensions relating to mobility patterns among hunter-gatherers.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway.
+This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
+These tendencies are already well-established when it comes to formal tool types, but the have been given less focus in terms of entire assemblages these notions have previously Precisely what behavioural implication the development in the use of these types have, however, is unclear
+
+The results of considering variables that are have been associated with mobility patterns gave rise to two competing hypotheses.
+These pertain to whether the degree of curation associated with the assemblages as captured by the changing ratio of primary to secondarily worked lithics over time is the result of changing raw material availability or mobility patterns.
+The orthogonal relationship between core to chip frequency would in this framework be expected to be capturing the opposite relationship.
+If degree of curation mainly reflects changes in raw-material quality and availability, this would then indicate that mobility patterns are comparable throughout the Mesolithic as the core to chip frequency is not time-dependent.
+And vice verca --- if the changing mobility patterns is the main driver of the transition from curated to expedient technological organisation, as operationalised here, While this would be in line with findings from a recent evaluation of the locational patterns associated with the sites (Roalkvam, 2020), there is no readily available empirical evidence, no smoking gun, se findings are the results of an exploratory analysis, and are therefore perhaps best seen as potentially laying the framework for later targeted strategic test of these competing hypotheses.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-andrefsky1994}{}%
+Andrefsky, W., 1994. Raw-material availability and the organization of technology. American Antiquity 59, 21--34. \url{https://doi.org/10.2307/3085499}
+
+\leavevmode\hypertarget{ref-bailey2007}{}%
+Bailey, G., 2007. Time perspectives, palimpsests and the archaeology of time. Journal of Anthropological Archaeology 26, 198--223. \url{https://doi.org/10.1016/j.jaa.2006.08.002}
+
+\leavevmode\hypertarget{ref-bamforth1986}{}%
+Bamforth, D.B., 1986. Technological Efficiency and Tool Curation. American Antiquity 51, 38--50. \url{https://doi.org/10.2307/280392}
+
+\leavevmode\hypertarget{ref-bang-andersen2003}{}%
+Bang-Andersen, S., 2003.
+Southwest Norway at the Pleistocene/Holocene Transition: Landscape Development, Colonization, Site Types, Settlement Patterns.
+Norwegian Archaeological Review 36, 5--25.
+\url{https://doi.org/10.1080/00293650307293}
+
+\leavevmode\hypertarget{ref-barton1991}{}%
+Barton, C.M., 1991. Retouched Tools, Fact or Fiction? Paradigms for Interpreting Paleolithic Chipped Stone, in: Clark, G.A. (Ed.),. University of Pennsylvania Press, pp. 143--163.
+
+\leavevmode\hypertarget{ref-barton1999}{}%
+Barton, C.M., Bernabeu, J., Aura, J.E., García, O., 1999.
+Land-Use Dynamics and Socioeconomic Change: An Example from the Polop Alto Valley.
+American Antiquity 64, 609--634.
+\url{https://doi.org/10.2307/2694208}
+
+\leavevmode\hypertarget{ref-barton2014}{}%
+Barton, C.M., Riel-Salvatore, J., 2014. The formation of lithic assemblages. Journal of Archaeological Science 46, 334--352. \url{https://doi.org/10.1016/j.jas.2014.03.031}
+
+\leavevmode\hypertarget{ref-baxter1994}{}%
+Baxter, M.J., 1994. Exploratory multivariate analysis in archaeology. Percheron Press, New York.
+
+\leavevmode\hypertarget{ref-berg-hansen1999}{}%
+Berg-Hansen, I.M., 1999. The availability of flint at lista and jæren, southwestern norway, in: Boaz, J. (Ed.),. University of Oslo, Oslo, pp. 255--266.
+
+\leavevmode\hypertarget{ref-bevan2015}{}%
+Bevan, A., 2015. The data deluge. Antiquity 89, 1473--1484. \url{https://doi.org/10.15184/aqy.2015.102}
+
 \leavevmode\hypertarget{ref-Bicho2020}{}%
@@
 
+\leavevmode\hypertarget{ref-binford1979}{}%
+Binford, L.R., 1979. Organization and Formation Processes: Looking at Curated Technologies. Journal of Anthropological Research 35, 255--273.
+
+\leavevmode\hypertarget{ref-breivik2020}{}%
+Breivik, H.M., 2020. Diachronic trends among early mesolithic sites types? A study from the coast of central norway, in: Schülke, A. (Ed.),. Rutledge, London \& New York, pp. 121--146.
+
+\leavevmode\hypertarget{ref-breivik2018}{}%
+Breivik, H.M., Fossum, G., Solheim, S., 2018.
+Exploring human responses to climatic fluctuations and environmental diversity: Two stories from Mesolithic Norway.
+Quaternary International, Impacts of gradual and abrupt environmental changes on Late glacial to Middle Holocene cultural changes in Europe 465, 258--275.
+\url{https://doi.org/10.1016/j.quaint.2016.12.019}
+
+\leavevmode\hypertarget{ref-buxf8lviken1982}{}%
+Bølviken, E., Helskog, E., Helskog, K., Holm-Olsen, I.M., Solheim, L., Bertelsen, R., 1982.
+Correspondence analysis: An alternative to principal components.
+World Archaeology 14, 41--60.
+\url{https://doi.org/10.1080/00438243.1982.9979848}
+
+\leavevmode\hypertarget{ref-clark2017}{}%
+Clark, G.A., Barton, C.M., 2017.
+Lithics, landscapes \& la Longue-durée {{}} Curation \& expediency as expressions of forager mobility.
+Quaternary International, Prehistoric hunter-gatherers and farmers in the Adriatic and neighboring regions 450, 137--149.
+\url{https://doi.org/10.1016/j.quaint.2016.08.002}
+
+\leavevmode\hypertarget{ref-clark2006}{}%
+Clark, G.A., Riel-Salvatore, J., 2006. Observations on systematics in paleolithic archaeology, in: Hovers, E., Kuhn, S. (Eds.),. Springer, New York, pp. 29--56.
+
+\leavevmode\hypertarget{ref-damlien2016}{}%
+Damlien, H., 2016.
+Eastern pioneers in westernmost territories?
+Current perspectives on Mesolithic hunter{{}}gatherer large-scale interaction and migration within Northern Eurasia.
+Quaternary International, Holocene Hunter-Gatherer Archaeology of Northern Eurasia (Guest Editors: Peter Jordan and Andrzej Weber) 419, 5--16.
+\url{https://doi.org/10.1016/j.quaint.2014.02.023}
+
+\leavevmode\hypertarget{ref-dibble1995}{}%
+Dibble, H.L., 1995.
+Middle paleolithic scraper reduction: Background, clarification, and review of the evidence to date.
+Journal of Archaeological Method and Theory 2, 299--368.
+\url{https://doi.org/10.1007/BF02229003}
+
+\leavevmode\hypertarget{ref-dibble2017}{}%
+Dibble, H.L., Holdaway, S.J., Lin, S.C., Braun, D.R., Douglass, M.J., Iovita, R., McPherron, S.P., Olszewski, D.I., Sandgathe, D., 2017.
+Major Fallacies Surrounding Stone Artifacts and Assemblages.
+Journal of Archaeological Method and Theory 24, 813--851.
+\url{https://doi.org/10.1007/s10816-016-9297-8}
+
+\leavevmode\hypertarget{ref-eigeland2015}{}%
+Eigeland, L., 2015. Maskinmennesket i steinalderen. Endring og kontinuitet i steinteknologi fram mot neolitiseringen av øst-norge (PhD thesis). Oslo.
+
+\leavevmode\hypertarget{ref-frivoll2017}{}%
+Frivoll, A., 2017. Identifisering og klassifisering av littiske råmaterialer i sør- og østnorsk steinalderforskning. Reliabilitet av visuell klassifiseringsmetode (PhD thesis). Oslo.
+
+\leavevmode\hypertarget{ref-gluxf8rstad2011}{}%
+Glørstad, H., 2011. The Nøstvet axe, in: Davis, V., Edmonds, M. (Eds.),. Oxbow Books, pp. 21--36.
+
+\leavevmode\hypertarget{ref-gluxf8rstad2010}{}%
+Glørstad, H., 2010. The structure and history of the late mesolithic societies in the oslo fjord area 6300-3800 BC. Bricoleur Press, Lindome.
+
+\leavevmode\hypertarget{ref-helskog1976}{}%
+Helskog, K., Indrelid, S., Mikkelsen, E., 1976. Morfologisk klassifisering av slåtte steinartefakter. Universitetets Oldsaksamling Årbok 1972-74, 9--40.
+
+\leavevmode\hypertarget{ref-jaksland2014}{}%
+Jaksland, L., 2014. Kulturhistorisk sammenstilling, in: Jaksland, L., Persson, P. (Eds.),. Museum of Cultural History, University of Oslo, Oslo, pp. 11--46.
+
+\leavevmode\hypertarget{ref-jelinek1976}{}%
+Jelinek, A.J., 1976. Form, function and style in lithic analysis, in: Cleland, C.E. (Ed.),. Academic Press, New York, pp. 19--33.
+
+\leavevmode\hypertarget{ref-juxf8rgensen2017}{}%
+Jørgensen, E.K., 2017.
+Om vegetasjonsforstyrrelser: Konsekvenser for bevaringen av arkeologisk kontekstinformasjon i norske jordsmonn.
+Viking 80, 157180--157180.
+\url{https://doi.org/10.5617/viking.5477}
+
+\leavevmode\hypertarget{ref-kuhn1992}{}%
+Kuhn, S.L., 1992. On Planning and Curated Technologies in the Middle Paleolithic. Journal of Anthropological Research 48, 185--214.
+
+\leavevmode\hypertarget{ref-manninen2014}{}%
+Manninen, M.A., Knutsson, K., 2014.
+Lithic raw material diversification as an adaptive strategy{{}}Technology, mobility, and site structure in Late Mesolithic northernmost Europe.
+Journal of Anthropological Archaeology 33, 84--98.
+\url{https://doi.org/10.1016/j.jaa.2013.12.001}
+
+\leavevmode\hypertarget{ref-mansrud2016}{}%
+Mansrud, A., Eymundsson, C., 2016. Socialized landscapes? Lithic clusters, hearths and relocation rituals at Middle Mesolithic sites in Eastern Norway. Fennoscandia archaeologica 33, 27--55.
+
+\leavevmode\hypertarget{ref-nuxe6ruxf8y2018}{}%
+Nærøy, A.J., 2018. Early Mesolithic spatial conformity in southern Norway. Journal of Archaeological Science: Reports 18, 905--912. \url{https://doi.org/10.1016/j.jasrep.2017.10.021}
+
+\leavevmode\hypertarget{ref-parry1987}{}%
+Parry, W.J., Kelly, R.L., 1987. Expedient core technology and sedentism, in: Johnson, J.K., Morrow, C.A. (Eds.),. Westview Press, Boulder \& London, pp. 285--308.
+
+\leavevmode\hypertarget{ref-perreault2019}{}%
+Perreault, C., 2019. The quality of the archaeological record. The University of Chicago Press, Chicago \& London.
+
+\leavevmode\hypertarget{ref-reitan2016}{}%
+Reitan, G., 2016. Mesolittisk kronologi i Sørøst-Norge {{}} et forslag til justering. Viking 79, 23--51. \url{https://doi.org/10.5617/viking.3903}
+
+\leavevmode\hypertarget{ref-roalkvam2020}{}%
+Roalkvam, I., 2020.
+Algorithmic Classification and Statistical Modelling of Coastal Settlement Patterns in Mesolithic South-Eastern Norway.
+Journal of Computer Applications in Archaeology 3, 288307.
+\url{https://doi.org/10.5334/jcaa.60}
+
+\leavevmode\hypertarget{ref-romundset2018}{}%
+Romundset, A., Lakeman, T.R., Høgaas, F., 2018.
+Quantifying variable rates of postglacial relative sea level fall from a cluster of 24 isolation basins in southern Norway.
+Quaternary Science Reviews 197, 175--192.
+\url{https://doi.org/10.1016/j.quascirev.2018.07.041}
+
+\leavevmode\hypertarget{ref-shott1996}{}%
+Shott, M.J., 1996. An Exegesis of the Curation Concept. Journal of Anthropological Research 52, 259--280. \url{https://doi.org/10.1086/jar.52.3.3630085}
+
+\leavevmode\hypertarget{ref-shott2005}{}%
+Shott, M.J., Sillitoe, P., 2005. Use life and curation in New Guinea experimental used flakes. Journal of Archaeological Science 32, 653--663. \url{https://doi.org/10.1016/j.jas.2004.11.012}
+
+\leavevmode\hypertarget{ref-skar1986}{}%
+Skar, B., Coulson, S., 1986. Evidence of behaviour from refitting - a case study. Norwegian Archaeological Review 19, 90--102.
+
+\leavevmode\hypertarget{ref-smith2015}{}%
+Smith, G.M., 2015.
+Modeling the influences of raw material availability and functional efficiency on obsidian projectile point curation: A Great Basin example.
+Journal of Archaeological Science: Reports 3, 112--121.
+\url{https://doi.org/10.1016/j.jasrep.2015.06.010}
+
+\leavevmode\hypertarget{ref-solheim2020}{}%
+Solheim, S., Damlien, H., Fossum, G., 2020.
+Technological transitions and human-environment interactions in Mesolithic southeastern Norway, 11~500{{}}6000~cal.
+BP.
+Quaternary Science Reviews 246, 106501.
+\url{https://doi.org/10.1016/j.quascirev.2020.106501}
+
+\leavevmode\hypertarget{ref-solheim2018}{}%
+Solheim, S., Fossum, G., Knutsson, H., 2018.
+Use-wear analysis of Early Mesolithic flake axes from South-eastern Norway.
+Journal of Archaeological Science: Reports 17, 560--570.
+\url{https://doi.org/10.1016/j.jasrep.2017.12.017}
+
+\leavevmode\hypertarget{ref-suxf8rensen1979}{}%
+Sørensen, R., 1979. Late Weichselian deglaciation in the Oslofjord area, south Norway. Boreas 8, 241--246. https://doi.org/\url{https://doi.org/10.1111/j.1502-3885.1979.tb00806.x}
+
+\leavevmode\hypertarget{ref-suxf8rensen2014}{}%
+Sørensen, R., Henningsmoen, K.E., Høeg, H.I., Gälman, V., 2014.
+Holocene landhevningsstudier i søndre vestfold og sørøstre telemark- revidert kurve, in: Melvold, S., Persson, P.
+(Eds.),.
+Portal, Kristiansand, pp.
+36--47.
+
+\leavevmode\hypertarget{ref-viken2018}{}%
+Viken, S., 2018.
+Sagene B1.
+En tidligmesolittisk basisboplass med én boligstruktur og spor etter flere samtidige hushold, in: Reitan, G., Sundström, L.
+(Eds.),.
+Cappelen Damm Akademisk, Oslo, pp.
+131--166.
+
 \end{CSLReferences}
@@
 
-This report was generated on 2021-03-26 16:50:59 using the following computational environment and dependencies:
+This report was generated on 2021-05-09 12:32:08 using the following computational environment and dependencies:
 
@@
 #>  setting  value                       
-#>  version  R version 4.0.4 (2021-02-15)
+#>  version  R version 4.0.5 (2021-03-31)
 #>  os       Linux Mint 19.3             
@@
 #>  tz       Europe/Oslo                 
-#>  date     2021-03-26                  
+#>  date     2021-05-09                  
 #> 
@@
 #>  backports       1.2.0    2020-11-02 [1] CRAN (R 4.0.3)
+#>  bitops          1.0-6    2013-08-17 [1] CRAN (R 4.0.3)
 #>  bookdown        0.21     2020-10-13 [1] CRAN (R 4.0.3)
@@
 #>  cellranger      1.1.0    2016-07-27 [1] CRAN (R 4.0.3)
+#>  checkmate       2.0.0    2020-02-06 [1] CRAN (R 4.0.3)
+#>  class           7.3-18   2021-01-24 [4] CRAN (R 4.0.3)
+#>  classInt        0.4-3    2020-04-07 [1] CRAN (R 4.0.3)
 #>  cli             2.2.0    2020-11-20 [1] CRAN (R 4.0.3)
-#>  cluster         2.1.1    2021-02-14 [4] CRAN (R 4.0.3)
+#>  cluster         2.1.2    2021-04-17 [4] CRAN (R 4.0.5)
+#>  codetools       0.2-18   2020-11-04 [4] CRAN (R 4.0.3)
 #>  colorspace      1.4-1    2019-03-18 [1] CRAN (R 4.0.3)
@@
 #>  curl            4.3      2019-12-02 [1] CRAN (R 4.0.3)
-#>  data.table      1.13.4   2020-12-08 [1] CRAN (R 4.0.3)
+#>  data.table      1.14.0   2021-02-21 [1] CRAN (R 4.0.5)
 #>  DBI             1.1.0    2019-12-15 [1] CRAN (R 4.0.3)
@@
 #>  DT              0.16     2020-10-13 [1] CRAN (R 4.0.3)
+#>  e1071           1.7-4    2020-10-14 [1] CRAN (R 4.0.3)
 #>  ellipsis        0.3.1    2020-05-15 [1] CRAN (R 4.0.3)
@@
 #>  generics        0.1.0    2020-10-31 [1] CRAN (R 4.0.3)
+#>  GGally        * 2.1.1    2021-03-08 [1] CRAN (R 4.0.5)
+#>  ggmap           3.0.0    2019-02-05 [1] CRAN (R 4.0.4)
 #>  ggplot2       * 3.3.2    2020-06-19 [1] CRAN (R 4.0.3)
@@
 #>  ggrepel         0.9.1    2021-01-15 [1] CRAN (R 4.0.3)
+#>  ggridges      * 0.5.2    2020-01-12 [1] CRAN (R 4.0.3)
 #>  ggsignif        0.6.0    2019-08-08 [1] CRAN (R 4.0.3)
+#>  ggsn            0.5.0    2019-02-18 [1] CRAN (R 4.0.4)
 #>  glue            1.4.2    2020-08-27 [1] CRAN (R 4.0.3)
+#>  gt            * 0.2.2    2020-08-05 [1] CRAN (R 4.0.3)
 #>  gtable          0.3.0    2019-03-25 [1] CRAN (R 4.0.3)
@@
 #>  httr            1.4.2    2020-07-20 [1] CRAN (R 4.0.3)
+#>  jpeg            0.1-8.1  2019-10-24 [1] CRAN (R 4.0.3)
 #>  jsonlite        1.7.1    2020-09-07 [1] CRAN (R 4.0.3)
+#>  KernSmooth      2.23-18  2020-10-29 [1] CRAN (R 4.0.3)
 #>  knitr           1.30     2020-09-22 [1] CRAN (R 4.0.3)
@@
 #>  magrittr        2.0.1    2020-11-17 [1] CRAN (R 4.0.3)
+#>  maptools        1.0-2    2020-08-24 [1] CRAN (R 4.0.3)
 #>  MASS            7.3-53.1 2021-02-12 [4] CRAN (R 4.0.3)
+#>  Matrix          1.3-2    2021-01-06 [4] CRAN (R 4.0.3)
 #>  memoise         1.1.0    2017-04-21 [1] CRAN (R 4.0.3)
+#>  mgcv            1.8-33   2020-08-27 [1] CRAN (R 4.0.3)
 #>  modelr          0.1.8    2020-05-19 [1] CRAN (R 4.0.3)
 #>  munsell         0.5.0    2018-06-12 [1] CRAN (R 4.0.3)
+#>  nlme            3.1-152  2021-02-04 [4] CRAN (R 4.0.3)
 #>  openxlsx        4.2.3    2020-10-27 [1] CRAN (R 4.0.3)
+#>  oxcAAR        * 1.1.0    2021-02-23 [1] CRAN (R 4.0.5)
 #>  patchwork     * 1.1.0    2020-11-09 [1] CRAN (R 4.0.3)
@@
 #>  pkgload         1.1.0    2020-05-29 [1] CRAN (R 4.0.3)
+#>  plyr            1.8.6    2020-03-03 [1] CRAN (R 4.0.3)
+#>  png             0.1-7    2013-12-03 [1] CRAN (R 4.0.3)
 #>  prettyunits     1.1.1    2020-01-24 [1] CRAN (R 4.0.3)
@@
 #>  R6              2.5.0    2020-10-28 [1] CRAN (R 4.0.3)
+#>  RColorBrewer    1.1-2    2014-12-07 [1] CRAN (R 4.0.3)
 #>  Rcpp            1.0.5    2020-07-06 [1] CRAN (R 4.0.3)
@@
 #>  reprex          0.3.0    2019-05-16 [1] CRAN (R 4.0.3)
+#>  reshape         0.8.8    2018-10-23 [1] CRAN (R 4.0.5)
+#>  RgoogleMaps     1.4.5.3  2020-02-12 [1] CRAN (R 4.0.4)
 #>  rio             0.5.26   2021-03-01 [1] CRAN (R 4.0.4)
+#>  rjson           0.2.20   2018-06-08 [1] CRAN (R 4.0.4)
 #>  rlang           0.4.10   2020-12-30 [1] CRAN (R 4.0.4)
 #>  rmarkdown       2.5      2020-10-21 [1] CRAN (R 4.0.3)
+#>  rnaturalearth * 0.1.0    2017-03-21 [1] CRAN (R 4.0.5)
 #>  rprojroot       2.0.2    2020-11-15 [1] CRAN (R 4.0.3)
@@
 #>  sessioninfo     1.1.1    2018-11-05 [1] CRAN (R 4.0.3)
+#>  sf            * 0.9-7    2021-01-06 [1] CRAN (R 4.0.4)
+#>  sp              1.4-5    2021-01-10 [1] CRAN (R 4.0.4)
 #>  stringi         1.5.3    2020-09-09 [1] CRAN (R 4.0.3)
@@
 #>  tibble        * 3.0.4    2020-10-12 [1] CRAN (R 4.0.3)
-#>  tidyr         * 1.1.2    2020-08-27 [1] CRAN (R 4.0.3)
+#>  tidyr         * 1.1.3    2021-03-03 [1] CRAN (R 4.0.4)
 #>  tidyselect      1.1.0    2020-05-11 [1] CRAN (R 4.0.3)
 #>  tidyverse     * 1.3.0    2019-11-21 [1] CRAN (R 4.0.3)
+#>  units           0.6-7    2020-06-13 [1] CRAN (R 4.0.3)
 #>  usethis         2.0.1    2021-02-10 [1] CRAN (R 4.0.4)
-#>  vctrs           0.3.5    2020-11-17 [1] CRAN (R 4.0.3)
+#>  vctrs           0.3.7    2021-03-29 [1] CRAN (R 4.0.4)
 #>  withr           2.3.0    2020-09-22 [1] CRAN (R 4.0.3)
@@
 #> Remote:   master @ origin (https://github.com/isakro/dialpastrepository.git)
-#> Head:     [75c1135] 2021-03-26: Initial
+#> Head:     [aa13e0e] 2021-05-07: Started assembling probabilistic sampling of dates for curation index plot
 \end{verbatim}

### 2021-05-12 analysis/paper/paper.Rmd
Further interpretation. Identified logistic base camps through the curation index?

@@
 This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
-This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space.
+This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time.
 However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g.
-@dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites.
+@dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites.
 The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
@@
 Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
+In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification.
 
-However, even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
-Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
+Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
+Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
 The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
@@
 To this end, the analysis follows two analytical avenues.
-The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports.
+The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports, which consequently assumes that these are behaviourally meaningful.
 The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.
@@
 Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
-While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.
+While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects.
 
@@
 Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
-These follow the original reports.", fig.height = 11, fig.fullwidth = TRUE}
+These follow the original reports.", fig.height = 10, fig.width = 12}
 # Create a bounding box around site points to limit the overview map
@@
 @perreault2019, 47].
-Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events.
-In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007].
+Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred.
+In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007].
 However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
@@
 @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g.
-@damlien2016; @solheim2020].
+@damlien2016; @berg-hansen2017; @solheim2020].
 Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;].
-There are also some more comprehensive evaluations of a larger number of entire lithic assemblages to get at potential site types and associated mobility patterns [@breivik2016; @viken2018c; @breivik2020].
-However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of site types.
-Each site could have had complex life histories and been associated with a wide range of tasks that in sum make up the assemblages.
-Finally, only a single study concerned with eight Middle-Mesolithic assemblages --- assemblages that are also treated here --- has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013].
-In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods.
+There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018c; @breivik2020].
+However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of different site types.
+Finally, only a single study concerned with eight Middle-Mesolithic sites has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013a].
+These sites are also included here.
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently structure and balance the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.
 
-In a series of case studies from Northern Norway, @bølviken1982 were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles [@baxter1994].
-In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project.
-As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at identifying potential variation in the
+## Methodology
 
-There have been a few studies that have conducted comprehensive analyses of artefact inventories and site features to get at variation in Mesolithic site types [@viken2018; @breivik2020].
-As Binford originally clarified, a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.
+As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied.
+Only parts of this process can sensibly be reported here.
+The data and R script used for the analysis is freely available as a research compendium following @marwick , allowing readers to explore the data and the analytical choices made [cf.
+@Marwick2017].
+However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly impacted by and selectively reported upon, following my own biases.
+It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies.
 
-## Methods
-
-The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods when the entirety of the assemblages are taken into account.
+The first part of the analysis conducted involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods.
 This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
 However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
-Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g.
+Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g.
 @barton1991; @dibble1995].
@@
 
-However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not take properly into account, and renders it difficult to align any identified patterns with specific behavioural dimensions [@dibble2017].
+However, for the most part we lack even a most basic understanding of what any individual lithic object has been used for [@dibble2017].
+This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions.
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
@@
 Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].
-Although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is therefore typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool or extending its use-life by the repeated rejuvination and modification of edges [e.g.
+
+However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g.
+@bamforth1986; @shott1996; @surovell2009, 9--13].
+The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g.
+@parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].
+Still, that the distinction can offer a useful analytical framework if clearly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted.
+For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g.
 @bamforth1986; @dibble1995; @shott2005].
-The continuum between curated and expedient technology has in turn been related to land-use and mobility strategies, as well as raw-material quality and availability [e.g.
-@parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].
+Furthermore, one concrete operationalisation of the terms have been forwarded by Barton and collegues [e.g.
+-@barton1999; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation.
+This is in turn mainly to follow from the accumulated nature of the land-use and mobility patterns associated with the sites [@barton2014].
+In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
+From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
+More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation.
+This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
+The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].
 
-While the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g.
-@bamforth1986; @shott1996; @surovell2009,9--13], following not least from the ambiguous definition first put forward by @binford1979.
-Some concrete operationalisation of the terms have, however, been suggested.
-Through a series of studies Barton and collegues [e.g.
--@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context, and is measure taken to reflect degree of curation associated with the assemblages.
-which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014].
-Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
-From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
-More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation.
-This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
-The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts.
-A further empirical expectation of an expendient tehcnological organisation is as a consequence also a higher number of cores and unretouched flakes and blades [@clark2017].
-
-This is the main measure employed here.
-Some additional variables suggested by @Bicho2020 to be related to mobility patterns are subsequently also explored using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
+These are the main measure employed here.
+However, as these measures are argued to predominantly be determined by mobility patterns, some additional variables suggested by @Bicho2020 to be related to mobility patterns are also explored.
+This is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
 @baxter1994, 100].
@@
 This pertains to number of site features such as fireplaces, cooking pits and dwelling structures.
-Feature count has been omitted as taphonomic loss is likely to lead to a chronological bias in their preservation.
+Feature count has been omitted as taphonomic loss is likely to have led to a chronological bias in their preservation.
 Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
@@
 This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].
-
-Within this framework, amount of retouch has been seen as a empirical indicator of degree of curation.
-While various methods for measuring the curation associated with individual tools have been forwarded, the approach taken here focuses on the assemblage considered as a whole.
 
@@
 Figure \@ref(fig:cor) displays the CA using the lithic count data.
-While no clear-cut clusters can be readily delineated, the general impression from the plots is that a chronological dimension is associated with the patterning in the data.
+The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
@@
          core_freq = sum(core, core_other) / artefact_count,
-         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
+         blanks_freq = sum(blade, microblade, flake, flake_other, fragment,
                            fragment_other, blade_other,  microblade_other) / artefact_count,
@@
                 flake_retouch_other) / artefact_count,
-         nonflint_freq = nonflint_count / artefact_count,
          lithic_dens = artefact_count / excavated_m3) %>%
@@
          core_freq = sum(core) / artefact_count,
-         blanks_freq = sum(blade, microblade, fragment, flake) / artefact_count,
+         blanks_freq = sum(blade, microblade, flake, fragment) / artefact_count,
          secondary_freq =  sum(axe_debitage, axe, chisel, 
@@
          "log10(secondary_freq)" = log10(secondary_freq),
-         "core_freq^(1/3)" = core_freq^(1/3),
-         "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
+         "core_freq^(1/3)" = core_freq^(1/3), .keep = "unused") %>% 
   relocate("Mean date cal. BCE*-1",
@@
    secondary_freq = "log10(secondary_freq)",
-   core_freq = "core_freq^(1/3)",
-   nonflint_freq = "nonflint_freq^(1/3)"
+   core_freq = "core_freq^(1/3)"
   )
@@
 Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020.
-While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data.
-hese last dimensions are capturing the most variation in the dataset presented here.
-Thus, if frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
-Furthermore, the fact that these
+There is a general temporal transition from older to younger sites along the second dimension of the plot, which is mainly defined by a negative correlation between the variables volumetric density of lithics and relative frequency of secondarily worked lithics, as well as relative frequency of cores.
+Almost orthogonal to this is the negative correlation between relative frequency of chips and blanks.
+While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear to be temporally determined.
+The investigation performed by @Bicho2020 indicated that in their data the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns.
+However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, combined with the fact that the relative frequency of cores align with the expectation from the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here.
+If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and in turn related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+To explore this proposition further, these two variables are subjected to more detailed scrutiny below.
 
@@
 However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically.
-Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly chronologically, the general pattern is the same.
+Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly contingent on chronology, the general pattern is the same.
 
@@
 The mean was then found for each site on these two normalised values.
-To account for the temporal uncertainty associated with the sites, a simulation based approach was also adopted [e.g.
+To account for the temporal uncertainty associated with the sites, a simulation-based approach was also adopted [e.g.
 @crema2012; @baxter2016; @orton2017].
-A LOESS curve was fit to the data for each simulation, where the age of each site was drawn from the date ranges associated with the sites.
-For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while dates for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date ranges.
-This simulation was repeated `r nsim` times, the results of which is visualed in .
-Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE before remaining relatively stable until the end of the Mesolithic.
+A LOESS curve was fit to the data for each simulation run, where the age of each site was drawn from the date ranges associated with the sites.
+For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while dates for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
+This simulation was repeated 1000 times, the results of which is visualed in .
+Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE, before remaining relatively stable until the end of the Mesolithic.
 It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site types and associated mobility patterns.
@@
   geom_smooth(method = "loess", col = "red") +
-  theme_classic() 
+  theme_classic()  + 
+  ggrepel::geom_label_repel(aes(label = site),box.padding   = 0.35, 
+                  point.padding = 0.5,
+                  segment.color = 'grey50') 
 #w3
@@
 
-```{r simage, cache = TRUE}
+```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index.
+The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their associated probability distributions given in Figure 1B.
+A LOESS curve has been fit to the distribution for each of the 1000 simulation runs."}
+
 agedata <- full_join(indexdata, rcarb_data, by = "site")
@@
 
-# Call to plot, grouping loess fitting by simulation runs
+# Call to plot, grouping loess fits by simulation runs
 ggplot(resultslong, aes(x = age, y = index)) + 
@@
 Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
-This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).
+This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural significance of the employed units of analysis unclear.
 
+Assuming for now that the curation index captures degree of mobility, the present data indicates two major modes with a generally higher degree of mobility until some time before 8000 BCE.
+This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic.
+The variation in degree of curation, which, as a note, is at it highest between c.
+8000 and 7000 BCE, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
+The five sites that have values on the curation index below -0.25, could in this perspective have predominantly functioned as base camps within a logistic settlement pattern.
+That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports, although on different and differing grounds than the curation measures employed here.
+Another case worth commenting on is sites such as Langangen Vestgård 1, which on the grounds of a large number of artefacts and the possibile presence of a dwelling structure was argued to reflect a more stable site location in the original report.
+However, the site could reflect the aggregation of stays which predominantly have been of shorter duration, while the possible dwelling structure, if taken as an indication of lower mobility, could represent a remnant from one or a few longer stays that constitute a smaller fraction of the use-life of the site as a whole.
+It is also interesting to note that some of the chronological patterns indicated in the CA does not appear to be related to the curation measures.
+This for example pertains to the apparent importance of the axe production sites.
+It has recently been suggested that the end of the Mesolithic sees a drop in population numbers and a subsequent increase in mobility as migrating populations arrived from southern Scandinavia.
+Whether the apparent absence of this pattern in the present data is a result of a strained sample size or a weaker or non-existent signal from any changes in mobility patterns in the later Mesolithic remains an open question.
+As it stands, however, the main hypothesis generated by the present data would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE and remain relatively stable throughout the rest of the period.
+Given the parallel increase in variation this would seem to correlate well with a transition from a residential to a logistical settlement system.
+
+As for the measures associated with curation and mobility patterns, relative frequency of secondarily worked lithics and volumentric density of lithics correspond most clearly to patterns previously proposed in the literature.
+However, as these were identified by employing PCA and identifying what variables most clearly correspond to a temporal dimension, there is a clear danger of circular reasoning here.
+Having established that these measures appears to capture a behavioural dimension of interest, this can, however, now be tested more rigorously by employing site data from a wider region.
 these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
@@
 As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.
-In attempt to determine if this might be the case, it was decided to introduce site locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020.
-Of the variables considered, this study found that the most consistently important environmental covariates for the location of sites in the region was related to the exposure of the sites, and failed to identify any diachronic variation in this pattern.
-As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similarly temporary , one could envisage that variablity in site functionality might be reflected in degree of exposure.
-However, introducing these variables does not clarify the picture, as both variables show negligible degrees of correlation with any of the variables pertaining to the assemblage data.
 
@@
 This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
-These tendencies are already well-established when it comes to formal tool types, but the have been given less focus in terms of entire assemblages these notions have previously Precisely what behavioural implication the development in the use of these types have, however, is unclear
+These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages.
+these notions have previously Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as measured here.
 

### 2021-05-13 analysis/paper/paper.Rmd
Curation index on both flint and non-flint. Highlighted sites of interest

@@
 Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;].
-There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018c; @breivik2020].
+There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018 ; @breivik2020].
 However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of different site types.
@@
 These sites are also included here.
-In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently structure and balance the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.
+In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently balance and structure the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.
 
@@
 
-As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied.
-Only parts of this process can sensibly be reported here.
-The data and R script used for the analysis is freely available as a research compendium following @marwick , allowing readers to explore the data and the analytical choices made [cf.
+As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored.
+Only parts of this process can sensibly be reported.
+However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly selectively reported upon and heavily impacted by my own biases.
+It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies.
+Furthermore, the employed data and R script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [cf.
 @Marwick2017].
-However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly impacted by and selectively reported upon, following my own biases.
-It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies.
 
@@
 @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].
-Still, that the distinction can offer a useful analytical framework if clearly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted.
+Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted.
 For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g.
@@
 -@barton1999; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation.
-This is in turn mainly to follow from the accumulated nature of the land-use and mobility patterns associated with the sites [@barton2014].
+This is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the sites [@barton2014].
 In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
@@
 
-```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
-# Create a table showing the eigenvalues of the components
-round(as_tibble(get_eig(pcalithics)), 2) %>% 
-  rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
-         "Cumulative variance (%)" = cumulative.variance.percent) %>% 
-  rownames_to_column(var = "Component") %>% 
-gt(rowname_col = "row", auto_align = FALSE)
-```
-
 ```{r var_contrib, fig.cap = "Contribution of variables to components.", fig.height = 3}
@@
 
-w2 <- ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
+w2 <- 
+  ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
                      fill = mean_date / 1000)) + 
@@
 
+minmaxflint <- flint2 %>% 
+   rownames_to_column(var = "site") %>% 
+  mutate(dens =  ((flint_dens - min(flint_dens)) / (max(flint_dens) - min(flint_dens))),
+         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))
+
 indexdata <- minmax %>% 
@@
 
+indexdataflint <- minmaxflint %>% 
+  rowwise() %>% 
+  mutate(index_flint = mean(c(dens * -1, secondary)))
+
+indexdata$index_flint <- indexdataflint$index_flint
+
 # Curation index plot simply using mean site age (not used)
 w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
+  geom_point() +
+  labs(title = "C",
+       x = "Mean date cal. BCE",
+       y = "Curation index") +
+  scale_x_reverse() + 
+  geom_smooth(method = "loess", col = "red") +
+  theme_classic()  + 
+  ggrepel::geom_label_repel(aes(label = site),box.padding   = 0.35, 
+                  point.padding = 0.5,
+                  segment.color = 'grey50')
+
+w4 <- ggplot(indexdata, aes(x = mean_date, y = index_flint)) + 
   geom_point() +
@@
 The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their associated probability distributions given in Figure 1B.
-A LOESS curve has been fit to the distribution for each of the 1000 simulation runs."}
+A LOESS curve has been fit to the distribution for each of the 1000 simulation runs.
+Each simulation is plotted with some transparency.
+Sites mentioned in the text are coloured.", fig.height = 10, fig.width = 12}
 
@@
 # Insert the curation index value for each site
-results[,"index"] <- agedata %>% 
+index_values <- agedata %>% 
   group_by(site) %>% 
@@
   ungroup() %>% 
-  select(index) %>% 
+  select(index, index_flint) %>% 
   c()
+
+results[,"index"] <- index_values$index
+results[,"index_flint"] <- index_values$index_flint
 
@@
 
+# Create column to hold whether site is to be coloured specifically in the plots
+# (pertains to sites discussed in the text)
+sitelabs <- c("Hovland 3", "Langangen Vestgård 1", "Hegna vest 4",
+              "Vallermyrene 4", "Gunnarsrød 6b", "Prestemoen 1")
+
+resultslong <- resultslong %>% 
+  mutate(site_col = ifelse(site %in% sitelabs, site, "other"))
+
 # Call to plot, grouping loess fits by simulation runs
-ggplot(resultslong, aes(x = age, y = index)) + 
-  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black", alpha = 0.05, se = TRUE) +
-  geom_point(size = 0.5, alpha = 0.005) +
-  xlab("Cal. BCE") +
-  ylab("Curation index") +
-  theme_classic()
+i1 <- ggplot(resultslong, aes(x = age, y = index, colour = site_col)) + 
+  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
+              alpha = 0.05, se = TRUE) +
+  geom_point(size = 0.5, alpha = 0.01) +
+  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
+                                             "Hegna vest 4","Gunnarsrød 6b",
+                                             "Prestemoen 1",  "Hovland 3",
+                                             "Vallermyrene 4"),
+                    values = c("Langangen Vestgård 1" = "red",
+                               "Hegna vest 4" = "blue",
+                               "Gunnarsrød 6b" = "violet",
+                               "Prestemoen 1" = "darkorange",
+                               "Hovland 3" = "green",
+                               "Vallermyrene 4"  = "cyan1", 
+                               "other" = "black"),
+                     guide = guide_legend(override.aes = list(size = 3,
+                                                              alpha = 1))) +
+  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
+  labs(title = "A", x = "Cal. BCE", y = "Curation index (all lithics)") +
+  theme_classic() +
+  theme(legend.position = "none")
 
+i2 <- ggplot(resultslong, aes(x = age, y = index_flint, colour = site_col)) + 
+  geom_smooth(aes(group = sim), method = "loess", fill = "grey", col = "black",
+              alpha = 0.05, se = TRUE) +
+  geom_point(size = 0.5, alpha = 0.01) +
+  scale_color_manual(name = NULL, breaks = c("Langangen Vestgård 1", 
+                                             "Hegna vest 4","Gunnarsrød 6b",
+                                             "Prestemoen 1",  "Hovland 3",
+                                             "Vallermyrene 4"),
+                    values = c("Langangen Vestgård 1" = "red",
+                               "Hegna vest 4" = "blue",
+                               "Gunnarsrød 6b" = "violet",
+                               "Prestemoen 1" = "darkorange",
+                               "Hovland 3" = "green",
+                               "Vallermyrene 4"  = "cyan1", 
+                               "other" = "black"),
+                     guide = guide_legend(override.aes = list(size = 3,
+                                                              alpha = 1))) +
+  scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
+  labs(title = "A", x = "Cal. BCE", y = "Curation index (flint)") +
+  theme_classic() +
+  theme(legend.position = "bottom")
+
+  i1 /
+  i2
 ```
@@
 This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic.
-The variation in degree of curation, which, as a note, is at it highest between c.
-8000 and 7000 BCE, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
-The five sites that have values on the curation index below -0.25, could in this perspective have predominantly functioned as base camps within a logistic settlement pattern.
-That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports, although on different and differing grounds than the curation measures employed here.
-Another case worth commenting on is sites such as Langangen Vestgård 1, which on the grounds of a large number of artefacts and the possibile presence of a dwelling structure was argued to reflect a more stable site location in the original report.
-However, the site could reflect the aggregation of stays which predominantly have been of shorter duration, while the possible dwelling structure, if taken as an indication of lower mobility, could represent a remnant from one or a few longer stays that constitute a smaller fraction of the use-life of the site as a whole.
+This pattern is evident in both the flint data and when all lithics are treated in aggregate.
+The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
+The five sites that have values on the curation index below c.
+-0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern.
+That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility.
+This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint.
+However, even if this renders it more problematic to directly align the values on the curation index with a continuum between a residental and logistic settlement system, the variation on the index after around 8000 BCE is still evident.
+Another case worth commenting on is Langangen Vestgård 1, which, on the grounds of a large number of artefacts and the possibile presence of a dwelling structure was argued to reflect a more stable site location in the original report [@melvold2014].
+However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of lower mobility, could represent a remnant from one or a few longer stays that constitute a smaller fraction of the use-life of the site as a whole.
+
+While it is possible the is to consider these patterns at a population level.
+From this perspective it seems clear that there is a marked drop in curation from just before 8000 BCE, which stabilises around 7000 BCE.
+This would correspond well with chronological where the end of the Early Mesolithic, or flake axe phase, is set to last until c.
+8200 BCE.
+The following phase linked to the production of microliths is in this framework set to last to around 7000 BCE, signifying the start of the production of the non-flint ground-pecked or round butted adzes.
+Later chronological transitions suggested at c.
+5600 and c.
+4500 BCE, are not reflected in fluctuations in the curation index.
+While this can be traced in the CA with the reduced importance of the non-flint categories, this could indicate that these changes are not related mobility patterns.
+
 It is also interesting to note that some of the chronological patterns indicated in the CA does not appear to be related to the curation measures.
@@
 It has recently been suggested that the end of the Mesolithic sees a drop in population numbers and a subsequent increase in mobility as migrating populations arrived from southern Scandinavia.
-Whether the apparent absence of this pattern in the present data is a result of a strained sample size or a weaker or non-existent signal from any changes in mobility patterns in the later Mesolithic remains an open question.
+It is an open question whether the apparent absence of this pattern in the present data is a result of a strained sample size, or a weaker or non-existent signal from any changes in mobility patterns in the later Mesolithic.
 As it stands, however, the main hypothesis generated by the present data would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE and remain relatively stable throughout the rest of the period.
-Given the parallel increase in variation this would seem to correlate well with a transition from a residential to a logistical settlement system.
+Even though the relative increase in Given the parallel increase in variation this would seem to correlate well with a transition from a residential to a logistical settlement system.
 
 As for the measures associated with curation and mobility patterns, relative frequency of secondarily worked lithics and volumentric density of lithics correspond most clearly to patterns previously proposed in the literature.
-However, as these were identified by employing PCA and identifying what variables most clearly correspond to a temporal dimension, there is a clear danger of circular reasoning here.
+However, as these were identified by employing PCA and identifying what variables most clearly correspond to a temporal dimension, there is a danger of circular reasoning.
 Having established that these measures appears to capture a behavioural dimension of interest, this can, however, now be tested more rigorously by employing site data from a wider region.

### 2021-05-14 analysis/paper/paper.Rmd
Some preliminary discussion of results

@@
 
-To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
-The values for artefact density was then made negative to reflect its relationship with degree of curation.
-The mean was then found for each site on these two normalised values.
-To account for the temporal uncertainty associated with the sites, a simulation-based approach was also adopted [e.g.
-@crema2012; @baxter2016; @orton2017].
-A LOESS curve was fit to the data for each simulation run, where the age of each site was drawn from the date ranges associated with the sites.
-For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while dates for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
-This simulation was repeated 1000 times, the results of which is visualed in .
-Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE, before remaining relatively stable until the end of the Mesolithic.
-It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site types and associated mobility patterns.
-
 ```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint.
@@
 ```
+
+To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
+The values for artefact density was then made negative to reflect its relationship with degree of curation.
+The mean was then found for each site on these two normalised values.
+To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g.
+@crema2012; @baxter2016; @orton2017].
+A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites.
+For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while age for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
+This simulation was repeated 1000 times, the results of which is visualised in \@ref(fig:simage).
+Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE.
+This stabilises around 7000 BCE without any further major fluctuations for the rest of the Mesolithic.
+It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.
 
@@
 
-```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index.
-The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their associated probability distributions given in Figure 1B.
+```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) all lithics in aggregate, and B) only the flint data.
+The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability distributions given in Figure 1B.
 A LOESS curve has been fit to the distribution for each of the 1000 simulation runs.
-Each simulation is plotted with some transparency.
+Each simulation run is plotted with some transparency.
 Sites mentioned in the text are coloured.", fig.height = 10, fig.width = 12}
@@
 
-The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time.
+The results of the CA does appear to align well with previous research that has been based on more informal treatment of the assemblages as a whole, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time.
 However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
@@
 This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural significance of the employed units of analysis unclear.
+The results of the CA can, however, been used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions, to nuance or explain discrepencies in htis data.
 
-Assuming for now that the curation index captures degree of mobility, the present data indicates two major modes with a generally higher degree of mobility until some time before 8000 BCE.
+Assuming for now that the curation index captures degree of mobility, the present data indicates a generally higher degree of mobility until some time before 8000 BCE.
 This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic.
 This pattern is evident in both the flint data and when all lithics are treated in aggregate.
-The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
+The variation in degree of curation when all lithics are considered, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
 The five sites that have values on the curation index below c.
@@
 This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint.
-However, even if this renders it more problematic to directly align the values on the curation index with a continuum between a residental and logistic settlement system, the variation on the index after around 8000 BCE is still evident.
-Another case worth commenting on is Langangen Vestgård 1, which, on the grounds of a large number of artefacts and the possibile presence of a dwelling structure was argued to reflect a more stable site location in the original report [@melvold2014].
-However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of lower mobility, could represent a remnant from one or a few longer stays that constitute a smaller fraction of the use-life of the site as a whole.
+Also referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages.
+As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
+The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps.
+The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel each other out.
+Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern.
+For the later Mesolithic, the high expediency on some sites when considering all lithics in aggregate does speak to variability in site use, but the nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential logistic continuum.
+A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].
+However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
 

### 2021-05-17 analysis/paper/paper.Rmd
Writing

@@
 The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
-Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
+Finally, analytic and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
 Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
@@
 @frivoll2017].
-While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [@eigeland2015, 370].
+While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g.
+@glørstad2011; @eigeland2015, 370].
 
@@
 
-As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored.
+As the title of the paper suggest, the data has here been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored.
 Only parts of this process can sensibly be reported.
@@
 There is a general temporal transition from older to younger sites along the second dimension of the plot, which is mainly defined by a negative correlation between the variables volumetric density of lithics and relative frequency of secondarily worked lithics, as well as relative frequency of cores.
-Almost orthogonal to this is the negative correlation between relative frequency of chips and blanks.
-While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear to be temporally determined.
+Almost orthogonal to this is the negative correlation between relative frequency of chips (artefacts of a size \>) and blanks (flakes, blades, microblades and fragments).
+While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear not to be temporally determined.
 The investigation performed by @Bicho2020 indicated that in their data the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns.
-However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, combined with the fact that the relative frequency of cores align with the expectation from the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here.
-If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and in turn related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, when combined with the fact that the relative frequency of cores align with the expectation of the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here.
+If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
 To explore this proposition further, these two variables are subjected to more detailed scrutiny below.
@@
 
-Figure \@ref(fig:wabi) displays the WABI as found for the site data.
-This indicates a negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+Figure \@ref(fig:wabi) clearly illustrates the negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
 Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.
-Variable raw-material availability can also impact these measures as [@manninen2014].
+As variable raw-material availability and workability can also impact these measures [@manninen2014].
 Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies.
 However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically.
-Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly contingent on chronology, the general pattern is the same.
+Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly temporally contingent, the general pattern is the same.
 
@@
 @crema2012; @baxter2016; @orton2017].
-A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites.
-For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while age for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
+A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites \@ref(fig:dist).
+For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
 This simulation was repeated 1000 times, the results of which is visualised in \@ref(fig:simage).
@@
 
-Assuming for now that the curation index captures degree of mobility, the present data indicates a generally higher degree of mobility until some time before 8000 BCE.
-This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic.
+The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises from around 7000 BCE for the rest of the Mesolithic.
 This pattern is evident in both the flint data and when all lithics are treated in aggregate.
@@
 The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps.
-The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel each other out.
-Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern.
-For the later Mesolithic, the high expediency on some sites when considering all lithics in aggregate does speak to variability in site use, but the nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential logistic continuum.
+The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel out.
+Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern, while the high expediency on some sites when considering all lithics in aggregate in the later parts of the Mesolithic does speak to variability in site use.
+The nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential-logistic continuum.
 A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].

### 2021-05-20 analysis/paper/paper.Rmd
Looked more concretely at some of the data reported by Barton et al. and compared with Meso S-E Norway. Seems its generally less curated here - which makes sense given both the coastal and meso setting.

@@
 @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites.
-The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials.
+The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as procurement strategies and access to raw materials.
 Finally, analytic and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined.
 Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
-In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification.
+In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification within larger unified projects.
 
@@
 
-A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@glørstad2010].
+A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@bjerck2010; @glørstad2010].
 Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
@@
 
-Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
-@helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or been concerned with technological processes associated with certain sub-categories of the site inventories, such as blades or axes [e.g.
+Studies concerned with chronological changes in the composition of lithic assemblages in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
+@helskog1976; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g.
 @damlien2016; @berg-hansen2017; @solheim2020].
-Some studies have involved in-depth analyses of handful of sites in terms of intrasite spatial organisation [@nærøy2018], and technological organisation [@eigeland2015; @mansrud2016;].
-There are also some more comprehensive evaluations of a larger number of sites with associated lithic assemblages to get at potential site types and related mobility patterns [@breivik2016; @viken2018 ; @breivik2020].
-However, the narratively driven analyses of these data makes the weighting of different variables unclear, and these studies draw their conclusions based on an arguably over-constraining predefined trichotomy of different site types.
-Finally, only a single study concerned with eight Middle-Mesolithic sites has employed a multivariate quantitative framework to help structure such a treatment [@solheim2013a].
-These sites are also included here.
-In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult, if not impossible, to consistently balance and structure the comparison of a larger number of sites with associated assemblages holding lithic artefacts numbering in the thousands.
+Studies that have involved the entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018; @breivik2020], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
+These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
+To my knowledege, only a single study in southern Norway, concerned with the analysis of eight Middle Mesolithic assemblages, has involved the use of a multivariate quantitative framework to structure the analysis [@solheim2013a].
+In sum, therefore, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult to consistently balance the comparison of a larger number of assemblages.
 
@@
 
-As the title of the paper suggest, the data has here been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored.
-Only parts of this process can sensibly be reported.
-However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly selectively reported upon and heavily impacted by my own biases.
-It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies.
-Furthermore, the employed data and R script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [cf.
-@Marwick2017].
+The relatively constrained geographical distribution of the analysed sites, the limited temporal spread over which they where investigated, as well as the methodological equivalency across projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
+However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
+While only parts of this process can sensibly be reported on here, the data and employed R programming script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves what Binford termed a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
+The process can still provide the identification of empirical patterns along the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses and lay the foundation for a deductive research agenda for which clear test implications can be derived [@clark2009, 29].
 
-The first part of the analysis conducted involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods.
-This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.
-However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
+The first part of the analysis involved employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+The purpose of this exercise was to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods.
+This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful.
+However, the approach taken was also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
 Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g.
@@
 While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
-An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
-While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.
+An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
+While largely intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.
 
-However, for the most part we lack even a most basic understanding of what any individual lithic object has been used for [@dibble2017].
-This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions.
+However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017].
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
@@
 Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
-As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind [see @clark2017, and below].
+This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions.
+As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below].
 The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979].
@@
 @bamforth1986; @dibble1995; @shott2005].
-Furthermore, one concrete operationalisation of the terms have been forwarded by Barton and collegues [e.g.
--@barton1999; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation.
+Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1999 and collegues [e.g.
+@barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation.
 This is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the sites [@barton2014].
@@
 
-# Theory/calculation
-
 Higher frequency of secondarily worked lithics
-
-# Results
 
@@
 ```
+# Results
+
+Figure \@ref(fig:cor) displays the CA using the lithic count data.
+The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
+The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
+Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
+
+The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
+It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves.
+The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+These findings are in large part in line with previous research [e.g.
+@bang-andersen2003], although highlight the importance of non-flint materials for the production of projectiles.
+
+The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
+The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
+Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
+This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
+@eigeland2015, 370, @glørstad2011].
 
@@
 
-Figure \@ref(fig:cor) displays the CA using the lithic count data.
-The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
-The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
-Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
-
-The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
-It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves.
-The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-These findings are in large part in line with previous research [e.g.
-@bang-andersen2003], although possibly highlighting the importance of the non-flint materials to a larger extent than what has traditionally been the case.
-
-The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
-While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
-Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
-This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
-@eigeland2015, 370, @glørstad2011].
-
 In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
@@
 base don the assumption that, this measure and has been shown to display a meaningful measure across a wide range of temporal, geographical and cultural contexts.
+
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020.
+There is a general temporal transition from older to younger sites along the second dimension of the plot, which is mainly defined by a negative correlation between the variables volumetric density of lithics and relative frequency of secondarily worked lithics, as well as relative frequency of cores.
+Almost orthogonal to this is the negative correlation between relative frequency of chips (artefacts of a size \>) and blanks (flakes, blades, microblades and fragments).
+While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear not to be temporally determined.
+The investigation performed by @Bicho2020 indicated that in their data the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns.
+However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, when combined with the fact that the relative frequency of cores align with the expectation of the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here.
+If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+To explore this proposition further, these two variables are subjected to more detailed scrutiny below.
 
@@
 wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
-wabicorflint <- cor(log10(flint2$flint_dens), log10(flint2$secondary_freq))
+# Langangen Vestgård 2 does not have any retouched lithics
+wabiflint <- filter(flint2, secondary_freq > 0)
+wabicorflint <- cor(log10(wabiflint$flint_dens), log10(wabiflint$secondary_freq))
+
+# Correlation with mean site date
+secfreqcor <- cor(lithics2$mean_date, lithics2$secondary_freq)
+denscor <- cor(lithics2$mean_date, lithics2$lithic_dens)
+
+secfreqcorflint <- cor(wabiflint$mean_date, wabiflint$secondary_freq)
+denscorflint <- cor(wabiflint$mean_date, wabiflint$flint_dens)
+
 ```
@@
 
-```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 7, fig.width = 10}
+```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity."}
+# , fig.height = 7, fig.width = 10
+
 # Rename the transformed variables again to avoid clutter in the plot. 
@@
 ```
-
-Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020.
-There is a general temporal transition from older to younger sites along the second dimension of the plot, which is mainly defined by a negative correlation between the variables volumetric density of lithics and relative frequency of secondarily worked lithics, as well as relative frequency of cores.
-Almost orthogonal to this is the negative correlation between relative frequency of chips (artefacts of a size \>) and blanks (flakes, blades, microblades and fragments).
-While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear not to be temporally determined.
-The investigation performed by @Bicho2020 indicated that in their data the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns.
-However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, when combined with the fact that the relative frequency of cores align with the expectation of the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here.
-If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
-To explore this proposition further, these two variables are subjected to more detailed scrutiny below.
 
@@
 
-Figure \@ref(fig:wabi) clearly illustrates the negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
-Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology.
-As variable raw-material availability and workability can also impact these measures [@manninen2014].
-Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies.
-However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically.
-Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly temporally contingent, the general pattern is the same.
+Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r = *`r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+The correlation is stronger between the mean site age and RFSL (*r = *`r round(secfreqcor, 2)`), than between mean site age and VDL (*r = *`r round(denscor, 2)`).
+As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data.
+While the negative correlation is slightly less marked when only the flint data is considered (*r = *`r round(wabicorflint, 2)`), the general pattern is the same.
+The relationship between mean site age and RFSF is even stronger (*r = *`r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the VDF is not temporally contingent (*r = *`r round(denscorflint, 2)`).
+This could perhaps indicate that when non-flint materials make up a higher share of the assemblages, as is especially the case for some of the later Mesolithic sites, expediency might to a larger extent be associated with the non-flint component of the assemblages rather than the flint material.
+This point is also returned to below where the temporal dimension of the relationship is explored further.
 
@@
 The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
+
 w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
@@
 
-To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
+To get at this temporal trend a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
 The values for artefact density was then made negative to reflect its relationship with degree of curation.
@@
 @crema2012; @baxter2016; @orton2017].
-A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites \@ref(fig:dist).
+A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist).
 For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
-This simulation was repeated 1000 times, the results of which is visualised in \@ref(fig:simage).
+This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot).
 Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE.
-This stabilises around 7000 BCE without any further major fluctuations for the rest of the Mesolithic.
-It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.
+This stabilises around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic.
+The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.
 
@@
 
-```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) all lithics in aggregate, and B) only the flint data.
-The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability distributions given in Figure 1B.
-A LOESS curve has been fit to the distribution for each of the 1000 simulation runs.
-Each simulation run is plotted with some transparency.
-Sites mentioned in the text are coloured.", fig.height = 10, fig.width = 12}
+```{r simage, cache = TRUE}
 
@@
 }
+```
+
+```{r simplot, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) all lithics, and B) flint.
+The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B.
+A LOESS curve has been fit to the distribution for each of the 1000 simulation runs.
+Each simulation run is plotted with some transparency.
+Sites mentioned in the text are given colour.", fig.height = 14, fig.width = 12}
 
@@
   scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
-  labs(title = "A", x = "Cal. BCE", y = "Curation index (flint)") +
+  labs(title = "B", x = "Cal. BCE", y = "Curation index (flint)") +
   theme_classic() +
@@
 
-The results of the CA does appear to align well with previous research that has been based on more informal treatment of the assemblages as a whole, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time.
+The results of the CA does appear to align well with previous research that has been based on more informal treatment of the assemblages as a whole, and indicates that the employed artefact categories are capturing a temporal component.
 However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
@@
 Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
-This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural significance of the employed units of analysis unclear.
-The results of the CA can, however, been used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions, to nuance or explain discrepencies in htis data.
+This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear.
+The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.
+
+While the relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here arrived at by means of an exploratory approach, some inferential merit can be achieved by invoking what has been termed consilience.
+Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30].
+Thus, while by no means a firm , the clear overlap in results presented here and those repeatedly reported by does to the general applicability of the measures, and offers an initial indication that they are capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
+While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all lithics ranges from `r round(max(lithics2$lithic_dens))` lithics per m^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia.
+Similarly, the frequency of retouched artefacts ranges from `r round(min(lithics2$secondary_freq)*100, 1)`--`r round(max(lithics2$secondary_freq)*100, 1)` % which is relatively low when compared to the same data.
+This could, in sum, indicate that the coastal Mesolithic sites of South-Eastern Norway have generally been associated with a low degree of mobility when compared to these context, a proposition that would be in line with both ethnographic and archaeological research that highlights the possible role marine resources can play.
 
@@
 This pattern is evident in both the flint data and when all lithics are treated in aggregate.
-The variation in degree of curation when all lithics are considered, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
+The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
 The five sites that have values on the curation index below c.
@@
 This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint.
-Also referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages.
+Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages.
 As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
@@
 A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].
-However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
+However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
 
@@
 
+In my view this alignes well with the framework of @clark2017, suggesting that shifts in mobility patterns has caused this empirical pattern.
+However, the role of non-flint material, variability of flint quality.
+Furthermore, precisely how this mobility pattern has.
+Here Binfords model of briefly invoked, but there exists of course a wide range of possible configurations and mobility strategies that have subsistence data, site locations
+
 The results of considering variables that are have been associated with mobility patterns gave rise to two competing hypotheses.

### 2021-05-20 analysis/paper/paper.Rmd
Do a CA using only the flint data

@@
 ```
+
 # Results
@@
 It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves.
-The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-These findings are in large part in line with previous research [e.g.
-@bang-andersen2003], although highlight the importance of non-flint materials for the production of projectiles.
+The non-flint material on the earliest, or among the earliest sites, appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+The single red point in the upper right quadrant of the row plot is Nedre Hobekk 2, which represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
+The use of metarhyolite and smoky quartz for the production of axes is present at other sites as well, but is evidently not as prominent a part of these assemblages.
+It should also be noted that the later Mesolithic sites that are also suppressing this pattern somewhat.
+In sum, the findings are in large part in line with previous research [e.g.
+@bang-andersen2003], although perhaps highlighting the relative importance of non-flint materials in south-eastern which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184]
 
 The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
-The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites.
+The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these latsites.
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
@@
 ```
-
-In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
-The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2.
 
@@
 
-```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity."}
-# , fig.height = 7, fig.width = 10
+```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 8}
+# fig.width = 10
 
@@
 
-Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r = *`r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
-The correlation is stronger between the mean site age and RFSL (*r = *`r round(secfreqcor, 2)`), than between mean site age and VDL (*r = *`r round(denscor, 2)`).
+Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+The correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`).
 As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data.
-While the negative correlation is slightly less marked when only the flint data is considered (*r = *`r round(wabicorflint, 2)`), the general pattern is the same.
-The relationship between mean site age and RFSF is even stronger (*r = *`r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the VDF is not temporally contingent (*r = *`r round(denscorflint, 2)`).
-This could perhaps indicate that when non-flint materials make up a higher share of the assemblages, as is especially the case for some of the later Mesolithic sites, expediency might to a larger extent be associated with the non-flint component of the assemblages rather than the flint material.
-This point is also returned to below where the temporal dimension of the relationship is explored further.
+While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same.
+The relationship between mean site age and RFSF is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the VDF is not temporally contingent (*r =* `r round(denscorflint, 2)`).
+As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.
 
@@
 
-To get at this temporal trend a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
+To get at this temporal trend, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
 The values for artefact density was then made negative to reflect its relationship with degree of curation.
@@
 Thus, while by no means a firm , the clear overlap in results presented here and those repeatedly reported by does to the general applicability of the measures, and offers an initial indication that they are capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
-While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all lithics ranges from `r round(max(lithics2$lithic_dens))` lithics per m^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia.
+While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all lithics ranges from `r round(max(lithics2$lithic_dens))` lithics per m\^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia.
 Similarly, the frequency of retouched artefacts ranges from `r round(min(lithics2$secondary_freq)*100, 1)`--`r round(max(lithics2$secondary_freq)*100, 1)` % which is relatively low when compared to the same data.
@@
 As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
-The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps.
+Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion seems to underplay.
+As the measure and general tendencies appear relatively robust to this effect, this it not explored further here.
+However, one possible approach could be to weight the curation index by proportion of non-flint material in the assemblages.
+the exlusion of this material The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps.
 The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel out.

### 2021-05-22 analysis/paper/paper.Rmd
Fixed temporal plot after flipping dates

@@
   # Exclude sites with less than 100 artefacts
-  filter(artefact_count > 100)
+  filter(artefact_count > 100) %>% 
+  arrange(mean_date) %>% 
+  rownames_to_column(var = "no")
+
 ```
@@
 # Assemble plot of date ranges
-temp_dist <- ggplot(datedata, aes(x = mean_date, y = reorder(site, mean_date))) +
+temp_dist <- ggplot(datedata, aes(x = -mean_date, y = reorder(no, -as.numeric(no)))) +
   geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
-                   yend = site, colour = excavation_project.x,
+                   yend = no, colour = excavation_project.x,
                    fill = excavation_project.x), size = 1.8) +
-  geom_ridgeline(aes(x = date_grid * -1, height = probabilities * 150,
+  geom_ridgeline(aes(x = date_grid, height = probabilities * 150,
                      fill = excavation_project.y, 
                      colour = excavation_project.x)) +
-  scale_x_reverse() +
   scale_fill_manual(values = project_colors,
@@
                             "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
-  scale_colour_manual(values = project_colors) +
+  scale_colour_manual(values = project_colors,
+                    labels = c("E18 Bommestad-Sky","E18 Brunlanes",
+                            "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
   labs(title = "B", fill = "Excavation project") +
@@
   theme_classic() +
-  theme(axis.title.y=element_blank(),
-        axis.text.y=element_blank(),
-        axis.ticks.y=element_blank(),
+  theme(axis.title.y = element_blank(),
+        axis.ticks.y = element_blank(),
         legend.position = "bottom") +
@@
   # Make site names the rownames
-  column_to_rownames(var = "site") %>% 
+  column_to_rownames(var = "n") %>% 
   # Deselect a few of the columns not be used
-  select(-n, -excavation_project, -excavated_m3, -dating_method)
+  select(-excavation_project, -excavated_m3, -dating_method)
+
+flint <- lithics %>% select(!(ends_with("other")))
 ```
@@
 # so is excluded here
-calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count, 
+calithics <- CA(sqrt(select(lithics, -site, -mean_date, -artefact_count, -flint_count, 
                        -nonflint_count, -reported_start_bc, -reported_end_bc,
-                       -askeladden_id),
+                       -askeladden_id)),
                 graph = FALSE)
 
-# Retrieve eigenvalues
+# Retrieve eigenvalues (for the text)
 eigval <- get_eigenvalue(calithics)
@@
 
-# Find date to hold the central tendency date
+# Find date to hold the central tendency date (for colouring row plot)
 centdate <- (sum(range(lithics$mean_date)) / 2) / 1000
@@
 Figure \@ref(fig:cor) displays the CA using the lithic count data.
+The square root has been taken of the data before running the analysis to reduce the impact of extreme values that were making the main cluster of sites in the bottom right of \@ref(fig:cor)A difficult to discern [cf.
+@lockyear2000].
 The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
@@
 The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
-It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves.
+It is also interesting that these sites to larger extent are characterised by core fragments, both in flint and non-flint materials, than cores themselves.
 The non-flint material on the earliest, or among the earliest sites, appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-The single red point in the upper right quadrant of the row plot is Nedre Hobekk 2, which represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
+The single yellow point in the upper right quadrant of the row plot is Nedre Hobekk 2, which represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
 The use of metarhyolite and smoky quartz for the production of axes is present at other sites as well, but is evidently not as prominent a part of these assemblages.
-It should also be noted that the later Mesolithic sites that are also suppressing this pattern somewhat.
-In sum, the findings are in large part in line with previous research [e.g.
+In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
 @bang-andersen2003], although perhaps highlighting the relative importance of non-flint materials in south-eastern which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184]
@@
 The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
-The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these latsites.
+The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these sites.
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
@@
                      repel = TRUE, shape = 21, pointsize = 2.9) +
-  scale_fill_gradient2(name = "ka BP", low="yellow", mid="orange", 
+  scale_fill_gradient2(name = "Mean age BCE", low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
 # Plot variables
-plt2 <- fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "B")
+plt2 <- fviz_ca_col(calithics, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B")
 
@@
 
-Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of any analysis is not simply the result of the employed units of analysis, and an overly naive trust in the fact that these units reflect any behaviourally meaningful categories.
+Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of an analysis is not simply the result of the employed units of analysis, and an overly naive trust in the fact that these units reflect any behaviourally meaningful categories.
 This realisation has led to a wide range of outside Scandinavian Mesolithic research, meaning a range of empirical measures associated with various mobility patterns are available.
@@
                        "chisel_other", "club_hatchet_other", 
-                       "grinding_slab_stone", "projectile_point_other", 
+                       "grinding_slab_stone", "projectile_other", 
                        "knife_other", "scraper_other", "drill_other", 
@@
                 axe_debitage_other, chisel_other, club_hatchet_other, 
-                grinding_slab_stone, projectile_point_other, knife_other,
+                grinding_slab_stone, projectile_other, knife_other,
                 scraper_other, drill_other, microblade_retouch_other, 
@@
   # Transform skewed variables and reorder
-  mutate("Mean date cal. BCE*-1" = mean_date * -1,
           "log10(lithic_dens)" = log10(lithic_dens),
@@
          "core_freq^(1/3)" = core_freq^(1/3), .keep = "unused") %>% 
-  relocate("Mean date cal. BCE*-1",
+  relocate("mean_date",
     "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
@@
 lithicspca <- lithicstr %>% 
-  mutate(mean_date =  `Mean date cal. BCE*-1` * -1, .keep = "unused") %>% 
   rename(
+   mean_date = "Mean date cal. BCE",
    lithic_dens = "log10(lithic_dens)",
@@
        y = "Curation index") +
-  scale_x_reverse() + 
   geom_smooth(method = "loess", col = "red") +

### 2021-05-24 analysis/paper/paper.Rmd
Added a CA excluding the non-flint material

@@
 
-This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
+This study employs multivariate exploratory statistics to analyse lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
 This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time.
@@
 
-```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
-Radiocarbon probability density functions are given as the posterior sum.
-Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
-These follow the original reports.", fig.height = 10, fig.width = 12}
+```{r 1dist, fig.height = 10, fig.width = 12}
 # Create a bounding box around site points to limit the overview map
@@
 
-(overview + sitemap) /
-(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))
+(overview + sitemap) 
+# (temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))
 
@@
 
-A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@bjerck2010; @glørstad2010].
-Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020].
+```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
+Radiocarbon age determintations are given as the sum of the posterior density esimates.
+Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
+These follow the original reports.", fig.height = 9, fig.width = 12}
+temp_dist
+```
+
+A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@bjerck2010].
+Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g.
+@åstveit2018; @breivik2018; @solheim2020].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
@@
 However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
-In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014].
+In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year towards the end of the Mesolithic [@sørensen2014].
 Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.
 
-The data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
+The dates associated with the sites are given.
+Where radiocarbon age determinations are available these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam].
+The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function.
+The resulting posterior density estimate was then summed for each site.
+
+The lithic data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
 While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions.
@@
 @berg-hansen1999].
-This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46].
-An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites.
-However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here [@eigeland2015, 45-53].
-Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46].
+Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53].
+Furthermore, the various non-flint raw materials that have been lumped together here have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
 Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories.
@@
 @glørstad2011; @eigeland2015, 370].
-
-Studies concerned with chronological changes in the composition of lithic assemblages in Norway have typically had a focus on chronological trends associated with formal tool types [e.g.
+Finally, while factors such as landscape changes through shoreline displacement still can have led to variable raw-material availability on the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.
+
+Studies concerned with chronological changes in the composition of lithic assemblages in Norway have typically had a focus on formal tool types [e.g.
 @helskog1976; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g.
@@
 These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
-To my knowledege, only a single study in southern Norway, concerned with the analysis of eight Middle Mesolithic assemblages, has involved the use of a multivariate quantitative framework to structure the analysis [@solheim2013a].
-In sum, therefore, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that makes it difficult to consistently balance the comparison of a larger number of assemblages.
+To my knowledege, only a single study dealing with Mesolithic material in southern Norway has involved the use of a multivariate quantitative framework to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a].
+In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that are difficult to up-scale and consistently balance in the comparison of a larger number of assemblages.
 
@@
 However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported on here, the data and employed R programming script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+While only parts of this process can sensibly be reported on here, the data and employed R programming script is freely available as a research compendium following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
 However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves what Binford termed a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
-The process can still provide the identification of empirical patterns along the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses and lay the foundation for a deductive research agenda for which clear test implications can be derived [@clark2009, 29].
-
-The first part of the analysis involved employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-The purpose of this exercise was to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods.
+The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses.
+This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].
+
+The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies --- studies that have employed more informally driven methods.
 This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful.
@@
 @barton1991; @dibble1995].
-Consequently, several artefact categories have here been collapased for the CA.
+Consequently, several artefact categories have here been collapsed for the CA.
 This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
@@
 As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below].
-The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1979].
+The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1979].
 An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement.
@@
 
-However, following not least from the ambiguous definition first put forward by @binford1979, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g.
+However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g.
 @bamforth1986; @shott1996; @surovell2009, 9--13].
@@
 @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation.
-This is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the sites [@barton2014].
-In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern.
+This is in turn mainly to follow from the accumulated nature of land-use, procurement and mobility patterns associated with the sites [@barton2014].
+In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern.
 From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
@@
 The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].
-
 These are the main measure employed here.
-However, as these measures are argued to predominantly be determined by mobility patterns, some additional variables suggested by @Bicho2020 to be related to mobility patterns are also explored.
-This is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
+However, as these measures are argued to predominantly be determined by mobility patterns, the investigation of @Bicho2020 has also led to the inclusion of relative frequency of chips in the analysis as well, as they argue that this is an important.
+The analysed variables are therefore relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched and ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per m³; as well as relative frequency of chips, defined as artefacts with size \< 0.1 cm; relative frequency of cores which are simply all artefacts classified as cores in the original reports, as well as relative frequency blanks, here constituted by the categories flakes, blades, micro-blades and fragments.
+The analysis is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
 @baxter1994, 100].
 
-A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here.
+A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here.
 This pertains to number of site features such as fireplaces, cooking pits and dwelling structures.
@@
   # Make site names the rownames
-  column_to_rownames(var = "n") %>% 
+  column_to_rownames(var = "no") %>% 
   # Deselect a few of the columns not be used
@@
 # so is excluded here
-calithics <- CA(sqrt(select(lithics, -site, -mean_date, -artefact_count, -flint_count, 
+calithics <- CA(select(lithics, -site, -mean_date, -artefact_count, -flint_count, 
                        -nonflint_count, -reported_start_bc, -reported_end_bc,
-                       -askeladden_id)),
+                       -askeladden_id),
+                graph = FALSE)
+
+caflint <- CA(select(flint, -site, -mean_date, -artefact_count, -flint_count, 
+                       -nonflint_count, -reported_start_bc, -reported_end_bc,
+                       -askeladden_id),
                 graph = FALSE)
@@
 Figure \@ref(fig:cor) displays the CA using the lithic count data.
-The square root has been taken of the data before running the analysis to reduce the impact of extreme values that were making the main cluster of sites in the bottom right of \@ref(fig:cor)A difficult to discern [cf.
-@lockyear2000].
 The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the clear horse-shoe shaped structure in the column plot [e.g.
+@lockyear2000], combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
@@
 The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
-It is also interesting that these sites to larger extent are characterised by core fragments, both in flint and non-flint materials, than cores themselves.
-The non-flint material on the earliest, or among the earliest sites, appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-The single yellow point in the upper right quadrant of the row plot is Nedre Hobekk 2, which represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite.
-The use of metarhyolite and smoky quartz for the production of axes is present at other sites as well, but is evidently not as prominent a part of these assemblages.
+It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves.
+The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014] .
+The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
 In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
-@bang-andersen2003], although perhaps highlighting the relative importance of non-flint materials in south-eastern which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184]
-
-The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles.
-The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these sites.
+@bang-andersen2003; @damlien2018b], although perhaps highlighting the relative importance of non-flint materials for some of the sites in south-eastern Norway which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184].
+
+The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages.
+Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these sites.
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
@@
 @eigeland2015, 370, @glørstad2011].
+
+With the exception of projectiles, microburins, macro tools and debitage from their production, which are more characteristic of earlier sites, the flint material is relatively similar across the sites, and most of the variation in the data is accounted for by variation in non-flint material.
+For the earliest sites this largely pertains to the presence of non-flint projectiles and core fragments, while it for later phases is the non-flint macro tools and associated debitage that is the main driver of the pattern.
+The comparatively limited impact of the flint material can possibly in part be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and due to technological nuances in the flint material that is temporally contingent but which are not recorded during a regular classification of the material.
+While the former pertains to the trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well, but does speak to the impact raw-material workability has for the composition of the assemblages.
 
@@
 ```{r cor, fig.cap = "Correspondence analysis using artefact count data. A) Object map, B) Variable map.", fig.height = 13}
-# Plot without site names tags
-plt1 <- fviz_ca_row(calithics, geom = "point", col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+# Plot sites
+plt1 <- fviz_ca_row(calithics, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
   theme(legend.position = "bottom")
-#guide(fill = guide_legend(reverse = TRUE)) # Reverses the legend but makes it
-                                            # discrete
+
+plt2 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "Mean age BCE", low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  theme(legend.position = "bottom")
 
 # Plot variables
-plt2 <- fviz_ca_col(calithics, col.col = "contrib",
+plt3 <- fviz_ca_col(calithics, col.col = "contrib",
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B")
 
+plt4 <- fviz_ca_col(caflint, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B")
+
 plt1 /
-plt2
-
-```
-
-#
+plt3
+
+```
+
+```{r flintcor, fig.height = 13}
+plt2 /
+plt4 
+```
 
@@
 
-Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020.
-There is a general temporal transition from older to younger sites along the second dimension of the plot, which is mainly defined by a negative correlation between the variables volumetric density of lithics and relative frequency of secondarily worked lithics, as well as relative frequency of cores.
-Almost orthogonal to this is the negative correlation between relative frequency of chips (artefacts of a size \>) and blanks (flakes, blades, microblades and fragments).
-While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear not to be temporally determined.
-The investigation performed by @Bicho2020 indicated that in their data the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns.
-However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, when combined with the fact that the relative frequency of cores align with the expectation of the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here.
-If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures that have been linked to mobility patterns.
+There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot.
+The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure @ref(fig:var-contrib)).
+Almost orthogonal to this is the strong negative correlation between relative frequency of chips and blanks.
+While there is a slight tendency for blanks to be more associated with younger sites, frequency of chips appears to be largely independent of time.
+However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have an methodological explanation.
+Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips, one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and to separate from naturally fragmented stone during excavation and classification, leading to a over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material.
+While this is not necessarily the entire explanation behind the relationship, this does render it difficult to Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well.
+That is not to say that cores may not be indicative or related to mobility patterns, but may require further analysis beyond their simple classification as such [@kitchel2021].
+
+Thus, while some secondary expectations of the @clark2017 model does not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
+The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material.
+Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
 To explore this proposition further, these two variables are subjected to more detailed scrutiny below.
@@
 lithics2 <- sites %>% 
-  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
+  select(-no, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
           -askeladden_id) %>% 
@@
                 flake_retouch_other) / artefact_count,
+         nonflint_freq = nonflint_count / artefact_count,
          lithic_dens = artefact_count / excavated_m3) %>%
@@
 flint2 <- sites %>% 
-  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
+  select(-no, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
          -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
@@
 wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
-# Langangen Vestgård 2 does not have any retouched lithics
+
+# Langangen Vestgård 2 does not have any retouched flint
 wabiflint <- filter(flint2, secondary_freq > 0)
@@
 
-```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
+```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these.
+The mean age of the sites has also been included to visualise overall temporal trends.", fig.height = 9.5, fig.width = 10}
 # Transform the skewed variables - sqrt() is used where there are zero values,
@@
   # Transform skewed variables and reorder
+  mutate(
          "log10(lithic_dens)" = log10(lithic_dens),
          "log10(secondary_freq)" = log10(secondary_freq),
-         "core_freq^(1/3)" = core_freq^(1/3), .keep = "unused") %>% 
+         "core_freq^(1/3)" = core_freq^(1/3),
+         "nonflint_freq^(1/3)" = nonflint_freq^(1/3), .keep = "unused") %>% 
   relocate("mean_date",
-    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
+    "log10(lithic_dens)", "nonflint_freq^(1/3)", "blanks_freq", "log10(secondary_freq)",
     "chip_freq", "core_freq^(1/3)" 
@@
 
-```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 8}
+```{r pca-analysis, fig.cap = "PCA using variables that have been related to mobility patterns.
+Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 8}
 # fig.width = 10
@@
   rename(
-   mean_date = "Mean date cal. BCE",
    lithic_dens = "log10(lithic_dens)",
    secondary_freq = "log10(secondary_freq)",
-   core_freq = "core_freq^(1/3)"
+   core_freq = "core_freq^(1/3)",
+   nonflint_freq = "nonflint_freq^(1/3)"
   )
@@
                   scale.unit = TRUE, graph = FALSE)
-fviz_pca(pcalithics, geom.ind = "point", fill.ind = lithicspca$mean_date / 1000, 
+
+fviz_pca(pcalithics, fill.ind = lithicspca$mean_date / 1000, 
           repel = TRUE, col.var = "black", pointshape = 21,  pointsize = 2.9,
@@
 
-```{r var_contrib, fig.cap = "Contribution of variables to components.", fig.height = 3}
+```{r var-contrib, fig.cap = "Contribution of variables to the components of the PCA.", fig.height = 3}
 # The plots below investigates the contribution of the variables to the
@@
 Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
-The correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`).
+The linear correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`).
 As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data.
@@
 
-To get at this temporal trend, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
+To get more directly at this temporal trend, a curation index based on VDL and RFSL was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
 The values for artefact density was then made negative to reflect its relationship with degree of curation.
@@
 Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE.
-This stabilises around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic.
-The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.
+This stabilises around 7500 BCE and remains stable without any major fluctuations for the rest of the Mesolithic.
+The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation associated mobility patterns.
 
@@
 
-```{r simplot, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) all lithics, and B) flint.
+```{r simplot, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) All lithics, and B) Flint.
 The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B.
@@
   scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
-  labs(title = "A", x = "Cal. BCE", y = "Curation index (all lithics)") +
+  labs(title = "A", x = "Cal. BCE", y = "Curation index (All lithics)") +
   theme_classic() +
@@
   scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
-  labs(title = "B", x = "Cal. BCE", y = "Curation index (flint)") +
+  labs(title = "B", x = "Cal. BCE", y = "Curation index (Flint)") +
   theme_classic() +
@@
 However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
-The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites.
+The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology.
 Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
@@
 
-While the relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here arrived at by means of an exploratory approach, some inferential merit can be achieved by invoking what has been termed consilience.
+While the relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here identified by means of an exploratory approach, some inferential merit can be achieved by invoking what has been termed consilience.
 Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30].
-Thus, while by no means a firm , the clear overlap in results presented here and those repeatedly reported by does to the general applicability of the measures, and offers an initial indication that they are capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
-While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all lithics ranges from `r round(max(lithics2$lithic_dens))` lithics per m\^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia.
+Thus, while no firm , the clear overlap in results presented here and those repeatedly reported by does speak to the general applicability of the measures, and offers an initial indication that they are capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
+While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all assemblages treated here ranges from `r round(max(lithics2$lithic_dens))` lithics per m\^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia.
 Similarly, the frequency of retouched artefacts ranges from `r round(min(lithics2$secondary_freq)*100, 1)`--`r round(max(lithics2$secondary_freq)*100, 1)` % which is relatively low when compared to the same data.
@@
 
-The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises from around 7000 BCE for the rest of the Mesolithic.
+The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises from around 7500 BCE for the rest of the Mesolithic.
 This pattern is evident in both the flint data and when all lithics are treated in aggregate.
-The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
+Furthermore, the variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
 The five sites that have values on the curation index below c.
@@
 Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages.
-As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
-Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion seems to underplay.
-As the measure and general tendencies appear relatively robust to this effect, this it not explored further here.
-However, one possible approach could be to weight the curation index by proportion of non-flint material in the assemblages.
+As can be seen in Figure \@ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
+Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion would underplay.
+As the measure and general tendencies appear relatively robust to this effect, this it not explored further here, however, one possible approach could be to weight the curation index by proportion of non-flint material in the assemblages.
+Another case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].
+However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
+
+While there are certainly nuances and idiosyncrasies in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust.
+It seems clear that there is a marked drop in curation from just before 8000 BCE, which stabilises around 7500 BCE.
+This would correspond well with chronological where the end of the Early Mesolithic, or flake axe phase, is set to last until c.
+8200 BCE.
+The following phase linked to the production of microliths is in this framework set to last to around 7500 BCE, signifying the start of the production of the non-flint ground-pecked or round butted adzes.
+Later chronological transitions suggested at c.
+5600 and c.
+4500 BCE, are not reflected in fluctuations in the curation index.
+While this can be traced in the CA with the reduced importance of the non-flint categories, this could indicate that these changes are not related mobility patterns.
+
 the exlusion of this material The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps.
@@
 The nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential-logistic continuum.
-A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].
-However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
-
-While it is possible the is to consider these patterns at a population level.
-From this perspective it seems clear that there is a marked drop in curation from just before 8000 BCE, which stabilises around 7000 BCE.
-This would correspond well with chronological where the end of the Early Mesolithic, or flake axe phase, is set to last until c.
-8200 BCE.
-The following phase linked to the production of microliths is in this framework set to last to around 7000 BCE, signifying the start of the production of the non-flint ground-pecked or round butted adzes.
-Later chronological transitions suggested at c.
-5600 and c.
-4500 BCE, are not reflected in fluctuations in the curation index.
-While this can be traced in the CA with the reduced importance of the non-flint categories, this could indicate that these changes are not related mobility patterns.
 

### 2021-05-26 analysis/paper/paper.Rmd
Going through some overall implications of the findings

@@
 # Assemble plot of date ranges
-temp_dist <- ggplot(datedata, aes(x = -mean_date, y = reorder(no, -as.numeric(no)))) +
+temp_dist <- ggplot(datedata, aes(x = -mean_date, y = reorder(no, -as.numeric(no)), label = no)) +
   geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
@@
 
-A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@bjerck2010].
+A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g.
+@bjerck2010].
 Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g.
@@
 
-The dates associated with the sites are given.
-Where radiocarbon age determinations are available these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam].
+The `r nrow(sites)` sites analysed here have been dated by reference to relative sea-level change, typology and/or radiocarbon dates.
+Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators.
+Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam].
 The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function.
-The resulting posterior density estimate was then summed for each site.
-
-The lithic data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
+The resulting posterior density estimates were then summed for each site.
+Radiocarbon data is provided in the supplementary data, and has also been collated and reported by @solheim2020a.
+
+The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
 While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions.
@@
 @damlien2016; @berg-hansen2017; @solheim2020].
-Studies that have involved the entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018; @breivik2020], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
+Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
 These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
-To my knowledege, only a single study dealing with Mesolithic material in southern Norway has involved the use of a multivariate quantitative framework to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a].
-In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that are difficult to up-scale and consistently balance in the comparison of a larger number of assemblages.
+To my knowledege, only a single study dealing with Mesolithic material in southern Norway has involved the use of a multivariate quantitative framework, which was used to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a].
+In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of assemblages.
 
@@
 
-The relatively constrained geographical distribution of the analysed sites, the limited temporal spread over which they where investigated, as well as the methodological equivalency across projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
+The relatively constrained geographical distribution of the analysed sites, the limited temporal spread over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
 However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported on here, the data and employed R programming script is freely available as a research compendium following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
 However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves what Binford termed a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
@@
 An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement.
-Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979].
-
-However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed, and no single definition has ever been reached [e.g.
+Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.
+
+However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed, and no single definition has ever been reached [e.g.
 @bamforth1986; @shott1996; @surovell2009, 9--13].
@@
 Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1999 and collegues [e.g.
-@barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages provide consistent results across a wide range of chronological and cultural context, and is a measure taken to reflect degree of curation.
-This is in turn mainly to follow from the accumulated nature of land-use, procurement and mobility patterns associated with the sites [@barton2014].
+@barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages, to lithic technology associated with both Neanderthals and modern humans [@kuhn2015].
+This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
 In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern.
@@
 The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].
-These are the main measure employed here.
-However, as these measures are argued to predominantly be determined by mobility patterns, the investigation of @Bicho2020 has also led to the inclusion of relative frequency of chips in the analysis as well, as they argue that this is an important.
-The analysed variables are therefore relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched and ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per m³; as well as relative frequency of chips, defined as artefacts with size \< 0.1 cm; relative frequency of cores which are simply all artefacts classified as cores in the original reports, as well as relative frequency blanks, here constituted by the categories flakes, blades, micro-blades and fragments.
+However, as these measures are argued to predominantly be determined by mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns [e.g.
+@Bicho2020; @kitchel2021].
+The use of local non-flint material has been taken to indicate reduced mobility and better familiarity with local .
+In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per m³; relative frequency of chips, defined as artefacts with size \< 0.1 cm; relative frequency of cores, which are simply all artefacts classified as cores in the original reports; relative frequency blanks, here constituted by the artefact categories flakes, blades, micro-blades and fragments; and finally relative frequency of non-flint material.
 The analysis is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
@@
 
-A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here.
-This pertains to number of site features such as fireplaces, cooking pits and dwelling structures.
-Feature count has been omitted as taphonomic loss is likely to have led to a chronological bias in their preservation.
+A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here.
+For the assemblage data itself this especially pertains to diversity in tool-types [e.g.
+@Bicho2020; @canessa2021] which has been omitted in light of the previously mentioned Frison effect.
+It also pertains to site specific aspects such as number of features which has been omitted as taphonomic loss is likely to have led to a chronological bias in their preservation that is difficult to account for.
 Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
-This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied.
+This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017].
 This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].
-
-Here retouch i At logistical basecamps, the extended time spent in a single location would imply a better control of raw material availability, and allow for longer trips to retrieve these.
-Higher availability of lithic raw materials should in turn reduce the necessity for extensive retouch, meaning the material expectation for logistical basecamps would be relatively high density of lithics and a low relative frequency of retouched artefacts.
-From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
-
-Higher frequency of secondarily worked lithics
 
@@
 
-flint <- lithics %>% select(!(ends_with("other")))
+flint <- lithics %>% select(!(ends_with("other")), -grinding_slab_stone, -hammerstone)
 ```
@@
 The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014] .
-The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
+Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
+However, as the site has been quite heavily impacted by modern disturbances The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
 In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
@@
 
-With the exception of projectiles, microburins, macro tools and debitage from their production, which are more characteristic of earlier sites, the flint material is relatively similar across the sites, and most of the variation in the data is accounted for by variation in non-flint material.
-For the earliest sites this largely pertains to the presence of non-flint projectiles and core fragments, while it for later phases is the non-flint macro tools and associated debitage that is the main driver of the pattern.
-The comparatively limited impact of the flint material can possibly in part be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and due to technological nuances in the flint material that is temporally contingent but which are not recorded during a regular classification of the material.
+As most of the variation in the data is accounted for by variation in non-flint material a second CA was run using only the flint data \@ref(fig:flintcor).
+There is clear temporal patterning in the flint data as well.
+This most marked for the very earliest sites which are pulled away from the main cluster especially the earliest sites are pulled away from the main cluster as the presence of projectiles, microburins, macro tools as well as the debitage from their production is characteristic of these sites.
+The temporal transition in the main cluster is more gradual, but clearly present, and is driven by a larger proportionof blades, flakes and small tools in the earliest assemblages.
+The comparatively limited impact of the flint material can in part be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and technological nuances in the flint material is likely temporally contingent but not recorded during a regular classification of the material.
 While the former pertains to the trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well, but does speak to the impact raw-material workability has for the composition of the assemblages.
@@
 
-```{r cor, fig.cap = "Correspondence analysis using artefact count data. A) Object map, B) Variable map.", fig.height = 13}
+```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Object map, B) Variable map. Note the differences on the axes between plots.", fig.height = 13}
 # Plot sites
@@
 
-plt2 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+# Plot variables
+plt2 <- fviz_ca_col(calithics, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B")
+
+# Plot sites for the flint data
+plt3 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 
-# Plot variables
-plt3 <- fviz_ca_col(calithics, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B")
-
+# Variables for the flint data
 plt4 <- fviz_ca_col(caflint, col.col = "contrib",
@@
 plt1 /
-plt3
-
-```
-
-```{r flintcor, fig.height = 13}
-plt2 /
+plt2
+
+```
+
+```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Object map, B) Variable map. Do not the differences on the axes between plots."}
+plt3 /
 plt4 
@@
 Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE.
-This stabilises around 7500 BCE and remains stable without any major fluctuations for the rest of the Mesolithic.
+This has stabilised by around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic.
 The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation associated mobility patterns.
@@
 ```{r simplot, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) All lithics, and B) Flint.
-The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B.
+The temporal uncertainty is handled by means of a simulation approach where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B.
 A LOESS curve has been fit to the distribution for each of the 1000 simulation runs.
@@
 
-The results of the CA does appear to align well with previous research that has been based on more informal treatment of the assemblages as a whole, and indicates that the employed artefact categories are capturing a temporal component.
+The results of the CA does appear to align well with previous research that has been based on more informal treatment of the assemblages as a whole, and the employed artefact categories are clearly capturing a temporal component.
+This could indicate that the aggregation of artefact categories that was done here might have been overly conservative.
 However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
-The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology.
-Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely.
+The developments evident in the results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology.
+Unpicking and aligning the significance of these chronological patterns with any specific behavioural and technological dimensions using the CA results is, however, another task entirely.
 This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear.
@@
 
-While the relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here identified by means of an exploratory approach, some inferential merit can be achieved by invoking what has been termed consilience.
+The relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here identified by means of an exploratory approach, and is in part investigated further because they align with suggestions from previous research, clearly representing a possible case of confirmation bias.
+However, some inferential merit can be achieved by invoking what has been termed consilience.
 Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30].
-Thus, while no firm , the clear overlap in results presented here and those repeatedly reported by does speak to the general applicability of the measures, and offers an initial indication that they are capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
-While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all assemblages treated here ranges from `r round(max(lithics2$lithic_dens))` lithics per m\^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia.
+Thus, the overlap in results presented here and those repeatedly reported from a range of different context does speak to the general applicability of the measures, and offers an initial indication that they might be capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
+While it is not within the scope of this paper to consider these comparisons to any wider extent, and so they must be considered highly speculative, the VDL for all assemblages treated here has a mean of `r round(mean(lithics2$lithic_dens))` lithics per m\\\^3 , with a standard deviation of `r round(sd(lithics2$lithic_dens))`.
+This appears to be relatively high when compared to Late Pleistocene sites from western Eurasia [@barton2011].
 Similarly, the frequency of retouched artefacts ranges from `r round(min(lithics2$secondary_freq)*100, 1)`--`r round(max(lithics2$secondary_freq)*100, 1)` % which is relatively low when compared to the same data.
-This could, in sum, indicate that the coastal Mesolithic sites of South-Eastern Norway have generally been associated with a low degree of mobility when compared to these context, a proposition that would be in line with both ethnographic and archaeological research that highlights the possible role marine resources can play.
-
-The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises from around 7500 BCE for the rest of the Mesolithic.
+This could, in sum, indicate that the coastal Mesolithic sites of South-Eastern Norway have generally been associated with a low degree of mobility when compared to these context, a proposition that would be in line with both ethnographic and archaeological research that highlights the possible role marine resources can play in, as well as a general trend towards more sedentary life through the Holocene [e.g.
+@perreault2019, 42] .
+
+Returning focus to the present study, the curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE for the rest of the Mesolithic.
 This pattern is evident in both the flint data and when all lithics are treated in aggregate.
-Furthermore, the variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
+Furthermore, the variation in degree of curation in Figure \@ref(fig:simplot)A could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
 The five sites that have values on the curation index below c.
--0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern.
-That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility.
+-0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern [sensu @binford1980].
+That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility [@eigeland2014b].
 This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint.
 Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages.
-As can be seen in Figure \@ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
-Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion would underplay.
-As the measure and general tendencies appear relatively robust to this effect, this it not explored further here, however, one possible approach could be to weight the curation index by proportion of non-flint material in the assemblages.
-Another case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].
-However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
+As can be seen in Figure @ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
+Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion would likely lead to its underestimation.
+One possible approach could be to weigh the curation index by proportion of non-flint material in the assemblages.
+This is not explored further here, however, as the overall tendencies are relatively robust to this effect.
+Another case also worth commenting on is Langangen Vestgård 1, which, on the grounds of an overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].
+However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a comparable duration to those on contemporary sites, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole [cf.
+@barton2014].
 
 While there are certainly nuances and idiosyncrasies in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust.
-It seems clear that there is a marked drop in curation from just before 8000 BCE, which stabilises around 7500 BCE.
-This would correspond well with chronological where the end of the Early Mesolithic, or flake axe phase, is set to last until c.
-8200 BCE.
-The following phase linked to the production of microliths is in this framework set to last to around 7500 BCE, signifying the start of the production of the non-flint ground-pecked or round butted adzes.
-Later chronological transitions suggested at c.
-5600 and c.
-4500 BCE, are not reflected in fluctuations in the curation index.
-While this can be traced in the CA with the reduced importance of the non-flint categories, this could indicate that these changes are not related mobility patterns.
+It seems clear that there is a marked drop in curation starting some time just before 8000 BCE, which has stabilised around 7000 BCE.
+This corresponds well with a chronological framework where the end of the Early Mesolithic, or Flake axe phase, is set to c.
+8200 BCE [@reitan2016].
+Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns.
+This corresponds very well with the findings reported here, where the earliest phase is characterised by relatively high and uniform values on the curation index.
+The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade and subsequently axe technology, which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021].
+The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021].
+Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is a serious issue [@solheim2018a; @solheim2020a; @nielsen2021].
+In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE, signifying the onset of the production of non-flint ground-pecked or round butted adzes.
+Referring back to the increasing expediency in the curation data between c.
+8200 and 7000 BCE, the microlith phase could thus represent a transitional phase where new living practices where either propagating through societies in South-Eastern Norway, or possibly a complete replacement of the population might even have been taking place --- a process that in light of the curation data would have concluded around 7000 BCE.
+While non-flint macro tools are present in the material before this, the first of the sites treated here that are heavily dominated by their production is Gunnarsrød 6b, dated to c.
+`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[1]]*-1`--`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[2]]*-1` BCE [@carrasco2014].
+The curation data remains stable from around 7000 BCE through the next typological transition at c.
+5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet phase.
+The Nøstvet phase has traditionally been seen as representing the onset of more stable mobility patterns in south-eastern Norway, and has been explicitly linked to a expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit somewhat vaguely defined [@eigeland2015, 127--130].
+In recent years it has been suggested that the transition to a decrease in mobility and more varied mobility patterns can be traced back to the Middle Mesolithic.
+The curation index employed here clearly supports this notion, and suggests that the mobility patterns of the Nøstvet-phase was well established in preceding phases.
+
+4500 BCE by reduced importance of the non-flint categories, these are not which could indicate that these changes are not related to major shifts in land-use and mobility patterns.
+Some explanations to this apparent discrepancy can be that fluctuations in population numbers simply do not affect mobility patterns in the setting considered here, or the signal may be too weak to be picked up considering the small sample size.
 

### 2021-05-26 analysis/paper/paper.Rmd
Moved over to PDF for now

@@
 output: 
-    bookdown::html_document2:
+    bookdown::pdf_document2:
+      toc: false
       fig_caption: yes
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/jas_reports.csl" # Insert path for the bib-style
 ---
-
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
 
@@
   comment = "#>",
+  fig.pos = "!h",
   fig.width = 9,
@@
 To this end, the analysis follows two analytical avenues.
-The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports, which consequently assumes that these are behaviourally meaningful.
-The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.
-
-## Archaeological context and material
+The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports.
+The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [@clark2017], which has not seen application in this context before.
+
+# Archaeological context and material
 
@@
 
-The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \ref\@(fig:dist)A].
+The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \@ref(fig:dist)A.
 The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
@@
 This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations.
-Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum.
+Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum [@melvold2014c; @koxvold2017].
 As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g.
 @clark2006; @dibble2017].
-In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \ref\@(fig:dist)B].
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \@ref(fig:dist)B].
 Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
@@
 
-```{r 1dist, fig.height = 10, fig.width = 12}
+```{r 1dist, fig.height = 13, fig.width = 12}
 # Create a bounding box around site points to limit the overview map
@@
 
-(overview + sitemap) 
-# (temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))
+(overview + sitemap) /
+(temp_dist + theme(plot.margin = unit(c(0,135,0,135), "pt")))
 
@@
 Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
-These follow the original reports.", fig.height = 9, fig.width = 12}
+These follow the original reports.", fig.height = 9, fig.width = 12, eval = FALSE}
 temp_dist
@@
 Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam].
-The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function.
+The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021].
 The resulting posterior density estimates were then summed for each site.
@@
 
-## Methodology
+# Methodology
 
@@
 However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed, and no single definition has ever been reached [e.g.
-@bamforth1986; @shott1996; @surovell2009, 9--13].
-The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies, as well as raw-material quality and availability [e.g.
+@bamforth1986; @nash1996; @shott1996; @surovell2009, 9--13].
+The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies and raw-material quality and availability [e.g.
 @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].
@@
 @bamforth1986; @dibble1995; @shott2005].
-Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1999 and collegues [e.g.
-@barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages, to lithic technology associated with both Neanderthals and modern humans [@kuhn2015].
+Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
+@barton1999; @riel-salvatore2004; @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@kuhn2015].
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
@@
 This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
-The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades [@clark2017].
-However, as these measures are argued to predominantly be determined by mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns [e.g.
-@Bicho2020; @kitchel2021].
-The use of local non-flint material has been taken to indicate reduced mobility and better familiarity with local .
-In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per m³; relative frequency of chips, defined as artefacts with size \< 0.1 cm; relative frequency of cores, which are simply all artefacts classified as cores in the original reports; relative frequency blanks, here constituted by the artefact categories flakes, blades, micro-blades and fragments; and finally relative frequency of non-flint material.
+The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades.
+These variables and underlying logic consitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.
+However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
+@breivik2016a; @reitan2016; @Bicho2020; @kitchel2021].
+The use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181] .
+In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
 The analysis is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
-@baxter1994, 100].
+@baxter1994, 100, @Bicho2020].
 
@@
 @Bicho2020; @canessa2021] which has been omitted in light of the previously mentioned Frison effect.
-It also pertains to site specific aspects such as number of features which has been omitted as taphonomic loss is likely to have led to a chronological bias in their preservation that is difficult to account for.
+Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation that is difficult to account for.
 Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
@@
 The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the clear horse-shoe shaped structure in the column plot [e.g.
-@lockyear2000], combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data.
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
@@
 Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
-However, as the site has been quite heavily impacted by modern disturbances The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
+However, as the site had been quite heavily impacted by modern disturbances, this led [@eigeland2014, 124] to suggest that the material might have been compromised.
+This could explain its position as an outlier in the plot.
+The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
 In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
-@bang-andersen2003; @damlien2018b], although perhaps highlighting the relative importance of non-flint materials for some of the sites in south-eastern Norway which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184].
+@bang-andersen2003; @damlien2018b], although perhaps highlighting the relative importance of non-flint materials for some of the sites in south-eastern Norway which, while also present, is typically not as marked in other parts of southern Norway [e.g.
+@breivik2018, 184].
 
@@
 
-As most of the variation in the data is accounted for by variation in non-flint material a second CA was run using only the flint data \@ref(fig:flintcor).
-There is clear temporal patterning in the flint data as well.
-This most marked for the very earliest sites which are pulled away from the main cluster especially the earliest sites are pulled away from the main cluster as the presence of projectiles, microburins, macro tools as well as the debitage from their production is characteristic of these sites.
-The temporal transition in the main cluster is more gradual, but clearly present, and is driven by a larger proportionof blades, flakes and small tools in the earliest assemblages.
-The comparatively limited impact of the flint material can in part be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and technological nuances in the flint material is likely temporally contingent but not recorded during a regular classification of the material.
-While the former pertains to the trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well, but does speak to the impact raw-material workability has for the composition of the assemblages.
+As most of the variation in the data is accounted for by variation in non-flint material this suppresses and makes it difficult to discern patterns in the flint data.
+A second CA was therefore run excluding the non-flint material [Figure @ref(fig:flintcor)].
+While not substantial, there is clear temporal patterning in the flint data as well.
+This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, the debitage from their production, and flakes characterises these sites.
+Slightly younger sites appear more impacted by core fragments and blades.
+The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades.
+The comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material.
+However, while the former pertains to the trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
+The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages.
 
@@
 
-Reducing the number of such inferential leaps by aggregating artefact groups could potentially subsume important variation, but it does also reduce the possibility that the conclusions of an analysis is not simply the result of the employed units of analysis, and an overly naive trust in the fact that these units reflect any behaviourally meaningful categories.
-This realisation has led to a wide range of outside Scandinavian Mesolithic research, meaning a range of empirical measures associated with various mobility patterns are available.
-
-It has, however, been suggested that access to raw-materials can be a more important decider for assemblage composition than mobility patterns , and that variable workability of these materials will impact factors such as the relative frequency of tools to debitage in the assemblages [e.g.
-@andrefsky1994; @manninen2014].
-In addition to attempts at accounting for this analytically (see below), the limited geographical distribution of the sites should alleviate the issue somewhat, as the overall availability of raw materials could be expected to be more comparable through time than if a wider region was under consideration.
-
-Of these the most well-estbalished is the.
-base don the assumption that, this measure and has been shown to display a meaningful measure across a wide range of temporal, geographical and cultural contexts.
-
-Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures that have been linked to mobility patterns.
+As some of the distributions for the variables were severely skewed, these were subjected to various transformations to Figure \@ref(fig:pca-analysis) displays a PCA using variations of the continuous measures that have been linked to mobility patterns.
 There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot.
@@
 While there is a slight tendency for blanks to be more associated with younger sites, frequency of chips appears to be largely independent of time.
-However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have an methodological explanation.
-Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips, one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and to separate from naturally fragmented stone during excavation and classification, leading to a over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material.
-While this is not necessarily the entire explanation behind the relationship, this does render it difficult to Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well.
-That is not to say that cores may not be indicative or related to mobility patterns, but may require further analysis beyond their simple classification as such [@kitchel2021].
-
-Thus, while some secondary expectations of the @clark2017 model does not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
+However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation.
+Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips, one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification.
+This could conceviably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material.
+While this is not necessarily the entire explanation behind the relationship, this does make it difficult to place much analytical weight on this pattern.
+Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well.
+That is not to say that cores may not be indicative or related to mobility patterns, but to get at this may require further analysis beyond their simple classification as cores [@kitchel2021].
+
+Thus, while some secondary expectations of the WABI does not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
 The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material.
-Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.
+Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases [e.g.
+@bergsvik2001; @bjerck2010; @glørstad2010].
 To explore this proposition further, these two variables are subjected to more detailed scrutiny below.
@@
 While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same.
-The relationship between mean site age and RFSF is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the VDF is not temporally contingent (*r =* `r round(denscorflint, 2)`).
+The relationship between mean site age and relative frequency of secondarily worked flint (RFSF) is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the volumetric density of flint (VDF) is not temporally contingent (*r =* `r round(denscorflint, 2)`).
 As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.
@@
 A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist).
-For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
-This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot).
+For sites with radiocarbon age determinations the dates were drawn from the associated summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
+This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot)A.
 Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE.
@@
 The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation associated mobility patterns.
+Figure \@ref(fig:simplot)B displays the result of running the same procedure on the flint data.
+The general pattern follows the same trajectory, but the result for some individual sites is markedly different.
+This is discussed below.
 
@@
   } else {
-    sdate <- sample(as.integer(site["reported_start_bc"]*-1):as.integer(site["reported_end_bc"]*-1), 1)
+    sdate <- sample(as.integer(site["reported_start_bc"]):as.integer(site["reported_end_bc"]), 1)
   }
@@
 
-# Specify number of simulations
+# Specify number of simulation runs
 nsim <- 1000
@@
 
-```{r simplot, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) All lithics, and B) Flint.
+```{r simplot, fig.cap = "Temporal variation in the curation index for A) All lithics, and B) Flint.
 The temporal uncertainty is handled by means of a simulation approach where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B.
@@
 
-The results of the CA does appear to align well with previous research that has been based on more informal treatment of the assemblages as a whole, and the employed artefact categories are clearly capturing a temporal component.
+The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component.
 This could indicate that the aggregation of artefact categories that was done here might have been overly conservative.
-However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'.
-The developments evident in the results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology.
-Unpicking and aligning the significance of these chronological patterns with any specific behavioural and technological dimensions using the CA results is, however, another task entirely.
+However, it also evidently clear, in the words of @kruskal1971, that 'time is not the only dimension'.
+The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by the deviances from these.
+Unpicking and aligning the significance of these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely.
 This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear.
@@
 
-The relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here identified by means of an exploratory approach, and is in part investigated further because they align with suggestions from previous research, clearly representing a possible case of confirmation bias.
+The relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here identified by means of an exploratory approach, and is in part investigated further because they align with suggestions from previous research --- clearly representing a possible case of confirmation bias.
 However, some inferential merit can be achieved by invoking what has been termed consilience.
 Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30].
-Thus, the overlap in results presented here and those repeatedly reported from a range of different context does speak to the general applicability of the measures, and offers an initial indication that they might be capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
-While it is not within the scope of this paper to consider these comparisons to any wider extent, and so they must be considered highly speculative, the VDL for all assemblages treated here has a mean of `r round(mean(lithics2$lithic_dens))` lithics per m\\\^3 , with a standard deviation of `r round(sd(lithics2$lithic_dens))`.
-This appears to be relatively high when compared to Late Pleistocene sites from western Eurasia [@barton2011].
-Similarly, the frequency of retouched artefacts ranges from `r round(min(lithics2$secondary_freq)*100, 1)`--`r round(max(lithics2$secondary_freq)*100, 1)` % which is relatively low when compared to the same data.
-This could, in sum, indicate that the coastal Mesolithic sites of South-Eastern Norway have generally been associated with a low degree of mobility when compared to these context, a proposition that would be in line with both ethnographic and archaeological research that highlights the possible role marine resources can play in, as well as a general trend towards more sedentary life through the Holocene [e.g.
-@perreault2019, 42] .
-
-Returning focus to the present study, the curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE for the rest of the Mesolithic.
+Thus, the overlap in results presented here and those repeatedly reported from a range of different context does speak to the general applicability and comparability offered by the measures, and gives an initial indication that they might be capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
+
+The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE for the rest of the Mesolithic.
 This pattern is evident in both the flint data and when all lithics are treated in aggregate.
@@
 Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages.
-As can be seen in Figure @ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
+As can be seen in Figure \@ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
 Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion would likely lead to its underestimation.
@@
 While there are certainly nuances and idiosyncrasies in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust.
-It seems clear that there is a marked drop in curation starting some time just before 8000 BCE, which has stabilised around 7000 BCE.
+It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE.
 This corresponds well with a chronological framework where the end of the Early Mesolithic, or Flake axe phase, is set to c.
 8200 BCE [@reitan2016].
-Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns.
-This corresponds very well with the findings reported here, where the earliest phase is characterised by relatively high and uniform values on the curation index.
-The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade and subsequently axe technology, which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021].
+Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
+@bjerck2010; @fuglestvedt2012].
+This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index.
+The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology, which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021].
 The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021].
-Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is a serious issue [@solheim2018a; @solheim2020a; @nielsen2021].
-In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE, signifying the onset of the production of non-flint ground-pecked or round butted adzes.
+Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021].
+In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE.
 Referring back to the increasing expediency in the curation data between c.
-8200 and 7000 BCE, the microlith phase could thus represent a transitional phase where new living practices where either propagating through societies in South-Eastern Norway, or possibly a complete replacement of the population might even have been taking place --- a process that in light of the curation data would have concluded around 7000 BCE.
-While non-flint macro tools are present in the material before this, the first of the sites treated here that are heavily dominated by their production is Gunnarsrød 6b, dated to c.
+8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices was propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.
+
+The Microlith phase is followed the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
+As is evident from the both CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c.
 `r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[1]]*-1`--`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[2]]*-1` BCE [@carrasco2014].
 The curation data remains stable from around 7000 BCE through the next typological transition at c.
-5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet phase.
-The Nøstvet phase has traditionally been seen as representing the onset of more stable mobility patterns in south-eastern Norway, and has been explicitly linked to a expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit somewhat vaguely defined [@eigeland2015, 127--130].
-In recent years it has been suggested that the transition to a decrease in mobility and more varied mobility patterns can be traced back to the Middle Mesolithic.
-The curation index employed here clearly supports this notion, and suggests that the mobility patterns of the Nøstvet-phase was well established in preceding phases.
-
-4500 BCE by reduced importance of the non-flint categories, these are not which could indicate that these changes are not related to major shifts in land-use and mobility patterns.
-Some explanations to this apparent discrepancy can be that fluctuations in population numbers simply do not affect mobility patterns in the setting considered here, or the signal may be too weak to be picked up considering the small sample size.
-
-the exlusion of this material The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps.
-The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel out.
-Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern, while the high expediency on some sites when considering all lithics in aggregate in the later parts of the Mesolithic does speak to variability in site use.
-The nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential-logistic continuum.
-
-It is also interesting to note that some of the chronological patterns indicated in the CA does not appear to be related to the curation measures.
-This for example pertains to the apparent importance of the axe production sites.
-It has recently been suggested that the end of the Mesolithic sees a drop in population numbers and a subsequent increase in mobility as migrating populations arrived from southern Scandinavia.
-It is an open question whether the apparent absence of this pattern in the present data is a result of a strained sample size, or a weaker or non-existent signal from any changes in mobility patterns in the later Mesolithic.
-As it stands, however, the main hypothesis generated by the present data would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE and remain relatively stable throughout the rest of the period.
-Even though the relative increase in Given the parallel increase in variation this would seem to correlate well with a transition from a residential to a logistical settlement system.
-
-As for the measures associated with curation and mobility patterns, relative frequency of secondarily worked lithics and volumentric density of lithics correspond most clearly to patterns previously proposed in the literature.
-However, as these were identified by employing PCA and identifying what variables most clearly correspond to a temporal dimension, there is a danger of circular reasoning.
-Having established that these measures appears to capture a behavioural dimension of interest, this can, however, now be tested more rigorously by employing site data from a wider region.
-these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies.
-The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway.
-However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips?
-Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pair is not correlated with these.
-As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous.
-
-This is arguably more established than the variables included bysimilarities between the variable plots of @Bicho2020 and the variables indicated in is striking.
-This pertains to the negative correlation between core and chip frequency, and the negative correlation between primary and secondarily worked lithics.
-This last
-
-These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at behavioural dimensions relating to mobility patterns among hunter-gatherers.
+5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet adze phase.
+While previously defined as having a slightly longer duration, the Nøstvet phase has traditionally been seen as representing the onset of more stable mobility patterns in south-eastern Norway [@jaksland2001;], and has been explicitly linked to an expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit with the term being somewhat vaguely invoked [@eigeland2015, 127--130].
+In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016].
+The curation index employed here clearly supports this notion, and suggests that the mobility patterns of the Nøstvet phase were well established in preceding periods.
+
+The subsequent Transverse arrowhead phase (c.
+4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged -and single-edged points [@reitan2016].
+It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021].
+The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns related to the sites.
+However, it also worth highlighting the limited sample size which could mean that this effect is either completely missed,
+
+As it stands the main hypotheses by the present data would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE.
+These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories.
+Given the quite dramatic fall in curation levels, and a parallel increase in variation, this would seem to correlate well with a transition from a residential to a logistical settlement system.
 
@@
 These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages.
-these notions have previously Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as measured here.
+Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as measured here.
 
@@
 Furthermore, precisely how this mobility pattern has.
-Here Binfords model of briefly invoked, but there exists of course a wide range of possible configurations and mobility strategies that have subsistence data, site locations
-
-The results of considering variables that are have been associated with mobility patterns gave rise to two competing hypotheses.
-These pertain to whether the degree of curation associated with the assemblages as captured by the changing ratio of primary to secondarily worked lithics over time is the result of changing raw material availability or mobility patterns.
-The orthogonal relationship between core to chip frequency would in this framework be expected to be capturing the opposite relationship.
-If degree of curation mainly reflects changes in raw-material quality and availability, this would then indicate that mobility patterns are comparable throughout the Mesolithic as the core to chip frequency is not time-dependent.
-And vice verca --- if the changing mobility patterns is the main driver of the transition from curated to expedient technological organisation, as operationalised here, While this would be in line with findings from a recent evaluation of the locational patterns associated with the sites [@roalkvam2020], there is no readily available empirical evidence, no smoking gun, se findings are the results of an exploratory analysis, and are therefore perhaps best seen as potentially laying the framework for later targeted strategic test of these competing hypotheses.
-
-<!-- The following line inserts a page break  -->
+Here Binfords model of briefly invoked, but there exists a wide range of possible configurations and mobility strategies that
 

### 2021-05-27 analysis/paper/paper.Rmd
Final edits before submission to phd-seminar

@@
 
-```{r 1dist, fig.height = 13, fig.width = 12}
+```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
+Radiocarbon age determintations are given as the sum of the posterior density esimates.
+Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
+These follow the original reports.", fig.height = 13, fig.width = 12}
 # Create a bounding box around site points to limit the overview map
@@
 
-```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
+```{r ldist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
 Radiocarbon age determintations are given as the sum of the posterior density esimates.
 Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
-These follow the original reports.", fig.height = 9, fig.width = 12, eval = FALSE}
+These follow the original reports.
+Site names corresponding to numbers are given in Table 1.", fig.height = 9, fig.width = 12, eval = FALSE}
 temp_dist
@@
 A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g.
-@bjerck2010].
+@bjerck2008].
 Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g.
@@
 
-The `r nrow(sites)` sites analysed here have been dated by reference to relative sea-level change, typology and/or radiocarbon dates.
+The `r nrow(sites)` sites analysed here have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table 1).
 Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators.
@@
 Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53].
-Furthermore, the various non-flint raw materials that have been lumped together here have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
+Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
 Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories.
@@
 @glørstad2011; @eigeland2015, 370].
-Finally, while factors such as landscape changes through shoreline displacement still can have led to variable raw-material availability on the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.
-
-Studies concerned with chronological changes in the composition of lithic assemblages in Norway have typically had a focus on formal tool types [e.g.
-@helskog1976; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g.
-@damlien2016; @berg-hansen2017; @solheim2020].
-Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
+Finally, while factors such as landscape changes through shoreline displacement still can have led to variable raw-material availability at the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.
+
+Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g.
+@ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g.
+@eymundsson2018; @damlien2016; @berg-hansen2017; @solheim2020].
+Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
 These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
-To my knowledege, only a single study dealing with Mesolithic material in southern Norway has involved the use of a multivariate quantitative framework, which was used to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a].
-In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, or to more subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of assemblages.
+To my knowledege, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was used to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
+In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
+
+```{r tab1}
+
+sites %>% 
+  mutate(strt_bce = reported_start_bc * -1, 
+         end_bce = reported_end_bc * -1,
+         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
+  select(no, site, dating_method, strt_bce, end_bce) %>% 
+  gt() %>% 
+    cols_label(
+    site = "Site name",
+    dating_method = "Dating method",
+    strt_bce = "Reported start (BCE)",
+    end_bce = "Reported end (BCE)",
+  ) %>% 
+  tab_header(
+    title = "Table 1. Analysed sites."
+  ) %>% 
+  tab_options(heading.title.font.size = 15)
+```
 
@@
 While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
-However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves what Binford termed a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
+However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
 The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses.
@@
 This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful.
-However, the approach taken was also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
-Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g.
-@barton1991; @dibble1995].
+However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
+Several scholars have built on this to argue that morphological variation in retouched lithics from the Paleolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g.
+@barton1991; @dibble1995; @barton2021].
 Consequently, several artefact categories have here been collapsed for the CA.
@@
 Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
-@barton1999; @riel-salvatore2004; @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@kuhn2015].
+@barton1999; @riel-salvatore2004; @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015].
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
@@
 @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021].
-The use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181] .
-In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
-The analysis is done using principal components analysis (PCA), following the continuous operationalisations of these measures [cf.
-@baxter1994, 100, @Bicho2020].
+The use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181], while chips .
+In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
+The analysis is done using principal components analysis (PCA), following the continuous measure of these variables [cf.
+@baxter1994, 100; @Bicho2020].
 
@@
 @Bicho2020; @canessa2021] which has been omitted in light of the previously mentioned Frison effect.
-Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation that is difficult to account for.
+Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation.
 Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
@@
 The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, and the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
@@
 This could explain its position as an outlier in the plot.
-The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
+The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b].
 In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
-@bang-andersen2003; @damlien2018b], although perhaps highlighting the relative importance of non-flint materials for some of the sites in south-eastern Norway which, while also present, is typically not as marked in other parts of southern Norway [e.g.
-@breivik2018, 184].
+@jaksland2014b; @breivik2018; @damlien2018b].
 
 The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages.
-Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these sites.
+Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among the sites.
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
 This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
-@eigeland2015, 370, @glørstad2011].
-
-As most of the variation in the data is accounted for by variation in non-flint material this suppresses and makes it difficult to discern patterns in the flint data.
-A second CA was therefore run excluding the non-flint material [Figure @ref(fig:flintcor)].
+@eigeland2015, 370; @glørstad2011; @reitan2016].
+
+As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data.
+A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)).
 While not substantial, there is clear temporal patterning in the flint data as well.
-This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, the debitage from their production, and flakes characterises these sites.
+This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites.
 Slightly younger sites appear more impacted by core fragments and blades.
 The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades.
-The comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material.
+Apart from the impact of core fragments, this must be considered very much in agreement with previous research [e.g.
+@solheim2017c, with references].
+The comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
+@damlien2016; @eigeland2015; @solheim2020] .
 However, while the former pertains to the trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
-The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages.
+The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf.
+@manninen2014].
 
@@
 
-```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Object map, B) Variable map. Note the differences on the axes between plots.", fig.height = 13}
+```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Object map, B) Variable map.", fig.height = 13}
 # Plot sites
@@
 plt2 <- fviz_ca_col(calithics, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B")
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
 
@@
 plt4 <- fviz_ca_col(caflint, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B")
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
 
@@
 
-```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Object map, B) Variable map. Do not the differences on the axes between plots."}
+```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Object map, B) Variable map."}
 plt3 /
@@
 
-As some of the distributions for the variables were severely skewed, these were subjected to various transformations to Figure \@ref(fig:pca-analysis) displays a PCA using variations of the continuous measures that have been linked to mobility patterns.
+Moving on the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed.
+These are displayed in the correlation matrix in in Figure \@ref(fig:cor-plot).
+Figure \@ref(fig:pca-analysis) displays the resulting PCA.
 There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot.
-The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure @ref(fig:var-contrib)).
+The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure \@ref(fig:var-contrib)).
 Almost orthogonal to this is the strong negative correlation between relative frequency of chips and blanks.
@@
 However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation.
-Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips, one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification.
-This could conceviably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material.
+Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification.
+This could conceivably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material.
 While this is not necessarily the entire explanation behind the relationship, this does make it difficult to place much analytical weight on this pattern.
@@
 Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases [e.g.
-@bergsvik2001; @bjerck2010; @glørstad2010].
+@bergsvik2001; @bjerck2008; @glørstad2010; @jaksland2001].
 To explore this proposition further, these two variables are subjected to more detailed scrutiny below.
@@
 lithics2 <- sites %>% 
-  select(-no, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
+  select(-site, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
           -askeladden_id) %>% 
@@
          -chip_other, -core_other, -core_fragment_other) %>% 
-  column_to_rownames("site")
+  column_to_rownames("no")
 
@@
 flint2 <- sites %>% 
-  select(-no, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
+  select(-site, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
          -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
@@
          -core_fragment, -chip, dating_method) %>% 
-  column_to_rownames("site")
+  column_to_rownames("no")
 
@@
 
-```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these.
+```{r cor-plot, fig.cap = "Correlation matrix showing transformation of skewed variables for the PCA, and the linear correlation between these.
 The mean age of the sites has also been included to visualise overall temporal trends.", fig.height = 9.5, fig.width = 10}
@@
 ```{r pca-analysis, fig.cap = "PCA using variables that have been related to mobility patterns.
-Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 8}
+Note that details on the transformation of the variables has been left out of the plot for clarity but follow those given in Figure 4.", fig.height = 8}
 # fig.width = 10
@@
          title = "") + 
-  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+  scale_fill_gradient2(name = "Mean age BCE",  low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
 While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same.
-The relationship between mean site age and relative frequency of secondarily worked flint (RFSF) is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the volumetric density of flint (VDF) is not temporally contingent (*r =* `r round(denscorflint, 2)`).
+The relationship between mean site age and relative frequency of secondarily worked flint (RFSF) is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint (VDF) is not temporally contingent (*r =* `r round(denscorflint, 2)`).
 As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.
@@
        x = "Volumetric density of lithics (log10)", 
-       y = "Relative frequency of secondarily worked lithics (log10)") +
-  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+       y = "% secondarily worked lithics (log10)") +
+  scale_fill_gradient2(name = "Mean age BCE",  low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
   scale_x_log10() +
-  theme_classic()
+  theme_classic() +
+  theme(legend.position = "bottom")
 
@@
        x = "Volumetric density of flint (log10)", 
-       y = "Relative frequency of secondarily worked flint (log10)") +
-  scale_fill_gradient2(name = "ka BP",  low="yellow", mid="orange", 
+       y = "% secondarily worked flint (log10)") +
+  scale_fill_gradient2(name = "Mean age BCE",  low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
 To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g.
-@crema2012; @baxter2016; @orton2017].
+@baxter2016; @crema2012; @orton2017].
 A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist).
@@
 
-# Curation index plot simply using mean site age (not used)
+# Curation index plot simply using mean site age (not in use)
 w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
@@
 The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component.
-This could indicate that the aggregation of artefact categories that was done here might have been overly conservative.
-However, it also evidently clear, in the words of @kruskal1971, that 'time is not the only dimension'.
-The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by the deviances from these.
-Unpicking and aligning the significance of these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely.
+One possible implication of this could be that the aggregation of artefact categories might have been overly conservative.
+However, it also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'.
+The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern.
+Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely.
 This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear.
@@
 
-The relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here identified by means of an exploratory approach, and is in part investigated further because they align with suggestions from previous research --- clearly representing a possible case of confirmation bias.
+The relevance of the relationship between frequency of secondarily worked lithics and volumetric density of lithics was here identified by means of an exploratory approach, and is in part investigated further because they align with suggestions from previous research --- clearly representing a possible case of both confirmation bias and circular reasoning.
 However, some inferential merit can be achieved by invoking what has been termed consilience.
@@
 The five sites that have values on the curation index below c.
--0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern [sensu @binford1980].
+-0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern [*sensu* @binford1980].
 That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility [@eigeland2014b].
@@
 Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
-@bjerck2010; @fuglestvedt2012].
+@bjerck2008; @fuglestvedt2012].
 This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index.
@@
 5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet adze phase.
-While previously defined as having a slightly longer duration, the Nøstvet phase has traditionally been seen as representing the onset of more stable mobility patterns in south-eastern Norway [@jaksland2001;], and has been explicitly linked to an expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit with the term being somewhat vaguely invoked [@eigeland2015, 127--130].
+While previously defined as having a slightly longer duration, the Nøstvet phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in south-eastern Norway [e.g.
+@jaksland2001; @lindblom1984], and has been explicitly linked to an expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit with the term being somewhat vaguely invoked [@eigeland2015, 127--130].
 In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016].
@@
 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged -and single-edged points [@reitan2016].
-It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021].
+It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c.
+4300 BCE [@nielsen2021].
 The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns related to the sites.
-However, it also worth highlighting the limited sample size which could mean that this effect is either completely missed,
-
-As it stands the main hypotheses by the present data would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE.
-These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories.
-Given the quite dramatic fall in curation levels, and a parallel increase in variation, this would seem to correlate well with a transition from a residential to a logistical settlement system.
+However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic here, which could mean that the effect could be missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
+
+As it stands, however, the main hypotheses resulting from the analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE.
+These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
+Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a residential to a logistical settlement system [@barton2011; @binford1980].
 

### 2021-06-08 analysis/paper/paper.Rmd
Corrections after seminar

@@
 The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports.
-The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [@clark2017], which has not seen application in this context before.
+The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [@clark2017], which aligns properties of the assemblages with has not seen application in this context before.
 
@@
 
-The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \@ref(fig:dist)A.
+The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway (Figure \@ref(fig:dist)A).
 The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
-The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts.
+The sites included in the analysis consist of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts.
 The institution responsible for these excavations was the Museum of Cultural History in Oslo.
@@
 @clark2006; @dibble2017].
-In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \@ref(fig:dist)B].
+In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@jaksland2014; @solheim2013b, see also Figure \@ref(fig:dist)B].
 Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
@@
 ```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis.
-Radiocarbon age determintations are given as the sum of the posterior density esimates.
+Radiocarbon age determinations are given as the sum of the posterior density esimates.
 Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
@@
 @perreault2019, 47].
-Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred.
-In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007].
+Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good control of when these accumulation events occurred.
+In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of losing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007].
 However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
@@
 @bamforth1986; @dibble1995; @shott2005].
-Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
+
+One concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
 @barton1999; @riel-salvatore2004; @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015].
@@
 More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation.
-This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
+This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
 The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades.
@@
 In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
-The analysis is done using principal components analysis (PCA), following the continuous measure of these variables [cf.
-@baxter1994, 100; @Bicho2020].
+Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a [cf.
+@baxter1994, 100].
 
@@
 The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, and the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
 The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
@@
 Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
-However, as the site had been quite heavily impacted by modern disturbances, this led [@eigeland2014, 124] to suggest that the material might have been compromised.
+However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014, 124 to suggest that the material might have been compromised.
 This could explain its position as an outlier in the plot.
@@
 
-Moving on the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed.
+Moving on to the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed.
 These are displayed in the correlation matrix in in Figure \@ref(fig:cor-plot).
@@
 While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same.
-The relationship between mean site age and relative frequency of secondarily worked flint (RFSF) is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint (VDF) is not temporally contingent (*r =* `r round(denscorflint, 2)`).
+The relationship between mean site age and relative frequency of secondarily worked flint is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint is not temporally contingent (*r =* `r round(denscorflint, 2)`).
 As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.
@@
 One possible implication of this could be that the aggregation of artefact categories might have been overly conservative.
-However, it also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'.
+However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'.
 The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern.
@@
 
-While there are certainly nuances and idiosyncrasies in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust.
+While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust.
 It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE.
@@
 Referring back to the increasing expediency in the curation data between c.
-8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices was propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.
+8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices were propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.
 
@@
 4300 BCE [@nielsen2021].
-The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns related to the sites.
-However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic here, which could mean that the effect could be missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
-
-As it stands, however, the main hypotheses resulting from the analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE.
+The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns in the material treated here.
+However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
+
+As it stands, however, the main hypotheses resulting from the analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and unform degrees of mobility, which then drop before leveling off at around 7000 BCE.
 These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
-Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a residential to a logistical settlement system [@barton2011; @binford1980].
+Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@barton2011; @binford1980].
 
@@
 These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages.
-Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as measured here.
+Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as operationalised here.
 

### 2021-06-10 analysis/paper/paper.Rmd
Implementing suggestions from PhD-seminar.

@@
 Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.
-In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification within larger unified projects.
+In an attempt to limit the influence of some potentially confounding effects, the material chosen for analysis has a constrained geographical distribution, and stems from recent investigations that have employed comparable methods for excavation and classification within larger unified projects.
 
@@
 The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports.
-The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [@clark2017], which aligns properties of the assemblages with has not seen application in this context before.
+The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [e.g.
+@clark2017], which has been employed in other contexts to align properties of the assemblages with land-use and mobility patterns.
 
@@
 These follow the original reports.
-Site names corresponding to numbers are given in Table 1.", fig.height = 9, fig.width = 12, eval = FALSE}
+Site numbering follow that given in Table 1.", fig.height = 9, fig.width = 12, eval = FALSE}
 temp_dist
@@
 In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year towards the end of the Mesolithic [@sørensen2014].
-Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.
+Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, it is worth bearing in mind that this could vary depending on when and where they were in use, potentially reducing the degree to which their composition can be directly compared.
 
@@
 @glørstad2011; @eigeland2015, 370].
-Finally, while factors such as landscape changes through shoreline displacement still can have led to variable raw-material availability at the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.
+Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.
 
@@
 @ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g.
-@eymundsson2018; @damlien2016; @berg-hansen2017; @solheim2020].
-Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or invovled extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
+@berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020].
+Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
 These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
-To my knowledege, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was used to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
+To my knowledege, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
 In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
@@
 
-The relatively constrained geographical distribution of the analysed sites, the limited temporal spread over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
+The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
 However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
@@
 As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below].
-The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1979].
+The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1977; @binford1979].
 An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement.
@@
 The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies and raw-material quality and availability [e.g.
-@parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017].
+@andrefsky1994; @parry1987; @kuhn1992; @smith2015; @clark2017].
 Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted.
@@
 
-One concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
-@barton1999; @riel-salvatore2004; @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015].
+Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
+@barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2013; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015; @riel-salvatore2008].
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
@@
 The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades.
-These variables and underlying logic consitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.
+These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.
+
 However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
-@breivik2016a; @reitan2016; @Bicho2020; @kitchel2021].
-The use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181], while chips .
+@breivik2016a; @reitan2016; @Bicho2020; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
 In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
-Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a [cf.
+Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative [cf.
 @baxter1994, 100].
 
-A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here.
-For the assemblage data itself this especially pertains to diversity in tool-types [e.g.
-@Bicho2020; @canessa2021] which has been omitted in light of the previously mentioned Frison effect.
+A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g.
+@breivik2016; @Bicho2020].
+For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the previously mentioned Frison effect.
 Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation.
-Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded.
+Similarly, the number of activity areas, effectively number of artefact clusters, however defined, has also been disregarded.
 This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017].
-This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].
+This pertains for example to the impact of bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].
 
@@
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
-The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
+The earliest sites tend to be located in the upper left corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
 Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
@@
 Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
-However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014, 124 to suggest that the material might have been compromised.
+However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014[124] to suggest that the material might have been compromised.
 This could explain its position as an outlier in the plot.
-The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b].
+The use of metarhyolite for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b].
 In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
@@
 A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)).
-While not substantial, there is clear temporal patterning in the flint data as well.
+While not as substantial, there is clear temporal patterning in the flint data as well.
 This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites.
@@
 @damlien2016; @eigeland2015; @solheim2020] .
-However, while the former pertains to the trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
+However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
 The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf.
@@
 
-```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Object map, B) Variable map.", fig.height = 13}
+```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Row plot, B) Column plot.", fig.height = 13}
 # Plot sites
@@
 
-```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Object map, B) Variable map."}
+```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Row plot, B) Column plot."}
 plt3 /
@@
 However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation.
-Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification.
+Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot)), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification.
 This could conceivably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material.
@@
 
-```{r cor-plot, fig.cap = "Correlation matrix showing transformation of skewed variables for the PCA, and the linear correlation between these.
-The mean age of the sites has also been included to visualise overall temporal trends.", fig.height = 9.5, fig.width = 10}
+```{r cor-plot, fig.cap = "Correlation matrix showing transformation of skewed variables for the PCA.
+The mean age of the sites has also been included to visualise overall temporal trends.
+Cells below the diagonal display the bivariate distributions with a fitted OLS-regression.
+The cells above the diagonal display and are coloured by the corresponding Pearsons correlation coefficient.", fig.height = 9.5, fig.width = 10}
 # Transform the skewed variables - sqrt() is used where there are zero values,
@@
 ```{r pca-analysis, fig.cap = "PCA using variables that have been related to mobility patterns.
-Note that details on the transformation of the variables has been left out of the plot for clarity but follow those given in Figure 4.", fig.height = 8}
+Note that details on the transformation of the variables has been left out of the plot for clarity, but follow those given in Figure 4.", fig.height = 8}
 # fig.width = 10
@@
 
-```{r var-contrib, fig.cap = "Contribution of variables to the components of the PCA.", fig.height = 3}
+```{r var-contrib, fig.cap = "Contribution of variables to the components of the PCA.
+The dotted red line indicates the expected contribution from each variable given a uniform distribution of impact.", fig.height = 3}
 # The plots below investigates the contribution of the variables to the
@@
 As can be seen in Figure \@ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
-Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion would likely lead to its underestimation.
+Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exclusion would likely lead to its underestimation.
 One possible approach could be to weigh the curation index by proportion of non-flint material in the assemblages.
@@
 This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index.
-The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology, which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021].
+The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021].
 The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021].
@@
 The Microlith phase is followed the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
-As is evident from the both CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c.
+As is evident from both the CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c.
 `r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[1]]*-1`--`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[2]]*-1` BCE [@carrasco2014].
@@
 These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages.
-Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as operationalised here.
+Precisely what behavioural implication the development in the use of these types have, however, is less clear, but appears to follow a different, and more complex trajectory than that of curation, as operationalised here.
 

### 2021-06-11 analysis/paper/paper.Rmd
Began trying to tackle the conclusion

@@
 
-The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway.
+The results of the CA align more or less with results of previous research in south-eastern Norway.
 This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
 These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages.
-Precisely what behavioural implication the development in the use of these types have, however, is less clear, but appears to follow a different, and more complex trajectory than that of curation, as operationalised here.
-
-In my view this alignes well with the framework of @clark2017, suggesting that shifts in mobility patterns has caused this empirical pattern.
+Precisely what behavioural implication the development in the use and occurrences of the tool and debitage categories are less clear, but appears to follow a different, and more complex trajectory than that of curation, as operationalised herge.
+
+The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in mobility patterns has caused this empirical pattern, in line with the framework of @clark2017.
+Another perspective would be that this is not surprising at all, and that its previously demonstrated relevance across a wide range of contexts rather points to its pervasive relevance for the organisation of lithic technology, and that there as a consequence should be little reason to think that a the setting of Mesolithic south-eastern Norway should be any different.
+In any case, if its applicability and relevance for this setting is accepted, the relationship does hold great potential for large scale comparative studies in Mesolithic Norway and beyond.
+Futhermore, while the temporal trends associated with the curation index was here narratively associated with the most immediate chronological trends that have frivolously been associated with the Mesolithic in south-eastern Norway, the explicit quanitiication offers the possibility of more formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales.
+
 However, the role of non-flint material, variability of flint quality.

### 2021-06-14 analysis/paper/paper.Rmd
Writing conclusion and abstract

--- 2021-06-10 analysis/paper/paper.Rmd
@@
 csl: "../templates/jas_reports.csl" # Insert path for the bib-style
+abstract: |
+  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway.
+To provide analytical control pertaining to factors such as raw-material variability and excavation and classification practices, the sites chosen for analysis have all been excavated relatively recently, and have a constrained geographical distribution.
+Reaching the conclusion that these measures are applicable in a Norwegian Mesolithic setting.
+While this certainly holds potential for larger scale comparative studies, it also sets the stage for a more targeted and rigorous model evaluation outside an exploratory setting.
+
+keywords: |
+  Mesolithic Norway; Lithic assemblages; Multivariate statistics;
 ---
+
+Keywords: `r rmarkdown::metadata$keywords`
 
@@
 @glørstad2011; @eigeland2015, 370].
-Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.
+Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.
 
@@
 Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
-@barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2013; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015; @riel-salvatore2008].
+@barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2013; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008].
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
@@
 In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
-Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative [cf.
+Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative as compared to the CA [cf.
 @baxter1994, 100].
@@
 The comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
-@damlien2016; @eigeland2015; @solheim2020] .
+@damlien2016; @eigeland2015; @solheim2020].
 However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
@@
 The linear correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`).
-As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data.
+As variable non-flint availability and workability has also been suggested to potentially impact these dimensions [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data.
 While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same.
@@
 This has stabilised by around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic.
-The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation associated mobility patterns.
+The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in associated mobility patterns.
 Figure \@ref(fig:simplot)B displays the result of running the same procedure on the flint data.
@@
 Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
-@bjerck2008; @fuglestvedt2012].
+@bjerck2008; @breivik2016; @fuglestvedt2012].
 This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index.
@@
 
-The Microlith phase is followed the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
+The Microlith phase is followed by the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
 As is evident from both the CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c.
@@
 The subsequent Transverse arrowhead phase (c.
-4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged -and single-edged points [@reitan2016].
+4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016].
 It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c.
@@
 The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns in the material treated here.
-However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
-
-As it stands, however, the main hypotheses resulting from the analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and unform degrees of mobility, which then drop before leveling off at around 7000 BCE.
+However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
+
+As it stands, the main hypotheses resulting from the present analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and uniform degrees of mobility, which then drop before levelling off at around 7000 BCE.
 These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
@@
 
-The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway.
+The results of the CA align more or less with results of previous research in south-eastern Norway.
 This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
 These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages.
-Precisely what behavioural implication the development in the use of these types have, however, is less clear, but appears to follow a different, and more complex trajectory than that of curation, as operationalised here.
-
-In my view this alignes well with the framework of @clark2017, suggesting that shifts in mobility patterns has caused this empirical pattern.
-However, the role of non-flint material, variability of flint quality.
-Furthermore, precisely how this mobility pattern has.
-Here Binfords model of briefly invoked, but there exists a wide range of possible configurations and mobility strategies that
+Precisely what behavioural implication the development in the occurrences of the tool and debitage categories have are less clear, but appears to follow a different and more complex development over time than that of curation, as operationalised here.
+
+The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns is the main driver behind this empirical pattern --- in line with the framework of @barton2011.
+Another perspective would be that this is not surprising at all [cf.
+@kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different.
+If its applicability and relevance for this setting is accepted, the relationship does hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
+Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with Mesolithic south-eastern Norway.
+The explicit quantification does, however, offer the possibility of more formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.
+
+Having reached the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting can only constitute a first analytical step.
+As @elster2015[12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached.
+This, however, can only constitute the absolute minimum of what is required of a proposed explanation.
+Subsequent steps should be to probe and challenge this explanatory framework.
 

### 2021-06-15 analysis/paper/paper.Rmd
Final edits before sending the text off to read-through

@@
 title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
 author:
@@
 institute:
-  - UiO: University of Oslo
+  - UiO: University of Oslo, Department of Archaeology, History and Conservation
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
   This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway.
-To provide analytical control pertaining to factors such as raw-material variability and excavation and classification practices, the sites chosen for analysis have all been excavated relatively recently, and have a constrained geographical distribution.
-Reaching the conclusion that these measures are applicable in a Norwegian Mesolithic setting.
-While this certainly holds potential for larger scale comparative studies, it also sets the stage for a more targeted and rigorous model evaluation outside an exploratory setting.
+To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution.
+The assemblages were explored following two strains of analysis.
+The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages.
+The findings pertaining to the artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research.
+Furthermore, the chronological trends associated with variables taken from the so-called whole assemblage behavioural index, originally developed for characterising Paleolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature.
+This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside an exploratory setting, which might ultimately yield a powerful comparative tool for the analysis of Mesolithic assemblages.
 
 keywords: |
-  Mesolithic Norway; Lithic assemblages; Multivariate statistics;
+  Mesolithic Norway; Lithic assemblages; Mobility; Multivariate statistics
 ---
@@
 The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [e.g.
-@clark2017], which has been employed in other contexts to align properties of the assemblages with land-use and mobility patterns.
+@clark2017], which has been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.
 
@@
 Some categories in the original reports have therefore been combined in the dataset.
-This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm.
-In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
+This for example pertains to the category narrow-blades (width > 8mm and < 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
 Narrow-blades were combined with the blade category here.
@@
 While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g.
-@glørstad2011; @eigeland2015, 370].
+@eigeland2015, 370; @glørstad2011].
 Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.
@@
 However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
-However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
+While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
 The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses.
@@
 Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
-@barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2013; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008].
+@barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008].
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
@@
 In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
-Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative as compared to the CA [cf.
-@baxter1994, 100].
+Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [cf.
+@baxter1994, 71--77, 103].
 
 A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g.
-@breivik2016; @Bicho2020].
+@Bicho2020; @breivik2016a].
 For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the previously mentioned Frison effect.
@@
 In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
-@jaksland2014b; @breivik2018; @damlien2018b].
+@breivik2018; @jaksland2014b; @damlien2018b].
 
@@
 Moving on to the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed.
-These are displayed in the correlation matrix in in Figure \@ref(fig:cor-plot).
+These are displayed in the correlation matrix in Figure \@ref(fig:cor-plot).
 Figure \@ref(fig:pca-analysis) displays the resulting PCA.
@@
 Cells below the diagonal display the bivariate distributions with a fitted OLS-regression.
-The cells above the diagonal display and are coloured by the corresponding Pearsons correlation coefficient.", fig.height = 9.5, fig.width = 10}
+The cells above the diagonal display and are coloured by the corresponding Pearson's correlation coefficient.", fig.height = 9.5, fig.width = 10}
 # Transform the skewed variables - sqrt() is used where there are zero values,
@@
 ```{r simplot, fig.cap = "Temporal variation in the curation index for A) All lithics, and B) Flint.
-The temporal uncertainty is handled by means of a simulation approach where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B.
+The temporal uncertainty is handled by means of a simulation approach where the site ages are drawn from their respective age determination probability density functions given in Figure 1B.
 A LOESS curve has been fit to the distribution for each of the 1000 simulation runs.
@@
 The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component.
-One possible implication of this could be that the aggregation of artefact categories might have been overly conservative.
+One possible implication of this close correspondence could be that the aggregation of artefact categories might have been overly conservative.
 However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'.
@@
 This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
-These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages.
+These tendencies are already well-established when it comes to the formal tool types and some debitage categories, but have been given less focus in light of entire assemblages.
 Precisely what behavioural implication the development in the occurrences of the tool and debitage categories have are less clear, but appears to follow a different and more complex development over time than that of curation, as operationalised here.
 
-The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns is the main driver behind this empirical pattern --- in line with the framework of @barton2011.
+The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns are the main drivers behind this empirical pattern --- in line with the framework of @barton2011.
 Another perspective would be that this is not surprising at all [cf.
 @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different.
-If its applicability and relevance for this setting is accepted, the relationship does hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
-Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with Mesolithic south-eastern Norway.
-The explicit quantification does, however, offer the possibility of more formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.
-
-Having reached the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting can only constitute a first analytical step.
+However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting by means of an exploratory analysis can only constitute a first analytical step.
 As @elster2015[12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached.
 This, however, can only constitute the absolute minimum of what is required of a proposed explanation.
-Subsequent steps should be to probe and challenge this explanatory framework.
+Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses.
+The relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
+Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway.
+The explicit quantification does, however, offer the possibility to conduct formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.
 

### 2021-06-25 analysis/paper/paper.Rmd
Minor edits

@@
 The assemblages were explored following two strains of analysis.
-The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages.
-The findings pertaining to the artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research.
+The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely.
+The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, research that has been based on more subjectively driven methods.
 Furthermore, the chronological trends associated with variables taken from the so-called whole assemblage behavioural index, originally developed for characterising Paleolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature.
-This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside an exploratory setting, which might ultimately yield a powerful comparative tool for the analysis of Mesolithic assemblages.
+This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting, and which might ultimately yield a powerful comparative tool for the analysis of Mesolithic assemblages.
 
@@
 @åstveit2018; @breivik2018; @solheim2020].
-In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
-@sørensen1979; @romundset2018].
+In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.@romundset2018; @sørensen1979].
 The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf.
@@
 
+```{r supp1, cache = FALSE}
+sites %>% 
+  mutate(strt_bce = reported_start_bc * -1, 
+         end_bce = reported_end_bc * -1,
+         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
+  select(no, site, dating_method, strt_bce, end_bce) %>% 
+  gt() %>% 
+    cols_label(
+    site = "Site name",
+    dating_method = "Dating method",
+    strt_bce = "Reported start (BCE)",
+    end_bce = "Reported end (BCE)",
+  ) %>% 
+  tab_header(
+    title = "Table 1. Radiocarbon dates."
+  ) %>% 
+  tab_options(heading.title.font.size = 15)
+```
+
+\newpage
 ### Colophon

### 2021-07-06 analysis/paper/paper.Rmd
Changes after getting feedback from Ingrid. Also added supplementary material

--- 2021-06-15 analysis/paper/paper.Rmd
@@
 The assemblages were explored following two strains of analysis.
-The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages.
-The findings pertaining to the artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research.
+The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely.
+The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research---research that has been based on more subjectively driven methods.
 Furthermore, the chronological trends associated with variables taken from the so-called whole assemblage behavioural index, originally developed for characterising Paleolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature.
-This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside an exploratory setting, which might ultimately yield a powerful comparative tool for the analysis of Mesolithic assemblages.
+This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting.
+This might ultimately yield a powerful comparative tool for more extensive analyses of Mesolithic assemblages.
 
@@
 
-```{r libraries, eval = TRUE}
+```{r libraries, eval = TRUE, results = 'hide'}
 library(tidyverse)
@@
 library(factoextra)
+library(networkD3)
+library(webshot)
 library(sf)
@@
 
-#plot.margin = unit(c(0,135,0,135)
-# top right, borrom , l
-
 ```
@@
 Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g.
-@åstveit2018; @breivik2018; @solheim2020].
+@åstveit2018; @breivik2018; @møller1987; @solheim2020].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
-@sørensen1979; @romundset2018].
+@romundset2018; @sørensen1979].
 The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf.
@@
 Some categories in the original reports have therefore been combined in the dataset.
-This for example pertains to the category narrow-blades (width > 8mm and < 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
+This for example pertains to the category narrow-blades (width \> 8mm and \< 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
 Narrow-blades were combined with the blade category here.
@@
 @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020].
-Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@eigeland2015; @mansrud2016].
+Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@fuglestvedt2007; @eigeland2015; @mansrud2016].
 These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
@@
 In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
-Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [cf.
-@baxter1994, 71--77, 103].
+Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].
 
@@
 lithics <- sites %>%
-  # Combine flint projectiles into a single variable
+  # Combine variables
   mutate(
@@
 Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
-However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014[124] to suggest that the material might have been compromised.
+However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
 This could explain its position as an outlier in the plot.
 The use of metarhyolite for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b].
-In sum, the findings for the earliest sites are in large part in line with previous research [e.g.
-@breivik2018; @jaksland2014b; @damlien2018b].
+In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
 
@@
 The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades.
-Apart from the impact of core fragments, this must be considered very much in agreement with previous research [e.g.
+Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g.
 @solheim2017c, with references].
-The comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
-@damlien2016; @eigeland2015; @solheim2020].
+A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007].
+Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
+@eigeland2015; @fuglestvedt2007; @damlien2016; @solheim2020].
 However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
@@
                      repel = TRUE, shape = 21, pointsize = 2.9) +
-  scale_fill_gradient2(name = "Mean age BCE", low="yellow", mid="orange", 
+  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
 lithics2 <- sites %>% 
-  select(-site, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
+  select(-beach_flint_flint_nodule, -hammerstone, -excavation_project,
           -askeladden_id) %>% 
@@
 flint2 <- sites %>% 
-  select(-site, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
+  select(-beach_flint_flint_nodule, -hammerstone, -excavation_project, 
          -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
@@
 lithicstr <- lithics2 %>%
+  select(-site) %>% 
   # Transform skewed variables and reorder
@@
          title = "") + 
-  scale_fill_gradient2(name = "Mean age BCE",  low="yellow", mid="orange", 
+  scale_fill_gradient2(name = "Mean age kyr BCE",  low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
        y = "% secondarily worked lithics (log10)") +
-  scale_fill_gradient2(name = "Mean age BCE",  low="yellow", mid="orange", 
+  scale_fill_gradient2(name = "Mean age kyr BCE",  low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
        y = "% secondarily worked flint (log10)") +
-  scale_fill_gradient2(name = "Mean age BCE",  low="yellow", mid="orange", 
+  scale_fill_gradient2(name = "Mean age kyr BCE",  low="yellow", mid="orange", 
                         high="red", midpoint = centdate) +
@@
 minmax <- lithics2 %>%  
-   rownames_to_column(var = "site") %>% 
   mutate(dens =  ((lithic_dens - min(lithic_dens)) / (max(lithic_dens) - min(lithic_dens))),
@@
 minmaxflint <- flint2 %>% 
-   rownames_to_column(var = "site") %>% 
   mutate(dens =  ((flint_dens - min(flint_dens)) / (max(flint_dens) - min(flint_dens))),
@@
                   segment.color = 'grey50')
-#w3
+
 ```
@@
   theme_classic() +
-  theme(legend.position = "none")
+  theme(legend.position = "none",
+        text=element_text(size=16))
 
@@
                                "other" = "black"),
-                     guide = guide_legend(override.aes = list(size = 3,
+                     guide = guide_legend(override.aes = list(size = 5,
                                                               alpha = 1))) +
@@
   theme_classic() +
-  theme(legend.position = "bottom")
+  theme(legend.position = "bottom",
+        text=element_text(size=16))
 
@@
 It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE.
-This corresponds well with a chronological framework where the end of the Early Mesolithic, or Flake axe phase, is set to c.
-8200 BCE [@reitan2016].
+This corresponds well with a chronological framework where the Early Mesolithic, or Flake axe phase, is defined as lasting from c.
+9200--8200 BCE [@reitan2016].
+The beginning of the phase is set to start with the first human occupation in Norway, which is widely held as originating from South-Scandinavian and North-European regions, and which is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016].
 Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
@@
 The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021].
-The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021].
+The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021].
 Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021].
@@
 Referring back to the increasing expediency in the curation data between c.
-8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices were propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.
+8200 and 7000 BCE, the Microlith phase could thus represent a transitional period where migrating people and new living practices were propagating through societies in south-eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.
 
@@
 
-The results of the CA align more or less with results of previous research in south-eastern Norway.
+The results of the CA align fairly well with results of previous research in south-eastern Norway.
 This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
@@
 @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different.
-However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting by means of an exploratory analysis can only constitute a first analytical step.
-As @elster2015[12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached.
+However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting, reached by means of an exploratory analysis can only constitute a first analytical step.
+As @elster2015 [12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached.
 This, however, can only constitute the absolute minimum of what is required of a proposed explanation.
 Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses.
-The relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
+The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
 Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway.
@@
 :::
+
+\newpage
+
+# Supplementary material A. Radiocarbon dates.
+
+```{r supp1, cache = FALSE}
+radiocarbon %>% 
+  select(-excavation_project) %>% 
+  gt() %>% 
+    cols_label(
+    site = "Site name",
+    material = "Material",
+    lab_code = "Lab code",
+    c14_bp = "C14-age",
+    error = "Error"
+  )
+```
+
+# Supplementary material B. Aggregation of variables for the correspondence analysis.
+
+```{r supp2, cache = FALSE}
+links <- data.frame(value = 1, categories = names(sites %>% select(core:flake_retouch_other))) %>% 
+  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point", "transverse_point", 
+                                   "single_edged_point", "hognipen_point", "projectile_undefined"), "projectile",
+                          ifelse(categories %in% c("scraper", "knife", "burin", "drill", "blade_retouch",  
+                                                     "microblade_retouch", "chip_retouch", "fragment_retouch", 
+                                                   "flake_retouch"), "small_tool",
+                                 ifelse(categories %in% c("axe", "chisel"), "macro_tool", 
+                                 ifelse(categories %in% c("axe_other", "chisel_other", "club_hatchet_other"), "macro_tool_other", 
+                                 ifelse(categories %in% c("knife_other", "scraper_other", "drill_other", 
+                                                            "microblade_retouch_other", "blade_retouch_other", "fragment_retouch_other",
+                                                            "flake_retouch_other"), "small_tool_other", paste0(categories, " ")))))))
+
+nodes <- data.frame(
+  name=c(as.character(links$categories), 
+         as.character(links$new_categories)) %>% unique()
+)
+
+links$IDsource <- match(links$categories, nodes$name)-1 
+links$IDtarget <- match(links$new_categories, nodes$name)-1 
+nodes$group <- as.factor(c("none"))
+
+my_color <- 'd3.scaleOrdinal() .domain([none"]) .range(["white"])'
+
+sn <- sankeyNetwork(Links = links, Nodes = nodes,
+              Source = "IDsource", Target = "IDtarget",
+              NodeID = "name", Value = "value",
+              sinksRight=TRUE, nodeWidth = 0, fontSize = 10)
+
+htmlwidgets::saveWidget(sn, here::here('analysis/figures/sn1.html'))
+webshot::webshot(here::here('analysis/figures/sn1.html'), here::here('analysis/figures/sn1.pdf'), delay = 0.2)
+```
 

### 2021-07-06 analysis/paper/paper.Rmd
Minor edits

@@
               NodeID = "name", Value = "value",
-              sinksRight=TRUE, nodeWidth = 0, fontSize = 10)
+              sinksRight=TRUE, nodeWidth = 0, fontSize = 14)
 

### 2021-07-07 analysis/paper/paper.Rmd
Minor edits to text and added color to Slinkey plot

@@
 library(networkD3)
-library(webshot)
 library(sf)
@@
 However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
-@breivik2016a; @reitan2016; @Bicho2020; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
+@Bicho2020; @breivik2016a; @reitan2016; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
 In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
@@
 Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
-@eigeland2015; @fuglestvedt2007; @damlien2016; @solheim2020].
+@damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020].
 However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
@@
 ```{r supp1, cache = FALSE}
+# Supplementary table of radiocarbon dates
 radiocarbon %>% 
@@
 ```{r supp2, cache = FALSE}
-links <- data.frame(value = 1, categories = names(sites %>% select(core:flake_retouch_other))) %>% 
-  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point", "transverse_point", 
-                                   "single_edged_point", "hognipen_point", "projectile_undefined"), "projectile",
-                          ifelse(categories %in% c("scraper", "knife", "burin", "drill", "blade_retouch",  
-                                                     "microblade_retouch", "chip_retouch", "fragment_retouch", 
+# Most of the code for assembling the Sankey plot has been taken directly from 
+# various walkthroughs at https://www.r-graph-gallery.com
+
+# Create two columns: old and new variable names
+links <- data.frame(value = 1,
+        categories = names(sites %>% select(core:flake_retouch_other))) %>% 
+  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point",
+                                 "transverse_point", "single_edged_point", 
+                                 "hognipen_point", "projectile_undefined"), 
+                                 "projectile",
+                          ifelse(categories %in% c("scraper", "knife", "burin",
+                                 "drill", "blade_retouch", "microblade_retouch",
+                                 "chip_retouch", "fragment_retouch", 
                                  "flake_retouch"), "small_tool",
-                                 ifelse(categories %in% c("axe", "chisel"), "macro_tool", 
-                                 ifelse(categories %in% c("axe_other", "chisel_other", "club_hatchet_other"), "macro_tool_other", 
-                                 ifelse(categories %in% c("knife_other", "scraper_other", "drill_other", 
-                                                            "microblade_retouch_other", "blade_retouch_other", "fragment_retouch_other",
-                                                            "flake_retouch_other"), "small_tool_other", paste0(categories, " ")))))))
-
+                          ifelse(categories %in% c("axe", "chisel"),
+                                 "macro_tool", 
+                          ifelse(categories %in% c("axe_other", "chisel_other",
+                                 "club_hatchet_other"), "macro_tool_other", 
+                          ifelse(categories %in% c("knife_other", 
+                                 "scraper_other", "drill_other", 
+                                 "microblade_retouch_other", 
+                                 "blade_retouch_other", 
+                                 "fragment_retouch_other",
+                                 "flake_retouch_other"), "small_tool_other",
+                                 paste0(categories, " ")))))))
+
+# Set up nodes and links for the plot
 nodes <- data.frame(
@@
 nodes$group <- as.factor(c("none"))
-
-my_color <- 'd3.scaleOrdinal() .domain([none"]) .range(["white"])'
-
+links$group <- as.factor(links$new_categories)
+
+# Specify colours for the plot (this can be assembled using a combination of 
+# e.g. unique(links$new_connection), rep() and paste(), but I couldn't be 
+# bothered so I just copy+pasted and set it up manually).
+my_colour <-'d3.scaleOrdinal() .domain(["core", "core_fragment", "blade", "microblade", "chip", "fragment", "flake", "axe_debitage", "macro_tool", "projectile", "microburin", "small_tool", "beach_flint_flint_nodule", "macro_tool_other", "axe_debitage_other", "grinding_slab_stone", "projectile_other", "small_tool_other", "hammerstone", "flake_other", "fragment_other", "chip_other", "blade_other", "microblade_other", "core_other", "core_fragment_other,"]) .range(["lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "#ff8086", "#79d49f", "lightgrey", "#8eb8fa", "lightgrey", "#a88c7d", "lightgrey", "lightgrey", "lightgrey", "#ffee8c",  "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey"])'
+
+# Create plot
 sn <- sankeyNetwork(Links = links, Nodes = nodes,
               Source = "IDsource", Target = "IDtarget",
-              NodeID = "name", Value = "value",
-              sinksRight=TRUE, nodeWidth = 0, fontSize = 14)
-
+              NodeID = "name", Value = "value", colourScale = my_colour,
+              sinksRight = FALSE, LinkGroup = "new_categories", 
+              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)
+
+# This has to be saved as html widget and then retrieved as a pdf or png using
+# webshot.
 htmlwidgets::saveWidget(sn, here::here('analysis/figures/sn1.html'))
-webshot::webshot(here::here('analysis/figures/sn1.html'), here::here('analysis/figures/sn1.pdf'), delay = 0.2)
+webshot::webshot(here::here('analysis/figures/sn1.html'), 
+                 here::here('analysis/figures/sn1.pdf'), delay = 0.2)
 ```

### 2021-07-09 analysis/paper/paper.Rmd
Minor edits to text

@@
 
-However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
+However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
 @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
@@
 This could explain its position as an outlier in the plot.
-The use of metarhyolite for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b].
+The use of metarhyolite for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
 In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
@@
 @solheim2017c, with references].
-A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007].
+A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007; see also @bjerck2017].
 Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.

### 2021-07-22 analysis/paper/paper.Rmd
Started major restructuring after feedback from David

@@
 The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely.
-The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research---research that has been based on more subjectively driven methods.
-Furthermore, the chronological trends associated with variables taken from the so-called whole assemblage behavioural index, originally developed for characterising Paleolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature.
+The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods.
+Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Index, originally developed for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature.
 This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting.
@@
 The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports.
-The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [e.g.
+The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Index [e.g.
 @clark2017], which has been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.
@@
 
-The `r nrow(sites)` sites analysed here have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table 1).
-Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators.
-Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam].
-The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021].
-The resulting posterior density estimates were then summed for each site.
-Radiocarbon data is provided in the supplementary data, and has also been collated and reported by @solheim2020a.
-
 The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
@@
 ```
-
-# Methodology
-
-The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
-However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
-However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon.
-The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses.
-This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].
-
-The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+# Multivariate analysis of lithic assemblages
+
 The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies --- studies that have employed more informally driven methods.
@@
 However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
-Several scholars have built on this to argue that morphological variation in retouched lithics from the Paleolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g.
+Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g.
 @barton1991; @dibble1995; @barton2021].
-Consequently, several artefact categories have here been collapsed for the CA.
-This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
-That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading.
-These have all been combined into the single category "small flint tools".
-(A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
-While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
-An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
-While largely intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.
 
@@
 Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
-This has major implications that the above-outlined analysis does not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions.
 As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below].
@@
 @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
+
+# Methodology
+
+The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
+However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
+While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium on GitHub (https://github.com/isakro/exploring-assemblages-se-norway) and Zenodo ([placeholder]), following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data---data that is both selectively and subjectively reported upon.
+The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses.
+This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].
+
+The `r nrow(sites)` analysed sites have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table 1).
+Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators.
+Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam].
+The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021].
+The resulting posterior density estimates were then summed for each site.
+Radiocarbon data is provided in the supplementary data, and has also been collated and reported by @solheim2020a.
+
+The first part of the analysis of the lithic data involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+As this above-mentioned Frison effect, several artefact categories have been collapsed for the CA.
+(A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
+While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
+An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
+
 In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.

### 2021-07-23 analysis/paper/paper.Rmd
Working through David's comments

@@
 Some categories in the original reports have therefore been combined in the dataset.
-This for example pertains to the category narrow-blades (width \> 8mm and \< 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
+For example, this pertains to the category narrow-blades (width \> 8mm and \< 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
 Narrow-blades were combined with the blade category here.
@@
 These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
-To my knowledege, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
+To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
 In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
@@
 The general pattern follows the same trajectory, but the result for some individual sites is markedly different.
-This is discussed below.
 
@@
 
-The relevance of the relationship between frequency of secondarily worked lithics and volumetric density of lithics was here identified by means of an exploratory approach, and is in part investigated further because they align with suggestions from previous research --- clearly representing a possible case of both confirmation bias and circular reasoning.
-However, some inferential merit can be achieved by invoking what has been termed consilience.
-Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30].
-Thus, the overlap in results presented here and those repeatedly reported from a range of different context does speak to the general applicability and comparability offered by the measures, and gives an initial indication that they might be capturing the social dimensions of interest also in a Norwegian Mesolithic setting.
-
 The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE for the rest of the Mesolithic.

### 2021-07-25 analysis/paper/paper.Rmd
Reworking structure and attempting to number sites in overview map

@@
 ```
+The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c.
+9200--8200 BCE [@reitan2016].
+The phase is set to start with the first human occupation in Norway---widely held as originating from southern Scandinavian and northern European regions.
+This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016].
+Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
+@bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018].
+The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021].
+The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021].
+Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021].
+In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE.
+The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
+Following @reitan2016, the next typological transition at c.
+5600 BCE signifies the onset of the Nøstvet Adze Phase.
+While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g.
+@jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016].
+The subsequent Transverse Arrowhead Phase (c.
+4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016].
+It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c.
+4300 BCE [@nielsen2021].
 
@@
 # Combine spatial point data with assemblage data (holds excavation project
-# for colouring points in the maps)
-sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id"))
+# for colouring points in the maps). Set up labels to aggregate. There might be
+# a way to avoid having to do this manually, for example by using coordinates?
+sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id")) %>% 
+  mutate(no = as.numeric(no)) %>% 
+    mutate(no_m = ifelse(no %in% c(1:10, 20) & !duplicated(no %in% c(1:10, 20)),
+                         "1-10, 20",
+           #ifelse(no %in% c(7, 9:10, 20) & !duplicated(no %in% c(7, 9:10, 20)), "7, 9-10, 20",
+           ifelse(no %in% 11:12 & !duplicated(no %in% 11:12), "11-12", 
+           ifelse(no %in% 14:17 & !duplicated(no %in% 14:17), "14-17", 
+           ifelse(no %in% 21:22 & !duplicated(no %in% 21:22), "21-22", 
+           ifelse(no %in% 25:27 & !duplicated(no %in% 25:27), "25-27",
+           ifelse(no %in% c(19, 28:29, 23) & 
+                !duplicated(no %in% c(19, 28:29, 23)),
+                "19, 23, 28-29",
+           ifelse(no %in% c(32:33, 35) & 
+                !duplicated(no %in% c(32:33, 35)),
+                "32-33, 35",
+           #ifelse(no %in% 32:33 & !duplicated(no %in% 32:33), "32-33",
+           ifelse(no %in% c(24, 34, 37:40)  &
+                    !duplicated(no %in% c(24, 34, 37:40)), 
+                  "24, 34, 37-40",
+            ifelse(no %in% c(42:43, 45:46, 53)  &
+                    !duplicated(no %in% c(42:43, 45:46, 53)), 
+                  "42-43, 45-46, 53",
+           #ifelse(no %in% 42:43 & !duplicated(no %in% 42:43), "42-43",
+           #ifelse(no %in% 45:46 & !duplicated(no %in% 45:46), "45-46",
+           ifelse(no %in% 48:50 & !duplicated(no %in% 48:50), "48-50",
+           ifelse(no %in% c(31, 47, 51:52) & !duplicated(no %in% c(31, 47, 51:52)), "31, 47, 51-52",
+           ifelse(no %in% c(13, 30, 36,
+                            41, 44, 54), no, "")))))))))))))
 
@@
   geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
-  geom_sf(data = sitejoin, aes(fill = excavation_project),
-          size = 2.25, shape=21, 
+  geom_sf(data = sitejoin, aes(fill = excavation_project, label = no),
+          size = 2.25, shape = 21, 
           colour = "black") +
+  ggrepel::geom_text_repel(data = sitejoin, aes(label = no_m,
+                                                 geometry = geom),
+                                                 stat = "sf_coordinates",
+                           box.padding = 0.2, 
+                           max.overlaps = getOption("ggrepel.max.overlaps",
+                                                    default = 54)) +
   labs(fill = "Excavation project") +
@@
 
-Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g.
-@ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g.
-@berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020].
-Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@fuglestvedt2007; @eigeland2015; @mansrud2016].
-These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
-To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
-In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
-
 ```{r tab1}
@@
 ```
-# Multivariate analysis of lithic assemblages
-
-The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies --- studies that have employed more informally driven methods.
+# The analysis of lithic assemblages
+
+Studies concerned with chronological changes in the composition of lithic assemblages in southern Norway have typically had a focus on morphological variation among artefacts [e.g.
+@ballin1999; @bjerck1986; @reitan2016] or been concerned with technological processes associated with certain sub-categories of the site inventories, such as the production of blades or axes [e.g.
+@berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020].
+Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@fuglestvedt2007; @eigeland2015; @mansrud2016].
+These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
+To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
+In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
+
+The aim of the first of part of the analysis is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies of the Mesolithic in southern Norway that have employed more informally driven methods.
 This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful.
@@
 The first part of the analysis of the lithic data involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
-As this above-mentioned Frison effect, several artefact categories have been collapsed for the CA.
-(A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
+As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA.
+A full overview of the aggregated variables and their constituent parts is provided in the supplementary material.
+An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
 While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
-An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
-
-In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
+
+In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
 Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].
@@
 @Bicho2020; @breivik2016a].
-For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the previously mentioned Frison effect.
+For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the Frison effect.
 Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation.
@@
 
-Figure \@ref(fig:cor) displays the CA using the lithic count data.
-The general impression from the plots is that a chronological dimension is associated with the patterning in the data.
+The general impression from the CA is that a chronological dimension is associated with the patterning in the data (Figure \@ref(fig:cor)).
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
@@
 resultslong <- resultslong %>% 
-  mutate(site_col = ifelse(site %in% sitelabs, site, "other"))
+  mutate(site_col = ifelse(site %in% sitelabs, site, "Other sites"))
 
@@
                                              "Prestemoen 1",  "Hovland 3",
-                                             "Vallermyrene 4"),
+                                             "Vallermyrene 4", "Other sites"),
                     values = c("Langangen Vestgård 1" = "red",
@@
                                "Vallermyrene 4"  = "cyan1", 
-                               "other" = "black"),
+                               "Other sites" = "black"),
                      guide = guide_legend(override.aes = list(size = 3,
@@
                                "Vallermyrene 4"  = "cyan1", 
-                               "other" = "black"),
+                               "Other sites" = "black"),
                      guide = guide_legend(override.aes = list(size = 5,
@@
 While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust.
-It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE.
-This corresponds well with a chronological framework where the Early Mesolithic, or Flake axe phase, is defined as lasting from c.
-9200--8200 BCE [@reitan2016].
-The beginning of the phase is set to start with the first human occupation in Norway, which is widely held as originating from South-Scandinavian and North-European regions, and which is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016].
-Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
+The curation index is relatively high and uniform until some time before 8000 BCE.
+This corresponds well with the view that the Early Mesolithic is characterised by a relatively high and uniform degree of mobility [e.g.
 @bjerck2008; @breivik2016; @fuglestvedt2012].
-This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index.
-The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021].
-The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021].
-Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021].
-In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE.
-Referring back to the increasing expediency in the curation data between c.
-8200 and 7000 BCE, the Microlith phase could thus represent a transitional period where migrating people and new living practices were propagating through societies in south-eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.
-
-The Microlith phase is followed by the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
-As is evident from both the CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c.
-`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[1]]*-1`--`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[2]]*-1` BCE [@carrasco2014].
-The curation data remains stable from around 7000 BCE through the next typological transition at c.
-5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet adze phase.
-While previously defined as having a slightly longer duration, the Nøstvet phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in south-eastern Norway [e.g.
-@jaksland2001; @lindblom1984], and has been explicitly linked to an expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit with the term being somewhat vaguely invoked [@eigeland2015, 127--130].
-In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016].
-The curation index employed here clearly supports this notion, and suggests that the mobility patterns of the Nøstvet phase were well established in preceding periods.
-
-The subsequent Transverse arrowhead phase (c.
-4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016].
-It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c.
-4300 BCE [@nielsen2021].
-The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns in the material treated here.
-However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
+This is followed by a marked increase in expedience, which has stabilised by around 7000 BCE.
+Again, this corresponds well with the employed chronological framework.
+Referring back to the demographic changes that are to take place around this transition, the Microlith phase could thus represent a period where migrating people and new living practices were propagating through societies in south-eastern Norway---a process that in light of the curation data would have concluded around 7000 BCE.
+
+The curation index then remains stable for the rest of the Mesolithic.
+This suggests that the transition to mobility patterns traditionally ascribed to the Nøstvet Phase can indeed be traced back to the Microlith Phase [@solheim2016].
+The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns in the material treated here.
+However, it is worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
 

### 2021-07-25 analysis/paper/paper.Rmd
Starting to seperate results and discussion more clearly

@@
 9200--8200 BCE [@reitan2016].
-The phase is set to start with the first human occupation in Norway---widely held as originating from southern Scandinavian and northern European regions.
+The phase is set to start with the first human occupation in Norway which is widely held as originating from southern Scandinavian and northern European regions.
 This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016].
@@
 # Combine spatial point data with assemblage data (holds excavation project
-# for colouring points in the maps). Set up labels to aggregate. There might be
+# for colouring points in the maps). Set up aggregate labels. There might be
 # a way to avoid having to do this manually, for example by using coordinates?
@@
   mutate(no = as.numeric(no)) %>% 
-    mutate(no_m = ifelse(no %in% c(1:10, 20) & !duplicated(no %in% c(1:10, 20)),
-                         "1-10, 20",
-           #ifelse(no %in% c(7, 9:10, 20) & !duplicated(no %in% c(7, 9:10, 20)), "7, 9-10, 20",
+    mutate(no_m = ifelse(no %in% c(1:10) & !duplicated(no %in% c(1:10)),
+                         "1-10",
            ifelse(no %in% 11:12 & !duplicated(no %in% 11:12), "11-12", 
            ifelse(no %in% 14:17 & !duplicated(no %in% 14:17), "14-17", 
-           ifelse(no %in% 21:22 & !duplicated(no %in% 21:22), "21-22", 
-           ifelse(no %in% 25:27 & !duplicated(no %in% 25:27), "25-27",
+           ifelse(no %in% c(21:22, 44) & !duplicated(no %in% c(21:22, 44)),
+                  "21-22, 44", 
+           ifelse(no %in% c(25:27, 30) & !duplicated(no %in% c(25:27, 30)),
+                  "25-27, 30",
+           ifelse(no %in% c(36, 41) &
+                    !duplicated(no %in% c(36, 41)),
+                  "36, 41",        
            ifelse(no %in% c(19, 28:29, 23) & 
@@
                 "32-33, 35",
-           #ifelse(no %in% 32:33 & !duplicated(no %in% 32:33), "32-33",
-           ifelse(no %in% c(24, 34, 37:40)  &
-                    !duplicated(no %in% c(24, 34, 37:40)), 
-                  "24, 34, 37-40",
-            ifelse(no %in% c(42:43, 45:46, 53)  &
-                    !duplicated(no %in% c(42:43, 45:46, 53)), 
-                  "42-43, 45-46, 53",
-           #ifelse(no %in% 42:43 & !duplicated(no %in% 42:43), "42-43",
-           #ifelse(no %in% 45:46 & !duplicated(no %in% 45:46), "45-46",
-           ifelse(no %in% 48:50 & !duplicated(no %in% 48:50), "48-50",
-           ifelse(no %in% c(31, 47, 51:52) & !duplicated(no %in% c(31, 47, 51:52)), "31, 47, 51-52",
-           ifelse(no %in% c(13, 30, 36,
-                            41, 44, 54), no, "")))))))))))))
+           ifelse(no %in% c(37:40)  &
+                    !duplicated(no %in% c(37:40)), 
+                  "37-40",
+           ifelse(no %in% c(24, 34)  &
+                    !duplicated(no %in% c(24, 34)), 
+                  "24, 34",
+           ifelse(no %in% c(42:43)  &
+                    !duplicated(no %in% c(42:43)), 
+                  "42-43",
+          ifelse(no %in% c(45:46, 53)  &
+                    !duplicated(no %in% c(45:46, 53)), 
+                  "45-46, 53",
+           ifelse(no %in% c(48:50, 54) & !duplicated(no %in% c(48:50, 54)),
+                  "48-50, 54",
+           ifelse(no %in% c(31, 47, 51:52) &
+                    !duplicated(no %in% c(31, 47, 51:52)),
+                  "31, 47, 51-52",
+           ifelse(no %in% c(13, 20), no, ""))))))))))))))))
 
@@
 The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
-However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
-This could explain its position as an outlier in the plot.
-The use of metarhyolite for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
-In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
 
@@
 Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
-This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
-@eigeland2015, 370; @glørstad2011; @reitan2016].
 
@@
 The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades.
-Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g.
-@solheim2017c, with references].
-A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007; see also @bjerck2017].
-Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
-@damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020].
-However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
-The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf.
-@manninen2014].
 
@@
 
+Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
+However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
+This could explain its position as an outlier in the plot.
+This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
+@eigeland2015, 370; @glørstad2011; @reitan2016].
+Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g.
+@solheim2017c, with references].
+A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007; see also @bjerck2017].
+Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
+@damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020].
+However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
+The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf.
+@manninen2014].
+The use of metarhyolite for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
+In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
+
 The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component.

### 2021-07-26 analysis/paper/paper.Rmd
Introducing logistic/residential a bit more throughly

@@
 ```
+
 The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c.
@@
 The phase is set to start with the first human occupation in Norway which is widely held as originating from southern Scandinavian and northern European regions.
-This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016].
+This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @berg-hansen2018; @fuglestvedt2012; @glørstad2016].
 Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
@@
 ```
+
 # The analysis of lithic assemblages
@@
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
+
 In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern.
@@
 However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
-@Bicho2020; @breivik2016a; @reitan2016; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
+@Bicho2020; @breivik2016a; @reitan2016; @kitchel2021]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
+
+This relationship has been related to the continuum defined by @binford1980 between residentially mobile foragers and logistically mobile collectors.
+Residential mobility involves the movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps from where task groups venture to retrieve resources, which can lead to a wider spectrum of site types and associated mobility patterns.
+While this framwork can offer
 
@@
 However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium on GitHub (https://github.com/isakro/exploring-assemblages-se-norway) and Zenodo ([placeholder]), following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium on GitHub (<https://github.com/isakro/exploring-assemblages-se-norway>) and Zenodo ([placeholder]), following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
 However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data---data that is both selectively and subjectively reported upon.
@@
 The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
+However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
+This could explain its position as an outlier in the plot.
 
@@
 
-Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
-However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
-This could explain its position as an outlier in the plot.
-This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g.
+The results of the CA does appear to align well with previous research [e.g.
+@solheim2017c, with references].
+In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [@bjerck2017; @breivik2018; @fuglestvedt2007; @eymundsson2018; @jaksland2014b; @damlien2018b].
+The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [e.g.
+@damlien2016].
+When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones.
+The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016].
+However, in agreement with the literature, this is evidently not as prominent a part of these assemblages [@reitan2016; @damlien2018b].
+
+The flint material of the later sites are to larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g.
+@glørstad2010, 161--163; @solheim2020].
+A more fragmented flint material, as indicated by the relative importance of flint chips and fragments, is also a previously noted property of some later Mesolithic, as well as early Neolithic sites [e.g.
+@fossum2017; @stokke2018a].
+The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages.
+It was noted above that despite the low number of the absolute earliest sites, these appear to not be impacted by this material.
+This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g.
 @eigeland2015, 370; @glørstad2011; @reitan2016].
-Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g.
-@solheim2017c, with references].
-A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007; see also @bjerck2017].
+
 Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
@@
 However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
-The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf.
+The overall pattern does nonetheless speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf.
 @manninen2014].
-The use of metarhyolite for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages.
-In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
-
-The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component.
-One possible implication of this close correspondence could be that the aggregation of artefact categories might have been overly conservative.
+
+One possible implication of the fact that the employed artefact categories are so clearly capturing a temporal component could consequently be that the aggregation of artefact categories might have been overly conservative.
 However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'.
@@
 The curation index then remains stable for the rest of the Mesolithic.
-This suggests that the transition to mobility patterns traditionally ascribed to the Nøstvet Phase can indeed be traced back to the Microlith Phase [@solheim2016].
+This suggests that the transition to mobility patterns traditionally ascribed to the Nøstvet Phase can indeed be traced back to the Microlith Phase [cf.
+@solheim2016].
 The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns in the material treated here.

### 2021-07-29 analysis/paper/paper.Rmd
Cutting down text

@@
 bibliography: references.bib
-csl: "../templates/jas_reports.csl" # Insert path for the bib-style
+csl: "../templates/saa.csl" # Insert path for the bib-style
 abstract: |
@@
 The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods.
-Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Index, originally developed for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature.
+Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g.
+Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature.
 This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting.
-This might ultimately yield a powerful comparative tool for more extensive analyses of Mesolithic assemblages.
+Furthermore, this supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.
 
@@
 The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports.
-The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Index [e.g.
-@clark2017], which has been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.
+The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g.
+@clark2017], which have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.
 
@@
 The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c.
-9200--8200 BCE [@reitan2016].
-The phase is set to start with the first human occupation in Norway which is widely held as originating from southern Scandinavian and northern European regions.
-This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @berg-hansen2018; @fuglestvedt2012; @glørstad2016].
+9200--8200 BCE [@reitan2016] and is set to start with the first recorded human presence in Norway.
 Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
-@bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018].
-The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021].
-The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021].
-Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021].
-In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE.
-The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
-Following @reitan2016, the next typological transition at c.
+@bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018c].
+Around the transition to the subsequent Middle Mesolithic, or Microlith Phase, at 8200 BCE pervasive changes in blade and axe technology takaes place [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021].
+The Microlith Phase is defined as lasting until around 7000 BCE, which is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
+The next typological transition at c.
 5600 BCE signifies the onset of the Nøstvet Adze Phase.
-While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g.
-@jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016].
+While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns [e.g.
+@jaksland2001; @lindblom1984].
+In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016].
 The subsequent Transverse Arrowhead Phase (c.
-4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016].
+4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of new flint projectiles [@reitan2016].
 It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c.
 4300 BCE [@nielsen2021].
-
-The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway (Figure \@ref(fig:dist)A).
-The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
-The sites included in the analysis consist of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts.
-The institution responsible for these excavations was the Museum of Cultural History in Oslo.
-This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations.
-Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum [@melvold2014c; @koxvold2017].
-As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g.
-@clark2006; @dibble2017].
-In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@jaksland2014; @solheim2013b, see also Figure \@ref(fig:dist)B].
-Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages.
-While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects.
 
@@
 @bjerck2008].
-Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g.
-@åstveit2018; @breivik2018; @møller1987; @solheim2020].
+Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@åstveit2018; @breivik2018; @møller1987; @solheim2020].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
@@
 In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of losing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007].
-However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space.
-In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year towards the end of the Mesolithic [@sørensen2014].
-Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, it is worth bearing in mind that this could vary depending on when and where they were in use, potentially reducing the degree to which their composition can be directly compared.
-
-The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
-While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions.
-Some categories in the original reports have therefore been combined in the dataset.
-For example, this pertains to the category narrow-blades (width \> 8mm and \< 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites.
-Narrow-blades were combined with the blade category here.
-Furthermore, the artefact data have here been divided into flint and non-flint materials.
-Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g.
+
+The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway (Figure \@ref(fig:dist)A).
+The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
+The sites included in the analysis consist of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts.
+The institution responsible for these excavations was the Museum of Cultural History in Oslo.
+This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations.
+Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum [@melvold2014c; @koxvold2017].
+As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [cf.
+@clark2006; @dibble2017].
+
+The lithic data analysed is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
+The artefact data have been divided into flint and non-flint materials.
+Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g.
 @berg-hansen1999].
-This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46].
-Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53].
+This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46].
+Thus, while flint is treated as a unified category here, the variability in quality could have been substantial.
 Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools.
@@
 @frivoll2017].
-While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g.
-@eigeland2015, 370; @glørstad2011].
 Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.
@@
 
-The aim of the first of part of the analysis is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies of the Mesolithic in southern Norway that have employed more informally driven methods.
-This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful.
+The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods.
+This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful.
 However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded.
@@
 @barton1991; @dibble1995; @barton2021].
+Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading.
 
@@
 Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
-As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below].
+As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [@barton2011; @clark2017, and below].
 The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1977; @binford1979].
 An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement.
-Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.
-
-However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed, and no single definition has ever been reached [e.g.
+Curated technological organisation, on the other hand, has been related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.
+
+However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed [e.g.
 @bamforth1986; @nash1996; @shott1996; @surovell2009, 9--13].
-The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies and raw-material quality and availability [e.g.
-@andrefsky1994; @parry1987; @kuhn1992; @smith2015; @clark2017].
 Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted.
-For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g.
+For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool---or extending its use-life---by the repeated rejuvenation and modification of edges [e.g.
 @bamforth1986; @dibble1995; @shott2005].
 
-Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g.
+One concrete operationalisation of the terms has been forwarded by @barton1998 and collegues [e.g.
 @barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008].
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
+Furthermore, the relationship between curated and expedient technological organisation has been related to the continuum defined by @binford1980 between residentially mobile foragers and logistically mobile collectors [@riel-salvatore2004; @clark2017; see also @bamforth1986; @binford1977].
+Residential mobility involves the movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps that are moved less often and from where task-groups venture on targeted forays to retrieve specific resources.
+A higher degree of logistic as opposed to residential mobility thus involves a wider range of site types and associated mobility patterns [@binford1980].
 
@@
 The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades.
-These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.
-
-However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
+These variables and underlying logic constitute what has been termed Whole Assemblage Behavioural Indicators [WABI, @clark2017], and is the main framework adopted here.
+
+As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
 @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
 
-This relationship has been related to the continuum defined by @binford1980 between residentially mobile foragers and logistically mobile collectors.
-Residential mobility involves the movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps from where task groups venture to retrieve resources, which can lead to a wider spectrum of site types and associated mobility patterns.
-While this framwork can offer
-
 # Methodology
 
-The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified.
-However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium on GitHub (<https://github.com/isakro/exploring-assemblages-se-norway>) and Zenodo ([placeholder]), following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
-However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data---data that is both selectively and subjectively reported upon.
-The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses.
-This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].
+The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
+While only parts of this process can sensibly be reported upon, the data and employed R programming script is freely available as a research compendium at [URL placeholder], following @marwick2018b.
 
@@
 The resulting posterior density estimates were then summed for each site.
-Radiocarbon data is provided in the supplementary data, and has also been collated and reported by @solheim2020a.
-
-The first part of the analysis of the lithic data involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+Radiocarbon data is provided in the supplementary data (see also @solheim2020a).
+
+The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
 As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA.
-A full overview of the aggregated variables and their constituent parts is provided in the supplementary material.
-An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
+This for example pertains to flint tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
+These have all been combined into the single category "small flint tools".
+(A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
 While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
 
-In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
-Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].
+Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
+Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77].
 
@@
 @Bicho2020; @breivik2016a].
-For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the Frison effect.
-Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation.
+For the assemblage data itself this especially pertains to diversity in tool-types [@canessa2021], which has been omitted in light of the above-metioned Frison effect.
+Number of features on the sites has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation.
 Similarly, the number of activity areas, effectively number of artefact clusters, however defined, has also been disregarded.
 This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017].
-This pertains for example to the impact of bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].
+This pertains for example to bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have impacted several of the sites treated here [@jørgensen2017; @darmark2018e].
 
@@
 The general impression from the CA is that a chronological dimension is associated with the patterning in the data (Figure \@ref(fig:cor)).
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120].
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120].
 The earliest sites tend to be located in the upper left corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
-Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
+The sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
 
@@
 The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014].
+
+The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages.
+Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among the sites.
+Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious early case with an assemblage characterised by axe production in metarhyolite [@eigeland2014].
 However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
 This could explain its position as an outlier in the plot.
-
-The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages.
-Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among the sites.
-Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
+Finally, although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
 
@@
 While not as substantial, there is clear temporal patterning in the flint data as well.
-This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites.
+This is most marked for the earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools and debitage from their production characterises these sites.
 Slightly younger sites appear more impacted by core fragments and blades.
@@
 
-Moving on to the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed.
-These are displayed in the correlation matrix in Figure \@ref(fig:cor-plot).
+Moving on to the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed (Figure \@ref(fig:cor-plot)).
 Figure \@ref(fig:pca-analysis) displays the resulting PCA.
-There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot.
+There is a general temporal transition from the upper left to the bottom right of the plot.
 The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure \@ref(fig:var-contrib)).
@@
 While there is a slight tendency for blanks to be more associated with younger sites, frequency of chips appears to be largely independent of time.
-However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation.
+However, this almost suspiciously strong negative correlation can perhaps have a practical explanation.
 Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot)), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification.
 This could conceivably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material.
-While this is not necessarily the entire explanation behind the relationship, this does make it difficult to place much analytical weight on this pattern.
-Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well.
-That is not to say that cores may not be indicative or related to mobility patterns, but to get at this may require further analysis beyond their simple classification as cores [@kitchel2021].
+While this is not necessarily the entire explanation, this does make it difficult to place much analytical weight on this pattern.
+Relative frequency of cores is not especially impactful in the PCA, and appears to be independent of the temporal dimension as well.
+That is not to say that cores may not be indicative of or related to mobility patterns, but to get at this may require further analysis beyond their simple classification as cores [@kitchel2021].
 
@@
 
-Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites.
+There is a strong negative correlation between the two variables (*r =* `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher VDL and a lower RFSL than older sites (Figure \@ref(fig:wabi)A).
 The linear correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`).
-As variable non-flint availability and workability has also been suggested to potentially impact these dimensions [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data.
-While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same.
+Variable non-flint availability and workability has also been suggested to potentially impact these dimensions [cf.
+@manninen2014], but while the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same (Figure \@ref(fig:wabi)B).
 The relationship between mean site age and relative frequency of secondarily worked flint is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint is not temporally contingent (*r =* `r round(denscorflint, 2)`).
@@
 @baxter2016; @crema2012; @orton2017].
-A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist).
-For sites with radiocarbon age determinations the dates were drawn from the associated summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
-This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot)A.
+A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from their respective date ranges as provided in Figure \@ref(fig:dist).
+For sites with radiocarbon age determinations the dates were drawn from the summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
+This simulation was repeated 1000 times (Figure \@ref(fig:simplot)A).
 Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE.
-This has stabilised by around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic.
-The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in associated mobility patterns.
+This has stabilised by around 7000 BCE and remains stable for the rest of the Mesolithic.
+The variation in degree of curation is also markedly higher after 8000 BCE.
 Figure \@ref(fig:simplot)B displays the result of running the same procedure on the flint data.
-The general pattern follows the same trajectory, but the result for some individual sites is markedly different.
+The general pattern follows the same trajectory, but the result for some individual sites is noticeably different.
 
@@
 
-The results of the CA does appear to align well with previous research [e.g.
+The results of the CA appear to align well with previous research [e.g.
 @solheim2017c, with references].
-In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [@bjerck2017; @breivik2018; @fuglestvedt2007; @eymundsson2018; @jaksland2014b; @damlien2018b].
-The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [e.g.
-@damlien2016].
+In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf.
+@bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
+The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016].
 When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones.
 The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016].
-However, in agreement with the literature, this is evidently not as prominent a part of these assemblages [@reitan2016; @damlien2018b].
+However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.
 
 The flint material of the later sites are to larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g.
-@glørstad2010, 161--163; @solheim2020].
+@solheim2020].
 A more fragmented flint material, as indicated by the relative importance of flint chips and fragments, is also a previously noted property of some later Mesolithic, as well as early Neolithic sites [e.g.
@@
 The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages.
-It was noted above that despite the low number of the absolute earliest sites, these appear to not be impacted by this material.
-This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g.
-@eigeland2015, 370; @glørstad2011; @reitan2016].
-
-Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g.
-@damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020].
-However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007].
-The overall pattern does nonetheless speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf.
-@manninen2014].
-
-One possible implication of the fact that the employed artefact categories are so clearly capturing a temporal component could consequently be that the aggregation of artefact categories might have been overly conservative.
+It was noted above that this material does not seem to impact the latest sites, which could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g.
+@glørstad2011; @reitan2016].
+
+One implication of the fact that the employed artefact categories are so clearly capturing a temporal component could be that the aggregation of artefact categories might have been overly conservative.
 However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'.
-The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern.
+The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some significant deviances from the overall pattern.
 Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely.
@@
 
-The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE for the rest of the Mesolithic.
+The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises around 7000 BCE.
 This pattern is evident in both the flint data and when all lithics are treated in aggregate.
-Furthermore, the variation in degree of curation in Figure \@ref(fig:simplot)A could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE.
+Furthermore, the increased variation in degree of curation after around 8000 BCE could indicate that these sites were associated with a more varied mobility pattern.
 The five sites that have values on the curation index below c.
--0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern [*sensu* @binford1980].
+-0.25 could in this perspective have predominantly functioned as base-camps within a logistic settlement pattern.
 That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility [@eigeland2014b].
@@
 As can be seen in Figure \@ref(fig:simplot)B, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages.
-Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exclusion would likely lead to its underestimation.
-One possible approach could be to weigh the curation index by proportion of non-flint material in the assemblages.
+Thus, the degree of expediency for assemblages dominated by non-flint might be somewhat exaggerated when the non-flint material is included, while its exclusion would likely lead to its underestimation.
+One possible approach could be to weigh the curation index by the proportion of non-flint material in the assemblages.
 This is not explored further here, however, as the overall tendencies are relatively robust to this effect.
+
 Another case also worth commenting on is Langangen Vestgård 1, which, on the grounds of an overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014].
-However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a comparable duration to those on contemporary sites, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole [cf.
+However, the relatively high value on the curation index could mean that the site reflects the aggregation of stays which predominantly have been of a comparable duration to those on contemporary sites, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole [cf.
 @barton2014].
@@
 The curation index is relatively high and uniform until some time before 8000 BCE.
-This corresponds well with the view that the Early Mesolithic is characterised by a relatively high and uniform degree of mobility [e.g.
+This corresponds well with the view that the Early Mesolithic is characterised by a high and uniform degree of mobility [e.g.
 @bjerck2008; @breivik2016; @fuglestvedt2012].
@@
 The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns in the material treated here.
-However, it is worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
+However, it is worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed.
 
@@
 These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
-Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@barton2011; @binford1980].
+Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system.
 
@@
 
-The results of the CA align fairly well with results of previous research in south-eastern Norway.
+The results of the CA align well with results of previous research in south-eastern Norway.
 This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
@@
 
-The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns are the main drivers behind this empirical pattern --- in line with the framework of @barton2011.
+The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns are the main drivers behind this empirical pattern---in line with the framework of @barton2011.
 Another perspective would be that this is not surprising at all [cf.
 @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different.
-However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting, reached by means of an exploratory analysis can only constitute a first analytical step.
+However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a relatively controlled empirical setting, reached by means of an exploratory analysis can only constitute a first analytical step.
 As @elster2015 [12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached.
@@
 The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
-Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway.
+Furthermore, the temporal trends of the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway.
 The explicit quantification does, however, offer the possibility to conduct formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.

### 2021-07-29 analysis/paper/paper.Rmd
Cutting down text. Still need table of chronological phases

@@
   arrange(mean_date) %>% 
-  rownames_to_column(var = "no")
+  tibble::rownames_to_column(var = "no")
 
@@
 @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018c].
-Around the transition to the subsequent Middle Mesolithic, or Microlith Phase, at 8200 BCE pervasive changes in blade and axe technology takaes place [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021].
+Around the transition to the subsequent Middle Mesolithic or Microlith Phase at c.
+8200 BCE, pervasive changes in blade and axe technology occur [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021].
 The Microlith Phase is defined as lasting until around 7000 BCE, which is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016].
@@
 # a way to avoid having to do this manually, for example by using coordinates?
-sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id")) %>% 
-  mutate(no = as.numeric(no)) %>% 
-    mutate(no_m = ifelse(no %in% c(1:10) & !duplicated(no %in% c(1:10)),
-                         "1-10",
-           ifelse(no %in% 11:12 & !duplicated(no %in% 11:12), "11-12", 
-           ifelse(no %in% 14:17 & !duplicated(no %in% 14:17), "14-17", 
-           ifelse(no %in% c(21:22, 44) & !duplicated(no %in% c(21:22, 44)),
-                  "21-22, 44", 
-           ifelse(no %in% c(25:27, 30) & !duplicated(no %in% c(25:27, 30)),
-                  "25-27, 30",
-           ifelse(no %in% c(36, 41) &
-                    !duplicated(no %in% c(36, 41)),
-                  "36, 41",        
-           ifelse(no %in% c(19, 28:29, 23) & 
-                !duplicated(no %in% c(19, 28:29, 23)),
-                "19, 23, 28-29",
-           ifelse(no %in% c(32:33, 35) & 
-                !duplicated(no %in% c(32:33, 35)),
-                "32-33, 35",
-           ifelse(no %in% c(37:40)  &
-                    !duplicated(no %in% c(37:40)), 
-                  "37-40",
-           ifelse(no %in% c(24, 34)  &
-                    !duplicated(no %in% c(24, 34)), 
-                  "24, 34",
-           ifelse(no %in% c(42:43)  &
-                    !duplicated(no %in% c(42:43)), 
-                  "42-43",
-          ifelse(no %in% c(45:46, 53)  &
-                    !duplicated(no %in% c(45:46, 53)), 
-                  "45-46, 53",
-           ifelse(no %in% c(48:50, 54) & !duplicated(no %in% c(48:50, 54)),
-                  "48-50, 54",
-           ifelse(no %in% c(31, 47, 51:52) &
-                    !duplicated(no %in% c(31, 47, 51:52)),
-                  "31, 47, 51-52",
-           ifelse(no %in% c(13, 20), no, ""))))))))))))))))
+sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id"))
 
@@
   geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
-  geom_sf(data = sitejoin, aes(fill = excavation_project, label = no),
+  geom_sf(data = sitejoin, aes(fill = excavation_project),
           size = 2.25, shape = 21, 
           colour = "black") +
-  ggrepel::geom_text_repel(data = sitejoin, aes(label = no_m,
-                                                 geometry = geom),
-                                                 stat = "sf_coordinates",
-                           box.padding = 0.2, 
-                           max.overlaps = getOption("ggrepel.max.overlaps",
-                                                    default = 54)) +
   labs(fill = "Excavation project") +
@@
   ) %>% 
-  tab_options(heading.title.font.size = 15)
+  tab_options(heading.title.font.size = 13, heading.align = "left")
 ```
@@
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120].
-The earliest sites tend to be located in the upper left corner of plot A, with increasingly younger sites towards the bottom along the second dimension.
-The sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.
-
-The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
+
+The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
 It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves.
 The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
-
 The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages.
-Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among the sites.
-Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious early case with an assemblage characterised by axe production in metarhyolite [@eigeland2014].
+Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that it is an early assemblage characterised by axe production in metarhyolite [@eigeland2014].
 However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
@@
 
-Thus, while some secondary expectations of the WABI does not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
+Thus, while some secondary expectations of the WABI do not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
 The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material.
@@
 
-```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint.
+```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts for A) All lithics, B) Flint.
 The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
@@
 When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones.
-The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016].
+The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016].
 However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.
@@
 The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages.
-It was noted above that this material does not seem to impact the latest sites, which could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g.
+It was noted above that this material does not seem to impact the latest sites, which would indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g.
 @glørstad2011; @reitan2016].
@@
 The curation index is relatively high and uniform until some time before 8000 BCE.
-This corresponds well with the view that the Early Mesolithic is characterised by a high and uniform degree of mobility [e.g.
-@bjerck2008; @breivik2016; @fuglestvedt2012].
+This corresponds well with the view that the Early Mesolithic is characterised by a high and uniform degree of mobility.
 This is followed by a marked increase in expedience, which has stabilised by around 7000 BCE.
@@
 These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
-Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system.
+Although the precise nature of this transition would require further consideration, the fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system.
 
@@
 
-\newpage
-
-### Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
-
-The current Git commit details are:
-
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```
+<!-- \newpage -->
+
+<!-- ### Colophon -->
+
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies: -->
+
+<!-- ```{r colophon, cache = FALSE} -->
+<!-- # which R packages and versions? -->
+<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
+<!-- ``` -->
+
+<!-- The current Git commit details are: -->
+
+<!-- ```{r} -->
+<!-- # what commit is this file at?  -->
+<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
+<!-- ``` -->

### 2021-07-30 analysis/paper/paper.Rmd
Created table holding chronological framework.

@@
 4300 BCE [@nielsen2021].
+```{r tab1}
+period <- c('Early Mesolithic, "Fosna Phase"', 'Middle Mesolithic, "Tørkop Phase"', 'Late Mesolithic, "Nøstvet Phase"', 'Late Mesolithic, "Kjeøy Phase"',
+            'Flake Axe Phase', 'Microlith Phase', 'Pecked Adze Phase', 'Nøstvet Adze Phase', 'Transverse Arrowhead Phase')
+duration <- c("9500–8250 BCE", "8250–6350 BCE", "6350–4650 BCE", "4650–3800 BCE", "9200–8200 BCE", "8200–7000 BCE", "7000–5600 BCE", "5600–4500 BCE", "4500–3900 BCE")
+
+df <- data.frame(period, duration)
+
+gt(df) %>% 
+  tab_row_group(
+    label = "Glørstad (2004)",
+    rows = 1:4
+  ) %>% 
+  tab_row_group(
+    label = "Reitan (2016)",
+    rows = 5:9
+  ) %>% 
+  tab_header(
+    title = "Table 1.
+Chronological frameworks.
+Glørstad's (2004) divison of phases reflects the traditional and more established frameworks, to which Reitan (2016) has recently suggested considerable changes."
+  ) %>% 
+  tab_options(column_labels.hidden = TRUE,
+              row_group.font.weight = "bold", 
+              heading.title.font.size = 13,
+              heading.align = "left")
+
+```
 
@@
 These follow the original reports.
-Site numbering follow that given in Table 1.", fig.height = 9, fig.width = 12, eval = FALSE}
+Site numbering follow that given in Table 2.", fig.height = 9, fig.width = 12, eval = FALSE}
 temp_dist
@@
 
-```{r tab1}
+```{r tab2}
 
@@
   tab_header(
-    title = "Table 1. Analysed sites."
+    title = "Table 2. Analysed sites."
   ) %>% 
@@
 To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g.
-@baxter2016; @crema2012; @orton2017].
+@crema2012; @orton2017].
 A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from their respective date ranges as provided in Figure \@ref(fig:dist).

### 2021-08-01 README.Rmd
Removed site labels from CA and PCA plots. Some additional plots and tables exploring the components of the PCA
Final edits

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Exploring the composition of lithic assemblages in Mesolithic
+south-eastern Norway"
 ```
 
-# dialpastrepository
+# exploring-assemblages-se-norway
 
@@
 
-This repository contains the data and code for our paper:
+<!-- This repository contains the data and code for the paper: -->
 
-> Authors, (YYYY). *`r Title`*. Name of journal/book
-> <https://doi.org/xxx/xxx>
+<!-- > Authors, (YYYY). *`r Title`*. Name of journal/book -->
+<!-- > <https://doi.org/xxx/xxx> -->
 
-Our pre-print is online here:
+Pre-print is online here:
 
-> Authors, (YYYY). *`r Title`*. Name of journal/book, Accessed
+>  Roalkvam, Isak (2021). *`r Title`*. Accessed
 > `r format(Sys.Date(), "%d %b %Y")`. Online at
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). *Compendium of R code and
+> Roalkvam, Isak (`r format(Sys.Date(), "%Y")`). *Compendium of R code and
 > data for `r Title`*. Accessed `r format(Sys.Date(), "%d %b %Y")`.
@@
     generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
+    `paper.pdf`, suitable for reading (the code is replaced by figures
     and tables in this file)

### 2021-08-01 analysis/paper/paper.Rmd
Final edits

--- 2021-07-29 analysis/paper/paper.Rmd
@@
 library(patchwork)
-library(gt)
+library(knitr)
+library(kableExtra)
 library(FactoMineR)
@@
 4300 BCE [@nielsen2021].
+
+```{r tab1}
+
+# Table 1. Overfiew of chronological framework.
+Period <- c('Early Mesolithic, Fosna Phase', 'Middle Mesolithic, Tørkop Phase', 'Late Mesolithic, Nøstvet Phase', 'Late Mesolithic, Kjeøy Phase',
+            'Flake Axe Phase', 'Microlith Phase', 'Pecked Adze Phase', 'Nøstvet Adze Phase', 'Transverse Arrowhead Phase')
+Duration <- c("9500–8200 BCE", "8200–6300 BCE", "6300–4600 BCE", "4600–3800 BCE", "9200–8200 BCE", "8200–7000 BCE", "7000–5600 BCE", "5600–4500 BCE", "4500–3900 BCE")
+framework <- c(rep("Glørstad (2010)", 4), rep("Reitan (2016)", 5))
+
+df <- data.frame(Period, Duration, framework)
+
+kable(select(df, -framework), booktabs = TRUE, 
+      caption = "Chronological framework. Glørstad's (2010) divison of phases reflects the more established framework, to which Reitan (2016) has recently suggested considerable changes.", 
+      col.names = NULL) %>% 
+pack_rows(index = c("Glørstad (2004)" = 4, "Reitan (2016)" = 5))
+
+```
 
@@
 
+# Call to plot (using patchwork)
 (overview + sitemap) /
@@
 
-```{r tab1}
-
+```{r tab2}
+# Table 2. Names of analysed sites, dating method used, and reported site age.
 sites %>% 
@@
   select(no, site, dating_method, strt_bce, end_bce) %>% 
-  gt() %>% 
-    cols_label(
-    site = "Site name",
-    dating_method = "Dating method",
-    strt_bce = "Reported start (BCE)",
-    end_bce = "Reported end (BCE)",
-  ) %>% 
-  tab_header(
-    title = "Table 1. Analysed sites."
-  ) %>% 
-  tab_options(heading.title.font.size = 13, heading.align = "left")
+  kable(caption = "Analysed sites.", col.names = c("no",
+                    "Site name",
+                    "Dating method",
+                    "Reported start (BCE)",
+                    "Reported end (BCE)"),
+        booktabs = TRUE, linesep = c(""))
 ```
@@
 # familiar with the sites)
-fviz_ca_row(calithics, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+fviz_ca_row(calithics, col.row = "black",
+            fill = lithics$mean_date / 1000, title = "A",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 # Plot sites
-plt1 <- fviz_ca_row(calithics, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+plt1 <- fviz_ca_row(calithics, col.row = "black",
+                    fill = lithics$mean_date / 1000, title = "A",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 plt2 <- fviz_ca_col(calithics, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B") +
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "B") +
   labs(colour = "Contribution to total inertia") +
@@
 # Plot sites for the flint data
-plt3 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000, title = "A",
+plt3 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000,
+                    title = "A",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 plt4 <- fviz_ca_col(caflint, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, title = "B") +
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "B") +
   labs(colour = "Contribution to total inertia") +
@@
           -askeladden_id) %>% 
-  # The reason for not using .keep below is for the option to 
-  #retain artefact_count
+  # The reason for not using .keep below is to 
+  # retain artefact_count
   mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
@@
 
+# VDL and RFSL for all lithics
 wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
@@
 
-# Correlation with mean site date
+# Correlation with mean site date, all lithics
 secfreqcor <- cor(lithics2$mean_date, lithics2$secondary_freq)
@@
 
+# Correlation with mean site date, flint
 secfreqcorflint <- cor(wabiflint$mean_date, wabiflint$secondary_freq)
@@
 Note that details on the transformation of the variables has been left out of the plot for clarity, but follow those given in Figure 4.", fig.height = 8}
-# fig.width = 10
-
 # Rename the transformed variables again to avoid clutter in the plot. 
-# Specify in figure caption that the transformed variables are used 
+# Specify in figure caption that the transformed variables are used.
 lithicspca <- lithicstr %>% 
@@
 # Supplementary table of radiocarbon dates
-radiocarbon %>% 
-  select(-excavation_project) %>% 
-  gt() %>% 
-    cols_label(
-    site = "Site name",
-    material = "Material",
-    lab_code = "Lab code",
-    c14_bp = "C14-age",
-    error = "Error"
-  )
+
+select(radiocarbon, -excavation_project) %>% 
+  kable(booktabs = TRUE, longtable = TRUE,
+      col.names = c("Site name", "Material", "Lab code", "C14-age", "Error"),
+      linesep = c(""))
 ```

### 2021-08-01 README.Rmd
Adjusting readme

@@
     illustrations
--   [:file_folder:
-    supplementary-materials](/analysis/supplementary-materials):
-    Supplementary materials including notes and other documents prepared
-    and collected during the analysis.
 

### 2021-08-01 analysis/paper/paper.Rmd
Added OSF link

@@
 The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported upon, the data and employed R programming script is freely available as a research compendium at [URL placeholder], following @marwick2018b.
+While only parts of this process can sensibly be reported upon, the data and employed R programming script is freely available as a research compendium at https://osf.io/ehjfc/, following @marwick2018b.
 

### 2021-08-02 analysis/paper/paper.Rmd
Minor edits

--- 2021-08-01 analysis/paper/paper.Rmd
@@
 This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting.
-Furthermore, this supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.
+Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.
 
@@
 The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c.
-9200--8200 BCE [@reitan2016] and is set to start with the first recorded human presence in Norway.
+9300--8200 BCE (Table \@ref(tab:tab2)), and is set to start with the first recorded human presence in Norway [@reitan2016].
 Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
@@
 
-# Table 1. Overfiew of chronological framework.
+# Table. Overfiew of chronological framework.
 Period <- c('Early Mesolithic, Fosna Phase', 'Middle Mesolithic, Tørkop Phase', 'Late Mesolithic, Nøstvet Phase', 'Late Mesolithic, Kjeøy Phase',
             'Flake Axe Phase', 'Microlith Phase', 'Pecked Adze Phase', 'Nøstvet Adze Phase', 'Transverse Arrowhead Phase')
-Duration <- c("9500–8200 BCE", "8200–6300 BCE", "6300–4600 BCE", "4600–3800 BCE", "9200–8200 BCE", "8200–7000 BCE", "7000–5600 BCE", "5600–4500 BCE", "4500–3900 BCE")
+Duration <- c("9500–8200 BCE", "8200–6300 BCE", "6300–4600 BCE", "4600–3800 BCE", "9300–8200 BCE", "8200–7000 BCE", "7000–5600 BCE", "5600–4500 BCE", "4500–3900 BCE")
 framework <- c(rep("Glørstad (2010)", 4), rep("Reitan (2016)", 5))
@@
 kable(select(df, -framework), booktabs = TRUE, 
-      caption = "Chronological framework. Glørstad's (2010) divison of phases reflects the more established framework, to which Reitan (2016) has recently suggested considerable changes.", 
+      caption = "Chronological framework. Glørstad's (2010) divison of phases reflects the more traditional framework, to which Reitan (2016) has recently suggested considerable changes.", 
       col.names = NULL) %>% 
-pack_rows(index = c("Glørstad (2004)" = 4, "Reitan (2016)" = 5))
+pack_rows(index = c("Glørstad (2010)" = 4, "Reitan (2016)" = 5))
 
@@
                     "Reported end (BCE)"),
-        booktabs = TRUE, linesep = c(""))
+        booktabs = TRUE, longtable = TRUE, linesep = c(""))
 ```
@@
 One concrete operationalisation of the terms has been forwarded by @barton1998 and collegues [e.g.
-@barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008].
+@barton1999; @barton2013; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008].
 This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].
 Furthermore, the relationship between curated and expedient technological organisation has been related to the continuum defined by @binford1980 between residentially mobile foragers and logistically mobile collectors [@riel-salvatore2004; @clark2017; see also @bamforth1986; @binford1977].
-Residential mobility involves the movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps that are moved less often and from where task-groups venture on targeted forays to retrieve specific resources.
-A higher degree of logistic as opposed to residential mobility thus involves a wider range of site types and associated mobility patterns [@binford1980].
-
-In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern.
+Residential mobility involves the relatively frequent movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps that are moved less often and from where smaller task-groups venture on targeted forays to retrieve specific resources.
+A higher degree of logistic as opposed to residential mobility thus also involves a wider range of site types and associated mobility patterns [@binford1980].
+
+Furthermore, in this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern.
 From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017].
@@
 The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported upon, the data and employed R programming script is freely available as a research compendium at [URL placeholder], following @marwick2018b.
-
-The `r nrow(sites)` analysed sites have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table 1).
+While only parts of this process can sensibly be reported upon, the data and employed R programming script [@rcoreteam] is freely available as a research compendium at https://osf.io/ehjfc/, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+
+The `r nrow(sites)` analysed sites have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table \@ref(tab:tab2)).
 Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators.
-Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam].
+Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R.
 The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021].
 The resulting posterior density estimates were then summed for each site.
-Radiocarbon data is provided in the supplementary data (see also @solheim2020a).
+Radiocarbon data are provided in the supplementary data [see also @solheim2020a].
 
@@
 To get more directly at this temporal trend, a curation index based on VDL and RFSL was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1.
-The values for artefact density was then made negative to reflect its relationship with degree of curation.
+The value for artefact density was then made negative to reflect its relationship with degree of curation.
 The mean was then found for each site on these two normalised values.
@@
 A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from their respective date ranges as provided in Figure \@ref(fig:dist).
-For sites with radiocarbon age determinations the dates were drawn from the summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range.
-This simulation was repeated 1000 times (Figure \@ref(fig:simplot)A).
+For sites with radiocarbon age determinations the dates were drawn from the summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range (Figure \@ref(fig:simplot)).
+This simulation was repeated 1000 times.
 Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE.
@@
 Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely.
-This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear.
+This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages have been used for, leaving the behavioural and social significance of the employed units of analysis unclear.
 The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.
@@
 
-```{r supp1, cache = FALSE}
+```{r supp1}
 # Supplementary table of radiocarbon dates
@@
 
-```{r supp2, cache = FALSE}
+```{r supp2}
 # Most of the code for assembling the Sankey plot has been taken directly from 

### 2021-08-03 analysis/paper/paper.Rmd
Minor edits to references and readme

@@
 ---
-title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
+title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
 header-includes:
@@
 @berg-hansen2017; @eymundsson2018; @damlien2016; @solheim2020].
-Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [@fuglestvedt2007; @eigeland2015; @mansrud2016].
+Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g.
+@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g.
+@fuglestvedt2007; @eigeland2015; @mansrud2016].
 These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
@@
 The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported upon, the data and employed R programming script [@rcoreteam] is freely available as a research compendium at https://osf.io/ehjfc/, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+While only parts of this process can sensibly be reported upon, the data and employed R programming script [@rcoreteam] is freely available as a research compendium at <https://osf.io/ehjfc/>, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
 
@@
 
-```{r var-contrib, fig.cap = "Contribution of variables to the components of the PCA.
+```{r var-contrib, fig.cap = "Contribution of variables to the first two components of the PCA.
 The dotted red line indicates the expected contribution from each variable given a uniform distribution of impact.", fig.height = 3}
@@
 In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf.
-@bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].
+@bjerck2017; @breivik2018; @fuglestvedt2009; @jaksland2014b; @damlien2018b].
 The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016].
@@
 <!-- ```{r colophon, cache = FALSE} -->
+
 <!-- # which R packages and versions? -->
+
 <!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
+
 <!-- ``` -->
@@
 <!-- ```{r} -->
+
 <!-- # what commit is this file at?  -->
+
 <!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
+
 <!-- ``` -->

### 2021-08-04 analysis/paper/paper.Rmd
Preprint

@@
 keywords: |
-  Mesolithic Norway; Lithic assemblages; Mobility; Multivariate statistics
+  Mesolithic Scandinavia; Multivariate Statistics; Mobility; Whole Assemblage Behavioural Indicators
 ---
@@
 
-This study employs multivariate exploratory statistics to analyse lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway.
+This study employs multivariate exploratory statistics to analyse lithic assemblages associated with Mesolithic sites located in south-eastern Norway.
 This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time.
@@
 The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c.
-9300--8200 BCE (Table \@ref(tab:tab2)), and is set to start with the first recorded human presence in Norway [@reitan2016].
+9400--8200 BCE (Table \@ref(tab:tab2)), and is set to start with the first recorded human presence in Norway [@damlien2018b].
 Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
@@
 It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c.
-4300 BCE [@nielsen2021].
+4300 BCE [@nielsen2021, cf.
+@solheim2018a; @solheim2020a].
 
@@
 
-# Table. Overfiew of chronological framework.
+# Table. Overview of chronological framework.
 Period <- c('Early Mesolithic, Fosna Phase', 'Middle Mesolithic, Tørkop Phase', 'Late Mesolithic, Nøstvet Phase', 'Late Mesolithic, Kjeøy Phase',
@@
 
-The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway (Figure \@ref(fig:dist)A).
+The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution (Figure \@ref(fig:dist)A).
 The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b].
@@
 
-The lithic data analysed is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
+The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
 The artefact data have been divided into flint and non-flint materials.
@@
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
-While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g.
-@solheim2018].
-Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017].
 As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [@barton2011; @clark2017, and below].
@@
 This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch.
-The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades.
+The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of unexhausted cores and unretouched flakes and blades.
 These variables and underlying logic constitute what has been termed Whole Assemblage Behavioural Indicators [WABI, @clark2017], and is the main framework adopted here.
@@
 
-Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
+Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 1 mm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
 Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77].
@@
 
-The general impression from the CA is that a chronological dimension is associated with the patterning in the data (Figure \@ref(fig:cor)).
+The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)).
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120].
@@
 The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage.
-It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves.
-The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.
+These assemblages are also to a larger extent characterised by core fragments, both in flint and non-flint materials, rather than cores.
+The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both projectiles and non-flint blades are important constituents of the assemblages at these sites.
 The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages.
 Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that it is an early assemblage characterised by axe production in metarhyolite [@eigeland2014].
-However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised.
-This could explain its position as an outlier in the plot.
+However, the site had been quite heavily impacted by modern disturbances that could have impacted the lithic material and which could explain its position as an outlier in the plot [@eigeland2014].
 Finally, although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
@@
 @solheim2016].
-The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns in the material treated here.
+The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns.
 However, it is worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed.
@@
 These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
-Although the precise nature of this transition would require further consideration, the fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system.
+The fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@binford1980].
+This indicates, in turn, that the measures represent an empirical link between technological organisation and economic behaviour and mobility patterns [@riel-salvatore2004].
 
@@
 
-The results of the CA align well with results of previous research in south-eastern Norway.
-This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories.
+The results of the CA align well with results of previous research in south-eastern Norway, indicating that meaningful chronological patterning is associated with the employed artefact categories.
 These tendencies are already well-established when it comes to the formal tool types and some debitage categories, but have been given less focus in light of entire assemblages.
@@
 The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
-Furthermore, the temporal trends of the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway.
+Furthermore, the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway.
 The explicit quantification does, however, offer the possibility to conduct formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.

### 2021-08-04 analysis/paper/paper.Rmd
Submitted version

@@
       institute: [UiO]
-      correspondence: true
+      correspondence: false
 institute:
@@
 keywords: |
-  Mesolithic Scandinavia; Multivariate Statistics; Mobility; Whole Assemblage Behavioural Indicators
+  Mesolithic Scandinavia; Multivariate statistics; Mobility strategies; Whole Assemblage Behavioural Indicators
 ---
+## Highlights {-}
+
+* Multivariate exploratory analysis of Mesolithic assemblages in south-eastern Norway
+* Explores relevance of artefact categories established in Norwegian archaeology 
+* Explores variables associated with mobility patterns in lithic assemblage studies
+* Draws on the Whole Assemblage Behavioural Indicators (WABI)
+* Relevance for Mesolithic Norway supports the notion that WABI are widely applicable
 
@@
 
-# References
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+# Declaration of interest {-}
+
+The author has no conflicts of interest to declare.
+
+# References {-}
 

### 2021-08-04 analysis/paper/paper.Rmd
Minor edits

@@
 institute:
-  - UiO: University of Oslo, Department of Archaeology, History and Conservation
+  - UiO: University of Oslo, Department of Archaeology, Conservation and History
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2021-08-07 README.Rmd
Minor edits to references and readme
Added link to pre-print and osf repository

@@
 # Please put your title here to include it in the file below.
-Title <- "Exploring the composition of lithic assemblages in Mesolithic
-south-eastern Norway"
+Title <- "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
 ```
@@
 > `r format(Sys.Date(), "%d %b %Y")`. Online at
-> <https://doi.org/xxx/xxx>
+> <https://doi.org/10.31235/osf.io/cqaps>
 
@@
 > data for `r Title`*. Accessed `r format(Sys.Date(), "%d %b %Y")`.
-> Online at <https://doi.org/xxx/xxx>
+> Online at <https://doi.org/10.17605/osf.io/ehjfc>
 

### 2021-12-21 analysis/paper/paper.Rmd
Made png versions of the figures
Changing supplementary material for peer review

@@
 * Multivariate exploratory analysis of Mesolithic assemblages in south-eastern Norway
-* Explores relevance of artefact categories established in Norwegian archaeology 
+* Explores relevance of established artefact categories in Norwegian archaeology 
 * Explores variables associated with mobility patterns in lithic assemblage studies
@@
 
-```{r libraries, eval = TRUE, results = 'hide'}
+```{r libraries, results = 'hide'}
 library(tidyverse)
@@
 ```{r get-functions, include = FALSE}
+
+# Load custom plot functions
 source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
@@
 
-```{r get-data, eval = TRUE, results = 'hide'}
+```{r get-data, results = 'hide'}
+# Read in data
 assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
@@
 
-# gis <- read.csv(here::here('../../site_data.csv'))
-# gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
-# gisjoin <- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
 ```
@@
 ```{r wrangling}
+
+# Some initial data preparation
 sites <- assemblages %>% 
@@
     # Find total artefact count and flint count for each site
-    artefact_count = sum(c_across(core:flake_retouch_other)),
+    artefact_count = sum(c_across(core:flake_retouch_nonflint)),
     flint_count = sum(c_across(core:flake_retouch)),
-    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
+    nonflint_count = sum(c_across(axe_nonflint:flake_retouch_nonflint))) %>%
   # Exclude sites with less than 100 artefacts
@@
 ```{r c14-modelling, cache = TRUE}
-# define function to model radiocarbon dates. Only assumption is that 
+# Define function to model radiocarbon dates. Only assumption is that 
 # dates from a single site belong to a single group
@@
 The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c.
-9400--8200 BCE (Table \@ref(tab:tab2)), and is set to start with the first recorded human presence in Norway [@damlien2018b].
+9300--8200 BCE (Table \@ref(tab:tab2)), and is set to start with the first recorded human presence in Norway [@damlien2018b].
 Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
-@bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018c].
+@bjerck2008; @breivik2016; @fuglestvedt2012; @nærøy2018; but see @viken2018c].
 Around the transition to the subsequent Middle Mesolithic or Microlith Phase at c.
@@
 It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c.
-4300 BCE [@nielsen2021, cf.
-@solheim2018a; @solheim2020a].
+4300 BCE [@nielsen2021; cf.
+@solheim2020a; @solheim2018a].
 
@@
 Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators.
-These follow the original reports.", fig.height = 13, fig.width = 12}
+These follow the original reports.
+Site numbers match those provided in Table 2.", fig.height = 13, fig.width = 12}
 # Create a bounding box around site points to limit the overview map
@@
                             "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet")) +
-  labs(title = "B", fill = "Excavation project") +
-  xlab("Cal. BCE") +
+  labs(title = "B", fill = "Excavation project",
+       x = "cal BCE", y = "Site number") +
   theme_classic() +
-  theme(axis.title.y = element_blank(),
-        axis.ticks.y = element_blank(),
+  theme(axis.ticks.y = element_blank(),
         legend.position = "bottom") +
@@
 
-The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types.
+The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` variables representing different debitage and tool types.
 The artefact data have been divided into flint and non-flint materials.
@@
          end_bce = reported_end_bc * -1,
-         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
+         dating_method = ifelse(dating_method == "shore_typo",
+                                "Shoreline/typology", "Radiocarbon")) %>% 
   select(no, site, dating_method, strt_bce, end_bce) %>% 
@@
 @fuglestvedt2007; @eigeland2015; @mansrud2016].
-These studies are, however, based on narratively driven methods, leaving the weighting of the different variables for the final interpretations unclear.
+These studies are, however, based on non-quantitative and less formal methods, leaving the weighting of the different variables for the final interpretations unclear.
 To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
@@
 @barton1991; @dibble1995; @barton2021].
-Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading.
+Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)) .
+This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
+That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading.
+An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
+While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
+
+```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis.
+Column on the left shows the data as originally compiled, the columns on the right shows how these have been aggregated for the analysis.", fig.height = 7}
+# Most of the code for assembling the Sankey plot has been taken directly from 
+# various walkthroughs at https://www.r-graph-gallery.com
+
+# Create two columns: old and new variable names
+links <- data.frame(value = 1,
+        categories = names(sites %>% select(core:flake_retouch_nonflint))) %>% 
+  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point",
+                                 "transverse_point", "single_edged_point", 
+                                 "hognipen_point", "projectile_undefined"), 
+                                 "projectile",
+                          ifelse(categories %in% c("scraper", "knife", "burin",
+                                 "drill", "blade_retouch", "microblade_retouch",
+                                 "chip_retouch", "fragment_retouch", 
+                                 "flake_retouch"), "small_tool",
+                          ifelse(categories %in% c("axe", "chisel"),
+                                 "macro_tool", 
+                          ifelse(categories %in% c("axe_nonflint", "chisel_nonflint",
+                                 "club_hatchet_nonflint"), "macro_tool_nonflint", 
+                          ifelse(categories %in% c("knife_nonflint", 
+                                 "scraper_nonflint", "drill_nonflint", 
+                                 "microblade_retouch_nonflint", 
+                                 "blade_retouch_nonflint", 
+                                 "fragment_retouch_nonflint",
+                                 "flake_retouch_nonflint"), "small_tool_nonflint",
+                                 paste0(categories, " ")))))))
+
+# Set up nodes and links for the plot
+nodes <- data.frame(
+  name=c(as.character(links$categories), 
+         as.character(links$new_categories)) %>% unique()
+)
+
+links$IDsource <- match(links$categories, nodes$name)-1 
+links$IDtarget <- match(links$new_categories, nodes$name)-1 
+nodes$group <- as.factor(c("none"))
+links$group <- as.factor(links$new_categories)
+
+# Specify colours for the plot (this can be assembled using a combination of 
+# e.g. unique(links$new_connection), rep() and paste(), but I couldn't be 
+# bothered so I just copy+pasted and set it up manually).
+my_colour <-'d3.scaleOrdinal() .domain(["core", "core_fragment", "blade", "microblade", "chip", "fragment", "flake", "axe_debitage", "macro_tool", "projectile", "microburin", "small_tool", "beach_flint_flint_nodule", "macro_tool_nonflint", "axe_debitage_nonflint", "grinding_slab_stone", "projectile_nonflint", "small_tool_nonflint", "hammerstone", "flake_nonflint", "fragment_nonflint", "chip_nonflint", "blade_nonflint", "microblade_nonflint", "core_nonflint", "core_fragment_nonflint,"]) .range(["lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "#ff8086", "#79d49f", "lightgrey", "#8eb8fa", "lightgrey", "#a88c7d", "lightgrey", "lightgrey", "lightgrey", "#ffee8c",  "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey"])'
+
+# Create plot
+sn <- sankeyNetwork(Links = links, Nodes = nodes,
+              Source = "IDsource", Target = "IDtarget",
+              NodeID = "name", Value = "value", colourScale = my_colour,
+              sinksRight = FALSE, LinkGroup = "new_categories", 
+              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)
+
+# This has to be saved as html widget and then retrieved as a pdf or png using
+# webshot.
+htmlwidgets::saveWidget(sn, here::here('analysis/figures/sn1.html'))
+webshot::webshot(here::here('analysis/figures/sn1.html'), 
+                 here::here('analysis/figures/sn1.pdf'), delay = 0.2)
+```
 
@@
 
-As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g.
-@Bicho2020; @breivik2016a; @reitan2016; @kitchel2021]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
+As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns [e.g.
+@Bicho2020; @kitchel2021] and are of central importance in Norwegian Stone Age archaeology [e.g.
+@breivik2016a; @reitan2016]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
 
@@
 The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data.
-While only parts of this process can sensibly be reported upon, the data and employed R programming script [@rcoreteam] is freely available as a research compendium at <https://osf.io/ehjfc/>, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
+While only parts of this process can sensibly be reported upon, all data and employed R programming scripts [@rcoreteam] are freely available as a research compendium at <https://osf.io/ehjfc/>, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017].
 
@@
 The resulting posterior density estimates were then summed for each site.
-Radiocarbon data are provided in the supplementary data [see also @solheim2020a].
 
@@
 As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA.
-This for example pertains to flint tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
-These have all been combined into the single category "small flint tools".
-(A full overview of the aggregated variables and their constituent parts is provided in the supplementary material).
-While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
+Some additional configurations and ways to aggregate the variables are also available as supplementary material to the paper.
 
@@
     macro_tool = axe + chisel, 
-    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
-    small_tool_other = knife_other + scraper_other + drill_other + 
-      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
-      flake_retouch_other,
+    macro_tool_nonflint = axe_nonflint + chisel_nonflint + club_hatchet_nonflint,
+    small_tool_nonflint = knife_nonflint + scraper_nonflint + drill_nonflint + 
+      microblade_retouch_nonflint + blade_retouch_nonflint + fragment_retouch_nonflint +
+      flake_retouch_nonflint,
     .keep = "unused") %>%
@@
 
-flint <- lithics %>% select(!(ends_with("other")), -grinding_slab_stone, -hammerstone)
+flint <- lithics %>% select(!(ends_with("nonflint")), -grinding_slab_stone, -hammerstone)
 ```
@@
 The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)).
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120].
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000].
+The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance also means that the structure of the data is
 
@@
 As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data.
-A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)).
+A second CA was therefore performed, excluding the non-flint material (Figure \@ref(fig:flintcor)).
 While not as substantial, there is clear temporal patterning in the flint data as well.
@@
 
-```{r cor1, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 8, fig.width = 10, eval = FALSE}
-# Plot with site names tags (probably most immediately relevant for an audience 
-# familiar with the sites)
-fviz_ca_row(calithics, col.row = "black",
-            fill = lithics$mean_date / 1000, title = "A",
-                     repel = TRUE, shape = 21, pointsize = 2.9) +
-  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
-                        high="red", midpoint = centdate) +
-  theme(legend.position = "bottom") 
-```
-
-```{r cor, fig.cap = "Correspondence analysis using the artefact count data. A) Row plot, B) Column plot.", fig.height = 13}
+```{r cor, fig.cap = "Correspondence analysis using the artefact count data.
+A) Row plot, B) Column plot.
+Points close together are more similar.
+By evaluating how the variables are distributed on the column plot it is possible to say how these define the two axes, in turn making it possible to relate the distribution of the sites in the row plot to the variables.
+As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns can be made (see Lockyear 2000 for a pedagogical case study).", fig.height = 13}
 # Plot sites
@@
 
-blanks_lithics <- c(blanks_flint, "flake_other", "fragment_other", 
-                     "blade_other",  "microblade_other")
+blanks_lithics <- c(blanks_flint, "flake_nonflint", "fragment_nonflint", 
+                     "blade_nonflint",  "microblade_nonflint")
 
@@
 
-secondary_lithics <- c(secondary_flint, "axe_other", "axe_debitage_other",
-                       "chisel_other", "club_hatchet_other", 
-                       "grinding_slab_stone", "projectile_other", 
-                       "knife_other", "scraper_other", "drill_other", 
-                       "microblade_retouch_other", "blade_retouch_other",
-                       "fragment_retouch_other", "flake_retouch_other")
+secondary_lithics <- c(secondary_flint, "axe_nonflint", "axe_debitage_nonflint",
+                       "chisel_nonflint", "club_hatchet_nonflint", 
+                       "grinding_slab_stone", "projectile_nonflint", 
+                       "knife_nonflint", "scraper_nonflint", "drill_nonflint", 
+                       "microblade_retouch_nonflint", "blade_retouch_nonflint",
+                       "fragment_retouch_nonflint", "flake_retouch_nonflint")
 
@@
   # retain artefact_count
-  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
-         core_freq = sum(core, core_other) / artefact_count,
-         blanks_freq = sum(blade, microblade, flake, flake_other, fragment,
-            fragment_other, blade_other,  microblade_other) / artefact_count,
-         secondary_freq =  sum(axe_debitage, axe, chisel, 
+  mutate(chip_prop =  sum(chip, chip_nonflint) / artefact_count,
+         core_prop = sum(core, core_nonflint) / artefact_count,
+         blanks_prop = sum(blade, microblade, flake, flake_nonflint, fragment,
+            fragment_nonflint, blade_nonflint,  microblade_nonflint) / artefact_count,
+         secondary_prop =  sum(axe_debitage, axe, chisel, 
                 microlith, tanged_point, transverse_point, single_edged_point,
@@
                 knife, burin, drill, blade_retouch, microblade_retouch, 
-                chip_retouch, fragment_retouch, flake_retouch, axe_other,
-                axe_debitage_other, chisel_other, club_hatchet_other, 
-                grinding_slab_stone, projectile_other, knife_other,
-                scraper_other, drill_other, microblade_retouch_other, 
-                blade_retouch_other, fragment_retouch_other,
-                flake_retouch_other) / artefact_count,
-         nonflint_freq = nonflint_count / artefact_count,
+                chip_retouch, fragment_retouch, flake_retouch, axe_nonflint,
+                axe_debitage_nonflint, chisel_nonflint, club_hatchet_nonflint, 
+                grinding_slab_stone, projectile_nonflint, knife_nonflint,
+                scraper_nonflint, drill_nonflint, microblade_retouch_nonflint, 
+                blade_retouch_nonflint, fragment_retouch_nonflint,
+                flake_retouch_nonflint) / artefact_count,
+         nonflint_prop = nonflint_count / artefact_count,
          lithic_dens = artefact_count / excavated_m3) %>%
@@
          -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
-         -chip_other, -core_other, -core_fragment_other) %>% 
+         -chip_nonflint, -core_nonflint, -core_fragment_nonflint) %>% 
   column_to_rownames("no")
@@
          -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
-  select(!ends_with("other")) %>% 
-  mutate(chip_freq =  sum(chip) / artefact_count,
-         core_freq = sum(core) / artefact_count,
-         blanks_freq = sum(blade, microblade, flake, fragment) / artefact_count,
-         secondary_freq =  sum(axe_debitage, axe, chisel, 
+  select(!ends_with("nonflint")) %>% 
+  mutate(chip_prop =  sum(chip) / artefact_count,
+         core_prop = sum(core) / artefact_count,
+         blanks_prop = sum(blade, microblade, flake, fragment) / artefact_count,
+         secondary_prop =  sum(axe_debitage, axe, chisel, 
                 microlith, tanged_point, transverse_point, single_edged_point,
@@
 # VDL and RFSL for all lithics
-wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_freq))
+wabicor <- cor(log10(lithics2$lithic_dens), log10(lithics2$secondary_prop))
 
 # Langangen Vestgård 2 does not have any retouched flint
-wabiflint <- filter(flint2, secondary_freq > 0)
-wabicorflint <- cor(log10(wabiflint$flint_dens), log10(wabiflint$secondary_freq))
+wabiflint <- filter(flint2, secondary_prop > 0)
+wabicorflint <- cor(log10(wabiflint$flint_dens), log10(wabiflint$secondary_prop))
 
 # Correlation with mean site date, all lithics
-secfreqcor <- cor(lithics2$mean_date, lithics2$secondary_freq)
+secfreqcor <- cor(lithics2$mean_date, lithics2$secondary_prop)
 denscor <- cor(lithics2$mean_date, lithics2$lithic_dens)
@@
 # Correlation with mean site date, flint
-secfreqcorflint <- cor(wabiflint$mean_date, wabiflint$secondary_freq)
+secfreqcorflint <- cor(wabiflint$mean_date, wabiflint$secondary_prop)
 denscorflint <- cor(wabiflint$mean_date, wabiflint$flint_dens)
@@
          "log10(lithic_dens)" = log10(lithic_dens),
-         "log10(secondary_freq)" = log10(secondary_freq),
-         "core_freq^(1/3)" = core_freq^(1/3),
-         "nonflint_freq^(1/3)" = nonflint_freq^(1/3), .keep = "unused") %>% 
+         "log10(secondary_prop)" = log10(secondary_prop),
+         "core_prop^(1/3)" = core_prop^(1/3),
+         "nonflint_prop^(1/3)" = nonflint_prop^(1/3), .keep = "unused") %>% 
   relocate("mean_date",
-    "log10(lithic_dens)", "nonflint_freq^(1/3)", "blanks_freq", "log10(secondary_freq)",
-    "chip_freq", "core_freq^(1/3)" 
+    "log10(lithic_dens)", "nonflint_prop^(1/3)", "blanks_prop", "log10(secondary_prop)",
+    "chip_prop", "core_prop^(1/3)" 
   )
@@
    lithic_dens = "log10(lithic_dens)",
-   secondary_freq = "log10(secondary_freq)",
-   core_freq = "core_freq^(1/3)",
-   nonflint_freq = "nonflint_freq^(1/3)"
+   secondary_prop = "log10(secondary_prop)",
+   core_prop = "core_prop^(1/3)",
+   nonflint_prop = "nonflint_prop^(1/3)"
   )
@@
 
-w1 <- ggplot(lithics2, aes(lithic_dens, secondary_freq * 100, 
+w1 <- ggplot(lithics2, aes(lithic_dens, secondary_prop * 100, 
                      fill = mean_date / 1000)) + 
@@
 w2 <- 
-  ggplot(flint2, aes(flint_dens, secondary_freq * 100, 
+  ggplot(flint2, aes(flint_dens, secondary_prop * 100, 
                      fill = mean_date / 1000)) + 
@@
 ```{r index}
-# Create curation index
+# Create curation index using min-max normalisation
+
+# First normalise the two variables
 minmax <- lithics2 %>%  
   mutate(dens =  ((lithic_dens - min(lithic_dens)) / (max(lithic_dens) - min(lithic_dens))),
-         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))
+         secondary = ((secondary_prop - min(secondary_prop)) / (max(secondary_prop) - min(secondary_prop))))
 
@@
   mutate(dens =  ((flint_dens - min(flint_dens)) / (max(flint_dens) - min(flint_dens))),
-         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))
-
+         secondary = ((secondary_prop - min(secondary_prop)) / (max(secondary_prop) - min(secondary_prop))))
+
+# Then find the mean while making dens negative
 indexdata <- minmax %>% 
@@
 indexdata$index_flint <- indexdataflint$index_flint
-
-# Curation index plot simply using mean site age (not in use)
-w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
-  geom_point() +
-  labs(title = "C",
-       x = "Mean date cal. BCE",
-       y = "Curation index") +
-  geom_smooth(method = "loess", col = "red") +
-  theme_classic()  + 
-  ggrepel::geom_label_repel(aes(label = site),box.padding   = 0.35, 
-                  point.padding = 0.5,
-                  segment.color = 'grey50')
-
-w4 <- ggplot(indexdata, aes(x = mean_date, y = index_flint)) + 
-  geom_point() +
-  labs(title = "C",
-       x = "Mean date cal. BCE",
-       y = "Curation index") +
-  geom_smooth(method = "loess", col = "red") +
-  theme_classic()  + 
-  ggrepel::geom_label_repel(aes(label = site),box.padding   = 0.35, 
-                  point.padding = 0.5,
-                  segment.color = 'grey50')
-
 ```
@@
   scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
-  labs(title = "A", x = "Cal. BCE", y = "Curation index (All lithics)") +
+  labs(title = "A", x = "cal BCE", y = "Curation index (All lithics)") +
   theme_classic() +
@@
   scale_x_continuous(breaks = c(seq(-9000,-4000, 1000))) +
-  labs(title = "B", x = "Cal. BCE", y = "Curation index (Flint)") +
+  labs(title = "B", x = "cal BCE", y = "Curation index (Flint)") +
   theme_classic() +
   theme(legend.position = "bottom",
-        text=element_text(size=16))
+        text = element_text(size=16))
 
@@
 This, however, can only constitute the absolute minimum of what is required of a proposed explanation.
-Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses.
+Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses [e.g.
+@clark2009; @perreault2019].
 The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
@@
 
-# Supplementary material A. Radiocarbon dates.
+# Supplementary material
 
 ```{r supp1}
-# Supplementary table of radiocarbon dates
-
-select(radiocarbon, -excavation_project) %>% 
-  kable(booktabs = TRUE, longtable = TRUE,
-      col.names = c("Site name", "Material", "Lab code", "C14-age", "Error"),
-      linesep = c(""))
-```
-
-# Supplementary material B. Aggregation of variables for the correspondence analysis.
-
-```{r supp2}
-# Most of the code for assembling the Sankey plot has been taken directly from 
-# various walkthroughs at https://www.r-graph-gallery.com
-
-# Create two columns: old and new variable names
-links <- data.frame(value = 1,
-        categories = names(sites %>% select(core:flake_retouch_other))) %>% 
-  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point",
-                                 "transverse_point", "single_edged_point", 
-                                 "hognipen_point", "projectile_undefined"), 
-                                 "projectile",
-                          ifelse(categories %in% c("scraper", "knife", "burin",
-                                 "drill", "blade_retouch", "microblade_retouch",
-                                 "chip_retouch", "fragment_retouch", 
-                                 "flake_retouch"), "small_tool",
-                          ifelse(categories %in% c("axe", "chisel"),
-                                 "macro_tool", 
-                          ifelse(categories %in% c("axe_other", "chisel_other",
-                                 "club_hatchet_other"), "macro_tool_other", 
-                          ifelse(categories %in% c("knife_other", 
-                                 "scraper_other", "drill_other", 
-                                 "microblade_retouch_other", 
-                                 "blade_retouch_other", 
-                                 "fragment_retouch_other",
-                                 "flake_retouch_other"), "small_tool_other",
-                                 paste0(categories, " ")))))))
-
-# Set up nodes and links for the plot
-nodes <- data.frame(
-  name=c(as.character(links$categories), 
-         as.character(links$new_categories)) %>% unique()
-)
-
-links$IDsource <- match(links$categories, nodes$name)-1 
-links$IDtarget <- match(links$new_categories, nodes$name)-1 
-nodes$group <- as.factor(c("none"))
-links$group <- as.factor(links$new_categories)
-
-# Specify colours for the plot (this can be assembled using a combination of 
-# e.g. unique(links$new_connection), rep() and paste(), but I couldn't be 
-# bothered so I just copy+pasted and set it up manually).
-my_colour <-'d3.scaleOrdinal() .domain(["core", "core_fragment", "blade", "microblade", "chip", "fragment", "flake", "axe_debitage", "macro_tool", "projectile", "microburin", "small_tool", "beach_flint_flint_nodule", "macro_tool_other", "axe_debitage_other", "grinding_slab_stone", "projectile_other", "small_tool_other", "hammerstone", "flake_other", "fragment_other", "chip_other", "blade_other", "microblade_other", "core_other", "core_fragment_other,"]) .range(["lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "#ff8086", "#79d49f", "lightgrey", "#8eb8fa", "lightgrey", "#a88c7d", "lightgrey", "lightgrey", "lightgrey", "#ffee8c",  "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey"])'
-
-# Create plot
-sn <- sankeyNetwork(Links = links, Nodes = nodes,
-              Source = "IDsource", Target = "IDtarget",
-              NodeID = "name", Value = "value", colourScale = my_colour,
-              sinksRight = FALSE, LinkGroup = "new_categories", 
-              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)
-
-# This has to be saved as html widget and then retrieved as a pdf or png using
-# webshot.
-htmlwidgets::saveWidget(sn, here::here('analysis/figures/sn1.html'))
-webshot::webshot(here::here('analysis/figures/sn1.html'), 
-                 here::here('analysis/figures/sn1.pdf'), delay = 0.2)
+# # Supplementary table of radiocarbon dates
+# 
+# select(radiocarbon, -excavation_project) %>% 
+#   kable(booktabs = TRUE, longtable = TRUE,
+#       col.names = c("Site name", "Material", "Lab code", "C14-age", "Error"),
+#       linesep = c(""))
+```
+
+```{r, fig.height = 15, fig.cap = "Correspondence analysis collapsing artefact types irrespective of raw-material and including proportion of non-flint as its own variable.
+A) Row plot, B) Column plot."}
+
+# Reviewer 2 wanted non-flint as a separate column and the artefact categories
+# combined, irrespective of raw-material. 
+all_nonflint_prop <- sites %>%
+    mutate(
+    projectile = microlith + tanged_point + transverse_point + 
+    single_edged_point + hognipen_point + projectile_undefined + 
+      projectile_nonflint,
+    small_tool = scraper + knife + burin + drill + blade_retouch + 
+      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch +
+      knife_nonflint + scraper_nonflint + drill_nonflint + 
+      microblade_retouch_nonflint + blade_retouch_nonflint + 
+      fragment_retouch_nonflint +
+      flake_retouch_nonflint,
+    macro_tool = axe + chisel + axe_nonflint + 
+      chisel_nonflint + club_hatchet_nonflint,
+    core = core + core_nonflint,
+    core_fragment = core_fragment + core_fragment_nonflint,
+    axe_debitage = axe_debitage + axe_debitage_nonflint,
+    blade = blade + blade_nonflint,
+    microblade = microblade + microblade_nonflint,
+    fragment = fragment + fragment_nonflint,
+    flake = flake + flake_nonflint,
+    chip = chip + chip_nonflint,
+    nonflint_prop = nonflint_count / artefact_count,
+    .keep = "unused") %>% 
+    # Make site names the rownames
+    column_to_rownames(var = "no") %>% 
+    # Deselect a few of the columns not be used
+    select(-excavation_project, -excavated_m3, -dating_method)
+
+nonprop <- CA(select(all_nonflint_prop, -site, -mean_date, -flint_count, 
+                       -reported_start_bc, -reported_end_bc,
+                       -askeladden_id),
+                graph = FALSE)
+
+splt1 <- fviz_ca_row(nonprop, col.row = "black",
+            fill = all_nonflint_prop$mean_date / 1000, title = "A",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  theme(legend.position = "bottom")
+
+splt2 <- fviz_ca_col(nonprop, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "B") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
+
+splt1 /
+splt2
+
+```
+
+```{r, fig.height = 15, fig.cap = "Same as above, only that the proportion of non-flint is used as a supplementary column.
+The negligble difference between the plots indicates that the raw material variability is captured by the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated.
+A) Row plot, B) Column plot."}
+
+# Version of the above where the proportion of non-flint is simply added as a 
+# supplementary column, followng Reviewer 2. 
+moddat <- select(all_nonflint_prop, -site, -mean_date, -flint_count, 
+                       -reported_start_bc, -reported_end_bc,
+                       -askeladden_id)
+
+nonpropsup <- CA(moddat, 
+              col.sup = which(colnames(moddat) == "nonflint_prop"),  
+                graph = FALSE)
+
+splt3 <- fviz_ca_row(nonpropsup, col.row = "black",
+            fill = all_nonflint_prop$mean_date / 1000, title = "A",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  theme(legend.position = "bottom")
+
+splt4 <- fviz_ca_col(nonpropsup, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "B") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
+
+splt3 /
+splt4
+
+```
+
+```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot."}
+all_cats <- sites %>%
+  # Make site names the rownames
+  column_to_rownames(var = "no") %>% 
+  # Deselect a few of the columns not be used
+  select(-excavation_project, -excavated_m3, -dating_method)
+
+caallcats <- CA(select(all_cats, -site, -mean_date, -artefact_count, -flint_count, 
+                       -nonflint_count, -reported_start_bc, -reported_end_bc,
+                       -askeladden_id),
+                graph = FALSE)
+
+splt5 <- fviz_ca_row(caallcats, col.row = "black",
+            fill = all_cats$mean_date / 1000, title = "A",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  theme(legend.position = "bottom")
+
+splt6 <- fviz_ca_col(caallcats, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "B") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
+
+splt5/
+splt6 
 ```

### 2021-12-23 analysis/paper/paper.Rmd
Minor edits

@@
 )
+
+knitr::opts_knit$set(
+  eval.after = "fig.cap"
+)
 ```
@@
 The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g.
-@clark2017], which have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.
+@clark2017] and other dimensions that have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.
 
@@
                  anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
+  ggsn::north(sitejoin, scale = 0.2, symbol = 10,
+              anchor = c(x = anc[2] - 30000, y = anc[1]) + 5500) +
   coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
@@
 @breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g.
-@fuglestvedt2007; @eigeland2015; @mansrud2016].
+@eigeland2015; @fuglestvedt2007; @mansrud2016].
 These studies are, however, based on non-quantitative and less formal methods, leaving the weighting of the different variables for the final interpretations unclear.
@@
 ```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis.
-Column on the left shows the data as originally compiled, the columns on the right shows how these have been aggregated for the analysis.", fig.height = 7}
+The column on the left shows the data as originally compiled.
+The column on the right shows how these have been aggregated for the analysis.", fig.height = 7}
 # Most of the code for assembling the Sankey plot has been taken directly from 
@@
               sinksRight = FALSE, LinkGroup = "new_categories", 
-              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)
+              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 12)
 
@@
 As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA.
-Some additional configurations and ways to aggregate the variables are also available as supplementary material to the paper.
-
-Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 1 mm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
+A version of the CA using the original artefact categories, as well as some additional configurations and ways to aggregate the variables are also available in the supplementary material to the paper.
+
+Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as the total number of artefacts divided by total excavated m³ as given in the original reports; relative frequency of chips, defined as the proportion of artefacts with size \< 1 mm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
 Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77].
@@
 By evaluating how the variables are distributed on the column plot it is possible to say how these define the two axes, in turn making it possible to relate the distribution of the sites in the row plot to the variables.
-As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns can be made (see Lockyear 2000 for a pedagogical case study).", fig.height = 13}
+As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns across the two plots can be made.", fig.height = 13}
 # Plot sites
@@
 
-Thus, while some secondary expectations of the WABI do not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
+Thus, while some secondary expectations of the WABI as related to cores and blanks do not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
 The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material.
@@
 When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones.
-The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016].
+The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatchets and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016].
 However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.
@@
 One possible approach could be to weigh the curation index by the proportion of non-flint material in the assemblages.
-This is not explored further here, however, as the overall tendencies are relatively robust to this effect.
+This is not explored further here, however, as the overall tendencies appear robust to this effect.
 
@@
 
-While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust.
+While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear robust.
 The curation index is relatively high and uniform until some time before 8000 BCE.
@@
 As it stands, the main hypotheses resulting from the present analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and uniform degrees of mobility, which then drop before levelling off at around 7000 BCE.
-These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
+These then remain stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA.
 The fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@binford1980].
@@
 
-```{r, fig.height = 15, fig.cap = "Same as above, only that the proportion of non-flint is used as a supplementary column.
-The negligble difference between the plots indicates that the raw material variability is captured by the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated.
+```{r, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column.
+The negligble difference between this CA and that above indicates that the raw material variability is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated.
+These plots thus hide important variability that is captured in the ones presented in the main text.
 A) Row plot, B) Column plot."}
@@
 
-splt4 <- fviz_ca_col(nonpropsup, col.col = "contrib",
+splt4 <- fviz_ca_col(nonpropsup, col.col.sup = "black", col.col = "contrib",
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
@@
 
-```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot."}
+```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
 all_cats <- sites %>%
@@
 ```
-
-<!-- \newpage -->
-
-<!-- ### Colophon -->
-
-<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies: -->
-
-<!-- ```{r colophon, cache = FALSE} -->
-
-<!-- # which R packages and versions? -->
-
-<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
-
-<!-- ``` -->
-
-<!-- The current Git commit details are: -->
-
-<!-- ```{r} -->
-
-<!-- # what commit is this file at?  -->
-
-<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
-
-<!-- ``` -->

### 2021-12-24 analysis/paper/paper.Rmd
Minor edits to text

@@
 source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
-
-#setOxcalExecutablePath("/tmp/RtmpFVOcBJ")
 ```
@@
 The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g.
-@clark2017] and other dimensions that have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.
+@clark2017] and other factors that have been employed to align properties of lithic assemblages with land-use and mobility patterns.
 
@@
 
-# Some initial data preparation
+# Some initial preparation of the data
 sites <- assemblages %>% 
@@
 ```{r tab1}
-
 # Table. Overview of chronological framework.
@@
 @barton1991; @dibble1995; @barton2021].
-Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)) .
+Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)).
 This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
@@
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000].
-The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance also means that the structure of the data is
+The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots.
 
@@
          blanks_prop = sum(blade, microblade, flake, flake_nonflint, fragment,
-            fragment_nonflint, blade_nonflint,  microblade_nonflint) / artefact_count,
+            fragment_nonflint, blade_nonflint,
+            microblade_nonflint) / artefact_count,
          secondary_prop =  sum(axe_debitage, axe, chisel, 
@@
 wabiflint <- filter(flint2, secondary_prop > 0)
-wabicorflint <- cor(log10(wabiflint$flint_dens), log10(wabiflint$secondary_prop))
+wabicorflint <- cor(log10(wabiflint$flint_dens),
+                    log10(wabiflint$secondary_prop))
 
@@
   relocate("mean_date",
-    "log10(lithic_dens)", "nonflint_prop^(1/3)", "blanks_prop", "log10(secondary_prop)",
-    "chip_prop", "core_prop^(1/3)" 
+    "log10(lithic_dens)", "nonflint_prop^(1/3)", "blanks_prop",
+    "log10(secondary_prop)", "chip_prop", "core_prop^(1/3)" 
   )
@@
 
-```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts for A) All lithics, B) Flint.
-The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}
+```{r wabi, fig.cap = paste0("Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts for A) All lithics (\\textit{r} = ", round(wabicor, 2), "), B) Flint (\\textit{r} = ", round(wabicorflint, 2), ").
+The logarithm is taken to base 10 on all axes."), fig.height = 5, fig.width = 10}
 
@@
   } else {
-    sdate <- sample(as.integer(site["reported_start_bc"]):as.integer(site["reported_end_bc"]), 1)
+    sdate <- sample(
+      as.integer(
+        site["reported_start_bc"]):as.integer(site["reported_end_bc"]), 1)
   }
@@
 
-# Supplementary material
-
-```{r supp1}
-# # Supplementary table of radiocarbon dates
-# 
-# select(radiocarbon, -excavation_project) %>% 
-#   kable(booktabs = TRUE, longtable = TRUE,
-#       col.names = c("Site name", "Material", "Lab code", "C14-age", "Error"),
-#       linesep = c(""))
-```
+Supplementary material
 
@@
 
-# Reviewer 2 wanted non-flint as a separate column and the artefact categories
-# combined, irrespective of raw-material. 
+# A reviewer wanted artefact types combined irrespective of raw-material and 
+# non-flint as supplementary column. I start by including it as a normal 
+# column.
 all_nonflint_prop <- sites %>%
@@
 
-caallcats <- CA(select(all_cats, -site, -mean_date, -artefact_count, -flint_count, 
-                       -nonflint_count, -reported_start_bc, -reported_end_bc,
-                       -askeladden_id),
+caallcats <- CA(select(all_cats, -site, -mean_date, -artefact_count,
+                       -flint_count, -nonflint_count, -reported_start_bc,
+                       -reported_end_bc, -askeladden_id),
                 graph = FALSE)

### 2022-01-04 analysis/paper/paper.Rmd
Edits after review

@@
    - \usepackage{lineno}
+   - \usepackage{afterpage}
    - \linenumbers
@@
 The assemblages were explored following two strains of analysis.
-The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely.
+The first of these entailed the use of artefact categories that are in established use within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely.
 The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods.
@@
 # Load custom plot functions
-source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
+source(here::here("analysis/script/plot_functions.R"),
+       local = knitr::knit_global())
 ```
@@
 @bjerck2008].
-Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@åstveit2018; @breivik2018; @møller1987; @solheim2020].
+Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@åstveit2018; @breivik2018; @møller1987; @solheim2020a].
 In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g.
@@
 @frivoll2017].
-Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.
+Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, for example by impacting accessibility by means of watercraft, the relatively constrained geographical distribution of the sites hopefully counteracts some environmentally given sources of variation.
 
@@
 ```{r variable-aggregation, fig.cap = "Aggregation of variables for the correspondence analysis.
-The column on the left shows the data as originally compiled.
+The column on the left shows the variables as originally compiled.
 The column on the right shows how these have been aggregated for the analysis.", fig.height = 7}
@@
               sinksRight = FALSE, LinkGroup = "new_categories", 
-              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 12)
+              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)
 
@@
 Cells below the diagonal display the bivariate distributions with a fitted OLS-regression.
-The cells above the diagonal display and are coloured by the corresponding Pearson's correlation coefficient.", fig.height = 9.5, fig.width = 10}
+The cells above the diagonal display and are coloured by the corresponding Pearson's correlation coefficient.", fig.height = 10.1, fig.width = 10.1}
 # Transform the skewed variables - sqrt() is used where there are zero values,
@@
 
-The flint material of the later sites are to larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g.
+The flint material of the later sites is to a larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g.
 @solheim2020].
@@
 \newpage
-
-Supplementary material
 
@@
                         high="red", midpoint = centdate) +
-  theme(legend.position = "bottom")
+  ggtitle("Supplementary material") +
+  theme(legend.position = "bottom",
+        plot.title = element_text(size = 23, face = "bold",
+                                  family = "serif"))
 
@@
 ```{r, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column.
-The negligble difference between this CA and that above indicates that the raw material variability is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated.
+The negligble difference between this CA and that above indicates that the flint/non-flint distinction is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated.
 These plots thus hide important variability that is captured in the ones presented in the main text.
@@
 # Version of the above where the proportion of non-flint is simply added as a 
-# supplementary column, followng Reviewer 2. 
+# supplementary column, following suggestion from the reviewer. 
 moddat <- select(all_nonflint_prop, -site, -mean_date, -flint_count, 
@@
 
-splt4 <- fviz_ca_col(nonpropsup, col.col.sup = "black", col.col = "contrib",
+splt4 <- fviz(nonpropsup, element = "col", col.col.sup = "black",
+             color = "contrib", 
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,

### 2022-01-06 analysis/paper/paper.Rmd
Accommodating peer review

@@
 * Multivariate exploratory analysis of Mesolithic assemblages in south-eastern Norway
-* Explores relevance of established artefact categories in Norwegian archaeology 
+* Explores patterns related to established artefact categories in Norwegian archaeology
 * Explores variables associated with mobility patterns in lithic assemblage studies
@@
 Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015].
-Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site.
+Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages.
+This would in turn give analytical access to the organisation of lithic technology and variation in past behaviour, adaptation and demographic development [see for example @andrefsky2009; @binford1979; @barton2011; @dibble2017; @rezek2020].
 The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern.
@@
 Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g.
-@bjerck2008; @breivik2016; @fuglestvedt2012; @nærøy2018; but see @viken2018c].
+@bjerck2008; @breivik2016; @fuglestvedt2012; @nærøy2018; but see @åstveit2014; @viken2018c].
 Around the transition to the subsequent Middle Mesolithic or Microlith Phase at c.
@@
 To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
-In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to subjectively and narratively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
+In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to less formal and subjectively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
 
@@
 That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading.
-An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
 While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
+An underlying assumption of the aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes, which, consequently, is largely intuitively determined.
 
@@
 
-```{r pca-analysis, fig.cap = "PCA using variables that have been related to mobility patterns.
+```{r pca-analysis, fig.cap = "PCA biplot resulting from variables that have been related to mobility patterns.
 Note that details on the transformation of the variables has been left out of the plot for clarity, but follow those given in Figure 4.", fig.height = 8}
@@
 
+# Perform PCA with standardised variables (scale.unit = TRUE)
 pcalithics <- PCA(select(lithicspca, -reported_start_bc, -dating_method,
@@
 Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses [e.g.
-@clark2009; @perreault2019].
+@clark2009, 29--30; @perreault2019].
 The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
@@
 \newpage
+
+```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
+all_cats <- sites %>%
+  # Make site names the rownames
+  column_to_rownames(var = "no") %>% 
+  # Deselect a few of the columns not be used
+  select(-excavation_project, -excavated_m3, -dating_method)
+
+caallcats <- CA(select(all_cats, -site, -mean_date, -artefact_count,
+                       -flint_count, -nonflint_count, -reported_start_bc,
+                       -reported_end_bc, -askeladden_id),
+                graph = FALSE)
+
+splt1 <- fviz_ca_row(caallcats, col.row = "black",
+            fill = all_cats$mean_date / 1000, title = "A",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
+  ggtitle("Supplementary material") +
+  theme(legend.position = "bottom",
+        plot.title = element_text(size = 23, face = "bold",
+                                  family = "serif"))
+
+splt2 <- fviz_ca_col(caallcats, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "B") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
+
+splt1/
+splt2 
+```
 
@@
 
-splt1 <- fviz_ca_row(nonprop, col.row = "black",
+splt3 <- fviz_ca_row(nonprop, col.row = "black",
             fill = all_nonflint_prop$mean_date / 1000, title = "A",
@@
                         high="red", midpoint = centdate) +
-  ggtitle("Supplementary material") +
-  theme(legend.position = "bottom",
-        plot.title = element_text(size = 23, face = "bold",
-                                  family = "serif"))
-
-splt2 <- fviz_ca_col(nonprop, col.col = "contrib",
+  theme(legend.position = "bottom")
+
+splt4 <- fviz_ca_col(nonprop, col.col = "contrib",
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
@@
 
-splt1 /
-splt2
+splt3 /
+splt4
 
@@
 
-splt3 <- fviz_ca_row(nonpropsup, col.row = "black",
+splt5 <- fviz_ca_row(nonpropsup, col.row = "black",
             fill = all_nonflint_prop$mean_date / 1000, title = "A",
@@
 
-splt4 <- fviz(nonpropsup, element = "col", col.col.sup = "black",
+# Seems to be a bug to change colour of supplementary column using 
+# fviz_ca_row(): 
+splt6 <- fviz(nonpropsup, element = "col", pointshape = 17,
+              col.col.sup = "black",
              color = "contrib", 
@@
 
-splt3 /
-splt4
-
-```
-
-```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
-all_cats <- sites %>%
-  # Make site names the rownames
-  column_to_rownames(var = "no") %>% 
-  # Deselect a few of the columns not be used
-  select(-excavation_project, -excavated_m3, -dating_method)
-
-caallcats <- CA(select(all_cats, -site, -mean_date, -artefact_count,
-                       -flint_count, -nonflint_count, -reported_start_bc,
-                       -reported_end_bc, -askeladden_id),
-                graph = FALSE)
-
-splt5 <- fviz_ca_row(caallcats, col.row = "black",
-            fill = all_cats$mean_date / 1000, title = "A",
-                     repel = TRUE, shape = 21, pointsize = 2.9) +
-  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
-                        high="red", midpoint = centdate) +
-  theme(legend.position = "bottom")
-
-splt6 <- fviz_ca_col(caallcats, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
-             title = "B") +
-  labs(colour = "Contribution to total inertia") +
-  theme(legend.position = "bottom")
-
-splt5/
+splt5 /
 splt6
-```
+
+```

### 2022-01-06 README.Rmd
Trying to set up docker and travis

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/isakro/dialpastrepository/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/isakro/exploring-assemblages-se-norway/master?urlpath=rstudio)
 
@@
 The simplest way to explore the text, code and data is to click on
-[binder](https://mybinder.org/v2/gh/isakro/dialpastrepository/master?urlpath=rstudio)
+[binder](https://mybinder.org/v2/gh/isakro/exploring-assemblages-se-norway/master?urlpath=rstudio)
 to open an instance of RStudio in your browser, which will have the

### 2022-01-09 analysis/paper/paper.Rmd
Accommodating comments from reviewers

@@
 While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
-An underlying assumption of the aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes, which, consequently, is largely intuitively determined.
+An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
 
@@
 
-Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as the total number of artefacts divided by total excavated m³ as given in the original reports; relative frequency of chips, defined as the proportion of artefacts with size \< 1 mm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
+Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as the total number of artefacts divided by total excavated m³ as taken from the original reports; relative frequency of chips, defined as the proportion of artefacts with size \< 1mm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material.
 Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77].
@@
 A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g.
-@Bicho2020; @breivik2016a].
+@bergsvik1995, 116; @Bicho2020; @breivik2016a].
 For the assemblage data itself this especially pertains to diversity in tool-types [@canessa2021], which has been omitted in light of the above-metioned Frison effect.
@@
 
-```{r pca-analysis, fig.cap = "PCA biplot resulting from variables that have been related to mobility patterns.
-Note that details on the transformation of the variables has been left out of the plot for clarity, but follow those given in Figure 4.", fig.height = 8}
+```{r pca-analysis, fig.cap = "PCA biplot resulting from analysing variables that have been related to mobility patterns.
+Note that details on the transformation of the variables has been left out of the plot for clarity, but follow those given in Figure 5.", fig.height = 8}
 # Rename the transformed variables again to avoid clutter in the plot. 

### 2022-01-10 analysis/paper/paper.Rmd
Revisions and replies to reviewers

@@
 For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life.
+This has major implications that the above-outlined analysis does
+not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions.
 As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [@barton2011; @clark2017, and below].
 The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1977; @binford1979].
-An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement.
+An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvenation, resulting in high rates of tool replacement.
 Curated technological organisation, on the other hand, has been related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.
@@
 As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns [e.g.
-@Bicho2020; @kitchel2021] and are of central importance in Norwegian Stone Age archaeology [e.g.
+@Bicho2020; @kitchel2021] and are of central importance to Norwegian Stone Age archaeology [e.g.
 @breivik2016a; @reitan2016]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].
@@
 This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000].
-The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots.
+The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots and that these therefore are likely reflect true patterning in the data.
 

### 2022-01-10 analysis/paper/paper.Rmd
Revisions and replies to reviewers

@@
 
-```{r, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
+```{r supplementary1, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
 all_cats <- sites %>%
@@
 
-```{r, fig.height = 15, fig.cap = "Correspondence analysis collapsing artefact types irrespective of raw-material and including proportion of non-flint as its own variable.
+```{r supplementary2, fig.height = 15, fig.cap = "Correspondence analysis collapsing artefact types irrespective of raw-material and including proportion of non-flint as its own variable.
 A) Row plot, B) Column plot."}
@@
 
-```{r, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column.
+```{r supplementary3, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column.
 The negligble difference between this CA and that above indicates that the flint/non-flint distinction is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated.

### 2022-01-19 analysis/paper/paper.Rmd
Minor adjustment following peer review

@@
 
-The lithic data analysed are based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` variables representing different debitage and tool types.
-The artefact data have been divided into flint and non-flint materials.
+The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))`debitage and tool types.
+These represent artefact categories that have been used consistently across the reports.
+Consequently, sub-categories that have only been used in the classification of some inventories have been omitted.
+This for example pertains to what blanks have been used for the production of tools, and thus distinctions such as scrapers produced from flakes or blades, which has only been noted in some of the reports.
+Furthermore, the artefact data have been divided into flint and non-flint materials.
 Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g.
@@
 @eigeland2015; @fuglestvedt2007; @mansrud2016].
-These studies are, however, based on non-quantitative and less formal methods, leaving the weighting of the different variables for the final interpretations unclear.
+These studies are, however, based on non-quantitative or uni- and bivariate methods, leaving the weighting of the many variables for the final interpretations unclear.
 To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
-In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to less formal and subjectively driven methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
+In sum then, previous studies have typically either been limited to a small number of sites, to a subset of the inventories, to morphological characteristics, or to methods that are difficult to scale and consistently balance in the comparison of a larger number of artefact categories and assemblages.
 
@@
 Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)).
-This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch.
-That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading.
+This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch which have been combined into the category 'small flint tools'.
 While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
@@
                      title = "Contribution of variables to Component 2")
-dim3 <- fviz_contrib(pcalithics, choice = "var", axes = 3,
-                     color = "black", fill = "gold2", 
-                     title = "Contribution of variables to Component 3")
-dim4 <- fviz_contrib(pcalithics, choice = "var", axes = 4,
-                     color = "black", fill = "gold2", 
-                     title = "Contribution of variables to Component 4")
 
@@
 (dim1 | dim2) 
-#(dim3 | dim4)
 ```

### 2022-01-20 analysis/paper/paper.Rmd
Minor adjustment following peer review

@@
 
-Thus, while some secondary expectations of the WABI as related to cores and blanks do not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
+Thus, while the relative frequency of blades and cores does not appear to meet the expectations of the WABI, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself.
 The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material.

### 2022-01-24 analysis/paper/paper.Rmd
Final edits before returning to JAS:Reports

@@
    - \usepackage{afterpage}
+   - \usepackage{fancyhdr}
    - \linenumbers
@@
 ---
+
+<!--- Page numbers in margin to avoid conflict with figure captions --->
+\fancypagestyle{plain}{%
+  \fancyhf{}
+  \fancyfoot[R]{\leavevmode\rlap{\thepage~}}
+}
+\pagestyle{plain}
+
 ## Highlights {-}
@@
   guides(fill =  guide_legend(title.position = "top", title.hjust = 0.5),
-         colour = FALSE)
+         colour = "none")
 
@@
 
-The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))`debitage and tool types.
+The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_nonflint))` debitage and tool types.
 These represent artefact categories that have been used consistently across the reports.
 Consequently, sub-categories that have only been used in the classification of some inventories have been omitted.
-This for example pertains to what blanks have been used for the production of tools, and thus distinctions such as scrapers produced from flakes or blades, which has only been noted in some of the reports.
+This for example pertains to what blanks have been used for the production of formal tools, which has only been noted in some of the reports.
 Furthermore, the artefact data have been divided into flint and non-flint materials.
@@
 Studies that have involved entire assemblages have either been concerned with general compositional traits such as relative frequency of various tool types and raw-materials [e.g.
-@breivik2016; @viken2018c; @breivik2020; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g.
+@breivik2020; @breivik2016; @viken2018c; @reitan2016], or involved extremely in-depth studies of technological organisation associated with a handful of assemblages [e.g.
 @eigeland2015; @fuglestvedt2007; @mansrud2016].
-These studies are, however, based on non-quantitative or uni- and bivariate methods, leaving the weighting of the many variables for the final interpretations unclear.
+These studies are, however, based on non-quantitative or descriptive uni- and bivariate methods, leaving the weighting of the many variables for the final interpretations unclear.
 To my knowledge, only a single study dealing with the composition of Mesolithic assemblages in southern Norway has involved the use of a multivariate quantitative framework, which was employed to structure the analysis of eight Middle Mesolithic assemblages [@solheim2013a; see @glørstad2010, 145--146 for a spatial application].
@@
 Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)).
-This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch which have been combined into the category 'small flint tools'.
-While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
+Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading.
 An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
+While aggregating artefact categories in this manner could subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.
 
@@
 
-The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports.
+The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports [e.g.
+@baxter1994; @shennan1997].
 As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA.
@@
 The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)).
-This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000].
+This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)B), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)A, @baxter1994, 119--120; @lockyear2000].
 The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots and that these therefore are likely reflect true patterning in the data.
@@
 Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that it is an early assemblage characterised by axe production in metarhyolite [@eigeland2014].
-However, the site had been quite heavily impacted by modern disturbances that could have impacted the lithic material and which could explain its position as an outlier in the plot [@eigeland2014].
+However, the site had been quite heavily impacted by modern disturbances that could have impacted the lithic material and which could explain its position as an outlier in the plot.
 Finally, although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.
@@
 ```{r cor, fig.cap = "Correspondence analysis using the artefact count data.
-A) Row plot, B) Column plot.
+A) Column plot (variables), B) Row plot (sites).
 Points close together are more similar.
@@
 As these are symmetrical plots, only general statements concerning the interrelation between the rows and the columns across the two plots can be made.", fig.height = 13}
+# Plot variables
+plt1 <- fviz_ca_col(calithics, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "A") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
+
 # Plot sites
-plt1 <- fviz_ca_row(calithics, col.row = "black",
-                    fill = lithics$mean_date / 1000, title = "A",
+plt2 <- fviz_ca_row(calithics, col.row = "black",
+                    fill = lithics$mean_date / 1000, title = "B",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 
-# Plot variables
-plt2 <- fviz_ca_col(calithics, col.col = "contrib",
+# Variables for the flint data
+plt3 <- fviz_ca_col(caflint, col.col = "contrib",
              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
-             title = "B") +
+             title = "A") +
   labs(colour = "Contribution to total inertia") +
@@
 
-# Plot sites for the flint data
-plt3 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000,
-                    title = "A",
+# Sites for the flint data
+plt4 <- fviz_ca_row(caflint, col.row = "black", fill = lithics$mean_date / 1000,
+                    title = "B",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 
-# Variables for the flint data
-plt4 <- fviz_ca_col(caflint, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
-             title = "B") +
-  labs(colour = "Contribution to total inertia") +
-  theme(legend.position = "bottom")
-
 plt1 /
@@
 
-```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Row plot, B) Column plot."}
+```{r flintcor, fig.height = 13, fig.cap = "Correspondence analysis using the flint data. A) Column plot, B) Row plot."}
 plt3 /
@@
 Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses [e.g.
-@clark2009, 29--30; @perreault2019].
+@clark2009, 29--30; @perreault2019, 1--22].
 The empirical relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond.
@@
 
-```{r supplementary1, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Row plot, B) Column plot.", fig.width = 10}
+```{r supplementary1, fig.height = 18, fig.cap = "Correspondence analysis using all original artefact categories. A) Column plot, B) Row plot.", fig.width = 10}
 all_cats <- sites %>%
@@
 
-splt1 <- fviz_ca_row(caallcats, col.row = "black",
-            fill = all_cats$mean_date / 1000, title = "A",
-                     repel = TRUE, shape = 21, pointsize = 2.9) +
-  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
-                        high="red", midpoint = centdate) +
+splt1 <- fviz_ca_col(caallcats, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE) +
+  labs(colour = "Contribution to total inertia") +
   ggtitle("Supplementary material") +
@@
 
-splt2 <- fviz_ca_col(caallcats, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
-             title = "B") +
-  labs(colour = "Contribution to total inertia") +
+splt2 <- fviz_ca_row(caallcats, col.row = "black",
+            fill = all_cats$mean_date / 1000, title = "B",
+                     repel = TRUE, shape = 21, pointsize = 2.9) +
+  scale_fill_gradient2(name = "Mean age kyr BCE", low="yellow", mid="orange", 
+                        high="red", midpoint = centdate) +
   theme(legend.position = "bottom")
@@
 ```{r supplementary2, fig.height = 15, fig.cap = "Correspondence analysis collapsing artefact types irrespective of raw-material and including proportion of non-flint as its own variable.
-A) Row plot, B) Column plot."}
+A) Column plot, B) Row plot."}
 
@@
 
-splt3 <- fviz_ca_row(nonprop, col.row = "black",
-            fill = all_nonflint_prop$mean_date / 1000, title = "A",
+splt3 <- fviz_ca_col(nonprop, col.col = "contrib",
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "A") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
+
+splt4 <- fviz_ca_row(nonprop, col.row = "black",
+            fill = all_nonflint_prop$mean_date / 1000, title = "B",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 
-splt4 <- fviz_ca_col(nonprop, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
-             title = "B") +
-  labs(colour = "Contribution to total inertia") +
-  theme(legend.position = "bottom")
-
 splt3 /
@@
 ```{r supplementary3, fig.height = 15, fig.cap = "Same as above, only that here the proportion of non-flint is used as a supplementary column.
+A) Column plot, B) Row plot.
 The negligble difference between this CA and that above indicates that the flint/non-flint distinction is integrated in the different artefact types, and therefore that the effect of artefact types and raw material cannot be seperated.
-These plots thus hide important variability that is captured in the ones presented in the main text.
-A) Row plot, B) Column plot."}
+These plots thus hide important variability that is captured in the ones presented in the main text."}
 
@@
 
-splt5 <- fviz_ca_row(nonpropsup, col.row = "black",
-            fill = all_nonflint_prop$mean_date / 1000, title = "A",
+# Seems to be a bug to change colour of supplementary column using 
+# fviz_ca_row(): 
+splt5 <- fviz(nonpropsup, element = "col", pointshape = 17,
+              col.col.sup = "black",
+             color = "contrib", 
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "A") +
+  labs(colour = "Contribution to total inertia") +
+  theme(legend.position = "bottom")
+
+splt6 <- fviz_ca_row(nonpropsup, col.row = "black",
+            fill = all_nonflint_prop$mean_date / 1000, title = "B",
                      repel = TRUE, shape = 21, pointsize = 2.9) +
@@
 
-# Seems to be a bug to change colour of supplementary column using 
-# fviz_ca_row(): 
-splt6 <- fviz(nonpropsup, element = "col", pointshape = 17,
-              col.col.sup = "black",
-             color = "contrib", 
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
-             title = "B") +
-  labs(colour = "Contribution to total inertia") +
-  theme(legend.position = "bottom")
-
 splt5 /

### 2022-02-08 analysis/paper/paper.Rmd
Setting up word document

@@
 output: 
-    bookdown::pdf_document2:
+    bookdown::word_document2:
       toc: false
@@
 csl: "../templates/saa.csl" # Insert path for the bib-style
+always_allow_html: true
 abstract: |

### 2022-02-08 analysis/paper/paper.Rmd
Cleaned out some unnecessary files

--- 2022-01-24 analysis/paper/paper.Rmd
@@
 splt1 <- fviz_ca_col(caallcats, col.col = "contrib",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE) +
+             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE,
+             title = "A") +
   labs(colour = "Contribution to total inertia") +
-  ggtitle("Supplementary material") +
-  theme(legend.position = "bottom",
-        plot.title = element_text(size = 23, face = "bold",
-                                  family = "serif"))
+  theme(legend.position = "bottom")
 

### 2022-02-19 analysis/paper/paper.Rmd
Minor edits

@@
 The first of these entailed the use of artefact categories that are in established use within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely.
-The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods.
+The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research.
 Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g.
@@
 @barton1991; @dibble1995; @barton2021].
-Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)).
-Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading.
+Consequently, artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading (Figure \@ref(fig:variable-aggregation)).
 An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes.
@@
 
-# A reviewer wanted artefact types combined irrespective of raw-material and 
-# non-flint as supplementary column. I start by including it as a normal 
-# column.
+# Based on suggestion from reviewer: artefact types combined irrespective of 
+# raw-material and  non-flint as supplementary column. I start by including 
+# it as a normal column.
 all_nonflint_prop <- sites %>%
@@
 # Version of the above where the proportion of non-flint is simply added as a 
-# supplementary column, following suggestion from the reviewer. 
+# supplementary column, following suggestion from reviewer. 
 moddat <- select(all_nonflint_prop, -site, -mean_date, -flint_count, 

### 2022-02-19 README.Rmd
Minor edits
Included ORCID

@@
 
-<!-- This repository contains the data and code for the paper: -->
+This repository contains the data and code for the paper:
 
-<!-- > Authors, (YYYY). *`r Title`*. Name of journal/book -->
-<!-- > <https://doi.org/xxx/xxx> -->
+> Roalkvam, Isak <a href="https://orcid.org/0000-0001-6974-1374"> <img alt="ORCID logo" src="https://info.orcid.org/wp-content/uploads/2019/11/orcid_16x16.png" width="16" height="16" /></a> (2022).
+`r Title`.
+*Journal of Archaeological Science: Reports*
+> <https://doi.org/10.1016/j.jasrep.2022.103371>
 
@@
 
->  Roalkvam, Isak (2021). *`r Title`*. Accessed
+>  Roalkvam, Isak (2022). *`r Title`*. Accessed
 > `r format(Sys.Date(), "%d %b %Y")`. Online at
 > <https://doi.org/10.31235/osf.io/cqaps>
-
-### How to cite
 
